### 2018-02-05 paper/draft1.tex
add Hogg's outline

\documentclass[onecolumn]{aastex62}

\usepackage{color}
\usepackage{amsmath}
\usepackage{enumitem}

\definecolor{bcolor}{RGB}{0, 51, 153}
\hypersetup{linkcolor=bcolor,citecolor=green,filecolor=cyan,urlcolor=bcolor}

\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}

\received{\color{red} MM/DD/YY}
\revised{\color{red} MM/DD/YY}
\accepted{\color{red} MM/DD/YY}
\submitjournal{\color{red} JOURNAL}

\shorttitle{Birky \& Hogg}
\shortauthors{Birky \& Hogg}

\begin{document}

\title{DATA DRIVEN MODELS FOR APOGEE M DWARFS}

\correspondingauthor{Jessica Birky}
\email{jbirky@ucsd.edu}

\author[0000-0002-7961-6881]{Jessica Birky}
\affil{Center for Astrophysics and Space Science, University of California San Diego, La Jolla, CA 92093, USA}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}
\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726
Broadway, New York, NY 10003, USA}
\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}

\begin{abstract}

The Cannon (Ness et al. 2015) is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (APOGEE; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mu m$) spectra of giant stars to estimate stellar labels (Teff, logg, [Fe/H], and chemical abundances) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ (Schmidt et al. 2016), motivates the extension of this approach to M dwarf stars. Using a training set of 51 M dwarfs with spectral types ranging M0-M9 obtained from SDSS optical spectra, we demonstrate that The Cannon can infer spectral types to a precision of 0.6 types. We then use 30 M dwarfs ranging 3072 $<$ Teff $<$ 4131K, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 44K and 0.05 dex respectively. Additionally we discuss the extension of a model to other labels, and the scientific objectives a data-driven pipeline could enable.

\end{abstract}

\keywords{stars: Mdwarfs -- stellar parameters}

%%==========================================================================================
\section{Introduction} \label{sec:intro}

\begin{enumerate}
\item[-] One paragraph about why M-dwarfs are so important in astronomy right now.

\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them. We might need help here!

\item[-] The idea in general of The Cannon, and the fact that it could be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects. Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all!
\end{enumerate}

%%==========================================================================================
\section{Data} \label{sec:data}

\begin{enumerate}
\item[-] Describe the APOGEE data. Note that it concentrates on giant stars but gets many M-dwarfs nonetheless.

\item[-] Describe the spectral-type training data. One paragraph plus citations. Include any cuts we apply.

\item[-] Describe the Teff, Fe/H training data. Same; Include any cuts.
\end{enumerate}

%%==========================================================================================
\section{The Cannon} \label{sec:cannon}
A few-paragraph summary of what The Cannon does, with references to Ness, Casey, Ho.

%%==========================================================================================
\section{Experimental Results} \label{sec:results}
\begin{enumerate}
\item[-] The spectral-type Cannon. Show the results on your AAS poster, and also show plots of derivatives wrt type, with maybe blow-ups on interesting features.

\item[-] The parameter Cannon. Same.
\end{enumerate}

%%==========================================================================================
\section{Discussion} \label{sec:discussion}

\begin{enumerate}
\item[-] Brief summary of what we did and found

\item[-] Things we learned along the way: For example, not all training sets worked well in our experiments, probably because they were too small or had too little dynamic range.

\item[-] Some discussion of precision and accuracy: The Cannon can produce precise results, but is only as accurate as its training set.

\item[-] Some discussion of whether the derivatives are sensible or surprising.

\item[-] Other assumptions / issues?
\end{enumerate}

%%==========================================================================================
\end{document}

### 2018-03-19 paper/draft1.tex
add ouline details

@@
-\documentclass[onecolumn]{aastex62}
+\documentclass[modern]{aastex62}
 
@@
 \definecolor{bcolor}{RGB}{0, 51, 153}
-\hypersetup{linkcolor=bcolor,citecolor=green,filecolor=cyan,urlcolor=bcolor}
+\hypersetup{linkcolor=bcolor,citecolor=bcolor,filecolor=cyan,urlcolor=bcolor}
 
@@
 
-The Cannon (Ness et al.
-2015) is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (APOGEE; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mu m$) spectra of giant stars to estimate stellar labels (Teff, logg, [Fe/H], and chemical abundances) to precisions higher than the model-grid pipeline.
-The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ (Schmidt et al.
-2016), motivates the extension of this approach to M dwarf stars.
+The Cannon \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (APOGEE; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mu m$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H], and chemical abundances) to precisions higher than the model-grid pipeline.
+The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars.
 Using a training set of 51 M dwarfs with spectral types ranging M0-M9 obtained from SDSS optical spectra, we demonstrate that The Cannon can infer spectral types to a precision of 0.6 types.
-We then use 30 M dwarfs ranging 3072 $<$ Teff $<$ 4131K, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 44K and 0.05 dex respectively.
+We then use 30 M dwarfs ranging $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 44K and 0.05 dex respectively.
 Additionally we discuss the extension of a model to other labels, and the scientific objectives a data-driven pipeline could enable.
@@
 \begin{enumerate}
-\item[-] One paragraph about why M-dwarfs are so important in astronomy right now.
+\item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (Trappist 1, other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) making them an interesting probe of Milky way populations.
+\color{red}{Need citations!}
+\end{enumerate}
 
-\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them. We might need help here!
+\subsection{M dwarf Challenge}
+\begin{enumerate}
+\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources...
+other catalogs?
+-- \color{red}{Need citations!}\color{black})
 
-\item[-] The idea in general of The Cannon, and the fact that it could be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects.
+\item[-] Why is there a problem labeling M dwarfs?
+Generally, two kinds of approaches to labeling: (1) model the spectra using atmospheric models (BTSettl, PHOENIX, KURUCZ, MARCS, ATLAS, etc.) find some places in literature where people have tried these on M dwarfs (\color{red}{Need citations...
+}\color{black} look at Burgasser, Blake...) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\color{red}{Need citations...
+}\color{black} look at Allard, Homeier...); (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (use photometries) -- however we want to use \emph{all} of the data!
+
+% \item[-] The idea in general of The Cannon, and the fact that it could be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects.
 Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all!
+
+\item[-] Why The Cannon approach?
+Can be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects; Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all!
+
 \end{enumerate}
@@
 
+\subsection{APOGEE Survey}
 \begin{enumerate}
-\item[-] Describe the APOGEE data. Note that it concentrates on giant stars but gets many M-dwarfs nonetheless.
+\item[-] Describe the APOGEE data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).
 
-\item[-] Describe the spectral-type training data. One paragraph plus citations. Include any cuts we apply.
+\item[-] Some M dwarfs in APOGEE proposed as ancilliary targets (Blake/Covey sources), plus an unknown number observed unintentionally \citep{Desphande:2013}.
 
-\item[-] Describe the Teff, Fe/H training data. Same; Include any cuts.
+\item[-] Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
+\end{enumerate}
+
+\subsection{Training Samples}
+\begin{enumerate}
+\item[-] Describe the spectral-type training data \citep{West:2011}. One paragraph plus citations. Include any cuts we apply. \color{red}{Make diagnostic plots.}\color{black}
+
+\item[-] Describe the Teff, Fe/H training data \citep{Mann:2015}. Same; Include any cuts. \color{red}{Make diagnostic plots.}\color{black}
 \end{enumerate}
@@
 %%==========================================================================================
-\section{The Cannon} \label{sec:cannon}
-A few-paragraph summary of what The Cannon does, with references to Ness, Casey, Ho.
+\section{Methods} \label{sec:cannon}
+
+\subsection{The Cannon: A Data-Driven Approach}
+\begin{enumerate}
+\item[-] A few-paragraph summary of what The Cannon does, with references (\citealt{Ness:2015}, \citealt{Casey:2016}, \citealt{Ho:2017a}, \citealt{Ho:2017b})
+\end{enumerate}
+
+\subsection{Data-Driven Model}
+\begin{enumerate}
+\item[-] Describe assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label. 
+\item[-] Briefly describe modeling procedure (training step, test step, quadratic labels, etc.), refer to \citealt{Ness:2015} for full description.
+\item[-] Briefly describe model inputs and outputs (1D spectral type model and 2D teff/metallicity and label vectors for those). Mention other models we've tested.
+\end{enumerate}
 
@@
 \section{Experimental Results} \label{sec:results}
-\begin{enumerate}
-\item[-] The spectral-type Cannon. Show the results on your AAS poster, and also show plots of derivatives wrt type, with maybe blow-ups on interesting features.
+% \begin{enumerate}
+% \item[-] The spectral-type Cannon. Show the results on your AAS poster, and also show plots of derivatives wrt type, with maybe blow-ups on interesting features.
 
-\item[-] The parameter Cannon. Same.
-\end{enumerate}
+% \item[-] The parameter Cannon. Same.
+% \end{enumerate}
+
+\subsection{Spectral Type Model}
+\begin{figure}[ht]
+\plotone{figures/west_spt_self_test.png}
+\caption{Spectral type label self test.}
+\end{figure}
+
+\begin{figure}[ht]
+\plotone{figures/1_Spectral_Sequence.pdf}
+\plotone{figures/2_Spectral_Sequence.pdf}
+\plotone{figures/3_Spectral_Sequence.pdf}
+\caption{ Spectral sequence of dwarfs in training set M0-M9; chip 1 of APOGEE
+spectrum with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.}
+\end{figure}
+
+\subsection{Temperature/Metallicity Model}
+\begin{figure}[ht]
+\plottwo{figures/mann_teff_self_test.png}{figures/mann_feh_self_test.png}
+\caption{Temperature and Metallicity label self test.}
+\end{figure}
+
+\begin{figure}[ht]
+\plotone{figures/mann_model_demo.png}
+\caption{\textit{Top two plots:} Mann-trained model for varying temperatures; \textit{bottom two plots:} Mann-trained model for varying metallicities.}
+\end{figure}
 
@@
 
-\item[-] Some discussion of whether the derivatives are sensible or surprising.
+\item[-] The Cannon as a test of evaluating training parameter accuracy?
+How to analyze this?...
+\color{red}{Maybe add training on theoretical models?
+Could be a test of seeing how well The Cannon could derive labels if there was a "ground truth" and tell us limitations of The Cannon model (i.e.
+we could say, for input spectra with zero label uncertainty, The Cannon can reproduce flux at each pixel with less than $XX\%$ error (comment on how well The Cannon reconstructs important features), and can reproduce labels to $XX\%$ uncertainty).
+Compare this to Mann and other traiing sets.
+Not hard to add, but having some issues with continuum normalization.}\color{black}
+
+\item[-] Some discussion of whether the derivatives are sensible or surprising. \color{red}{Add derivative plots.}\color{black}
+
+\item[-] Analysis of specific features?
+\color{red}{Can easily plot some line lists, but not sure where to go with the analysis...
+What features are important for M dwarfs?
+Compare derivative plots of specific lines between theory-trained models and data-trained models?}\color{black}
 
@@
 %%==========================================================================================
+
+\bibliographystyle{aasjournal}
+\bibliography{ref}
+
 \end{document}

### 2018-03-26 paper/draft1/draft1.tex
add derivative plots

--- 2018-03-19 paper/draft1.tex
@@
 
-\begin{enumerate}
-\item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (Trappist 1, other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) making them an interesting probe of Milky way populations.
-\color{red}{Need citations!}
-\end{enumerate}
+% \begin{enumerate}
+% \item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (The CARMENES search for exoplanets around M dwarfs \citealt{Trifonov:2018}; some famous examples: Trappist 1 (M8) -- \citealt{Gillon:2017}; Kepler-186, Kepler-138...
+other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
+% \end{enumerate}
+Very low mass (VLM) are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass 
+ are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
 
@@
 \begin{enumerate}
-\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources...
-other catalogs?
--- \color{red}{Need citations!}\color{black})
+\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources; \citealt{Deacon:2014} -- \color{red}{Need citations!}\color{black})
 
-\item[-] Why is there a problem labeling M dwarfs?
-Generally, two kinds of approaches to labeling: (1) model the spectra using atmospheric models (BTSettl, PHOENIX, KURUCZ, MARCS, ATLAS, etc.) find some places in literature where people have tried these on M dwarfs (\color{red}{Need citations...
+\item[-] Why is there a problem labeling M dwarfs? Generally, two kinds of approaches to labeling:
+
+ (1) model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\color{red}{Need citations...
 }\color{black} look at Burgasser, Blake...) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\color{red}{Need citations...
-}\color{black} look at Allard, Homeier...); (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (use photometries) -- however we want to use \emph{all} of the data!
+}\color{black} \citealt{Allard:2013});
+
+ (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (\citealt{Casagrande:2008} estimated effective temperatures, bolometric luminosities, and metallicities using optical and infrared photometries) -- however we want to use \emph{all} of the data!
 
@@
 
+\item[-] Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
+
 \end{enumerate}
+
+A number of M dwarf studies
 
@@
 \subsection{APOGEE Survey}
-\begin{enumerate}
-\item[-] Describe the APOGEE data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).
+% \begin{enumerate}
+% \item[-] Describe the APOGEE data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).
 
-\item[-] Some M dwarfs in APOGEE proposed as ancilliary targets (Blake/Covey sources), plus an unknown number observed unintentionally \citep{Desphande:2013}.
+% \item[-] Some M dwarfs in APOGEE proposed as ancilliary targets (Blake/Covey sources), plus an unknown number observed unintentionally \citep{Desphande:2013}.
+% \end{enumerate}
 
-\item[-] Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
-\end{enumerate}
+The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey IV (SDSS-IV) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
+While APOGEE is primarily targeted for bright stellar populations (descirbe cuts \citealt{Zasowski:2013}) with the objective of studying galactic composition and evolution, numerous cool, main sequence have also been observed either as targets proposed by the M dwarf ancilliary survey \citep{Desphande:2013}
+
+Fundamental parameters for each of these stars (including temperature, surface gravity, metallicity and 15 chemical abundances) are estimated by the APOGEE Stellar Parameter
+and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016})
 
@@
 \section{Experimental Results} \label{sec:results}
-% \begin{enumerate}
-% \item[-] The spectral-type Cannon. Show the results on your AAS poster, and also show plots of derivatives wrt type, with maybe blow-ups on interesting features.
-
-% \item[-] The parameter Cannon. Same.
-% \end{enumerate}
 
 \subsection{Spectral Type Model}
+
+\begin{enumerate}
+\item[-] We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and was able to achieve a precision of $\pm$0.8 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
+\item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure 2 displays
+\end{enumerate}
+
 \begin{figure}[ht]
 \plotone{figures/west_spt_self_test.png}
-\caption{Spectral type label self test.}
+\caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest}
 \end{figure}
@@
 \caption{ Spectral sequence of dwarfs in training set M0-M9; chip 1 of APOGEE
-spectrum with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.}
+spectrum with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
 \end{figure}
@@
 \plottwo{figures/mann_teff_self_test.png}{figures/mann_feh_self_test.png}
-\caption{Temperature and Metallicity label self test.}
+\caption{Temperature and Metallicity label self test.} \label{fig:mann_selftest}
 \end{figure}
@@
 \begin{figure}[ht]
-\plotone{figures/mann_model_demo.png}
-\caption{\textit{Top two plots:} Mann-trained model for varying temperatures; \textit{bottom two plots:} Mann-trained model for varying metallicities.}
+\plotone{figures/1_physical_model.png}
+\plotone{figures/2_physical_model.png}
+\plotone{figures/3_physical_model.png}
+\caption{\textit{Top two panels:} Mann-trained model for varying temperatures; \textit{bottom two panels:} Mann-trained model for varying metallicities.} \label{fig:mann_demo}
 \end{figure}
@@
 
-\item[-] The Cannon as a test of evaluating training parameter accuracy?
-How to analyze this?...
-\color{red}{Maybe add training on theoretical models?
-Could be a test of seeing how well The Cannon could derive labels if there was a "ground truth" and tell us limitations of The Cannon model (i.e.
-we could say, for input spectra with zero label uncertainty, The Cannon can reproduce flux at each pixel with less than $XX\%$ error (comment on how well The Cannon reconstructs important features), and can reproduce labels to $XX\%$ uncertainty).
-Compare this to Mann and other traiing sets.
-Not hard to add, but having some issues with continuum normalization.}\color{black}
-
 \item[-] Some discussion of whether the derivatives are sensible or surprising. \color{red}{Add derivative plots.}\color{black}
 
-\item[-] Analysis of specific features?
+\item[-] Comment on specific features?
 \color{red}{Can easily plot some line lists, but not sure where to go with the analysis...
@@
 
-\item[-] Other assumptions / issues?
+\item[-] The Cannon as a test of evaluating training parameter accuracy? 
 \end{enumerate}
 
+\begin{figure}[ht]
+\plotone{figures/west_spt_derivative.png}
+\caption{Derivative plots for spectral type model.} \label{fig:west_derivative}
+\end{figure}
+
+\begin{figure}[ht]
+\plotone{figures/mann_teff_derivative.png}
+\plotone{figures/mann_feh_derivative.png}
+\caption{Derivative plots for physical parameter model.
+\textit{Top:} Derivative of flux with repsect to temperature.
+\textit{Bottom:} Derivative of flux with repsect to metallicity.} \label{fig:mann_derivative}
+\end{figure}
+
 %%==========================================================================================
-
+\clearpage
 \bibliographystyle{aasjournal}

### 2018-03-26 paper/outline/outline.tex
add derivative plots

--- 2018-03-19 paper/draft1.tex
@@
 \begin{enumerate}
-\item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (Trappist 1, other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) making them an interesting probe of Milky way populations.
+\item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (The CARMENES search for exoplanets around M dwarfs \citealt{Trifonov:2018}; some famous examples: Trappist 1 (M8) -- \citealt{Gillon:2017}; Kepler-186, Kepler-138...
+other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
 \color{red}{Need citations!}
@@
 \begin{enumerate}
-\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources...
-other catalogs?
--- \color{red}{Need citations!}\color{black})
+\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources; \citealt{Deacon:2014} -- \color{red}{Need citations!}\color{black})
 
-\item[-] Why is there a problem labeling M dwarfs?
-Generally, two kinds of approaches to labeling: (1) model the spectra using atmospheric models (BTSettl, PHOENIX, KURUCZ, MARCS, ATLAS, etc.) find some places in literature where people have tried these on M dwarfs (\color{red}{Need citations...
+\item[-] Why is there a problem labeling M dwarfs? Generally, two kinds of approaches to labeling:
+
+ (1) model the spectra using atmospheric models (BTSettl, PHOENIX, KURUCZ, MARCS, ATLAS, etc.) find some places in literature where people have tried these on M dwarfs (\color{red}{Need citations...
 }\color{black} look at Burgasser, Blake...) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\color{red}{Need citations...
-}\color{black} look at Allard, Homeier...); (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (use photometries) -- however we want to use \emph{all} of the data!
+}\color{black} \citealt{Allard:2013});
+
+ (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (\citealt{Casagrande:2008} estimated effective temperatures, bolometric luminosities, and metallicities using optical and infrared photometries) -- however we want to use \emph{all} of the data!
 
@@
 %%==========================================================================================
+\section{Future Work} \label{sec:future}
 
+\begin{enumerate}
+\item[-] M dwarf pipeline: with The Cannon we can identify and label \emph{all} of the M dwarfs in APOGEE.
+\item[-] Describe ideal Cannon input (for example using FGK pairs for very accurate metallicites).
+\end{enumerate}
+
+%%==========================================================================================
+\section{Conclusion} \label{sec:conclusion}
+Reiterate findings.
+
+%%==========================================================================================
+\clearpage
 \bibliographystyle{aasjournal}

### 2018-03-26 paper/draft1/draft1.tex
apogee survey description

@@
 The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey IV (SDSS-IV) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-While APOGEE is primarily targeted for bright stellar populations (descirbe cuts \citealt{Zasowski:2013}) with the objective of studying galactic composition and evolution, numerous cool, main sequence have also been observed either as targets proposed by the M dwarf ancilliary survey \citep{Desphande:2013}
+Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure to fit ATLAS9 radiative transfer models to determine atmospheric parameters (temperature, surface gravity, metallicity) and 15 chemical abundances \citep{Meszaros:2012}.
 
-Fundamental parameters for each of these stars (including temperature, surface gravity, metallicity and 15 chemical abundances) are estimated by the APOGEE Stellar Parameter
-and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016})
+APOGEE is primarily targeted for bright stellar populations, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution.
+However numerous (specific estimate based on Gaia?) cool, main sequence sources have also been observed either accidentally, or as targets proposed by the M dwarf ancilliary survey \citep{Desphande:2013}.
 
 \subsection{Training Samples}
-\begin{enumerate}
-\item[-] Describe the spectral-type training data \citep{West:2011}. One paragraph plus citations. Include any cuts we apply. \color{red}{Make diagnostic plots.}\color{black}
 
-\item[-] Describe the Teff, Fe/H training data \citep{Mann:2015}. Same; Include any cuts. \color{red}{Make diagnostic plots.}\color{black}
-\end{enumerate}
+The Cannon uses a \emph{data-driven} approach for constructing a model which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels, a flexible number of parameters chosen by the user.
+The basic principle
+
+For the purpose of this study we have constructed two different training samples: first a one-dimensional \emph{spectral type model}, and second a two-dimensional \emph{physical parameter model} (the choice of training labels and dimensionality of our data set we discuss further in Section \ref{sec:discussion}).
+
+The spectral type sample consists of 51 sources from \citealt{West:2011}, spanning M0-M9
+
+% \begin{enumerate}
+% \item[-] Describe the spectral-type training data \citep{West:2011}. One paragraph plus citations. Include any cuts we apply. \color{red}{Make diagnostic plots.}\color{black}
+
+% \item[-] Describe the Teff, Fe/H training data \citep{Mann:2015}. Same; Include any cuts. \color{red}{Make diagnostic plots.}\color{black}
+% \end{enumerate}
+
+\subsection{Data Processing}
+Describe ASPCAP spectra files and Ness continuum normalization procedure.
 
@@
 \subsection{The Cannon: A Data-Driven Approach}
+
 \begin{enumerate}
@@
 \end{enumerate}
+
+Constructing such a model requires two steps: first a \emph{training step} in which a polynomial model is
+
+The key advantages of The Cannon being:
+Data-driven models take away the challenge of directly infering labels from a survey, instead we \emph{transfer labels} from another survery that is more accurate or easier to model.
 
@@
 \end{figure}
+
+\begin{enumerate}
+\item[-] For the physical parameter model we trained The Cannon on 30 M dwarfs with a two-dimensional with temperature/metallicity labels, and was able to achieve precisions of 44K/0.05dex respectively, more precise than both of the original training label uncertainties of 60K/0.08dex.
+
+\item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure 2 displays
+\end{enumerate}
 

### 2018-03-27 paper/draft1/draft1.tex
update data paragraph

--- 2018-03-26 paper/draft1/draft1.tex
@@
 \usepackage{enumitem}
+\usepackage{graphicx}
 
@@
 % \end{enumerate}
-Very low mass (VLM) are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass 
- are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
+Very low mass (VLM) are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
 
@@
 The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey IV (SDSS-IV) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-While APOGEE is primarily targeted for bright stellar populations (descirbe cuts \citealt{Zasowski:2013}) with the objective of studying galactic composition and evolution, numerous cool, main sequence have also been observed either as targets proposed by the M dwarf ancilliary survey \citep{Desphande:2013}
+Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [α/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
 
-Fundamental parameters for each of these stars (including temperature, surface gravity, metallicity and 15 chemical abundances) are estimated by the APOGEE Stellar Parameter
-and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016})
+APOGEE is primarily targeted for bright stellar populations, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution.
+However numerous cool, main sequence sources have also been observed either accidentally, or as targets proposed by the M dwarf ancilliary survey \citep{Desphande:2013} (specific estimate based on Gaia?).
 
 \subsection{Training Samples}
-\begin{enumerate}
-\item[-] Describe the spectral-type training data \citep{West:2011}. One paragraph plus citations. Include any cuts we apply. \color{red}{Make diagnostic plots.}\color{black}
 
-\item[-] Describe the Teff, Fe/H training data \citep{Mann:2015}. Same; Include any cuts. \color{red}{Make diagnostic plots.}\color{black}
-\end{enumerate}
+The Cannon is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:cannon}).
+The model in this sense is used to \emph{transfer} labels from spectra which we know parameters for to those which we do not.
+Hence this data-driven approach effectively removes the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (ideally very accurately measured) \emph{reference labels} that can be measured from other surveys/methods that are easier to infer from.
+
+For the purpose of this study we have constructed two different training samples: first a one-dimensional \emph{spectral type model}, and second a two-dimensional \emph{physical parameter model} which describes the temperature and metallicity.
+The choice of training labels, dimensionality of our data set, and requirements for a good training set we discuss further in Section \ref{sec:discussion}.
+
+The spectral type sample consists of 51 sources, spanning M0-M9 (Figure \ref{fig:train_dist}) cross-matched from the \citealt{West:2011} catalog of 78,841 M dwarfs from SDSS.
+For each source in the catalog, spectral types were determined both through an automated routine using The Hammer \citep{Covey:2007} and by visual inspection.
+
+The physical parameter sample consists of 30 sources distributed over $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 (Figure \ref{fig:train_dist}) drawn from \citealt{Mann:2015}.
+\color{red}{why logg not included as atmospheric parameter?}\color{black}
+
+\begin{figure}[ht]
+\plottwo{figures/west_train_dist.png}{figures/mann_train_dist.png}
+\caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest} \label{fig:train_dist}
+\end{figure}
+
+\subsection{Data Processing}
+Describe ASPCAP spectra files and Ness continuum normalization procedure.
 
@@
 \end{enumerate}
+
+Constructing such a model requires two steps: first the \emph{training step} in which the generative model is constructed at each pixel from the set of reference labels and fluxes, and second the \emph{test step} in which the model is applied to determine the labels of a spectrum
+
+% The key advantages of The Cannon being:
+% Data-driven models take away the challenge of directly infering labels from a survey, instead we \emph{transfer labels} from another survery that is more accurate or easier to model.
 

### 2018-03-30 paper/draft1/draft1 (copy).tex
new figures

--- 2018-03-27 paper/draft1/draft1.tex
@@
 
- (1) model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\color{red}{Need citations...
-}\color{black} look at Burgasser, Blake...) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\color{red}{Need citations...
-}\color{black} \citealt{Allard:2013});
+ (1) model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\citealt{Rajpurohit:2014}, \citealt{Rajpurohit:2018}) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\citealt{Allard:2013});
 
@@
 The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey IV (SDSS-IV) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [α/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
+Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
 
@@
 
-The spectral type sample consists of 51 sources, spanning M0-M9 (Figure \ref{fig:train_dist}) cross-matched from the \citealt{West:2011} catalog of 78,841 M dwarfs from SDSS.
-For each source in the catalog, spectral types were determined both through an automated routine using The Hammer \citep{Covey:2007} and by visual inspection.
+The spectral type sample consists of 51 sources, spanning M0-M9 (Figure \ref{fig:train_dist}) cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from SDSS.
+For each source in the catalog, spectral types were determined both through an automated routine using The Hammer \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
 
-The physical parameter sample consists of 30 sources distributed over $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 (Figure \ref{fig:train_dist}) drawn from \citealt{Mann:2015}.
+The physical parameter sample consists of 30 sources with reference labels distributed over $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 (Figure \ref{fig:train_dist}) drawn from \citealt{Mann:2015} (hereafter M15).
+The M15 catalog in total contains 183 sources selected from CONCH-SHELL \citep{Gaidos:2013} and \citealt{Lepine:2011} \color{red}{these have optical specs?}\color{black} with infrared spectra taken in follow up from the SpeX Spectograph.
+Effective temperatures are measured by comparison of BT-SETTL atmospheric models \citep{Allard:2011} to optical spectra.
+A subsample of 29 sources with measured angular diameters and bolometric fluxes
+
+Spectral types reported by M15 for the sources using optical molecular band indices (by the method of \citealt{Lepine:2013}) and ranged K7.3--M4.9 for the subsample we used, although these labels were not included in the model training.
+
 \color{red}{why logg not included as atmospheric parameter?}\color{black}
@@
 \subsection{Data Processing}
+
+APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar.
+
 Describe ASPCAP spectra files and Ness continuum normalization procedure.

### 2018-03-30 paper/draft1/draft1.tex
acknowlegments

@@
 \begin{enumerate}
-\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources; \citealt{Deacon:2014} -- \color{red}{Need citations!}\color{black})
+\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources...)
 
@@
 
-The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey IV (SDSS-IV) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [α/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
+The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III (SDSS-III; \citealt{Eisenstein:2011}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
+Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
 
@@
 
-The spectral type sample consists of 51 sources, spanning M0-M9 (Figure \ref{fig:train_dist}) cross-matched from the \citealt{West:2011} catalog of 78,841 M dwarfs from SDSS.
-For each source in the catalog, spectral types were determined both through an automated routine using The Hammer \citep{Covey:2007} and by visual inspection.
+The spectral type sample consists of 51 sources, spanning M0-M9 (Figure \ref{fig:train_dist}) cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from SDSS.
+For each source in the catalog, spectral types were determined both through an automated routine using The Hammer \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
 
-The physical parameter sample consists of 30 sources distributed over $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 (Figure \ref{fig:train_dist}) drawn from \citealt{Mann:2015}.
+The physical parameter sample consists of 30 sources with reference labels distributed over $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 (Figure \ref{fig:train_dist}) drawn from \citealt{Mann:2015} (hereafter M15).
+The M15 catalog in total contains 183 sources selected from CONCH-SHELL \citep{Gaidos:2013} and with infrared spectra taken in follow up from the SpeX Spectograph.
+Effective temperatures are measured by comparison of BT-SETTL atmospheric models \citep{Allard:2011} to optical spectra.
+A subsample of 29 sources with measured angular diameters and bolometric fluxes from long-baseline optical interferometry are used to callibrate effective temperature measurements.
+Relative uncertainty from the callibration sample is measured to be \color{red}{$\sim$30K (check this)}\color{black}, which added in quadrature to the scatter from model comparison gives a total reported uncertainty of $\sim$60K in T$_{\rm eff}$.
+
+Spectral types reported by M15 for the sources using optical molecular band indices (by the method of \citealt{Lepine:2013}) and ranged K7.3--M4.9 for the subsample we used, although these labels were not included in the model training.
+
 \color{red}{why logg not included as atmospheric parameter?}\color{black}
@@
 \subsection{Data Processing}
+
+APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample.
+
 Describe ASPCAP spectra files and Ness continuum normalization procedure.
@@
 \item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure 2 displays
+
+\item[-] Vsini and logg were not included as training parameters.
+Vsini is less import of a parameter as the sample consists of mainly old, slowly rotating sources (check what are ages?).
+Logg because?
 \end{enumerate}
@@
 %%==========================================================================================
+\acknowledgements
+We would like to acknowledge Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Hans Walter Rix (MPIA), Anna-Christina Eilers (MPIA), Anna Ho (Caltech), and Jackie Faherty (AMNH) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
+Software: The Cannon, numpy, matplotlib...
+
+\color{red}{HOGG: SDSS acknowledgments... Grants and stuff...}\color{black}
+
+%%==========================================================================================
 \clearpage

### 2018-03-30 paper/draft1/draft1.tex
add affil

@@
 
-\shorttitle{Birky \& Hogg}
-\shortauthors{Birky \& Hogg}
+\shorttitle{Birky et al.}
+\shortauthors{Birky et al.}
 
@@
 \affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
+
+\author[0000-0003-3654-1602]{Andrew Mann}
+\affil{Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY 10027, USA}
 
@@
 \begin{enumerate}
-\item[-] We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and was able to achieve a precision of $\pm$0.8 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
-\item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure 2 displays
+\item[-] We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.8 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
+\item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure  displays
 \end{enumerate}

### 2018-04-12 paper/draft1/draft1.tex
add cannon eqns

@@
 \usepackage{graphicx}
+\usepackage{mathrsfs}
+\usepackage[mathscr]{euscript}
 
 \definecolor{bcolor}{RGB}{0, 51, 153}
+\definecolor{gcolor}{RGB}{51, 153, 51}
 \hypersetup{linkcolor=bcolor,citecolor=bcolor,filecolor=cyan,urlcolor=bcolor}
@@
 \affil{Center for Astrophysics and Space Science, University of California San Diego, La Jolla, CA 92093, USA}
-\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}
+\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
 
 \author[0000-0003-2866-9403]{David W. Hogg}
-\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}
+\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
 \affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
@@
 
+\author[0000-0002-6523-9536]{Adam Burgasser}
+\affil{Center for Astrophysics and Space Science, University of California San Diego, La Jolla, CA 92093, USA}
+
 \begin{abstract}
@@
 We then use 30 M dwarfs ranging $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 44K and 0.05 dex respectively.
-Additionally we discuss the extension of a model to other labels, and the scientific objectives a data-driven pipeline could enable.
+Finally we demonstrate that the two-parameter Cannon model replicates the spectral features (\color{gcolor}{HOGG:}\color{black} report some pixel accracy?)
+
+to theoretical models of M dwarf atmospheres (PHOENIX, \citealt{Husser:2013}; BTSETTL, \citealt{Allard:2011}) and identify features of missing line strengths.
 
@@
 % \end{enumerate}
-Very low mass (VLM) are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
+Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
 
@@
 Relative uncertainty from the callibration sample is measured to be \color{red}{$\sim$30K (check this)}\color{black}, which added in quadrature to the scatter from model comparison gives a total reported uncertainty of $\sim$60K in T$_{\rm eff}$.
+Metallicities are measured
 
@@
 
-\color{red}{why logg not included as atmospheric parameter?}\color{black}
+We note that surface gravity (logg) and rotational velocity (vsini) parameters are not included as training labels because...
 
@@
 
-APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample.
+APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample \citealt{Perez:2016}.
 
@@
 \item[-] A few-paragraph summary of what The Cannon does, with references (\citealt{Ness:2015}, \citealt{Casey:2016}, \citealt{Ho:2017a}, \citealt{Ho:2017b})
+\item[-] Describe assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label. 
+\item[-] Briefly describe modeling procedure (training step, test step, quadratic labels, etc.), refer to \citealt{Ness:2015} for full description.
+\item[-] Briefly describe model inputs and outputs (1D spectral type model and 2D teff/metallicity and label vectors for those). Mention other models we've tested.
 \end{enumerate}
@@
 
-\subsection{Data-Driven Model}
-\begin{enumerate}
-\item[-] Describe assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label. 
-\item[-] Briefly describe modeling procedure (training step, test step, quadratic labels, etc.), refer to \citealt{Ness:2015} for full description.
-\item[-] Briefly describe model inputs and outputs (1D spectral type model and 2D teff/metallicity and label vectors for those). Mention other models we've tested.
-\end{enumerate}
+Following from \citealt{Ness:2015} and \citealt{Ho:2017a} 
+\begin{equation}
+	f_{n\lambda} = g(\ell_n | \theta_{\lambda}) + {\rm noise}
+\end{equation}
+
+\begin{equation}
+	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + [s_{\lambda}^{2} + \sigma_{n\lambda}^{2}] \,\xi_{n\lambda}
+\end{equation}
+
+Quadratic label vectors:
+\begin{equation}
+	\ell_{n} = [1, SPT, SPT^{2}]
+\end{equation}
+
+\begin{equation}
+	\ell_{n} = [1, T_{\rm eff}, [Fe/H], T_{\rm eff} \cdot [Fe/H], [Fe/H]^{2}]
+\end{equation}
 
@@
 \plotone{figures/3_physical_model.png}
-\caption{\textit{Top two panels:} Mann-trained model for varying temperatures; \textit{bottom two panels:} Mann-trained model for varying metallicities.} \label{fig:mann_demo}
+\caption{\textit{Top two panels:} Mann-trained model for varying temperatures; \textit{bottom two panels:} Mann-trained model for varying metallicities.
+\color{red}{Things to add: line list and model - data difference}\color{black}} \label{fig:mann_demo}
 \end{figure}
+
+\begin{enumerate}
+\item[-] Describe derivative plots and jacknife error estimation.
+\end{enumerate}
 
@@
 
-\item[-] Some discussion of precision and accuracy: The Cannon can produce precise results, but is only as accurate as its training set.
+\item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: The Cannon can produce precise results, but is only as accurate as its training set.}\color{black}
 
 \item[-] Some discussion of whether the derivatives are sensible or surprising. \color{red}{Add derivative plots.}\color{black}
-
-\item[-] Comment on specific features?
-\color{red}{Can easily plot some line lists, but not sure where to go with the analysis...
-What features are important for M dwarfs?
-Compare derivative plots of specific lines between theory-trained models and data-trained models?}\color{black}
 
@@
 We would like to acknowledge Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Hans Walter Rix (MPIA), Anna-Christina Eilers (MPIA), Anna Ho (Caltech), and Jackie Faherty (AMNH) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
-Software: The Cannon, numpy, matplotlib...
+\color{red}Software: The Cannon, numpy, matplotlib...
+\color{black}
 
-\color{red}{HOGG: SDSS acknowledgments... Grants and stuff...}\color{black}
+\color{gcolor}{HOGG: SDSS acknowledgments... Grants and stuff...}\color{black}
 

### 2018-04-20 paper/draft1/draft1.tex
update

@@
 \usepackage{graphicx}
-\usepackage{mathrsfs}
-\usepackage[mathscr]{euscript}
+\usepackage{mathrsfs}r
 
@@
 We then use 30 M dwarfs ranging $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 44K and 0.05 dex respectively.
-Finally we demonstrate that the two-parameter Cannon model replicates the spectral features (\color{gcolor}{HOGG:}\color{black} report some pixel accracy?)
-
-to theoretical models of M dwarf atmospheres (PHOENIX, \citealt{Husser:2013}; BTSETTL, \citealt{Allard:2011}) and identify features of missing line strengths.
+Finally we compare our temperature/metallicty model to theoretical models of M dwarf atmospheres (PHOENIX, \citealt{Husser:2013}; BTSETTL, \citealt{Allard:2011}) and identify features of missing line strengths.
 

### 2018-04-25 paper/draft1/draft1.tex
back up all files

@@
 \usepackage{graphicx}
-\usepackage{mathrsfs}r
+\usepackage{mathrsfs}
 
@@
 % \begin{enumerate}
-% \item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (The CARMENES search for exoplanets around M dwarfs \citealt{Trifonov:2018}; some famous examples: Trappist 1 (M8) -- \citealt{Gillon:2017}; Kepler-186, Kepler-138...
-other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
-% \end{enumerate}
 Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
+
+%define 'labels' - could be physical parameters or more empirical
 
@@
 \begin{enumerate}
-\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources...)
+\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources...
+Terrien, Mann, Newton, Dressing catalogs)
 
-\item[-] Why is there a problem labeling M dwarfs? Generally, two kinds of approaches to labeling:
+\item[-] Why is there a problem labeling M dwarfs? Generally, two kinds of approaches to determing physical parameters:
 
- (1) model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\citealt{Rajpurohit:2014}, \citealt{Rajpurohit:2018}) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\citealt{Allard:2013});
+ (1) model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\citealt{Rajpurohit:2014}, \citealt{Rajpurohit:2018}) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\citealt{Allard:2013})--models don't accurately reproduce spectra plus unquantified systematic errors;
 
  (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (\citealt{Casagrande:2008} estimated effective temperatures, bolometric luminosities, and metallicities using optical and infrared photometries) -- however we want to use \emph{all} of the data!
-
-% \item[-] The idea in general of The Cannon, and the fact that it could be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects.
-Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all!
 
@@
 \end{enumerate}
+
+%using empirical classifications/templates/standards - SPT, luminosity class - this is a relative system but does not provide an absolute standard for measuring physical parameters
+
+%brief importance statment about APOGEE - new instrumentation has allowed for new opportunity - high resolution and large datasets of Mdwarfs are now available
+
+%brief importance statment cannon - better than template matching in that it is using less sparsely sampled known mdwarfs
+
+%ISSUES: models dont accurately reproduce data; empirical standards are relative not absolute
+
+%photometric data, HR digram, comparison to evolutionary models
 

### 2018-07-02 paper/draft1/draft1.tex
update figures

@@
 
-\subsection{M dwarf Challenge}
 \begin{enumerate}
+
+\item[-] Define what parameters/classifications are important for Mdwarfs.
+We want to estimate physical parameters (such as Teff, logg, metallicity) and empirical parameters/classifications (SPT, etc.), which we will refer to as \emph{stellar labels}.
+
 \item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources...
@@
 
-\item[-] Why is there a problem labeling M dwarfs? Generally, two kinds of approaches to determing physical parameters:
+\item[-] brief importance statment about APOGEE - new instrumentation has allowed for new opportunity - high resolution and large datasets of Mdwarfs are now available
+
+\item[-] Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
+\citealt{Gilhool:2018}; \citealt{Skinner:2018}.
+
+\item[-] Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: models dont accurately reproduce data; empirical standards are relative not absolute.
 
@@
 
-\item[-] Why The Cannon approach?
-Can be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects; Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all!
+ (3) using empirical classifications/templates/standards - SPT, luminosity class - this is a relative system but does not provide an absolute standard for measuring physical parameters
 
-\item[-] Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
+\item[-] Why The Cannon approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016})
+
+ (1) Can be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects;
+
+ (2) Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all
+
+ (3) better than template matching in that it is using less sparsely sampled known mdwarfs
 
@@
 
-%using empirical classifications/templates/standards - SPT, luminosity class - this is a relative system but does not provide an absolute standard for measuring physical parameters
-
-%brief importance statment about APOGEE - new instrumentation has allowed for new opportunity - high resolution and large datasets of Mdwarfs are now available
-
-%brief importance statment cannon - better than template matching in that it is using less sparsely sampled known mdwarfs
-
-%ISSUES: models dont accurately reproduce data; empirical standards are relative not absolute
-
 %photometric data, HR digram, comparison to evolutionary models
-
-A number of M dwarf studies
 
@@
 
-The spectral type sample consists of 51 sources, spanning M0-M9 (Figure \ref{fig:train_dist}) cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from SDSS.
+The spectral type sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from SDSS.
 For each source in the catalog, spectral types were determined both through an automated routine using The Hammer \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
 
-The physical parameter sample consists of 30 sources with reference labels distributed over $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 (Figure \ref{fig:train_dist}) drawn from \citealt{Mann:2015} (hereafter M15).
+The physical parameter sample consists of 30 sources with reference labels distributed over $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 drawn from \citealt{Mann:2015} (hereafter M15).
 The M15 catalog in total contains 183 sources selected from CONCH-SHELL \citep{Gaidos:2013} and with infrared spectra taken in follow up from the SpeX Spectograph.
@@
 Relative uncertainty from the callibration sample is measured to be \color{red}{$\sim$30K (check this)}\color{black}, which added in quadrature to the scatter from model comparison gives a total reported uncertainty of $\sim$60K in T$_{\rm eff}$.
-Metallicities are measured
 
@@
 
-\begin{figure}[ht]
-\plottwo{figures/west_train_dist.png}{figures/mann_train_dist.png}
-\caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest} \label{fig:train_dist}
-\end{figure}
+% \begin{figure}[ht]
+% \plottwo{figures/west_train_dist.png}{figures/mann_train_dist.png}
+% \caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest} \label{fig:train_dist}
+% \end{figure}
 
@@
 
-APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample \citealt{Perez:2016}.
+We use the aspcapStar spectra (see \citealt{Perez:2016} for a complete description of the pipeline).
+
+APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample .
 
@@
 \begin{enumerate}
-\item[-] A few-paragraph summary of what The Cannon does, with references (\citealt{Ness:2015}, \citealt{Casey:2016}, \citealt{Ho:2017a}, \citealt{Ho:2017b})
-\item[-] Describe assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label. 
+\item[-] Describe the assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label. 
 \item[-] Briefly describe modeling procedure (training step, test step, quadratic labels, etc.), refer to \citealt{Ness:2015} for full description.
@@
 
-Following from \citealt{Ness:2015} and \citealt{Ho:2017a} 
+Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a model that assumes the flux at each pixel of the spectrum is parameterized as a function \textit{g} 
 \begin{equation}
@@
 
-\begin{figure}[ht]
-\plotone{figures/west_spt_self_test.png}
-\caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest}
-\end{figure}
-
-\begin{figure}[ht]
-\plotone{figures/1_Spectral_Sequence.pdf}
-\plotone{figures/2_Spectral_Sequence.pdf}
-\plotone{figures/3_Spectral_Sequence.pdf}
-\caption{ Spectral sequence of dwarfs in training set M0-M9; chip 1 of APOGEE
-spectrum with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
-\end{figure}
-
 \begin{enumerate}
@@
 \end{enumerate}
-
-\subsection{Temperature/Metallicity Model}
-\begin{figure}[ht]
-\plottwo{figures/mann_teff_self_test.png}{figures/mann_feh_self_test.png}
-\caption{Temperature and Metallicity label self test.} \label{fig:mann_selftest}
-\end{figure}
-
-\begin{figure}[ht]
-\plotone{figures/1_physical_model.png}
-\plotone{figures/2_physical_model.png}
-\plotone{figures/3_physical_model.png}
-\caption{\textit{Top two panels:} Mann-trained model for varying temperatures; \textit{bottom two panels:} Mann-trained model for varying metallicities.
-\color{red}{Things to add: line list and model - data difference}\color{black}} \label{fig:mann_demo}
-\end{figure}
 
@@
 
-\begin{figure}[ht]
-\plotone{figures/west_spt_derivative.png}
-\caption{Derivative plots for spectral type model.} \label{fig:west_derivative}
-\end{figure}
-
-\begin{figure}[ht]
-\plotone{figures/mann_teff_derivative.png}
-\plotone{figures/mann_feh_derivative.png}
-\caption{Derivative plots for physical parameter model.
-\textit{Top:} Derivative of flux with repsect to temperature.
-\textit{Bottom:} Derivative of flux with repsect to metallicity.} \label{fig:mann_derivative}
-\end{figure}
-
 %%==========================================================================================
@@
 %%==========================================================================================
+%% Figures
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=16cm]{figures/west_self_test_spt.png}
+\end{center}
+\caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest}
+\end{figure}
+
+\subsection{Temperature/Metallicity Model}
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=16cm]{figures/self_and_validation_test_teff.png}
+\includegraphics[width=16cm]{figures/self_and_validation_test_fe_h.png}
+\end{center}
+\caption{Temperature and Metallicity label self test.} \label{fig:mann_selftest}
+\end{figure}
+
+\begin{figure}[ht]
+\plotone{figures/1_Spectral_Sequence.pdf}
+\plotone{figures/2_Spectral_Sequence.pdf}
+\plotone{figures/3_Spectral_Sequence.pdf}
+\caption{ Spectral sequence of dwarfs in training set M0-M9; chip 1 of APOGEE
+spectrum with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=16cm]{figures/demo_derivatives_teff1.png}
+\includegraphics[width=16cm]{figures/demo_derivatives_teff2.png}
+\includegraphics[width=16cm]{figures/demo_derivatives_teff3.png}
+\end{center}
+\caption{\textit{Top two panels:} Mann-trained model for varying temperatures.} \label{fig:demo_teff}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=16cm]{figures/demo_derivatives_feh1.png}
+\includegraphics[width=16cm]{figures/demo_derivatives_feh2.png}
+\includegraphics[width=16cm]{figures/demo_derivatives_feh3.png}
+\end{center}
+\caption{\textit{Top two panels:} Mann-trained model for varying metallicities.} \label{fig:demo_feh}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=12cm]{figures/west_spt_derivative.png}
+\end{center}
+\caption{Derivative plots for spectral type model.} \label{fig:west_derivative}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=15cm]{figures/derivative_jackknife_teff.png}
+\includegraphics[width=15cm]{figures/derivative_jackknife_feh.png}
+\end{center}
+\caption{Derivative plots for physical parameter model.
+\textit{Top:} Derivative of flux with repsect to temperature.
+\textit{Bottom:} Derivative of flux with repsect to metallicity.} \label{fig:mann_derivative}
+\end{figure}
+
+%%==========================================================================================
 \clearpage

### 2018-07-03 paper/draft1/draft1.tex
add mann result table

@@
 \usepackage{mathrsfs}
+\usepackage{booktabs}
 
@@
 
+\newpage
+
 \begin{figure}[ht]
 \begin{center}
-\includegraphics[width=16cm]{figures/west_self_test_spt.png}
+\includegraphics[width=18cm]{figures/self_and_validation_test_spt.png}
 \end{center}
@@
 
-\subsection{Temperature/Metallicity Model}
 \begin{figure}[ht]
 \begin{center}
-\includegraphics[width=16cm]{figures/self_and_validation_test_teff.png}
-\includegraphics[width=16cm]{figures/self_and_validation_test_fe_h.png}
+\includegraphics[width=18cm]{figures/self_and_validation_test_teff.png}
+\includegraphics[width=18cm]{figures/self_and_validation_test_fe_h.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=12cm]{figures/west_spt_derivative.png}
+\includegraphics[width=16cm]{figures/derivative_jackknife_spt.png}
 \end{center}
-\caption{Derivative plots for spectral type model.} \label{fig:west_derivative}
+\caption{Derivative plots for spectral type model taken at the median training spectral type, SPT$=3$.} \label{fig:west_derivative}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=15cm]{figures/derivative_jackknife_teff.png}
-\includegraphics[width=15cm]{figures/derivative_jackknife_feh.png}
+\includegraphics[width=16cm]{figures/derivative_jackknife_teff.png}
+\includegraphics[width=16cm]{figures/derivative_jackknife_feh.png}
 \end{center}
 \caption{Derivative plots for physical parameter model.
-\textit{Top:} Derivative of flux with repsect to temperature.
-\textit{Bottom:} Derivative of flux with repsect to metallicity.} \label{fig:mann_derivative}
+\textit{Top:} Derivative of flux with repsect to temperature taken at the median training temperature, T$_{\rm eff}=3463K$.
+\textit{Bottom:} Derivative of flux with repsect to metallicity taken at the median training metallicity, [Fe/H]$=-0.03$ dex.} \label{fig:mann_derivative}
 \end{figure}
+
+%%==========================================================================================
+%% Tables
+
+\newpage
+
+% \begin{longrotatetable}
+\startlongtable
+\begin{deluxetable*}{ccc|ccc|ccc}
+\tablecaption{Cannon results for Temperature/Metallicity model \label{mann_results}}
+\tabletypesize{\scriptsize}
+\tablehead{
+% \colhead{Designation} & \colhead{RA} & \colhead{DEC} & \colhead{$T_{\rm eff}$} & \colhead{[Fe/H]} \\
+% \colhead{(2MASS)} & \colhead{(deg)} & \colhead{(deg)} & \colhead{(K)} & \colhead{(dex)}
+\multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Temperature (K)}} & \multicolumn{3}{c}{\underline{Metallicity (dex)}} \\
+	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
+    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}
+    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}  
+} 
+\startdata
+2M00182256+4401222 & 4.59542   & 44.02278  & 3603        & 3532       & 3519        & -0.3       & -0.27     & -0.24      \\
+2M00182549+4401376 & 4.60779   & 44.02734  & 3218        & 3527       & 3554        & -0.3       & -0.28     & -0.27      \\
+2M00285391+5022330 & 7.22488   & 50.37588  & 3207        & 3185       & 3188        & 0.11       & 0.05      & 0.02       \\
+2M00401001+0308050 & 10.04169  & 3.13473   & 3725        & 3783       & 3778        & 0.04       & 0.11      & 0.13       \\
+2M00580115+3919111 & 14.50482  & 39.31977  & 3157        & 3094       & 3101        & -0.07      & -0.01     & 0.03       \\
+2M01232542+1638384 & 20.8559   & 16.64401  & 3272        & 3218       & 3217        & 0.1        & 0.03      & 0.0        \\
+2M02001278+1303112 & 30.05402  & 13.05196  & 3080        & 3057       & 3074        & -0.16      & -0.1      & -0.01      \\
+2M02361535+0652191 & 39.06358  & 6.87167   & 3284        & 3249       & 3253        & -0.12      & -0.29     & -0.29      \\
+2M03044335+6144097 & 46.18104  & 61.73583  & 3500        & 3468       & 3467        & -0.12      & -0.26     & -0.27      \\
+2M03553688+5214291 & 58.90373  & 52.2414   & 3435        & 3381       & 3369        & -0.35      & -0.25     & -0.18      \\
+2M04125880+5236421 & 63.24499  & 52.61165  & 3100        & 3181       & 3200        & -0.04      & -0.01     & 0.01       \\
+2M04310001+3647548 & 67.75003  & 36.79855  & 3419        & 3371       & 3369        & 0.08       & 0.11      & 0.1        \\
+2M05030563+2122362 & 75.77354  & 21.37664  & 3155        & 3290       & 3503        & 0.06       & -0.15     & -0.36      \\
+2M05222053+3031097 & 80.58561  & 30.51933  & 3389        & 3417       & 3426        & 0.28       & 0.2       & 0.14       \\
+2M05312734-0340356 & 82.86417  & -3.67722  & 3801        & 3820       & 3854        & 0.49       & 0.49      & 0.37       \\
+2M05413073+5329239 & 85.37804  & 53.48987  & 3765        & 3752       & 3755        & 0.19       & 0.15      & 0.13       \\
+2M05420897+1229252 & 85.53833  & 12.48956  & 3250        & 3226       & 3235        & -0.22      & -0.34     & -0.31      \\
+2M06000351+0242236 & 90.01458  & 2.70657   & 3214        & 3164       & 3165        & 0.07       & 0.05      & 0.03       \\
+2M06011106+5935508 & 90.2961   & 59.59713  & 3340        & 3258       & 3255        & -0.09      & -0.05     & -0.03      \\
+2M06112610+1032599 & 92.8588   & 10.54998  & 3636        & 3711       & 3698        & -0.39      & -0.42     & -0.39      \\
+2M06544902+3316058 & 103.70411 & 33.26823  & 3448        & 3364       & 3359        & -0.02      & 0.04      & 0.06       \\
+2M07171706-0501031 & 109.32108 & -5.01754  & 3193        & 3159       & 3168        & -0.09      & -0.12     & -0.09      \\
+2M07272450+0513329 & 111.852   & 5.2263    & 3317        & 3285       & 3285        & -0.11      & -0.2      & -0.19      \\
+2M07444018+0333089 & 116.16744 & 3.55252   & 3217        & 3176       & 3174        & 0.23       & 0.26      & 0.15       \\
+2M08031949+5250387 & 120.83131 & 52.84402  & 3508        & 3622       & 3626        & -0.26      & -0.24     & -0.24      \\
+2M08083284+5304377 & 122.13679 & 53.07712  & 3398        & 3280       & 3273        & 0.0        & 0.08      & 0.09       \\
+2M08103429-1348514 & 122.64277 & -13.81427 & 3544        & 3517       & 3515        & 0.0        & 0.0       & 0.0        \\
+2M08155393+3136392 & 123.97459 & 31.61077  & 3753        & 3810       & 3821        & 0.16       & 0.07      & 0.03       \\
+2M08323599+4510175 & 128.14998 & 45.17153  & 3494        & 3505       & 3506        & -0.14      & -0.21     & -0.21      \\
+2M08370799+1507475 & 129.28319 & 15.12944  & 3596        & 3553       & 3552        & -0.17      & -0.26     & -0.27      \\
+2M08524084+2818589 & 133.17018 & 28.31634  & 3166        & 3213       & 3239        & 0.31       & 0.5       & 0.27       \\
+2M08585633+0828259 & 134.73479 & 8.47384   & 3200        & 3142       & 3147        & -0.04      & -0.09     & -0.08      \\
+2M09142298+5241125 & 138.595   & 52.68639  & 3920        & 3900       & 3879        & -0.01      & -0.02     & 0.01       \\
+2M09174473+4612246 & 139.4363  & 46.20685  & 3478        & 3548       & 3556        & 0.04       & -0.02     & -0.03      \\
+2M09214911+4330284 & 140.45447 & 43.50784  & 3170        & 3196       & 3206        & 0.16       & 0.23      & 0.19       \\
+2M09285333-0722148 & 142.22221 & -7.37095  & 3508        & 3445       & 3440        & -0.27      & -0.23     & -0.22      \\
+2M10112218+4927153 & 152.84254 & 49.45431  & 4131        & 4007       & 3946        & 0.24       & 0.22      & 0.23       \\
+2M10121768-0344441 & 153.07364 & -3.74563  & 3623        & 3619       & 3620        & 0.13       & 0.16      & 0.16       \\
+2M10193634+1952122 & 154.90117 & 19.87002  & 3370        & 3384       & 3389        & 0.15       & 0.24      & 0.21       \\
+2M10285555+0050275 & 157.23155 & 0.8411    & 3548        & 3517       & 3515        & -0.18      & -0.16     & -0.15      \\
+2M10331367+3409120 & 158.30687 & 34.15333  & 3500        & 3518       & 3518        & -0.15      & -0.19     & -0.2       \\
+2M10350859+3349499 & 158.7858  & 33.83055  & 3816        & 3906       & 3907        & -0.04      & 0.01      & 0.03       \\
+2M10453795+1833111 & 161.4081  & 18.55306  & 3673        & 3773       & 3780        & 0.14       & 0.13      & 0.13       \\
+2M10505201+0648292 & 162.71704 & 6.80826   & 3238        & 3258       & 3265        & 0.16       & 0.18      & 0.12       \\
+2M10520440+1359509 & 163.01762 & 13.99756  & 3298        & 3230       & 3228        & -0.13      & -0.11     & -0.07      \\
+2M11000432+2249592 & 165.01781 & 22.833    & 3463        & 3426       & 3423        & -0.11      & -0.1      & -0.09      \\
+2M11032023+3558117 & 165.83423 & 35.97054  & 3563        & 3682       & 3551        & -0.38      & -0.58     & -0.39      \\
+2M11052903+4331357 & 166.36992 & 43.52665  & 3619        & 3547       & 3525        & -0.37      & -0.34     & -0.3       \\
+2M11054316+1014093 & 166.42966 & 10.23605  & 3439        & 3395       & 3393        & -0.03      & -0.06     & -0.07      \\
+2M11091225-0436249 & 167.30106 & -4.60695  & 3661        & 3743       & 3748        & -0.03      & -0.09     & -0.1       \\
+2M11263270+2145262 & 171.63622 & 21.75719  & 3610        & 3642       & 3642        & -0.18      & -0.16     & -0.15      \\
+2M11273856+0358359 & 171.91071 & 3.97664   & 3920        & 3932       & 3963        & 0.43       & 0.44      & 0.31       \\
+2M11474074+0015201 & 176.9198  & 0.25553   & 3261        & 3235       & 3236        & 0.1        & 0.23      & 0.22       \\
+2M11474440+0048164 & 176.93491 & 0.80473   & 3192        & 3155       & 3157        & -0.02      & -0.06     & -0.06      \\
+2M11530522+1855480 & 178.27184 & 18.93002  & 3773        & 3760       & 3752        & -0.14      & -0.08     & -0.06      \\
+2M12192028+1323524 & 184.83449 & 13.39793  & 3926        & 3954       & 3945        & 0.0        & 0.13      & 0.18       \\
+2M12274471-0315006 & 186.93625 & -3.25018  & 3355        & 3399       & 3407        & 0.19       & 0.2       & 0.17       \\
+2M12390461+4702234 & 189.76958 & 47.03974  & 3492        & 3447       & 3445        & 0.05       & 0.06      & 0.06       \\
+2M12470102+4637334 & 191.75412 & 46.62587  & 3571        & 3543       & 3544        & 0.22       & 0.23      & 0.22       \\
+2M13113464+1306461 & 197.89433 & 13.11282  & 3459        & 3407       & 3402        & -0.1       & -0.03     & -0.01      \\
+2M13160127+1415504 & 199.00528 & 14.26403  & 3686        & 3768       & 3775        & -0.06      & -0.16     & -0.17      \\
+2M13295979+1022376 & 202.49895 & 10.37731  & 3727        & 3674       & 3669        & -0.09      & -0.13     & -0.13      \\
+2M13345147+3746195 & 203.71442 & 37.7721   & 3273        & 3247       & 3250        & 0.14       & 0.13      & 0.08       \\
+2M13400879+4346380 & 205.03676 & 43.77729  & 3332        & 3322       & 3325        & -0.19      & -0.18     & -0.14      \\
+2M13454354+1453317 & 206.43251 & 14.89139  & 3649        & 3688       & 3677        & -0.31      & -0.37     & -0.36      \\
+2M15290936+5724418 & 232.28886 & 57.41167  & 3660        & 3719       & 3723        & -0.09      & -0.14     & -0.14      \\
+2M15594729+4403595 & 239.94707 & 44.06654  & 3526        & 3614       & 3632        & 0.05       & 0.02      & 0.01       \\
+2M16252459+5418148 & 246.35249 & 54.30413  & 3475        & 3407       & 3400        & -0.35      & -0.25     & -0.22      \\
+2M16541202+1154529 & 253.5502  & 11.91464  & 3834        & 3760       & 3697        & -0.48      & -0.38     & -0.29      \\
+2M17093153+4340531 & 257.38138 & 43.68139  & 3284        & 3242       & 3242        & 0.2        & 0.25      & 0.17       \\
+2M17362594+6820220 & 264.10803 & 68.33932  & 3439        & 3487       & 3492        & 0.05       & -0.21     & -0.23      \\
+2M18161819+0131277 & 274.07575 & 1.5243    & 3391        & 3350       & 3347        & -0.25      & -0.19     & -0.17      \\
+2M18244689-0620311 & 276.1954  & -6.342    & 3412        & 3386       & 3384        & 0.01       & 0.07      & 0.08       \\
+2M18424666+5937499 & 280.69489 & 59.63015  & 3441        & 3386       & 3385        & -0.23      & -0.32     & -0.32      \\
+2M18424688+5937374 & 280.6958  & 59.62659  & 3345        & 3391       & 3400        & -0.3       & -0.35     & -0.33      \\
+2M18451027+0620158 & 281.29282 & 6.33774   & 3674        & 3684       & 3688        & 0.02       & 0.03      & 0.03       \\
+2M19142541+1755081 & 288.6059  & 17.91894  & 3725        & 3830       & 3835        & -0.08      & -0.09     & -0.09      \\
+2M19165526+0510086 & 289.23032 & 5.16909   & 3558        & 3536       & 3538        & 0.1        & 0.08      & 0.07       \\
+2M19500252+3235012 & 297.51022 & 32.58347  & 3385        & 3214       & 3215        & -0.18      & -0.31     & -0.28      \\
+2M19535443+4424541 & 298.47977 & 44.41508  & 2859        & 2801       & 2889        & -0.05      & 0.08      & 0.19       \\
+2M20032651+2952000 & 300.86075 & 29.86652  & 3144        & 3164       & 3178        & 0.06       & -0.03     & -0.09      \\
+2M20531977+6209156 & 313.33246 & 62.1545   & 3791        & 3728       & 3712        & -0.06      & 0.04      & 0.08       \\
+2M21065473+3844265 & 316.73013 & 38.74202  & 4021        & 4004       & 3922        & -0.22      & -0.23     & -0.11      \\
+2M22464980+4420030 & 341.70737 & 44.33405  & 3298        & 3236       & 3235        & 0.03       & 0.03      & 0.02       \\
+2M22563497+1633130 & 344.14517 & 16.55347  & 3720        & 3689       & 3688        & 0.21       & 0.29      & 0.28       \\
+2M23060482+6355339 & 346.52011 & 63.92622  & 3700        & 3770       & 3780        & 0.19       & 0.14      & 0.11       \\
+2M23315208+1956142 & 352.96732 & 19.93727  & 3353        & 3278       & 3274        & 0.03       & 0.05      & 0.05       \\
+2M23315244+1956138 & 352.9689  & 19.93719  & 3072        & 3266       & 3296        & 0.03       & 0.04      & 0.03       \\
+2M23491255+0224037 & 357.30206 & 2.40136   & 3646        & 3604       & 3560        & -0.45      & -0.46     & -0.4    
+\enddata
+\end{deluxetable*}
+% \end{longrotatetable}
 

### 2018-07-03 paper/draft1/create_tables.py

@@ -1,0 +1,27 @@
+import pandas as pd
+import numpy as np
+import os
+
+if __name__ == 'main':
+
+	file = 'data_files/mann_results.csv'
+	cont = pd.read_csv(file)
+
+	des = cont['ID']
+	ra  = cont['RA']
+	dec = cont['DEC']
+	trn_teff = cont['TRAIN_TEFF']
+	tst_teff = cont['TEST_TEFF']
+	crv_teff = cont['CROSS_TEFF']
+	trn_feh = cont['TRAIN_FEH']
+	tst_feh = cont['TEST_FEH']
+	crv_feh = cont['CROSS_FEH']
+		
+	with open('tables/mann_results.txt', 'w') as wf:
+		for i in range(len(des)):
+			line = "{} & {} & {} & {} & {} & {} & {} & {} & {} \\".format(des[i], ra[i], dec[i], \
+				trn_teff[i], tst_teff[i], crv_teff[i], \
+				trn_feh[i], tst_feh[i], crv_feh[i])
+
+			print(line)
+			wf.write(line)

### 2018-07-12 paper/draft1/draft1.tex
backup

@@
 \usepackage{booktabs}
+\usepackage{morefloats}
 
@@
 The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars.
-Using a training set of 51 M dwarfs with spectral types ranging M0-M9 obtained from SDSS optical spectra, we demonstrate that The Cannon can infer spectral types to a precision of 0.6 types.
-We then use 30 M dwarfs ranging $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 44K and 0.05 dex respectively.
+Using a training set of \textcolor{red}{51 M dwarfs}\textcolor{black} with spectral types ranging M0-M9 obtained from SDSS optical spectra, we demonstrate that The Cannon can infer spectral types to a precision of \textcolor{red}{0.6 types}\textcolor{black}.
+We then use \textcolor{red}{30 M dwarfs}\textcolor{black} ranging \textcolor{red}{$3072 < T_{\rm eff} < 4131K$}\textcolor{black}, and \textcolor{red}{-0.48 $<$ [Fe/H] $<$ 0.49}\textcolor{black} to train a two-parameter model precise to \textcolor{red}{44K and 0.05 dex}\textcolor{black} respectively.
 Finally we compare our temperature/metallicty model to theoretical models of M dwarf atmospheres (PHOENIX, \citealt{Husser:2013}; BTSETTL, \citealt{Allard:2011}) and identify features of missing line strengths.
@@
 % \begin{enumerate}
-Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
+Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
 
-%define 'labels' - could be physical parameters or more empirical
+Define what parameters/classifications are important for Mdwarfs.
+We want to estimate physical parameters (such as Teff, logg, metallicity) and empirical parameters/classifications (SPT, etc.), which we will refer to as \emph{stellar labels}.
+In the context of this paper we will
 
-\begin{enumerate}
+% Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them:
 
-\item[-] Define what parameters/classifications are important for Mdwarfs.
-We want to estimate physical parameters (such as Teff, logg, metallicity) and empirical parameters/classifications (SPT, etc.), which we will refer to as \emph{stellar labels}.
+Advances in instrumentation and the lauch of many large scale spectroscopic stellar surveys in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} M dwarf catalog with 70,0841 sources obtained from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and \citealt{Yi:2014} with  58,360 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012})
 
-\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources...
-Terrien, Mann, Newton, Dressing catalogs)
+APOGEE instrumentation has allowed for new opportunity - high resolution and large datasets of Mdwarfs are now available
 
-\item[-] brief importance statment about APOGEE - new instrumentation has allowed for new opportunity - high resolution and large datasets of Mdwarfs are now available
-
-\item[-] Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
+Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
 \citealt{Gilhool:2018}; \citealt{Skinner:2018}.
 
-\item[-] Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: models dont accurately reproduce data; empirical standards are relative not absolute.
+Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: models dont accurately reproduce data; empirical standards are relative not absolute.
 
@@
 
-\item[-] Why The Cannon approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016})
+Why The Cannon approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016})
 
@@
  (3) better than template matching in that it is using less sparsely sampled known mdwarfs
-
-\end{enumerate}
 
@@
 % \begin{enumerate}
-% \item[-] Describe the APOGEE data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).
+% Describe the APOGEE data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).
 
@@
 \begin{center}
-\includegraphics[width=18cm]{figures/self_and_validation_test_spt.png}
+\includegraphics[width=18cm]{figures/self_and_validation_test_spt_snr.png}
 \end{center}
-\caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest}
+\caption{Consistency test for the West-trained spectral type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:west_selftest}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=18cm]{figures/self_and_validation_test_teff.png}
-\includegraphics[width=18cm]{figures/self_and_validation_test_fe_h.png}
+\includegraphics[width=18cm]{figures/self_and_validation_test_teff_snr.png}
+\includegraphics[width=18cm]{figures/self_and_validation_test_fe_h_snr.png}
 \end{center}
-\caption{Temperature and Metallicity label self test.} \label{fig:mann_selftest}
+\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:mann_selftest}
 \end{figure}
 
+%---------------------------
 \begin{figure}[ht]
-\plotone{figures/1_Spectral_Sequence.pdf}
-\plotone{figures/2_Spectral_Sequence.pdf}
-\plotone{figures/3_Spectral_Sequence.pdf}
-\caption{ Spectral sequence of dwarfs in training set M0-M9; chip 1 of APOGEE
-spectrum with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
+\plotone{figures/Spectral_Sequence_1.pdf}
+\plotone{figures/Spectral_Sequence_2.pdf}
+\plotone{figures/Spectral_Sequence_3.pdf}
+\caption{ Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
 \end{figure}
 
+%---------------------------
 \begin{figure}[ht]
@@
 \end{center}
-\caption{\textit{Top two panels:} Mann-trained model for varying temperatures.} \label{fig:demo_teff}
+\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities.
+\textit{Third panel:} Derivative of the Cannon model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}.
+\label{fig:demo_teff}
 \end{figure}
@@
 \end{center}
-\caption{\textit{Top two panels:} Mann-trained model for varying metallicities.} \label{fig:demo_feh}
+\caption{\textit{Top two panels of each plot:} Mann-trained model for varying metallicities and similar temperatures.
+\textit{Third panel:} Derivative of the Cannon model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$.} \label{fig:demo_feh}
 \end{figure}
@@
 \tablehead{
-% \colhead{Designation} & \colhead{RA} & \colhead{DEC} & \colhead{$T_{\rm eff}$} & \colhead{[Fe/H]} \\
-% \colhead{(2MASS)} & \colhead{(deg)} & \colhead{(deg)} & \colhead{(K)} & \colhead{(dex)}
 \multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Temperature (K)}} & \multicolumn{3}{c}{\underline{Metallicity (dex)}} \\
@@
 
+\newpage
+
+\startlongtable
+\begin{deluxetable*}{ccc|ccc}
+\tablecaption{Cannon results for Spectral Type model \label{west_results}}
+\tabletypesize{\scriptsize}
+\tablehead{
+% \colhead{Designation} & \colhead{RA} & \colhead{DEC} & \colhead{$T_{\rm eff}$} & \colhead{[Fe/H]} \\
+% \colhead{(2MASS)} & \colhead{(deg)} & \colhead{(deg)} & \colhead{(K)} & \colhead{(dex)}
+\multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Spectral Type}} \\
+	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
+    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV} 
+} 
+\startdata
+2M03114974+0115158 & 47.95722  & 1.25442  & 1          & 0.5       & 0.8        \\
+2M03122509+0021585 & 48.1047   & 0.36621  & 7          & 7.3       & 6.8        \\
+2M03153109+3955161 & 48.87842  & 39.92346 & 2          & 0.3       & 0.2        \\
+2M03423963+0012102 & 55.66512  & 0.20284  & 4          & 4.8       & 4.9        \\
+2M03440466+0049332 & 56.01972  & 0.82795  & 2          & 0.4       & 0.4        \\
+2M04254701+1732407 & 66.44614  & 17.54455 & 5          & 5.7       & 5.7        \\
+2M04262170+1800009 & 66.59065  & 18.00024 & 5          & 5.7       & 5.8        \\
+2M06030979+2233260 & 90.78794  & 22.55808 & 1          & 1.5       & 1.6        \\
+2M07230160+4055426 & 110.75797 & 40.92653 & 4          & 1.6       & 1.4        \\
+2M08043605+5330565 & 121.14987 & 53.51342 & 1          & 0.5       & 0.6        \\
+2M08460531+1035309 & 131.5221  & 10.59194 & 7          & 7.1       & 6.9        \\
+2M08494430+1135327 & 132.43433 & 11.59268 & 0          & 0.6       & 0.9        \\
+2M09013404+0326287 & 135.39185 & 3.44134  & 0          & 1.7       & 2.1        \\
+2M09152918+4407461 & 138.87157 & 44.12966 & 3          & 4.5       & 4.6        \\
+2M09183649+2207022 & 139.65205 & 22.1173  & 3          & 3.0       & 3.0        \\
+2M09332262+2749021 & 143.34409 & 27.81724 & 3          & 3.4       & 3.3        \\
+2M09373349+5534057 & 144.38957 & 55.56828 & 6          & 6.8       & 6.7        \\
+2M10222276+1353383 & 155.59618 & 13.89607 & 3          & 2.3       & 2.2        \\
+2M10313413+3441535 & 157.89214 & 34.69814 & 1          & 2.5       & 2.6        \\
+2M10545293+4402506 & 163.72188 & 44.04674 & 4          & 4.9       & 5.0        \\
+2M11010030+2035056 & 165.25085 & 20.58252 & 2          & 1.0       & 1.0        \\
+2M11194647+0820356 & 169.94386 & 8.34307  & 8          & 7.7       & 7.1        \\
+2M11203609+0704135 & 170.15036 & 7.07021  & 6          & 6.5       & 6.5        \\
+2M11215827+0646252 & 170.4927  & 6.77355  & 5          & 6.1       & 6.1        \\
+2M11570299+2028436 & 179.26254 & 20.47885 & 2          & 3.6       & 3.7        \\
+2M12052751+1811555 & 181.36514 & 18.19978 & 1          & 3.0       & 3.2        \\
+2M12080810+3520281 & 182.03253 & 35.34103 & 7          & 7.1       & 7.0        \\
+2M12102152+4026185 & 182.5888  & 40.44021 & 4          & 3.1       & 2.9        \\
+2M12115216+3525458 & 182.9689  & 35.43142 & 4          & 1.5       & 1.4        \\
+2M12132318+3900529 & 183.34637 & 39.01468 & 7          & 7.4       & 6.9        \\
+2M12202436+2508225 & 185.1015  & 25.1396  & 6          & 6.6       & 6.5        \\
+2M12203634+2505351 & 185.15141 & 25.09318 & 4          & 3.6       & 3.4        \\
+2M12212701-0030560 & 185.36256 & -0.5156  & 0          & 1.7       & 1.9        \\
+2M12252076+2517082 & 186.3364  & 25.28559 & 7          & 7.7       & 7.6        \\
+2M12264289+0009059 & 186.67876 & 0.15164  & 0          & 2.3       & 2.5        \\
+2M12294016+2619567 & 187.4174  & 26.33245 & 7          & 4.4       & 4.1        \\
+2M12423245-0646077 & 190.63527 & -6.76885 & 2          & 3.0       & 3.0        \\
+2M12464541-0312524 & 191.68922 & -3.2146  & 4          & 4.3       & 4.3        \\
+2M12471099+1109566 & 191.79576 & 11.16566 & 3          & 3.8       & 3.9        \\
+2M12471599+2817425 & 191.81609 & 28.29519 & 1          & 0.9       & 1.0        \\
+2M12492657-0312032 & 192.36092 & -3.20097 & 3          & 3.7       & 3.8        \\
+2M12503440+4309482 & 192.64331 & 43.16345 & 4          & 4.9       & 4.9        \\
+2M12523816+1240586 & 193.15907 & 12.68297 & 4          & 4.8       & 4.8        \\
+2M12552141+4150425 & 193.83911 & 41.84522 & 3          & 3.7       & 3.8        \\
+2M12564117+4233175 & 194.17155 & 42.55485 & 3          & 3.7       & 3.8        \\
+2M13032161+4220407 & 195.83998 & 42.34466 & 2          & 3.4       & 3.4        \\
+2M13282079+3859398 & 202.08669 & 38.99444 & 2          & 3.5       & 3.6        \\
+2M13415860+1852278 & 205.49413 & 18.87437 & 4          & 4.7       & 4.8        \\
+2M13442970+5625445 & 206.12355 & 56.42916 & 5          & 6.4       & 6.4        \\
+2M13452385+1844222 & 206.34949 & 18.73947 & 4          & 4.3       & 4.3        \\
+2M13542316+2543152 & 208.59673 & 25.72078 & 4          & 5.0       & 5.1        \\
+2M13565847+3200294 & 209.24355 & 32.00805 & 4          & 4.7       & 4.7        \\
+2M14003808+3217085 & 210.15862 & 32.28572 & 3          & 4.4       & 4.5        \\
+2M14005977+3226109 & 210.2487  & 32.43644 & 7          & 6.9       & 6.6        \\
+2M14030704+5403356 & 210.77922 & 54.05996 & 5          & 5.8       & 5.9        \\
+2M14055603+0140310 & 211.48352 & 1.67535  & 2          & -0.4      & -0.5       \\
+2M14102952+4349477 & 212.62604 & 43.83093 & 3          & 3.1       & 3.1        \\
+2M14213414+0451188 & 215.39229 & 4.85523  & 0          & -1.7      & -1.3       \\
+2M14275590+5817281 & 216.98328 & 58.29095 & 3          & 3.9       & 4.0        \\
+2M14290377+4917403 & 217.26582 & 49.29448 & 4          & 5.5       & 5.5        \\
+2M14350795+4954080 & 218.78323 & 49.90218 & 0          & 1.1       & 1.4        \\
+2M14402293+1339230 & 220.09528 & 13.65579 & 7          & 7.6       & 7.5        \\
+2M14575608+2140337 & 224.48373 & 21.67605 & 1          & 1.6       & 1.7        \\
+2M14592059+3219293 & 224.83575 & 32.32482 & 1          & 0.4       & 0.5        \\
+2M15004819+2050499 & 225.20079 & 20.84723 & 3          & 2.6       & 2.5        \\
+2M15010818+2250020 & 225.28406 & 22.83383 & 9          & 8.5       & 6.1        \\
+2M15034431+2730327 & 225.93346 & 27.50773 & 3          & 0.3       & 0.2        \\
+2M15064250+3246099 & 226.6774  & 32.76876 & 2          & 4.3       & 4.4        \\
+2M15141711+0044474 & 228.5713  & 0.74654  & 7          & 5.8       & 5.4        \\
+2M15195523+0254040 & 229.98016 & 2.90112  & 0          & 0.4       & 0.8        \\
+2M15232142+3453119 & 230.8392  & 34.88651 & 4          & 4.8       & 4.9        \\
+2M15424779+4534214 & 235.69858 & 45.57528 & 3          & 2.1       & 2.1        \\
+2M15575183+0657068 & 239.4659  & 6.9519   & 4          & 5.5       & 5.5        \\
+2M15591175+0637577 & 239.79903 & 6.63268  & 3          & 4.4       & 4.4        \\
+2M16022228+0616424 & 240.59286 & 6.27846  & 3          & 2.3       & 2.3        \\
+2M16055380+2303058 & 241.47423 & 23.05165 & 7          & 6.9       & 6.7        \\
+2M16084282+2327406 & 242.17833 & 23.4614  & 4          & 4.0       & 4.0        \\
+2M16101653+4511112 & 242.56818 & 45.18571 & 0          & 0.3       & 0.6        \\
+2M16103232+2249116 & 242.63477 & 22.81992 & 7          & 7.3       & 7.2        \\
+2M16112965+2403267 & 242.87363 & 24.05741 & 2          & 3.1       & 3.2        \\
+2M16113384+2429072 & 242.89096 & 24.4854  & 4          & 5.5       & 5.6        \\
+2M16114429+2253294 & 242.93452 & 22.89149 & 5          & 4.3       & 4.2        \\
+2M16282719-0050428 & 247.11214 & -0.84372 & 3          & 2.1       & 2.1        \\
+2M16351077-0102156 & 248.79636 & -1.03922 & 3          & 0.5       & 0.4        \\
+2M16355440+3716239 & 248.97668 & 37.27332 & 3          & 3.4       & 3.5        \\
+2M16363188+3655364 & 249.13287 & 36.92676 & 2          & 1.1       & 1.1        \\
+2M16383117+3746233 & 249.62987 & 37.77329 & 4          & 4.8       & 4.9        \\
+2M16393137+3722527 & 249.88008 & 37.38122 & 0          & 3.2       & 3.3        \\
+2M16393824+3550337 & 249.90933 & 35.84268 & 2          & 3.1       & 3.2        \\
+2M16410962+3535407 & 250.29008 & 35.59465 & 3          & 3.0       & 2.9        \\
+2M16485665+3639205 & 252.23611 & 36.65564 & 3          & 3.0       & 2.9        \\
+2M17223434+5946359 & 260.64276 & 59.77846 & 4          & 2.5       & 2.4        \\
+2M19203934+3748048 & 290.16394 & 37.80135 & 0          & -2.0      & -2.1       \\
+2M19210112+3742134 & 290.25467 & 37.70374 & 0          & -2.2      & -2.4       \\
+2M21342546+0021006 & 323.60614 & 0.35243  & 2          & 1.1       & 1.0        \\
+2M23295172+4825211 & 352.4656  & 48.42256 & 1          & 0.3       & 0.4     
+\enddata
+\end{deluxetable*}
+
 %%==========================================================================================

### 2018-07-13 paper/draft1/draft1.tex
clean up files

@@
 
-Advances in instrumentation and the lauch of many large scale spectroscopic stellar surveys in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} M dwarf catalog with 70,0841 sources obtained from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and \citealt{Yi:2014} with  58,360 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012})
+Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,0841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} catalog with 58,360 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.
 
-APOGEE instrumentation has allowed for new opportunity - high resolution and large datasets of Mdwarfs are now available
+The launch of the (APOGEE; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III and IV missions (citations) contains the largest sample of M dwarfs observed with high resolution spectroscopy.
 
-Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
+A number of M dwarf studies have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
 \citealt{Gilhool:2018}; \citealt{Skinner:2018}.
@@
 % \plottwo{figures/west_train_dist.png}{figures/mann_train_dist.png}
-% \caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest} \label{fig:train_dist}
+% \caption{Spectral type model: training label vs. test label.} \label{fig:west_validation} \label{fig:train_dist}
 % \end{figure}
@@
 
-\begin{enumerate}
-\item[-] We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.8 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
-\item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure  displays
-\end{enumerate}
+We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.8 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
+Two tests of consistency were used to assess the validity of our derived test labels (Figure \ref{fig:west_validation}).
+The self-test comparing training labels and test labels where precision and bias of the model are calculated as the standard deviation and mean of the difference in training and test labels respectively.
 
-\begin{enumerate}
-\item[-] For the physical parameter model we trained The Cannon on 30 M dwarfs with a two-dimensional with temperature/metallicity labels, and was able to achieve precisions of 44K/0.05dex respectively, more precise than both of the original training label uncertainties of 60K/0.08dex.
+\subsection{Temperature/Metallicity Model}
 
-\item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure 2 displays
-
-\item[-] Vsini and logg were not included as training parameters.
-Vsini is less import of a parameter as the sample consists of mainly old, slowly rotating sources (check what are ages?).
-Logg because?
-\end{enumerate}
-
-\begin{enumerate}
-\item[-] Describe derivative plots and jacknife error estimation.
-\end{enumerate}
+For the physical parameter model we trained The Cannon on 88 M dwarfs with a two-dimensional with temperature/metallicity labels, with a precision of 44K/0.05dex respectively, more precise than both of the original training label uncertainties of 60K/0.08dex.
 
@@
 \acknowledgements
-We would like to acknowledge Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Hans Walter Rix (MPIA), Anna-Christina Eilers (MPIA), Anna Ho (Caltech), and Jackie Faherty (AMNH) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
+We would like to acknowledge Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Hans Walter Rix (MPIA), Anna-Christina Eilers (MPIA) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
 \color{red}Software: The Cannon, numpy, matplotlib...
@@
 \end{center}
-\caption{Consistency test for the West-trained spectral type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:west_selftest}
+\caption{Consistency test for the West-trained spectral type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:west_validation}
 \end{figure}
@@
 \end{center}
-\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:mann_selftest}
+\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:mann_validation}
 \end{figure}

### 2018-07-13 scripts/plot.py

@@ -1,906 +1,0 @@
-from __future__ import print_function, division
-import numpy as np
-import pandas as pd
-import matplotlib.pyplot as plt
-from matplotlib.backends.backend_pdf import PdfPages
-from matplotlib import rc
-rc('font', family='serif')
-from operator import itemgetter
-import os
-import math
-
-from astropy.table import Table
-from astropy.io import fits, ascii
-from astropy import units as u
-
-AP_PATH = os.environ['APOGEE_DATA']
-
-import apogee_tools as ap
-
-def plotRegion(**kwargs):
-
-    specs  = kwargs.get('specs')
-    params = kwargs.get('params')
-    unc    = kwargs.get('unc')
-
-    spc1, spc2 = specs[0], specs[1]
-    par1, par2 = params[0], params[1]
-    unc1, unc2 = unc[0], unc[1]
-
-    wave1, flux1, name1 = spc1.wave, spc1.flux, spc1.name
-    wave2, flux2, name2 = spc2.wave, spc2.flux, spc2.name
-
-    save = kwargs.get('save', False)
-    out  = kwargs.get('out', 'comparison.pdf')
-    xlim = kwargs.get('xlim', [wave1[0], wave1[-1]])
-
-    # x_range = xlim[1] - xlim[0]
-    # scale = math.ceil(math.log10(x_range))
-
-    fig = plt.figure(figsize=(10,4)) 
-    ax  = fig.add_subplot(1,1,1)
-
-    label1 = name1 + r'$, T_{eff}=%s \pm %s, logg=%s \pm %s, [Fe/H]=%s \pm %s$' %(par1[0], unc1[0], par1[1], unc1[1], par1[2], unc1[2])
-    label2 = name2 + r'$, T_{eff}=%s \pm %s, logg=%s \pm %s, [Fe/H]=%s \pm %s$' %(par2[0], unc2[0], par2[1], unc2[1], par2[2], unc2[2])                                                               
-
-    plt.plot(wave1, flux1, label=label1, alpha=.8, color='b')
-    plt.plot(wave2, flux2, label=label2, alpha=.8, color='g')
-
-    # major_ticks = np.arange(15000, 17000, 200)
-    # minor_ticks = np.arange(15000, 17000, 50)
[…]

### 2018-07-13 scripts/run_cannon.py

@@ -1,584 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-from operator import itemgetter
-from astropy.io import ascii
-from pathlib import Path
-import apogee_tools as ap
-
-from TheCannon import apogee, dataset, model
-
-import os
-os.environ["PATH"] += os.pathsep + '/usr/local/texlive/2016/bin/x86_64-darwin'
-
-
-def loadLabels(filename, **kwargs):
-	
-    data = ascii.read(filename)
-    ids  = data['ID']
-    inds = ids.argsort()
-    ids  = ids[inds]
-
-    lbl_names = kwargs.get('lbl_names', ['SPT'])
-
-    nlbl = len(lbl_names)
-    params = []
-
-    for i in range(nlbl):
-    	p = data[lbl_names[i]]
-    	p = p[inds]
-    	params.append(p)
-    
-    tr_label = np.vstack(params).T
-    
-    return tr_label
-
-
-def initializeTrainingSet(**kwargs):
-
-	"""
-	Read in training data set for The Cannon, and normalize the spectra
-
-	Input:  'data' : director of folder containing aspcap data
-		    'ref'  : csv file of reference labels
-
-	Output: ids, wl, tr_flux, tr_ivar, tr_label = training_set_info
-
-			'ids'      : spectrum ids
-			'wl'       : wavelength array 
-			'tr_flux'  : array of training fluxes
-			'tr_ivar'  : array of inverse variances
-			'tr_label' : array of training set labels
-
-			ds : dataset object for training set spectra
-	"""
-
-	# required
-	data_dir  = kwargs.get('data') 
-	lbl_file  = kwargs.get('ref', None)
-
-	# optional
-	e_vers = kwargs.get('eiler', False) # if True, will use Eiler's Cannon settings
-	uncert = kwargs.get('uncert', False) 
-	save_base = kwargs.get('save_base') #specifies name normalized flux will be save as
-	ds_ranges = kwargs.get('ds_ranges', [[371,3192], [3697,5997], [6461,8255]])
-
-	# main directory of cannon run
-	main_dir = Path(data_dir).parent
-
[…]

### 2018-07-13 scripts/spec.py

@@ -1,173 +1,0 @@
-import numpy as np
-import pandas as pd
-import matplotlib.pyplot as plt
-from operator import itemgetter
-import os
-import math
-
-from astropy.table import Table
-from astropy.io import fits, ascii
-from astropy import units as u
-
-AP_PATH = os.environ['APOGEE_DATA']
-
-import apogee_tools as ap
-
-from plot import *
-
-
-def compareToModels(**kwargs):
-
-	"""
-	This function compares a spectrum to two models and outputs chi-squared values
-	The default M dwarf model is 2M19213157+4317347 and 
-	the default M giant model is 2M00001653+5540107
-	"""
-
-	default_spec1 = ap.Spectrum(id='2M19213157+4317347', type='aspcap')
-	default_spec2 = ap.Spectrum(id='2M00001653+5540107', type='aspcap')
-
-	default_params = [['4079','4.75','-0.20'],['4080','1.57','-0.15']]
-	default_uncert = [['238', '0.15', '0.10 *'],['70', '0.08', '0.02']]
-
-	spec   = kwargs.get('spec')
-	slbl   = kwargs.get('slbl')
-	models = kwargs.get('models', [default_spec1, default_spec2])
-	params = kwargs.get('params', default_params)
-	uncert = kwargs.get('unc', default_uncert)
-	regs   = kwargs.get('regs', [[15650,15780], [16150,16280]]) #for plotting
-	plot   = kwargs.get('plot', False)
-
-	mdl1, mdl2 = models[0], models[1]
-	par1, par2 = params[0], params[1]
-	unc1, unc2 = uncert[0], uncert[1]
-
-	chi1, spec, mdl1 = ap.compareSpectra(spec, mdl1)
-	chi2, spec, mdl2 = ap.compareSpectra(spec, mdl2)
-
-	if plot == True:
-		plotChiComparison(spec=spec, models=models, params=params, unc=uncert, \
-			chis=[chi1, chi2], regs=regs, save=True, slbl=slbl)
-
-	print(spec.name, chi1, chi2)
-
-	return chi1, chi2
-
-
-def sortMStars(**kwargs):
-
-	"""
-	Enter up to 3 lists of spectrum apogee IDs, return list of chi-squared values 
-	for two models and plot (optional).
-
-	Input:  Enter 3 separate lists to plot in different colors:
[…]

### 2018-07-19 paper/draft1/draft1.tex
reorganize files
draft backup

--- 2018-07-03 paper/draft1/draft1.tex
@@
 
-\title{DATA DRIVEN MODELS FOR APOGEE M DWARFS}
+\title{APOGEE CATALOG OF 10,000+ M DWARFS MEASURED BY DATA-DRIVEN MODELS}
 
@@
 Using a training set of 51 M dwarfs with spectral types ranging M0-M9 obtained from SDSS optical spectra, we demonstrate that The Cannon can infer spectral types to a precision of 0.6 types.
-We then use 30 M dwarfs ranging $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 44K and 0.05 dex respectively.
-Finally we compare our temperature/metallicty model to theoretical models of M dwarf atmospheres (PHOENIX, \citealt{Husser:2013}; BTSETTL, \citealt{Allard:2011}) and identify features of missing line strengths.
+We then use 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 68K and 0.07 dex respectively.
+Finally we apply our model to determine spectral subtypes, temperatures and metallicities for 10,000+ M dwarfs in the APOGEE-Gaia cross match.
 
@@
 % \begin{enumerate}
-Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
+Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
 
-%define 'labels' - could be physical parameters or more empirical
+Define what parameters/classifications are important for M dwarfs.
+We want to estimate empirical parameters/classifications (SPT, etc.), physical parameters (such as Teff, logg, metallicity), which we will refer to as \emph{stellar labels}.
 
-\begin{enumerate}
+State why fundamental physical parameters are dificult to measure for M dwarfs.
+When temperatures and metallicities for a star are unavailable, it is often useful to classify a star in terms of spectral type.
+Define spectral type (this is a classification that is based on how a spectra looks, something to do with TiO band strengths, and it correlates with the temperature of a star).
 
-\item[-] Define what parameters/classifications are important for Mdwarfs.
-We want to estimate physical parameters (such as Teff, logg, metallicity) and empirical parameters/classifications (SPT, etc.), which we will refer to as \emph{stellar labels}.
+% Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them:
 
-\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources...
-Terrien, Mann, Newton, Dressing catalogs)
+Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,0841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.
 
-\item[-] brief importance statment about APOGEE - new instrumentation has allowed for new opportunity - high resolution and large datasets of Mdwarfs are now available
+The launch of the (APOGEE; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
 
-\item[-] Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
+A number of M dwarf studies have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
 \citealt{Gilhool:2018}; \citealt{Skinner:2018}.
 
-\item[-] Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: models dont accurately reproduce data; empirical standards are relative not absolute.
+Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: models dont accurately reproduce data; empirical standards are relative not absolute.
 
@@
 
-\subsection{Training Samples}
+\subsection{Training Sample Selection}
 
@@
 The model in this sense is used to \emph{transfer} labels from spectra which we know parameters for to those which we do not.
-Hence this data-driven approach effectively removes the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (ideally very accurately measured) \emph{reference labels} that can be measured from other surveys/methods that are easier to infer from.
+Hence this data-driven approach effectively removes the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (ideally very accurately measured) \emph{reference labels} that can be measured from other surveys/methods that are easier or more accurate to infer from.
 
@@
 
-% \begin{figure}[ht]
-% \plottwo{figures/west_train_dist.png}{figures/mann_train_dist.png}
-% \caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest} \label{fig:train_dist}
-% \end{figure}
+\subsection{Test Sample Selection}
 
@@
 
-APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample .
-
-Describe ASPCAP spectra files and Ness continuum normalization procedure.
+APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample.
 
@@
 
-\begin{enumerate}
-\item[-] Describe the assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label. 
-\item[-] Briefly describe modeling procedure (training step, test step, quadratic labels, etc.), refer to \citealt{Ness:2015} for full description.
-\item[-] Briefly describe model inputs and outputs (1D spectral type model and 2D teff/metallicity and label vectors for those). Mention other models we've tested.
-\end{enumerate}
+The Cannon is a fully empirical model, which applies no use of line lists or radiative transfer models.
+Describe the assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label.
 
-Constructing such a model requires two steps: first the \emph{training step} in which the generative model is constructed at each pixel from the set of reference labels and fluxes, and second the \emph{test step} in which the model is applied to determine the labels of a spectrum
+While The Cannon can in principle be trained on any physical or empirical labels.
 
-% The key advantages of The Cannon being:
-% Data-driven models take away the challenge of directly infering labels from a survey, instead we \emph{transfer labels} from another survery that is more accurate or easier to model.
+Inferring the label of a star with such a model requires two steps: first the \emph{training step} in which a generative model is constructed at each pixel from the set of reference labels and fluxes, and second the \emph{test step} in which the model is applied to determine the labels of a spectrum.
 
-Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a model that assumes the flux at each pixel of the spectrum is parameterized as a function \textit{g} 
+Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector \emph{l} and and coefficient vector \emph{$\theta$}.
+So for each star \emph{n}, at pixel wavelength \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
 \begin{equation}
-	f_{n\lambda} = g(\ell_n | \theta_{\lambda}) + {\rm noise}
+	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + {\rm noise}
 \end{equation}
 
-\begin{equation}
-	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + [s_{\lambda}^{2} + \sigma_{n\lambda}^{2}] \,\xi_{n\lambda}
-\end{equation}
+Where the noise model is ${\rm noise}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, which the root mean sum of (1) the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and (2) the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
+Quadratic label vectors:
+	\[\ell_{n} = [1, \, SPT, \, SPT^{2}] \]
+	\[\ell_{n} = [1, \, T_{\rm eff}, \, [Fe/H], \, T_{\rm eff} \cdot [Fe/H], \, [Fe/H]^{2}] \]
 
-Quadratic label vectors:
-\begin{equation}
-	\ell_{n} = [1, SPT, SPT^{2}]
-\end{equation}
-
-\begin{equation}
-	\ell_{n} = [1, T_{\rm eff}, [Fe/H], T_{\rm eff} \cdot [Fe/H], [Fe/H]^{2}]
-\end{equation}
+The training step is the stage
 
@@
 
-\subsection{Spectral Type Model}
+\subsection{Training Sample}
 
-\begin{enumerate}
-\item[-] We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.8 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
-\item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure  displays
-\end{enumerate}
+For each training sample
 
-\begin{enumerate}
-\item[-] For the physical parameter model we trained The Cannon on 30 M dwarfs with a two-dimensional with temperature/metallicity labels, and was able to achieve precisions of 44K/0.05dex respectively, more precise than both of the original training label uncertainties of 60K/0.08dex.
+Two tests of consistency were used to assess the validity of our derived test labels (Figure \ref{fig:west_validation}).
+The self-test comparing training labels and test labels where precision and bias of the model are calculated as the standard deviation and mean of the difference in training and test labels respectively.
 
-\item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure 2 displays
+Some note on what lines vary significantly with parameter?
 
-\item[-] Vsini and logg were not included as training parameters.
-Vsini is less import of a parameter as the sample consists of mainly old, slowly rotating sources (check what are ages?).
-Logg because?
-\end{enumerate}
+\subsubsection{Spectral Type Model}
+We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
+We note however, that the training sample is distributed heavily towards earlier type sources $<$M7, with only one M8 and one M9 source.
 
-\begin{enumerate}
-\item[-] Describe derivative plots and jacknife error estimation.
-\end{enumerate}
+\subsubsection{Temperature/Metallicity Model}
+For the physical parameter model we trained The Cannon on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 68K/0.07dex respectively, similar to the precision of the training uncertainties of 60K/0.08dex.
+
+We note for this model XX out of 87 sources show strong rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), however the model
+
+\subsection{Test Sample}
 
@@
 \begin{center}
-\includegraphics[width=18cm]{figures/self_and_validation_test_spt.png}
+\includegraphics[width=18cm]{figures/self_and_validation_test_spt_snr.png}
 \end{center}
-\caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest}
+\caption{Consistency test for the West-trained spectral type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:west_validation}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=18cm]{figures/self_and_validation_test_teff.png}
-\includegraphics[width=18cm]{figures/self_and_validation_test_fe_h.png}
+\includegraphics[width=18cm]{figures/self_and_validation_test_teff_snr.png}
+\includegraphics[width=18cm]{figures/self_and_validation_test_fe_h_snr.png}
 \end{center}
-\caption{Temperature and Metallicity label self test.} \label{fig:mann_selftest}
+\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:mann_validation}
 \end{figure}
 
+%---------------------------
 \begin{figure}[ht]
-\plotone{figures/1_Spectral_Sequence.pdf}
-\plotone{figures/2_Spectral_Sequence.pdf}
-\plotone{figures/3_Spectral_Sequence.pdf}
-\caption{ Spectral sequence of dwarfs in training set M0-M9; chip 1 of APOGEE
-spectrum with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
+\plotone{figures/Spectral_Sequence_1.pdf}
+\plotone{figures/Spectral_Sequence_2.pdf}
+\plotone{figures/Spectral_Sequence_3.pdf}
+\caption{ Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
 \end{figure}
 
+%---------------------------
 \begin{figure}[ht]
@@
 \end{center}
-\caption{\textit{Top two panels:} Mann-trained model for varying temperatures.} \label{fig:demo_teff}
+\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities.
+\textit{Third panel:} Derivative of the Cannon model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}.
+\label{fig:demo_teff}
 \end{figure}
 
+%---------------------------
 \begin{figure}[ht]
@@
 \end{center}
-\caption{\textit{Top two panels:} Mann-trained model for varying metallicities.} \label{fig:demo_feh}
+\caption{\textit{Top two panels of each plot:} Mann-trained model for varying metallicities and similar temperatures.
+\textit{Third panel:} Derivative of the Cannon model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$.} \label{fig:demo_feh}
 \end{figure}
 
+%---------------------------
 \begin{figure}[ht]
@@
 
+%---------------------------
+\begin{figure}[!htb]
+\minipage{0.32\textwidth}
+  \includegraphics[width=\linewidth]{figures/test_cmd_teff.png}
+\endminipage\hfill
+\minipage{0.32\textwidth}
+  \includegraphics[width=\linewidth]{figures/test_cmd_feh.png}
+\endminipage\hfill
+\minipage{0.32\textwidth}%
+  \includegraphics[width=\linewidth]{figures/test_cmd_chisq.png}
+\endminipage
+\end{figure}
+
 %%==========================================================================================
@@
 
-% \begin{longrotatetable}
-\startlongtable
-\begin{deluxetable*}{ccc|ccc|ccc}
-\tablecaption{Cannon results for Temperature/Metallicity model \label{mann_results}}
-\tabletypesize{\scriptsize}
-\tablehead{
-% \colhead{Designation} & \colhead{RA} & \colhead{DEC} & \colhead{$T_{\rm eff}$} & \colhead{[Fe/H]} \\
-% \colhead{(2MASS)} & \colhead{(deg)} & \colhead{(deg)} & \colhead{(K)} & \colhead{(dex)}
-\multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Temperature (K)}} & \multicolumn{3}{c}{\underline{Metallicity (dex)}} \\
-	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
-    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}
-    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}  
-} 
-\startdata
-2M00182256+4401222 & 4.59542   & 44.02278  & 3603        & 3532       & 3519        & -0.3       & -0.27     & -0.24      \\
-2M00182549+4401376 & 4.60779   & 44.02734  & 3218        & 3527       & 3554        & -0.3       & -0.28     & -0.27      \\
-2M00285391+5022330 & 7.22488   & 50.37588  & 3207        & 3185       & 3188        & 0.11       & 0.05      & 0.02       \\
-2M00401001+0308050 & 10.04169  & 3.13473   & 3725        & 3783       & 3778        & 0.04       & 0.11      & 0.13       \\
-2M00580115+3919111 & 14.50482  & 39.31977  & 3157        & 3094       & 3101        & -0.07      & -0.01     & 0.03       \\
-2M01232542+1638384 & 20.8559   & 16.64401  & 3272        & 3218       & 3217        & 0.1        & 0.03      & 0.0        \\
-2M02001278+1303112 & 30.05402  & 13.05196  & 3080        & 3057       & 3074        & -0.16      & -0.1      & -0.01      \\
-2M02361535+0652191 & 39.06358  & 6.87167   & 3284        & 3249       & 3253        & -0.12      & -0.29     & -0.29      \\
-2M03044335+6144097 & 46.18104  & 61.73583  & 3500        & 3468       & 3467        & -0.12      & -0.26     & -0.27      \\
-2M03553688+5214291 & 58.90373  & 52.2414   & 3435        & 3381       & 3369        & -0.35      & -0.25     & -0.18      \\
-2M04125880+5236421 & 63.24499  & 52.61165  & 3100        & 3181       & 3200        & -0.04      & -0.01     & 0.01       \\
-2M04310001+3647548 & 67.75003  & 36.79855  & 3419        & 3371       & 3369        & 0.08       & 0.11      & 0.1        \\
-2M05030563+2122362 & 75.77354  & 21.37664  & 3155        & 3290       & 3503        & 0.06       & -0.15     & -0.36      \\
-2M05222053+3031097 & 80.58561  & 30.51933  & 3389        & 3417       & 3426        & 0.28       & 0.2       & 0.14       \\
-2M05312734-0340356 & 82.86417  & -3.67722  & 3801        & 3820       & 3854        & 0.49       & 0.49      & 0.37       \\
-2M05413073+5329239 & 85.37804  & 53.48987  & 3765        & 3752       & 3755        & 0.19       & 0.15      & 0.13       \\
-2M05420897+1229252 & 85.53833  & 12.48956  & 3250        & 3226       & 3235        & -0.22      & -0.34     & -0.31      \\
-2M06000351+0242236 & 90.01458  & 2.70657   & 3214        & 3164       & 3165        & 0.07       & 0.05      & 0.03       \\
-2M06011106+5935508 & 90.2961   & 59.59713  & 3340        & 3258       & 3255        & -0.09      & -0.05     & -0.03      \\
-2M06112610+1032599 & 92.8588   & 10.54998  & 3636        & 3711       & 3698        & -0.39      & -0.42     & -0.39      \\
-2M06544902+3316058 & 103.70411 & 33.26823  & 3448        & 3364       & 3359        & -0.02      & 0.04      & 0.06       \\
-2M07171706-0501031 & 109.32108 & -5.01754  & 3193        & 3159       & 3168        & -0.09      & -0.12     & -0.09      \\
-2M07272450+0513329 & 111.852   & 5.2263    & 3317        & 3285       & 3285        & -0.11      & -0.2      & -0.19      \\
-2M07444018+0333089 & 116.16744 & 3.55252   & 3217        & 3176       & 3174        & 0.23       & 0.26      & 0.15       \\
-2M08031949+5250387 & 120.83131 & 52.84402  & 3508        & 3622       & 3626        & -0.26      & -0.24     & -0.24      \\
-2M08083284+5304377 & 122.13679 & 53.07712  & 3398        & 3280       & 3273        & 0.0        & 0.08      & 0.09       \\
-2M08103429-1348514 & 122.64277 & -13.81427 & 3544        & 3517       & 3515        & 0.0        & 0.0       & 0.0        \\
-2M08155393+3136392 & 123.97459 & 31.61077  & 3753        & 3810       & 3821        & 0.16       & 0.07      & 0.03       \\
-2M08323599+4510175 & 128.14998 & 45.17153  & 3494        & 3505       & 3506        & -0.14      & -0.21     & -0.21      \\
-2M08370799+1507475 & 129.28319 & 15.12944  & 3596        & 3553       & 3552        & -0.17      & -0.26     & -0.27      \\
-2M08524084+2818589 & 133.17018 & 28.31634  & 3166        & 3213       & 3239        & 0.31       & 0.5       & 0.27       \\
-2M08585633+0828259 & 134.73479 & 8.47384   & 3200        & 3142       & 3147        & -0.04      & -0.09     & -0.08      \\
-2M09142298+5241125 & 138.595   & 52.68639  & 3920        & 3900       & 3879        & -0.01      & -0.02     & 0.01       \\
-2M09174473+4612246 & 139.4363  & 46.20685  & 3478        & 3548       & 3556        & 0.04       & -0.02     & -0.03      \\
-2M09214911+4330284 & 140.45447 & 43.50784  & 3170        & 3196       & 3206        & 0.16       & 0.23      & 0.19       \\
-2M09285333-0722148 & 142.22221 & -7.37095  & 3508        & 3445       & 3440        & -0.27      & -0.23     & -0.22      \\
-2M10112218+4927153 & 152.84254 & 49.45431  & 4131        & 4007       & 3946        & 0.24       & 0.22      & 0.23       \\
-2M10121768-0344441 & 153.07364 & -3.74563  & 3623        & 3619       & 3620        & 0.13       & 0.16      & 0.16       \\
-2M10193634+1952122 & 154.90117 & 19.87002  & 3370        & 3384       & 3389        & 0.15       & 0.24      & 0.21       \\
-2M10285555+0050275 & 157.23155 & 0.8411    & 3548        & 3517       & 3515        & -0.18      & -0.16     & -0.15      \\
-2M10331367+3409120 & 158.30687 & 34.15333  & 3500        & 3518       & 3518        & -0.15      & -0.19     & -0.2       \\
-2M10350859+3349499 & 158.7858  & 33.83055  & 3816        & 3906       & 3907        & -0.04      & 0.01      & 0.03       \\
-2M10453795+1833111 & 161.4081  & 18.55306  & 3673        & 3773       & 3780        & 0.14       & 0.13      & 0.13       \\
-2M10505201+0648292 & 162.71704 & 6.80826   & 3238        & 3258       & 3265        & 0.16       & 0.18      & 0.12       \\
-2M10520440+1359509 & 163.01762 & 13.99756  & 3298        & 3230       & 3228        & -0.13      & -0.11     & -0.07      \\
-2M11000432+2249592 & 165.01781 & 22.833    & 3463        & 3426       & 3423        & -0.11      & -0.1      & -0.09      \\
-2M11032023+3558117 & 165.83423 & 35.97054  & 3563        & 3682       & 3551        & -0.38      & -0.58     & -0.39      \\
-2M11052903+4331357 & 166.36992 & 43.52665  & 3619        & 3547       & 3525        & -0.37      & -0.34     & -0.3       \\
-2M11054316+1014093 & 166.42966 & 10.23605  & 3439        & 3395       & 3393        & -0.03      & -0.06     & -0.07      \\
-2M11091225-0436249 & 167.30106 & -4.60695  & 3661        & 3743       & 3748        & -0.03      & -0.09     & -0.1       \\
-2M11263270+2145262 & 171.63622 & 21.75719  & 3610        & 3642       & 3642        & -0.18      & -0.16     & -0.15      \\
-2M11273856+0358359 & 171.91071 & 3.97664   & 3920        & 3932       & 3963        & 0.43       & 0.44      & 0.31       \\
-2M11474074+0015201 & 176.9198  & 0.25553   & 3261        & 3235       & 3236        & 0.1        & 0.23      & 0.22       \\
-2M11474440+0048164 & 176.93491 & 0.80473   & 3192        & 3155       & 3157        & -0.02      & -0.06     & -0.06      \\
-2M11530522+1855480 & 178.27184 & 18.93002  & 3773        & 3760       & 3752        & -0.14      & -0.08     & -0.06      \\
-2M12192028+1323524 & 184.83449 & 13.39793  & 3926        & 3954       & 3945        & 0.0        & 0.13      & 0.18       \\
-2M12274471-0315006 & 186.93625 & -3.25018  & 3355        & 3399       & 3407        & 0.19       & 0.2       & 0.17       \\
-2M12390461+4702234 & 189.76958 & 47.03974  & 3492        & 3447       & 3445        & 0.05       & 0.06      & 0.06       \\
-2M12470102+4637334 & 191.75412 & 46.62587  & 3571        & 3543       & 3544        & 0.22       & 0.23      & 0.22       \\
-2M13113464+1306461 & 197.89433 & 13.11282  & 3459        & 3407       & 3402        & -0.1       & -0.03     & -0.01      \\
-2M13160127+1415504 & 199.00528 & 14.26403  & 3686        & 3768       & 3775        & -0.06      & -0.16     & -0.17      \\
-2M13295979+1022376 & 202.49895 & 10.37731  & 3727        & 3674       & 3669        & -0.09      & -0.13     & -0.13      \\
-2M13345147+3746195 & 203.71442 & 37.7721   & 3273        & 3247       & 3250        & 0.14       & 0.13      & 0.08       \\
-2M13400879+4346380 & 205.03676 & 43.77729  & 3332        & 3322       & 3325        & -0.19      & -0.18     & -0.14      \\
-2M13454354+1453317 & 206.43251 & 14.89139  & 3649        & 3688       & 3677        & -0.31      & -0.37     & -0.36      \\
-2M15290936+5724418 & 232.28886 & 57.41167  & 3660        & 3719       & 3723        & -0.09      & -0.14     & -0.14      \\
-2M15594729+4403595 & 239.94707 & 44.06654  & 3526        & 3614       & 3632        & 0.05       & 0.02      & 0.01       \\
-2M16252459+5418148 & 246.35249 & 54.30413  & 3475        & 3407       & 3400        & -0.35      & -0.25     & -0.22      \\
-2M16541202+1154529 & 253.5502  & 11.91464  & 3834        & 3760       & 3697        & -0.48      & -0.38     & -0.29      \\
-2M17093153+4340531 & 257.38138 & 43.68139  & 3284        & 3242       & 3242        & 0.2        & 0.25      & 0.17       \\
-2M17362594+6820220 & 264.10803 & 68.33932  & 3439        & 3487       & 3492        & 0.05       & -0.21     & -0.23      \\
-2M18161819+0131277 & 274.07575 & 1.5243    & 3391        & 3350       & 3347        & -0.25      & -0.19     & -0.17      \\
-2M18244689-0620311 & 276.1954  & -6.342    & 3412        & 3386       & 3384        & 0.01       & 0.07      & 0.08       \\
-2M18424666+5937499 & 280.69489 & 59.63015  & 3441        & 3386       & 3385        & -0.23      & -0.32     & -0.32      \\
-2M18424688+5937374 & 280.6958  & 59.62659  & 3345        & 3391       & 3400        & -0.3       & -0.35     & -0.33      \\
-2M18451027+0620158 & 281.29282 & 6.33774   & 3674        & 3684       & 3688        & 0.02       & 0.03      & 0.03       \\
-2M19142541+1755081 & 288.6059  & 17.91894  & 3725        & 3830       & 3835        & -0.08      & -0.09     & -0.09      \\
-2M19165526+0510086 & 289.23032 & 5.16909   & 3558        & 3536       & 3538        & 0.1        & 0.08      & 0.07       \\
-2M19500252+3235012 & 297.51022 & 32.58347  & 3385        & 3214       & 3215        & -0.18      & -0.31     & -0.28      \\
-2M19535443+4424541 & 298.47977 & 44.41508  & 2859        & 2801       & 2889        & -0.05      & 0.08      & 0.19       \\
-2M20032651+2952000 & 300.86075 & 29.86652  & 3144        & 3164       & 3178        & 0.06       & -0.03     & -0.09      \\
-2M20531977+6209156 & 313.33246 & 62.1545   & 3791        & 3728       & 3712        & -0.06      & 0.04      & 0.08       \\
-2M21065473+3844265 & 316.73013 & 38.74202  & 4021        & 4004       & 3922        & -0.22      & -0.23     & -0.11      \\
-2M22464980+4420030 & 341.70737 & 44.33405  & 3298        & 3236       & 3235        & 0.03       & 0.03      & 0.02       \\
-2M22563497+1633130 & 344.14517 & 16.55347  & 3720        & 3689       & 3688        & 0.21       & 0.29      & 0.28       \\
-2M23060482+6355339 & 346.52011 & 63.92622  & 3700        & 3770       & 3780        & 0.19       & 0.14      & 0.11       \\
-2M23315208+1956142 & 352.96732 & 19.93727  & 3353        & 3278       & 3274        & 0.03       & 0.05      & 0.05       \\
-2M23315244+1956138 & 352.9689  & 19.93719  & 3072        & 3266       & 3296        & 0.03       & 0.04      & 0.03       \\
-2M23491255+0224037 & 357.30206 & 2.40136   & 3646        & 3604       & 3560        & -0.45      & -0.46     & -0.4    
-\enddata
-\end{deluxetable*}
-% \end{longrotatetable}
+% \startlongtable
+% \begin{deluxetable*}{ccc|ccc|ccc}
+% \tablecaption{Cannon results for Temperature/Metallicity model \label{mann_results}}
+% \tabletypesize{\scriptsize}
+% \tablehead{
+% \multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Temperature (K)}} & \multicolumn{3}{c}{\underline{Metallicity (dex)}} \\
+% 	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
+%     & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}
+%     & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}  
+% } 
+% \startdata
+% 2M00182256+4401222 & 4.59542   & 44.02278  & 3603        & 3532       & 3519        & -0.3       & -0.27     & -0.24      \\
+% 2M00182549+4401376 & 4.60779   & 44.02734  & 3218        & 3527       & 3554        & -0.3       & -0.28     & -0.27      \\
+% 2M00285391+5022330 & 7.22488   & 50.37588  & 3207        & 3185       & 3188        & 0.11       & 0.05      & 0.02       \\
+% 2M00401001+0308050 & 10.04169  & 3.13473   & 3725        & 3783       & 3778        & 0.04       & 0.11      & 0.13       \\
+% 2M00580115+3919111 & 14.50482  & 39.31977  & 3157        & 3094       & 3101        & -0.07      & -0.01     & 0.03       \\
+% 2M01232542+1638384 & 20.8559   & 16.64401  & 3272        & 3218       & 3217        & 0.1        & 0.03      & 0.0        \\
+% 2M02001278+1303112 & 30.05402  & 13.05196  & 3080        & 3057       & 3074        & -0.16      & -0.1      & -0.01      \\
+% 2M02361535+0652191 & 39.06358  & 6.87167   & 3284        & 3249       & 3253        & -0.12      & -0.29     & -0.29      \\
+% 2M03044335+6144097 & 46.18104  & 61.73583  & 3500        & 3468       & 3467        & -0.12      & -0.26     & -0.27      \\
+% 2M03553688+5214291 & 58.90373  & 52.2414   & 3435        & 3381       & 3369        & -0.35      & -0.25     & -0.18      \\
+% 2M04125880+5236421 & 63.24499  & 52.61165  & 3100        & 3181       & 3200        & -0.04      & -0.01     & 0.01       \\
+% 2M04310001+3647548 & 67.75003  & 36.79855  & 3419        & 3371       & 3369        & 0.08       & 0.11      & 0.1        \\
+% 2M05030563+2122362 & 75.77354  & 21.37664  & 3155        & 3290       & 3503        & 0.06       & -0.15     & -0.36      \\
+% 2M05222053+3031097 & 80.58561  & 30.51933  & 3389        & 3417       & 3426        & 0.28       & 0.2       & 0.14       \\
+% 2M05312734-0340356 & 82.86417  & -3.67722  & 3801        & 3820       & 3854        & 0.49       & 0.49      & 0.37       \\
+% 2M05413073+5329239 & 85.37804  & 53.48987  & 3765        & 3752       & 3755        & 0.19       & 0.15      & 0.13       \\
+% 2M05420897+1229252 & 85.53833  & 12.48956  & 3250        & 3226       & 3235        & -0.22      & -0.34     & -0.31      \\
+% 2M06000351+0242236 & 90.01458  & 2.70657   & 3214        & 3164       & 3165        & 0.07       & 0.05      & 0.03       \\
+% 2M06011106+5935508 & 90.2961   & 59.59713  & 3340        & 3258       & 3255        & -0.09      & -0.05     & -0.03      \\
+% 2M06112610+1032599 & 92.8588   & 10.54998  & 3636        & 3711       & 3698        & -0.39      & -0.42     & -0.39      \\
+% 2M06544902+3316058 & 103.70411 & 33.26823  & 3448        & 3364       & 3359        & -0.02      & 0.04      & 0.06       \\
+% 2M07171706-0501031 & 109.32108 & -5.01754  & 3193        & 3159       & 3168        & -0.09      & -0.12     & -0.09      \\
+% 2M07272450+0513329 & 111.852   & 5.2263    & 3317        & 3285       & 3285        & -0.11      & -0.2      & -0.19      \\
+% 2M07444018+0333089 & 116.16744 & 3.55252   & 3217        & 3176       & 3174        & 0.23       & 0.26      & 0.15       \\
+% 2M08031949+5250387 & 120.83131 & 52.84402  & 3508        & 3622       & 3626        & -0.26      & -0.24     & -0.24      \\
+% 2M08083284+5304377 & 122.13679 & 53.07712  & 3398        & 3280       & 3273        & 0.0        & 0.08      & 0.09       \\
+% 2M08103429-1348514 & 122.64277 & -13.81427 & 3544        & 3517       & 3515        & 0.0        & 0.0       & 0.0        \\
+% 2M08155393+3136392 & 123.97459 & 31.61077  & 3753        & 3810       & 3821        & 0.16       & 0.07      & 0.03       \\
+% 2M08323599+4510175 & 128.14998 & 45.17153  & 3494        & 3505       & 3506        & -0.14      & -0.21     & -0.21      \\
+% 2M08370799+1507475 & 129.28319 & 15.12944  & 3596        & 3553       & 3552        & -0.17      & -0.26     & -0.27      \\
+% 2M08524084+2818589 & 133.17018 & 28.31634  & 3166        & 3213       & 3239        & 0.31       & 0.5       & 0.27       \\
+% 2M08585633+0828259 & 134.73479 & 8.47384   & 3200        & 3142       & 3147        & -0.04      & -0.09     & -0.08      \\
+% 2M09142298+5241125 & 138.595   & 52.68639  & 3920        & 3900       & 3879        & -0.01      & -0.02     & 0.01       \\
+% 2M09174473+4612246 & 139.4363  & 46.20685  & 3478        & 3548       & 3556        & 0.04       & -0.02     & -0.03      \\
+% 2M09214911+4330284 & 140.45447 & 43.50784  & 3170        & 3196       & 3206        & 0.16       & 0.23      & 0.19       \\
+% 2M09285333-0722148 & 142.22221 & -7.37095  & 3508        & 3445       & 3440        & -0.27      & -0.23     & -0.22      \\
+% 2M10112218+4927153 & 152.84254 & 49.45431  & 4131        & 4007       & 3946        & 0.24       & 0.22      & 0.23       \\
+% 2M10121768-0344441 & 153.07364 & -3.74563  & 3623        & 3619       & 3620        & 0.13       & 0.16      & 0.16       \\
+% 2M10193634+1952122 & 154.90117 & 19.87002  & 3370        & 3384       & 3389        & 0.15       & 0.24      & 0.21       \\
+% 2M10285555+0050275 & 157.23155 & 0.8411    & 3548        & 3517       & 3515        & -0.18      & -0.16     & -0.15      \\
+% 2M10331367+3409120 & 158.30687 & 34.15333  & 3500        & 3518       & 3518        & -0.15      & -0.19     & -0.2       \\
+% 2M10350859+3349499 & 158.7858  & 33.83055  & 3816        & 3906       & 3907        & -0.04      & 0.01      & 0.03       \\
+% 2M10453795+1833111 & 161.4081  & 18.55306  & 3673        & 3773       & 3780        & 0.14       & 0.13      & 0.13       \\
+% 2M10505201+0648292 & 162.71704 & 6.80826   & 3238        & 3258       & 3265        & 0.16       & 0.18      & 0.12       \\
+% 2M10520440+1359509 & 163.01762 & 13.99756  & 3298        & 3230       & 3228        & -0.13      & -0.11     & -0.07      \\
+% 2M11000432+2249592 & 165.01781 & 22.833    & 3463        & 3426       & 3423        & -0.11      & -0.1      & -0.09      \\
+% 2M11032023+3558117 & 165.83423 & 35.97054  & 3563        & 3682       & 3551        & -0.38      & -0.58     & -0.39      \\
+% 2M11052903+4331357 & 166.36992 & 43.52665  & 3619        & 3547       & 3525        & -0.37      & -0.34     & -0.3       \\
+% 2M11054316+1014093 & 166.42966 & 10.23605  & 3439        & 3395       & 3393        & -0.03      & -0.06     & -0.07      \\
+% 2M11091225-0436249 & 167.30106 & -4.60695  & 3661        & 3743       & 3748        & -0.03      & -0.09     & -0.1       \\
+% 2M11263270+2145262 & 171.63622 & 21.75719  & 3610        & 3642       & 3642        & -0.18      & -0.16     & -0.15      \\
+% 2M11273856+0358359 & 171.91071 & 3.97664   & 3920        & 3932       & 3963        & 0.43       & 0.44      & 0.31       \\
+% 2M11474074+0015201 & 176.9198  & 0.25553   & 3261        & 3235       & 3236        & 0.1        & 0.23      & 0.22       \\
+% 2M11474440+0048164 & 176.93491 & 0.80473   & 3192        & 3155       & 3157        & -0.02      & -0.06     & -0.06      \\
+% 2M11530522+1855480 & 178.27184 & 18.93002  & 3773        & 3760       & 3752        & -0.14      & -0.08     & -0.06      \\
+% 2M12192028+1323524 & 184.83449 & 13.39793  & 3926        & 3954       & 3945        & 0.0        & 0.13      & 0.18       \\
+% 2M12274471-0315006 & 186.93625 & -3.25018  & 3355        & 3399       & 3407        & 0.19       & 0.2       & 0.17       \\
+% 2M12390461+4702234 & 189.76958 & 47.03974  & 3492        & 3447       & 3445        & 0.05       & 0.06      & 0.06       \\
+% 2M12470102+4637334 & 191.75412 & 46.62587  & 3571        & 3543       & 3544        & 0.22       & 0.23      & 0.22       \\
+% 2M13113464+1306461 & 197.89433 & 13.11282  & 3459        & 3407       & 3402        & -0.1       & -0.03     & -0.01      \\
+% 2M13160127+1415504 & 199.00528 & 14.26403  & 3686        & 3768       & 3775        & -0.06      & -0.16     & -0.17      \\
+% 2M13295979+1022376 & 202.49895 & 10.37731  & 3727        & 3674       & 3669        & -0.09      & -0.13     & -0.13      \\
+% 2M13345147+3746195 & 203.71442 & 37.7721   & 3273        & 3247       & 3250        & 0.14       & 0.13      & 0.08       \\
+% 2M13400879+4346380 & 205.03676 & 43.77729  & 3332        & 3322       & 3325        & -0.19      & -0.18     & -0.14      \\
+% 2M13454354+1453317 & 206.43251 & 14.89139  & 3649        & 3688       & 3677        & -0.31      & -0.37     & -0.36      \\
+% 2M15290936+5724418 & 232.28886 & 57.41167  & 3660        & 3719       & 3723        & -0.09      & -0.14     & -0.14      \\
+% 2M15594729+4403595 & 239.94707 & 44.06654  & 3526        & 3614       & 3632        & 0.05       & 0.02      & 0.01       \\
+% 2M16252459+5418148 & 246.35249 & 54.30413  & 3475        & 3407       & 3400        & -0.35      & -0.25     & -0.22      \\
+% 2M16541202+1154529 & 253.5502  & 11.91464  & 3834        & 3760       & 3697        & -0.48      & -0.38     & -0.29      \\
+% 2M17093153+4340531 & 257.38138 & 43.68139  & 3284        & 3242       & 3242        & 0.2        & 0.25      & 0.17       \\
+% 2M17362594+6820220 & 264.10803 & 68.33932  & 3439        & 3487       & 3492        & 0.05       & -0.21     & -0.23      \\
+% 2M18161819+0131277 & 274.07575 & 1.5243    & 3391        & 3350       & 3347        & -0.25      & -0.19     & -0.17      \\
+% 2M18244689-0620311 & 276.1954  & -6.342    & 3412        & 3386       & 3384        & 0.01       & 0.07      & 0.08       \\
+% 2M18424666+5937499 & 280.69489 & 59.63015  & 3441        & 3386       & 3385        & -0.23      & -0.32     & -0.32      \\
+% 2M18424688+5937374 & 280.6958  & 59.62659  & 3345        & 3391       & 3400        & -0.3       & -0.35     & -0.33      \\
+% 2M18451027+0620158 & 281.29282 & 6.33774   & 3674        & 3684       & 3688        & 0.02       & 0.03      & 0.03       \\
+% 2M19142541+1755081 & 288.6059  & 17.91894  & 3725        & 3830       & 3835        & -0.08      & -0.09     & -0.09      \\
+% 2M19165526+0510086 & 289.23032 & 5.16909   & 3558        & 3536       & 3538        & 0.1        & 0.08      & 0.07       \\
+% 2M19500252+3235012 & 297.51022 & 32.58347  & 3385        & 3214       & 3215        & -0.18      & -0.31     & -0.28      \\
+% 2M19535443+4424541 & 298.47977 & 44.41508  & 2859        & 2801       & 2889        & -0.05      & 0.08      & 0.19       \\
+% 2M20032651+2952000 & 300.86075 & 29.86652  & 3144        & 3164       & 3178        & 0.06       & -0.03     & -0.09      \\
+% 2M20531977+6209156 & 313.33246 & 62.1545   & 3791        & 3728       & 3712        & -0.06      & 0.04      & 0.08       \\
+% 2M21065473+3844265 & 316.73013 & 38.74202  & 4021        & 4004       & 3922        & -0.22      & -0.23     & -0.11      \\
+% 2M22464980+4420030 & 341.70737 & 44.33405  & 3298        & 3236       & 3235        & 0.03       & 0.03      & 0.02       \\
+% 2M22563497+1633130 & 344.14517 & 16.55347  & 3720        & 3689       & 3688        & 0.21       & 0.29      & 0.28       \\
+% 2M23060482+6355339 & 346.52011 & 63.92622  & 3700        & 3770       & 3780        & 0.19       & 0.14      & 0.11       \\
+% 2M23315208+1956142 & 352.96732 & 19.93727  & 3353        & 3278       & 3274        & 0.03       & 0.05      & 0.05       \\
+% 2M23315244+1956138 & 352.9689  & 19.93719  & 3072        & 3266       & 3296        & 0.03       & 0.04      & 0.03       \\
+% 2M23491255+0224037 & 357.30206 & 2.40136   & 3646        & 3604       & 3560        & -0.45      & -0.46     & -0.4        
+% \enddata
+% \end{deluxetable*}
+% % \end{longrotatetable}
+
+% \newpage
+
+% \startlongtable
+% \begin{deluxetable*}{ccc|ccc}
+% \tablecaption{Cannon results for Spectral Type model \label{west_results}}
+% \tabletypesize{\scriptsize}
+% \tablehead{
+% % \colhead{Designation} & \colhead{RA} & \colhead{DEC} & \colhead{$T_{\rm eff}$} & \colhead{[Fe/H]} \\
+% % \colhead{(2MASS)} & \colhead{(deg)} & \colhead{(deg)} & \colhead{(K)} & \colhead{(dex)}
+% \multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Spectral Type}} \\
+% 	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
+%     & \colhead{Training} & \colhead{Test} & \colhead{LOOCV} 
+% } 
+% \startdata
+% 2M03114974+0115158 & 47.95722  & 1.25442  & 1          & 0.5       & 0.8        \\
+% 2M03122509+0021585 & 48.1047   & 0.36621  & 7          & 7.3       & 6.8        \\
+% 2M03153109+3955161 & 48.87842  & 39.92346 & 2          & 0.3       & 0.2        \\
+% 2M03423963+0012102 & 55.66512  & 0.20284  & 4          & 4.8       & 4.9        \\
+% 2M03440466+0049332 & 56.01972  & 0.82795  & 2          & 0.4       & 0.4        \\
+% 2M04254701+1732407 & 66.44614  & 17.54455 & 5          & 5.7       & 5.7        \\
+% 2M04262170+1800009 & 66.59065  & 18.00024 & 5          & 5.7       & 5.8        \\
+% 2M06030979+2233260 & 90.78794  & 22.55808 & 1          & 1.5       & 1.6        \\
+% 2M07230160+4055426 & 110.75797 & 40.92653 & 4          & 1.6       & 1.4        \\
+% 2M08043605+5330565 & 121.14987 & 53.51342 & 1          & 0.5       & 0.6        \\
+% 2M08460531+1035309 & 131.5221  & 10.59194 & 7          & 7.1       & 6.9        \\
+% 2M08494430+1135327 & 132.43433 & 11.59268 & 0          & 0.6       & 0.9        \\
+% 2M09013404+0326287 & 135.39185 & 3.44134  & 0          & 1.7       & 2.1        \\
+% 2M09152918+4407461 & 138.87157 & 44.12966 & 3          & 4.5       & 4.6        \\
+% 2M09183649+2207022 & 139.65205 & 22.1173  & 3          & 3.0       & 3.0        \\
+% 2M09332262+2749021 & 143.34409 & 27.81724 & 3          & 3.4       & 3.3        \\
+% 2M09373349+5534057 & 144.38957 & 55.56828 & 6          & 6.8       & 6.7        \\
+% 2M10222276+1353383 & 155.59618 & 13.89607 & 3          & 2.3       & 2.2        \\
+% 2M10313413+3441535 & 157.89214 & 34.69814 & 1          & 2.5       & 2.6        \\
+% 2M10545293+4402506 & 163.72188 & 44.04674 & 4          & 4.9       & 5.0        \\
+% 2M11010030+2035056 & 165.25085 & 20.58252 & 2          & 1.0       & 1.0        \\
+% 2M11194647+0820356 & 169.94386 & 8.34307  & 8          & 7.7       & 7.1        \\
+% 2M11203609+0704135 & 170.15036 & 7.07021  & 6          & 6.5       & 6.5        \\
+% 2M11215827+0646252 & 170.4927  & 6.77355  & 5          & 6.1       & 6.1        \\
+% 2M11570299+2028436 & 179.26254 & 20.47885 & 2          & 3.6       & 3.7        \\
+% 2M12052751+1811555 & 181.36514 & 18.19978 & 1          & 3.0       & 3.2        \\
+% 2M12080810+3520281 & 182.03253 & 35.34103 & 7          & 7.1       & 7.0        \\
+% 2M12102152+4026185 & 182.5888  & 40.44021 & 4          & 3.1       & 2.9        \\
+% 2M12115216+3525458 & 182.9689  & 35.43142 & 4          & 1.5       & 1.4        \\
+% 2M12132318+3900529 & 183.34637 & 39.01468 & 7          & 7.4       & 6.9        \\
+% 2M12202436+2508225 & 185.1015  & 25.1396  & 6          & 6.6       & 6.5        \\
+% 2M12203634+2505351 & 185.15141 & 25.09318 & 4          & 3.6       & 3.4        \\
+% 2M12212701-0030560 & 185.36256 & -0.5156  & 0          & 1.7       & 1.9        \\
+% 2M12252076+2517082 & 186.3364  & 25.28559 & 7          & 7.7       & 7.6        \\
+% 2M12264289+0009059 & 186.67876 & 0.15164  & 0          & 2.3       & 2.5        \\
+% 2M12294016+2619567 & 187.4174  & 26.33245 & 7          & 4.4       & 4.1        \\
+% 2M12423245-0646077 & 190.63527 & -6.76885 & 2          & 3.0       & 3.0        \\
+% 2M12464541-0312524 & 191.68922 & -3.2146  & 4          & 4.3       & 4.3        \\
+% 2M12471099+1109566 & 191.79576 & 11.16566 & 3          & 3.8       & 3.9        \\
+% 2M12471599+2817425 & 191.81609 & 28.29519 & 1          & 0.9       & 1.0        \\
+% 2M12492657-0312032 & 192.36092 & -3.20097 & 3          & 3.7       & 3.8        \\
+% 2M12503440+4309482 & 192.64331 & 43.16345 & 4          & 4.9       & 4.9        \\
+% 2M12523816+1240586 & 193.15907 & 12.68297 & 4          & 4.8       & 4.8        \\
+% 2M12552141+4150425 & 193.83911 & 41.84522 & 3          & 3.7       & 3.8        \\
+% 2M12564117+4233175 & 194.17155 & 42.55485 & 3          & 3.7       & 3.8        \\
+% 2M13032161+4220407 & 195.83998 & 42.34466 & 2          & 3.4       & 3.4        \\
+% 2M13282079+3859398 & 202.08669 & 38.99444 & 2          & 3.5       & 3.6        \\
+% 2M13415860+1852278 & 205.49413 & 18.87437 & 4          & 4.7       & 4.8        \\
+% 2M13442970+5625445 & 206.12355 & 56.42916 & 5          & 6.4       & 6.4        \\
+% 2M13452385+1844222 & 206.34949 & 18.73947 & 4          & 4.3       & 4.3        \\
+% 2M13542316+2543152 & 208.59673 & 25.72078 & 4          & 5.0       & 5.1        \\
+% 2M13565847+3200294 & 209.24355 & 32.00805 & 4          & 4.7       & 4.7        \\
+% 2M14003808+3217085 & 210.15862 & 32.28572 & 3          & 4.4       & 4.5        \\
+% 2M14005977+3226109 & 210.2487  & 32.43644 & 7          & 6.9       & 6.6        \\
+% 2M14030704+5403356 & 210.77922 & 54.05996 & 5          & 5.8       & 5.9        \\
+% 2M14055603+0140310 & 211.48352 & 1.67535  & 2          & -0.4      & -0.5       \\
+% 2M14102952+4349477 & 212.62604 & 43.83093 & 3          & 3.1       & 3.1        \\
+% 2M14213414+0451188 & 215.39229 & 4.85523  & 0          & -1.7      & -1.3       \\
+% 2M14275590+5817281 & 216.98328 & 58.29095 & 3          & 3.9       & 4.0        \\
+% 2M14290377+4917403 & 217.26582 & 49.29448 & 4          & 5.5       & 5.5        \\
+% 2M14350795+4954080 & 218.78323 & 49.90218 & 0          & 1.1       & 1.4        \\
+% 2M14402293+1339230 & 220.09528 & 13.65579 & 7          & 7.6       & 7.5        \\
+% 2M14575608+2140337 & 224.48373 & 21.67605 & 1          & 1.6       & 1.7        \\
+% 2M14592059+3219293 & 224.83575 & 32.32482 & 1          & 0.4       & 0.5        \\
+% 2M15004819+2050499 & 225.20079 & 20.84723 & 3          & 2.6       & 2.5        \\
+% 2M15010818+2250020 & 225.28406 & 22.83383 & 9          & 8.5       & 6.1        \\
+% 2M15034431+2730327 & 225.93346 & 27.50773 & 3          & 0.3       & 0.2        \\
+% 2M15064250+3246099 & 226.6774  & 32.76876 & 2          & 4.3       & 4.4        \\
+% 2M15141711+0044474 & 228.5713  & 0.74654  & 7          & 5.8       & 5.4        \\
+% 2M15195523+0254040 & 229.98016 & 2.90112  & 0          & 0.4       & 0.8        \\
+% 2M15232142+3453119 & 230.8392  & 34.88651 & 4          & 4.8       & 4.9        \\
+% 2M15424779+4534214 & 235.69858 & 45.57528 & 3          & 2.1       & 2.1        \\
+% 2M15575183+0657068 & 239.4659  & 6.9519   & 4          & 5.5       & 5.5        \\
+% 2M15591175+0637577 & 239.79903 & 6.63268  & 3          & 4.4       & 4.4        \\
+% 2M16022228+0616424 & 240.59286 & 6.27846  & 3          & 2.3       & 2.3        \\
+% 2M16055380+2303058 & 241.47423 & 23.05165 & 7          & 6.9       & 6.7        \\
+% 2M16084282+2327406 & 242.17833 & 23.4614  & 4          & 4.0       & 4.0        \\
+% 2M16101653+4511112 & 242.56818 & 45.18571 & 0          & 0.3       & 0.6        \\
+% 2M16103232+2249116 & 242.63477 & 22.81992 & 7          & 7.3       & 7.2        \\
+% 2M16112965+2403267 & 242.87363 & 24.05741 & 2          & 3.1       & 3.2        \\
+% 2M16113384+2429072 & 242.89096 & 24.4854  & 4          & 5.5       & 5.6        \\
+% 2M16114429+2253294 & 242.93452 & 22.89149 & 5          & 4.3       & 4.2        \\
+% 2M16282719-0050428 & 247.11214 & -0.84372 & 3          & 2.1       & 2.1        \\
+% 2M16351077-0102156 & 248.79636 & -1.03922 & 3          & 0.5       & 0.4        \\
+% 2M16355440+3716239 & 248.97668 & 37.27332 & 3          & 3.4       & 3.5        \\
+% 2M16363188+3655364 & 249.13287 & 36.92676 & 2          & 1.1       & 1.1        \\
+% 2M16383117+3746233 & 249.62987 & 37.77329 & 4          & 4.8       & 4.9        \\
+% 2M16393137+3722527 & 249.88008 & 37.38122 & 0          & 3.2       & 3.3        \\
+% 2M16393824+3550337 & 249.90933 & 35.84268 & 2          & 3.1       & 3.2        \\
+% 2M16410962+3535407 & 250.29008 & 35.59465 & 3          & 3.0       & 2.9        \\
+% 2M16485665+3639205 & 252.23611 & 36.65564 & 3          & 3.0       & 2.9        \\
+% 2M17223434+5946359 & 260.64276 & 59.77846 & 4          & 2.5       & 2.4        \\
+% 2M19203934+3748048 & 290.16394 & 37.80135 & 0          & -2.0      & -2.1       \\
+% 2M19210112+3742134 & 290.25467 & 37.70374 & 0          & -2.2      & -2.4       \\
+% 2M21342546+0021006 & 323.60614 & 0.35243  & 2          & 1.1       & 1.0        \\
+% 2M23295172+4825211 & 352.4656  & 48.42256 & 1          & 0.3       & 0.4     
+% \enddata
+% \end{deluxetable*}
 

### 2018-07-23 paper/draft1/draft1.tex
update figures

@@
 
-While The Cannon can in principle be trained on any physical or empirical labels.
+The Cannon model can in principle be trained on any physical or empirical labels available.
+However, limitations to consider are that test labels from The Cannon are only accurate if the training labels accurate, and only precise if the training labels are measured using a consistent method.
+The Cannon does not extrapolate well outside the parameter space of the training sample.
 
@@
 
-For each training sample
-
-Two tests of consistency were used to assess the validity of our derived test labels (Figure \ref{fig:west_validation}).
+For each model, two tests of consistency were used to assess the validity of our derived test labels.
 The self-test comparing training labels and test labels where precision and bias of the model are calculated as the standard deviation and mean of the difference in training and test labels respectively.
@@
 
-We note for this model XX out of 87 sources show strong rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), however the model
+We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), however the model
+
+\color{gcolor}{HOGG: Some commentary on why $\chi^2$ is higher than expected.}\color{black}
 
 \subsection{Test Sample}
+Some validation of our test label results...
 
@@
 \begin{center}
-\includegraphics[width=18cm]{figures/self_and_validation_test_spt_snr.png}
+\includegraphics[width=15cm]{figures/self_and_validation_test_spt_snr.png}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=18cm]{figures/self_and_validation_test_teff_snr.png}
-\includegraphics[width=18cm]{figures/self_and_validation_test_fe_h_snr.png}
+\includegraphics[width=15cm]{figures/self_and_validation_test_teff_snr.png}
+\includegraphics[width=15cm]{figures/self_and_validation_test_feh_snr.png}
 \end{center}
-\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:mann_validation}
+\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.
+Sources with strong rotational broadening (2M07171706−0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380) are identified with red circles.} \label{fig:mann_validation}
 \end{figure}
@@
 \plotone{figures/Spectral_Sequence_3.pdf}
-\caption{ Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
+\caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
 \end{figure}
@@
 %---------------------------
-\begin{figure}[!htb]
-\minipage{0.32\textwidth}
-  \includegraphics[width=\linewidth]{figures/test_cmd_teff.png}
-\endminipage\hfill
-\minipage{0.32\textwidth}
-  \includegraphics[width=\linewidth]{figures/test_cmd_feh.png}
-\endminipage\hfill
-\minipage{0.32\textwidth}%
-  \includegraphics[width=\linewidth]{figures/test_cmd_chisq.png}
-\endminipage
+\begin{figure}[ht]
+\plottwo{figures/test_cmd_chisq.png}{figures/test_fit_distribution.png}
+\caption{The distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE-Gaia cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$.
+Note that there are $>$1,000 sources above the main sequence, many of which are pre-main sequence stars and main sequence binaries.}
+\plottwo{figures/test_cmd_teff.png}{figures/test_cmd_feh.png}
+\caption{Gaia color-magnitude diagram of 13,915 sources, selected with $\chi^2<200,000$ out of the total test sample.}
 \end{figure}
@@
 
-\newpage
+% \newpage
 

### 2018-07-25 paper/draft1/draft1.tex
update figures

@@
 We then use 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 68K and 0.07 dex respectively.
-Finally we apply our model to determine spectral subtypes, temperatures and metallicities for 10,000+ M dwarfs in the APOGEE-Gaia cross match.
+Finally we apply our model to determine spectral subtypes, temperatures and metallicities for 10,000+ M dwarfs in the survey and compare our results to APOGEE pipeline measurements.
 
@@
 We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
-We note however, that the training sample is distributed heavily towards earlier type sources $<$M7, with only one M8 and one M9 source.
+We note however, that the training sample is distributed mostly towards earlier type sources, with only one M8 and one M9 source.
 
@@
 We would like to acknowledge Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Hans Walter Rix (MPIA), Anna-Christina Eilers (MPIA) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
-\color{red}Software: The Cannon, numpy, matplotlib...
-\color{black}
-
-\color{gcolor}{HOGG: SDSS acknowledgments... Grants and stuff...}\color{black}
+\color{gcolor}{HOGG: SDSS acknowledgments...
+Grants and stuff...}\color{black}
 
@@
 \caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.
-Sources with strong rotational broadening (2M07171706−0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380) are identified with red circles.} \label{fig:mann_validation}
+Sources with strong rotational broadening (2M07171706−0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.} \label{fig:mann_validation}
 \end{figure}
@@
 \begin{figure}[ht]
-\plottwo{figures/test_cmd_chisq.png}{figures/test_fit_distribution.png}
-\caption{The distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE-Gaia cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$.
-Note that there are $>$1,000 sources above the main sequence, many of which are pre-main sequence stars and main sequence binaries.}
-\plottwo{figures/test_cmd_teff.png}{figures/test_cmd_feh.png}
-\caption{Gaia color-magnitude diagram of 13,915 sources, selected with $\chi^2<200,000$ out of the total test sample.}
+\plotone{figures/cannon_chi_dist.png}
+\caption{The distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE-Gaia cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$ \textit{(right)}, and the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}).}
+\plotone{figures/safe_selection.png}
+\caption{Safe sample selection.}
+\end{figure}
+
+%---------------------------
+\begin{figure}[ht]
+\plotone{figures/aspcap_cannon_validation.png}
+\caption{Validation test}
+\plotone{figures/aspcap_cannon_label_hist.png}
+\caption{Distribution of temperatures and metallicities from The Cannon compared to the ASPCAP pipeline.}
 \end{figure}
@@
 
-% \startlongtable
+% \startlongtable{}
 % \begin{deluxetable*}{ccc|ccc|ccc}

### 2018-07-26 paper/draft1/draft1.tex
backup

@@
 The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars.
-Using a training set of 51 M dwarfs with spectral types ranging M0-M9 obtained from SDSS optical spectra, we demonstrate that The Cannon can infer spectral types to a precision of 0.6 types.
-We then use 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 68K and 0.07 dex respectively.
-Finally we apply our model to determine spectral subtypes, temperatures and metallicities for 10,000+ M dwarfs in the survey and compare our results to APOGEE pipeline measurements.
+Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 we train a two-parameter model precise to 68K and 0.07 dex respectively.
+We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that The Cannon can infer spectral types to a precision of 0.6 types.
+Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,084 M dwarfs in the survey and compare our results to APOGEE pipeline measurements, and find our results to be significantly more precise and accurate than ASPCAP, with scatters of $\sim80-100K$ comapred to color-temperature relations from several literature sources.
 
@@
 
-\keywords{stars: Mdwarfs -- stellar parameters}
+% \keywords{stars: Mdwarfs -- stellar parameters}
 
@@
 % \begin{enumerate}
-Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
+Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}.
 
@@
 
-% Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them:
-
 Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,0841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.
@@
 The launch of the (APOGEE; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
+\color{gcolor}{What about APOGEE's wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?}\color{black}
 
@@
 
-Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: models dont accurately reproduce data; empirical standards are relative not absolute.
+Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)?
+Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.
 
@@
 
-Why The Cannon approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016})
+Why The Cannon approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}) \color{gcolor}{For Hogg to articulate...}\color{black}
 
@@
 
-The physical parameter sample consists of 30 sources with reference labels distributed over $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 drawn from \citealt{Mann:2015} (hereafter M15).
+The physical parameter sample consists of 87 sources with reference labels distributed over $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample.
 The M15 catalog in total contains 183 sources selected from CONCH-SHELL \citep{Gaidos:2013} and with infrared spectra taken in follow up from the SpeX Spectograph.
@@
 Relative uncertainty from the callibration sample is measured to be \color{red}{$\sim$30K (check this)}\color{black}, which added in quadrature to the scatter from model comparison gives a total reported uncertainty of $\sim$60K in T$_{\rm eff}$.
+Derivation of [Fe/H] and reported uncertainty are...
 
@@
 We note that surface gravity (logg) and rotational velocity (vsini) parameters are not included as training labels because...
-
-\subsection{Test Sample Selection}
+\color{gcolor}{MANN: comments to add? particularly about the extension sample?}\color{black}
 
@@
 
-We use the aspcapStar spectra (see \citealt{Perez:2016} for a complete description of the pipeline).
-
-APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample.
+APOGEE data products: ap1D, apVisit, apStar, and aspcapStar.
+We use the aspcapStar spectra, which contains the continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
+SNR of stars in our sample...
 
@@
 \end{equation}
+We use the noise model is ${\rm noise}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
 
-Where the noise model is ${\rm noise}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, which the root mean sum of (1) the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and (2) the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
 Quadratic label vectors:
@@
 
-The training step is the stage
+The training step description...
+
+Test step description...
 
@@
 
-For each model, two tests of consistency were used to assess the validity of our derived test labels.
-The self-test comparing training labels and test labels where precision and bias of the model are calculated as the standard deviation and mean of the difference in training and test labels respectively.
+For each model, two tests of consistency were used to assess the validity of our derived test labels: (1) the self-test comparing training labels and test labels, and (2) the leave-one-out cross validation test, in which we train a model on all sources but \emph{n}, then apply the \emph{N-1} source trained model to get the labels for \emph{n}.
+Precision (scatter) and bias of the model for each test are calculated as the standard deviation and mean of the difference in training and test labels respectively.
 
-Some note on what lines vary significantly with parameter?
+Another mode of analysis we can look at with The Cannon is how the derivative of the model changes with respect to given training parameter, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters.
+As discussed further in sections \ref{subsec:mann_results} and \ref{subsec:west_results}, we show the derivative plots of each of our two models and compute the error of the derivative at each pixel using a jackknife statistic:
+	\[\sigma_{\theta,m}^2 = \frac{N-1}{N} \sum^N_{n=1} (\theta_{/n}^m - \theta^m)^2 \]
+where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample, indexed by n, $\theta$ is the coeffient vector trainined on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.
 
-\subsubsection{Spectral Type Model}
-We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
-We note however, that the training sample is distributed mostly towards earlier type sources, with only one M8 and one M9 source.
+\subsubsection{Temperature/Metallicity Model \label{subsec:mann_results}}
+For the physical parameter model we trained The Cannon on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 68K/0.07dex respectively, similar to the original training sample uncertainties of 60K/0.08dex.
+We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$Gyr) and likely slowly rotating sources.
+While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources does not significantly
 
-\subsubsection{Temperature/Metallicity Model}
-For the physical parameter model we trained The Cannon on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 68K/0.07dex respectively, similar to the precision of the training uncertainties of 60K/0.08dex.
-
-We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), however the model
+Some note on what lines vary significantly with parameter...
 
@@
 
+\subsubsection{Spectral Type Model \label{subsec:west_results}}
+We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
+We note however, that the training sample is distributed heavily towards earlier type sources, with only one M8 and one M9 source.
+
 \subsection{Test Sample}
-Some validation of our test label results...
+Out of the total APOGEE DR14 catalog of 277,371 sources, we selected 254,478 sources which were in the cross match of Gaia DR2 (\citealt{Brown:2018}) and applied Gaia color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources.
+From there we applied the following selection procedure to identify a "safe" sample of M dwarfs:
 
@@
 \item[-] The Cannon as a test of evaluating training parameter accuracy? 
+\item[-] Discussion of validation tests and ground truth
 \end{enumerate}
@@
 \begin{center}
-\includegraphics[width=15cm]{figures/self_and_validation_test_spt_snr.png}
+\plotone{figures/self_and_validation_test_spt_snr.png}
 \end{center}
-\caption{Consistency test for the West-trained spectral type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:west_validation}
+\caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).} \label{fig:west_validation}
 \end{figure}
@@
 \begin{figure}[ht]
-\begin{center}
-\includegraphics[width=15cm]{figures/self_and_validation_test_teff_snr.png}
-\includegraphics[width=15cm]{figures/self_and_validation_test_feh_snr.png}
-\end{center}
-\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.
+\plotone{figures/self_and_validation_test_teff_snr.png}
+\plotone{figures/self_and_validation_test_feh_snr.png}
+\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (left), and leave-one-out cross validation (right).
 Sources with strong rotational broadening (2M07171706−0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.} \label{fig:mann_validation}
@@
 \caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities.
-\textit{Third panel:} Derivative of the Cannon model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}.
+\textit{Third panel of each plot:} Derivative of the Cannon model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}.
 \label{fig:demo_teff}
@@
 \plotone{figures/cannon_chi_dist.png}
-\caption{The distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE-Gaia cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$ \textit{(right)}, and the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}).}
+\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE-Gaia cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$.
+We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as "safe".
+\label{fig:chi_dist}}
+\end{figure}
+
+\begin{figure}[ht]
+\plotone{figures/cmd_selection.png}
+\caption{Color-magnitude cuts.}
+\end{figure}
+
+\begin{figure}[ht]
 \plotone{figures/safe_selection.png}

### 2018-07-30 paper/draft1/draft1.tex
update draft

@@
-\documentclass[modern]{aastex62}
+\documentclass[preprint]{aastex62}
 
@@
 
-\title{APOGEE CATALOG OF 10,000+ M DWARFS MEASURED BY DATA-DRIVEN MODELS}
+\title{APOGEE CATALOG OF 10,000+ M DWARFS MEASURED WITH DATA-DRIVEN MODELS}
 
@@
 % \begin{enumerate}
-Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}.
+Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}.
 
@@
 Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.
+Several approaches to to determining physical parameters and the limitations:
 
- (1) model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\citealt{Rajpurohit:2014}, \citealt{Rajpurohit:2018}) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\citealt{Allard:2013})--models don't accurately reproduce spectra plus unquantified systematic errors;
+\begin{enumerate}
+\item model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\citealt{Rajpurohit:2014}, \citealt{Rajpurohit:2018}) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\citealt{Allard:2013})--models don't accurately reproduce spectra plus unquantified systematic errors;
 
- (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (\citealt{Casagrande:2008} estimated effective temperatures, bolometric luminosities, and metallicities using optical and infrared photometries) -- however we want to use \emph{all} of the data!
+\item model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (\citealt{Casagrande:2008} estimated effective temperatures, bolometric luminosities, and metallicities using optical and infrared photometries) -- however we want to leverage the full wavelength range and resolution of our data.
 
- (3) using empirical classifications/templates/standards - SPT, luminosity class - this is a relative system but does not provide an absolute standard for measuring physical parameters
+\item using empirical classifications/templates/standards - SPT, luminosity class - this is a relative system but does not provide an absolute standard for measuring physical parameters
+\end{enumerate}
 
@@
 
- (1) Can be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects;
+\begin{enumerate}
+\item Can be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects;
 
- (2) Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all
+\item Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all
 
- (3) better than template matching in that it is using less sparsely sampled known mdwarfs
+\item better than template matching in that it is using less sparsely sampled known mdwarfs
+\end{enumerate}
 
@@
 
-The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III (SDSS-III; \citealt{Eisenstein:2011}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
+The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (SDSS-III/IV \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
 Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
@@
 APOGEE is primarily targeted for bright stellar populations, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution.
-However numerous cool, main sequence sources have also been observed either accidentally, or as targets proposed by the M dwarf ancilliary survey \citep{Desphande:2013} (specific estimate based on Gaia?).
+However numerous cool, main sequence sources have also been observed either as targets proposed by the M dwarf ancilliary survey (1,350 sources selected by proper motions from the  \citealt{Desphande:2013}), or randomly by APOGEE, which we quantify in this study to be over 10,000 in total.
 
@@
 
-Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector \emph{l} and and coefficient vector \emph{$\theta$}.
+Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector \emph{l} and and coefficient vector \emph{$\theta$}.
 So for each star \emph{n}, at pixel wavelength \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
@@
 As discussed further in sections \ref{subsec:mann_results} and \ref{subsec:west_results}, we show the derivative plots of each of our two models and compute the error of the derivative at each pixel using a jackknife statistic:
-	\[\sigma_{\theta,m}^2 = \frac{N-1}{N} \sum^N_{n=1} (\theta_{/n}^m - \theta^m)^2 \]
-where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample, indexed by n, $\theta$ is the coeffient vector trainined on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.
+\begin{equation}
+	\sigma_{\theta,m}^2 = \frac{N-1}{N} \sum^N_{n=1} (\theta_{/n}^m - \theta^m)^2 
+\end{equation}
+where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample, indexed by \emph{n}, $\theta$ is the coeffient vector trainined on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.
 
@@
 For the physical parameter model we trained The Cannon on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 68K/0.07dex respectively, similar to the original training sample uncertainties of 60K/0.08dex.
-We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$Gyr) and likely slowly rotating sources.
-While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources does not significantly
+We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr) and likely slowly rotating sources.
+While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.
 
@@
 We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
-We note however, that the training sample is distributed heavily towards earlier type sources, with only one M8 and one M9 source.
+We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
+As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates The Cannon does a poor job at extrapolating to labels outside of the training sample space.
 
-\subsection{Test Sample}
+\color{gcolor}{Further assessment of the reliability of our model for late types?...}\color{black}
+
+\subsection{Test Sample \label{subsec:test_selection}} 
 Out of the total APOGEE DR14 catalog of 277,371 sources, we selected 254,478 sources which were in the cross match of Gaia DR2 (\citealt{Brown:2018}) and applied Gaia color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources.
-From there we applied the following selection procedure to identify a "safe" sample of M dwarfs:
+From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.
+Although our catalog reports the temperatures, metallicities, and spectral types for XXX sources out of 14,000 with $\chi^2<200,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
+
+\begin{enumerate}
+\item \textbf{Known contaminant cut:} We remove known pre-main sequence sources from \citealt{Cottaar:2014} (280 sources), and known binaries from \citealt{ElBadry:2018} and \citealt{Skinner:2018} (58 sources).
+
+\item \textbf{K dwarf cut:} In 2MASS photometry we select $M_K > 4.3$ to remove likely K dwarfs, which appear reddened in Gaia $BP$, $RP$ photometry.
+
+\item \textbf{Pre-main sequence, multiple system and subdwarf (?) cut:} in Gaia and 2MASS we apply $.7<J-K<1.1$ and points in the selction region within the points $\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ of ($BP-RP$, $M_G$) to cut remaining likely pre-main sequence, multiple sources and subdwarfs.
+
+\item \textbf{Quality of fit cut:} We apply a $\chi^2$ cut of less than 100,000.
+
+\item \textbf{Training bound selection:} Because extrapolation with The Cannon is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<[Fe/H]<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
+\end{enumerate}
+
+The final selection for our ``safe'' is displayed in Figure \ref{fig:safe_selection}.
+We also note that applying our model requires very little computational demand: the time to run a model on all 14,827 sources was two minutes on a personal laptop.
+
+In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of ASPCAP temperature values.
 
@@
 \begin{enumerate}
-\item[-] Brief summary of what we did and found
+\item[-] Brief summary of what we did and found{}
 
@@
 
-\item[-] Some discussion of whether the derivatives are sensible or surprising. \color{red}{Add derivative plots.}\color{black}
+\item[-] Some discussion of whether the derivatives are sensible or surprising.
 
 \item[-] The Cannon as a test of evaluating training parameter accuracy?
+\color{gcolor}{HOGG: Should this be precision, not accuracy?
+For example adding a constant factor (say adding +1000 to every metallicity training value) does not change the scatter of the test labels}\color{black}
+
 \item[-] Discussion of validation tests and ground truth
@@
 \acknowledgements
-We would like to acknowledge Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Hans Walter Rix (MPIA), Anna-Christina Eilers (MPIA) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
+We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
 \color{gcolor}{HOGG: SDSS acknowledgments...
@@
 \begin{figure}[ht]
-\begin{center}
-\plotone{figures/self_and_validation_test_spt_snr.png}
-\end{center}
-\caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).} \label{fig:west_validation}
-\end{figure}
-
-\begin{figure}[ht]
 \plotone{figures/self_and_validation_test_teff_snr.png}
@@
 
-%---------------------------
 \begin{figure}[ht]
-\plotone{figures/Spectral_Sequence_1.pdf}
-\plotone{figures/Spectral_Sequence_2.pdf}
-\plotone{figures/Spectral_Sequence_3.pdf}
-\caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
+\begin{center}
+\plotone{figures/self_and_validation_test_spt_snr.png}
+\end{center}
+\caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).} \label{fig:west_validation}
 \end{figure}
@@
 \begin{center}
+\includegraphics[width=12cm]{figures/Spectral_Sequence_1.png}
+\includegraphics[width=12cm]{figures/Spectral_Sequence_2.png}
+\includegraphics[width=12cm]{figures/Spectral_Sequence_3.png}
+\end{center}
+\caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
+\end{figure}
+
+%---------------------------
+\begin{figure}[ht]
+\begin{center}
 \includegraphics[width=16cm]{figures/derivative_jackknife_spt.png}
@@
 
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[width=16cm]{figures/derivative_jackknife_teff.png}
-\includegraphics[width=16cm]{figures/derivative_jackknife_feh.png}
-\end{center}
-\caption{Derivative plots for physical parameter model.
+% \begin{figure}[ht]
+% \begin{center}
+% \includegraphics[width=16cm]{figures/derivative_jackknife_teff.png}
+% \includegraphics[width=16cm]{figures/derivative_jackknife_feh.png}
+% \end{center}
+% \caption{Derivative plots for physical parameter model.
 \textit{Top:} Derivative of flux with repsect to temperature taken at the median training temperature, T$_{\rm eff}=3463K$.
 \textit{Bottom:} Derivative of flux with repsect to metallicity taken at the median training metallicity, [Fe/H]$=-0.03$ dex.} \label{fig:mann_derivative}
-\end{figure}
+% \end{figure}
 
-%---------------------------
 \begin{figure}[ht]
@@
 \plotone{figures/cmd_selection.png}
-\caption{Color-magnitude cuts.}
+\caption{Gaia and 2MASS color-magnitude cuts for 12,037 sources with $\chi^2<100,000$.
+Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by Gaia.
+\label{fig:cmd_selection}}
 \end{figure}
@@
 \plotone{figures/safe_selection.png}
-\caption{Safe sample selection.}
+\caption{Final sample selection with 10,084 M dwarfs based on criteria described in Section \ref{subsec:test_selection}. \label{fig:safe_selection}}
+\end{figure}
+
+\begin{figure}[ht]
+\plotone{figures/aspcap_cannon_label_hist.png}
+\caption{Distribution of temperatures and metallicities for 10,084 sources from The Cannon compared to the ASPCAP pipeline. \label{fig:aspcap_cannon_label_hist}}
 \end{figure}
@@
 \plotone{figures/aspcap_cannon_validation.png}
-\caption{Validation test}
-\plotone{figures/aspcap_cannon_label_hist.png}
-\caption{Distribution of temperatures and metallicities from The Cannon compared to the ASPCAP pipeline.}
+\caption{Validation test \label{fig:aspcap_cannon_validation}}
 \end{figure}

### 2018-07-30 paper/draft1/create_tables.py

@@ -1,27 +1,0 @@
-import pandas as pd
-import numpy as np
-import os
-
-if __name__ == 'main':
-
-	file = 'data_files/mann_results.csv'
-	cont = pd.read_csv(file)
-
-	des = cont['ID']
-	ra  = cont['RA']
-	dec = cont['DEC']
-	trn_teff = cont['TRAIN_TEFF']
-	tst_teff = cont['TEST_TEFF']
-	crv_teff = cont['CROSS_TEFF']
-	trn_feh = cont['TRAIN_FEH']
-	tst_feh = cont['TEST_FEH']
-	crv_feh = cont['CROSS_FEH']
-		
-	with open('tables/mann_results.txt', 'w') as wf:
-		for i in range(len(des)):
-			line = "{} & {} & {} & {} & {} & {} & {} & {} & {} \\".format(des[i], ra[i], dec[i], \
-				trn_teff[i], tst_teff[i], crv_teff[i], \
-				trn_feh[i], tst_feh[i], crv_feh[i])
-
-			print(line)
-			wf.write(line)

### 2018-07-30 paper/draft1/draft1.tex
update draft

@@
 
-APOGEE data products: ap1D, apVisit, apStar, and aspcapStar.
-We use the aspcapStar spectra, which contains the continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
+APOGEE data products: {\tt\string ap1D}, {\tt\string apVisit}, {\tt\string apStar}, and {\tt\string aspcapStar}.
+We use the {\tt\string aspcapStar} spectra, which contains the continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
 SNR of stars in our sample...
@@
 
-Some note on what lines vary significantly with parameter...
+\color{gcolor}{Some note on what lines vary significantly with parameter...}\color{black}
 
@@
 
-In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of ASPCAP temperature values.
+In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of ASPCAP temperature values, using a combination of 2MASS and visual band photometries.
+\color{gcolor}{Brief description of the color-teff relations from \citealt{Casagrande:2008}, \citealt{Boyajian:2013}, and \citealt{Mann:2015}...}\color{black}.
+
+To obtain visual band magnitudes for a set of sources, we cross-matched the 10,084 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from APOGEE.
+In Figure \ref{fig:aspcap_cannon_validation} we repeat the temperature validation tests for each of the three relations and compare the ASPCAP temperatures for the same sources.
+\color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}
 

### 2018-08-05 presentations/meeting_8_15.tex
clean folders

\documentclass[compress]{beamer}
\usetheme{sthlm}

%-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=
%        LOADING BEAMER PACKAGES
%-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=

\usepackage{
booktabs,
datetime,
dtk-logos,
graphicx,
multicol,
pgfplots,
ragged2e,
tabularx,
tikz,
wasysym,
animate
}

\pgfplotsset{compat=1.8}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage{newpxtext,newpxmath}
\usepackage{listings}

\lstset{ %
language=[LaTeX]TeX,
basicstyle=\normalsize\ttfamily,
keywordstyle=,
numbers=left,
numberstyle=\tiny\ttfamily,
stepnumber=1,
showspaces=false,
showstringspaces=false,
showtabs=false,
breaklines=true,
frame=tb,
framerule=0.5pt,
tabsize=4,
framexleftmargin=0.5em,
framexrightmargin=0.5em,
xleftmargin=0.5em,
xrightmargin=0.5em
}

%-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=
%        LOADING TIKZ LIBRARIES
%-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=

\usetikzlibrary{
backgrounds,
mindmap
}

%-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=
%        BEAMER OPTIONS
%-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=

\setbeameroption{show notes}

%-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=
%        BEAMER COMMANDS
%-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=

%-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=
%
%	PRESENTATION INFORMATION
%
%-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=

\title{Data-driven cannon Models for APOGEE M dwarfs}
\subtitle{\textcolor{sthlmDarkGreen}{Milky Way Group Meeting}}
\date{\small{16 August 2017}}
\author{\texttt{Jessica Birky - UC San Diego}}
\institute{\texttt{Advisor: David Hogg - NYU/MPIA/Simons}}

\hypersetup{
pdfauthor = {Egmon Pereira: @Egmon},      
pdfsubject = {Teologia, },
pdfkeywords = {},  
pdfmoddate= {D:\pdfdate},          
pdfcreator = {WriteLaTeX}
}

\begin{document}

\begin{frame}
	\titlepage
\end{frame}

\begin{frame}{Motivation}

\Large Data-driven models are physically informative
\begin{enumerate} \large
	\item[•] The Cannon allows us to systematically search for features/indices that are sensitive to physical parameters of interest
	\pause
	\item[•] Cannon successes:
	\begin{enumerate}
		\item[•] Stellar parameters $T_{eff}$, logg, [Fe/H] (Ness 2015)
		\begin{enumerate}
			\item[•] Currently implemented into the APOGEE DR14 pipeline
		\end{enumerate}
		\item[•] Survey 1 $\Rightarrow$ Survey 2 (LAMOST $\Rightarrow$ APOGEE, Ho 2015)
	\end{enumerate}
	\pause
	\item[•] Potential ideas for M dwarf regime:
	\begin{enumerate}
		\item[•] Optical $\Rightarrow$ Infrared (K band $\Rightarrow$ H band)
		\item[•] Magnetic activity (H$_{\alpha}$), any indices that would be interesting
	\end{enumerate}
\end{enumerate}

\end{frame}

\begin{frame}{Motivation}

\Large Potential with the \textcolor{sthlmDarkGreen}{APOGEE} survey
\begin{enumerate} \large
	\item[•] Specifications: H band (1.5 - 1.7 $\mu$m), R $\sim$ 22,000
	\item[•] aspcap pipeline does not attempt to model the M dwarf spectra
\end{enumerate}

\pause
\begin{enumerate} \large
	\item[$\Rightarrow$] There are many M dwarfs (1400+, Desphande 2013), most which are not labeled or classified
\end{enumerate}
\end{frame}

\begin{frame}{Background: The Cannon}
\Large 
Assumptions: 
\begin{enumerate} \large
	\item[•] Spectral with identical labels have identical looking spectra
	\item[•] Flux at each wavelength pixel varies smoothly with change in parameter
\end{enumerate}

\pause
Requirements:
\begin{enumerate} \large
	\item[•] \textcolor{sthlmDarkGreen}{Training Set:} set of objects with known parameters (ideally consistent and very accurate) 
	\item[•] \textcolor{sthlmDarkGreen}{Test Set:} all objects (with unknown or unreliable parameters) that lie within the parameter span of the training set
\end{enumerate}
\end{frame}

\begin{frame}{Background: The Cannon}
\Large 
Cannon Model:
\[\textcolor{sthlmDarkGreen}{f_{n\lambda} = \theta^{T} \cdot l_{n} + noise} \]

\pause
$\theta$: Spectral model coefficients (at each $\lambda$) \\
1D Cannon Model: 
\begin{enumerate} \large
	\item[•] $l_{n} = [1, SPT, SPT^{2}]$ 
\end{enumerate}
3D Cannon Model: 
\begin{enumerate} \large
	\item[•] $l_{n} = [1, Teff, log g, [Fe/H], Teff^{2}, Teff \cdot log g, Teff \cdot [Fe/H], log g^{2}, log g \cdot [Fe/H], [Fe/H]^{2}]$ 
\end{enumerate}
\end{frame}

\begin{frame}{Project Scope}
\large Constructing a training set
\begin{enumerate} \large
	\item[•] Literature sample: 44 sources from Simbad \textcolor{sthlmDarkGreen}{\{Teff, [Fe/H]\}}
	\begin{enumerate}
		\item[•] Inconsistent training labels, inconsistent results
	\end{enumerate}

	\pause
	\item[•] Rajpurohit 2017: 45 sources \textcolor{sthlmDarkGreen}{\{Teff, logg, [Fe/H]\}}, or \textcolor{sthlmDarkGreen}{\{Spectral type\}}
	\begin{enumerate}
		\item[•] Temperatures concentrated 3000-3300K range
		\item[•] Labels derived are inconsistent with training labels
	\end{enumerate}

	\pause
	\item[•] West 2011: 69/70,841 sources matched \textcolor{sthlmDarkGreen}{\{Spectral type\}}
	\begin{enumerate}
		\item[•] Most consistent training set
	\end{enumerate}

\end{enumerate}

\pause 
\textcolor{red}{Objective:} Spectral type all APOGEE M dwarfs using data-driven models

\pause
\textcolor{sthlmDarkGreen}{Broad scope:} we hope in some way data-driven models can be used to improve physical ones.
\end{frame}

\begin{frame}{Selecting the Test Set}
Sorting by photometry
\begin{enumerate} \small
	\item[•] All APOGEE sources vs. 340 known M dwarfs (West 2011, Rajpurohit 2017, Desphande 2013, Aganze's brown dwarf xmatch)
\end{enumerate}
\begin{figure}
	\includegraphics[width=.7\textwidth, keepaspectratio]{photos/dwarfs_aspcap} 
\end{figure}
\end{frame}

\begin{frame}{Selecting the Test Set}
Comparison: sources labeled vs. not labeled by aspcap:
\begin{figure}
	\includegraphics[height=.45\textheight]{photos/aspcap_by_teff}  
	\includegraphics[height=.45\textheight]{photos/aspcap_by_logg}  
\end{figure}
\textcolor{sthlmDarkGreen}{Left}: all APOGEE sources colored by aspcap \textcolor{sthlmDarkGreen}{Teff} \\
\textcolor{sthlmDarkGreen}{Right}: all APOGEE sources colored by aspcap \textcolor{sthlmDarkGreen}{logg} 
\end{frame}

\begin{frame}{Selecting the Test Set}
Sorting by spectra
\begin{figure}
	\includegraphics[width=\textwidth, keepaspectratio]{photos/templates.pdf} 
	\caption{M dwarf vs. M giant}
\end{figure}
\end{frame}

\begin{frame}{Selecting the Test Set}
Template sorting and SNR cut
\begin{figure}
	\includegraphics[width=.45\textwidth]{photos/known_mdwarfs_sorted.png} 
	\includegraphics[width=.6\textwidth]{photos/Chi_Squared_Scatter_SNR.pdf} 
\end{figure}

\pause
Proceedure (in progress): 1) Color cut, 2) Sort by comparing to M dwarf and M giant templates, 3) Cut SNR > 100
\end{frame}

\begin{frame}{West Sample: Training with Spectral Types}
Sample Specifications
\begin{enumerate} \small
	\item[•] \textcolor{sthlmDarkGreen}{51 sources M0-M9} with SNR > 40
	\item[•] Consistent SPT labels obtained from SDSS spectra 
	\begin{enumerate} 
		\item[•] using The Hammer + visual inspection 
	\end{enumerate}
	\item[•] Spectral Types: not the most physically informative parameter, but useful start to indentifying M dwarfs and for later modelling Teff and [Fe/H]
\end{enumerate}

\begin{figure}
	\includegraphics[width=.5\textwidth]{photos/west_spt_dist} 
	\includegraphics[width=.5\textwidth]{photos/west_color_mag} 
\end{figure}
\end{frame}

\begin{frame}{West Spectral sequence}
\begin{figure}
	\includegraphics[width=.42\textwidth]{photos/desphande_spec_seq} 
	\includegraphics[width=.57\textwidth]{photos/west_spec_seq} 
\end{figure}
\small
\textcolor{sthlmDarkGreen}{Left}: Desphande 2013; Spectral Types apporoximated by binned V-J color
\hangindent=7.8mm Highlights: Spectral type sensitive regions identified by Desphande

\textcolor{sthlmDarkGreen}{Right}: 9 selected spectral with types from West 2011
\end{frame}

\begin{frame}{West Spectral sequence}
\begin{figure}
	\includegraphics[width=.5\textwidth]{photos/bochanski_spec_seq} 
	\includegraphics[width=.5\textwidth]{photos/west_sdss_spec_seq} 
\end{figure}
\end{frame}

\begin{frame}{West-trained sample: results}
\small \textcolor{sthlmDarkGreen}{SPT Scatter:} \textcolor{red}{0.617}
\begin{figure}
	\includegraphics[width=.55\textwidth]{photos/west_self_test} 
	\includegraphics[width=.45\textwidth]{photos/west_crv} 
\end{figure}
\small \textcolor{sthlmDarkGreen}{Left}: Self-consistency test 
\footnotesize
\hangindent=15mm -- train on 51 spectra, re-derive labels for all 51 spectra \\
\small
\textcolor{sthlmDarkGreen}{Right}: Cross-validation test
\footnotesize
\hangindent=15mm -- leave out spectrum n, train on 50 spectra, derive label for spectrum n; repeat for all 51 spectra \\
\end{frame}

\begin{frame}{West-trained sample: results}
Training with magnetic activity (including EW of H$_{\alpha}$) \\
\small \textcolor{sthlmDarkGreen}{SPT Scatter:} \textcolor{red}{0.556}
\begin{figure}
	\includegraphics[width=.5\textwidth]{photos/west_2_1} 
	\includegraphics[width=.5\textwidth]{photos/west_2_2} 
\end{figure}
\small \textcolor{sthlmDarkGreen}{Left}: Teff Self-consistency test; \textcolor{sthlmDarkGreen}{Right}: H$_{\alpha}$ Self-consistency test
\end{frame}

\begin{frame}{West-trained models: Spectral sequence}
\begin{figure}
	\includegraphics[width=.5\textwidth]{photos/desphande_spec_seq} 
	\includegraphics[width=.52\textwidth]{photos/cannon_spec_seq} 
\end{figure}
\small \textcolor{sthlmDarkGreen}{Left}: Desphande 2013; \textcolor{sthlmDarkGreen}{Right}: West-trained Cannon models
\end{frame}

\begin{frame}{Spectral sequence: data vs. Cannon models}
\begin{figure}
	\includegraphics[width=\textwidth]{photos/1_Spectral_Sequence.pdf} 
\end{figure}
\end{frame}

\begin{frame}{Spectral sequence: data vs. Cannon models}
\begin{figure}
	\includegraphics[height=.45\textheight]{photos/2_Spectral_Sequence.pdf}  \\
	\includegraphics[height=.45\textheight]{photos/3_Spectral_Sequence.pdf}  
\end{figure}
\end{frame}

\begin{frame}{Rajpurohit-trained Cannon Models}
Sample specifications
\begin{enumerate} 
	\item[•] \textcolor{sthlmDarkGreen}{45 sources} with labels \textcolor{sthlmDarkGreen}{\{Teff, logg, [Fe/H], Spectral type\}}
	\item[•] Labels derived from BT-Settl model grid (Homeier, unpublished)
\end{enumerate}

2 label results: \textcolor{sthlmDarkGreen}{\{Teff, [Fe/H], Spectral type\}}
\begin{figure}
	\includegraphics[height=.4\textheight]{photos/raj_2_1}  
	\includegraphics[height=.4\textheight]{photos/raj_2_2}   
\end{figure}
\end{frame}

\begin{frame}{Rajpurohit-trained Cannon Models}
3 label results: \textcolor{sthlmDarkGreen}{\{Teff, logg, [Fe/H], Spectral type\}}
\begin{figure}
	\includegraphics[height=.4\textheight]{photos/raj_3_1}  \\
	\includegraphics[height=.4\textheight]{photos/raj_3_2}  
	\includegraphics[height=.4\textheight]{photos/raj_3_3} 
\end{figure}
\end{frame}

\begin{frame}{Summary}
We have consistent results for spectral-typing APOGEE M dwarfs. \\

Current {Teff, [Fe/H]} training set does not span parameter range well enough
\begin{enumerate} 
	\item[•] But features are modelled well and may be useful for precise RVs/finding binaries
\end{enumerate}

Fixing systematic issues within The Cannon/aspcap may improve model fits
\begin{enumerate} 
	\item[•] Cannon optimizer is often converging to a local minimum
	\item[•] Continuum normalization producing artificial spikes in the data
	\item[•] Removing outliers in polynomial interpolation
\end{enumerate}
\end{frame}

\section{Model Fits}

\begin{frame}{Model fits: 2M11091225-0436249}
Rajpurohit-trained model synthesized to \textcolor{sthlmDarkGreen}{Training Labels}
\begin{figure}
	\includegraphics[width=\textwidth]{photos/train_fit}   
\end{figure}
Rajpurohit-trained model synthesized to \textcolor{sthlmDarkGreen}{Test Labels}
\begin{figure}
	\includegraphics[width=\textwidth]{photos/test_fit}   
\end{figure}
\small
Notice $\chi^{2}$ of test fit is lower than $\chi^{2}$ of training fit -- there is a bug in The Cannon optimizer, but at least the model coefficients are not skewed at the edge of the training set.
\end{frame}

\section{Future Work}

\begin{frame}{Future Work} \large
Would ideally like to train on 3 stellar parameters (Teff, logg, [Fe/H]) -- constructing a complete training data set is the biggest challenge
\begin{enumerate} \normalsize
	\item[•] Ideally my training set would be larger in number and span a larger parameter range
	\item[•] APOGEE was not designed to study cool main-sequence stars (rather giant and hotter stars) - APOGEE lacks benchmark sources for M dwarfs
\end{enumerate}

\pause
Solutions:
\begin{enumerate} \normalsize
	\item[•] Model more M dwarfs with unknown parameters
	\item[•] Obtain more known sources (plans for SpeX/NIRSPEC observing with Adam Burgasser)
\end{enumerate}
\end{frame}

\begin{frame}{Future Work/Questions} \large
Obtaining accurate metallicities: \textcolor{sthlmDarkGreen}{FGK-M dwarf pairs}
\begin{enumerate} \normalsize
	\item[•] Use FGK pairs as the training set (but only 5 overlap with APOGEE -- xmatch Newton + Neves samples with DR14)
	\pause
	\item[•] Use empirical calibrations derived in these studies
\end{enumerate}
\small
Terrien 2012: [Fe/H] = .34(EW$_{Ca}$) + .407(EW$_{K}$) + .436(H$_{2}$O - H) - 1.485
\begin{figure}
	\includegraphics[width=\textwidth]{photos/bonfils_table}
	\caption{M dwarf comparative study of photometric metallicity scales (Neves 2011)}   
\end{figure}
\end{frame}

\begin{frame}{Training parameters} 
The Cannon (in principle) can be trained on any label/parameter.
Parameters we have:
	\begin{center}
	\begin{tabular}{ |c|c|c| } 
	 \hline
	 Parameter & West Sample & Rajpurohit Sample \\ 
	 \hline
	 \hline
	 Spectra & APOGEE 	& APOGEE \\ 
	 		 & SDSS 	&  \\ 
	 \hline 
	 Stellar & Spectral Type & Spectral Type \\
	 Parameters & & Teff, logg, [Fe/H] \\
	 \hline
	 Photometry & JHK, ugriz & JHK \\
	 \hline
	 Distance   & Photometric estimates & -- \\
	 \hline
	 EW & H$_{\alpha}$, H$_{\beta}$, H$_{\gamma}$ & -- \\
	 measurements & TiO$_{5}$, TiO$_{2}$  & -- \\
	 \hline
	\end{tabular}
	\end{center}
\end{frame}

\section{End}

\end{document}

### 2018-08-05 paper/draft.tex
add makefile

--- 2018-07-30 paper/draft1/draft1.tex
@@
 \subsection{Test Sample \label{subsec:test_selection}} 
-Out of the total APOGEE DR14 catalog of 277,371 sources, we selected 254,478 sources which were in the cross match of Gaia DR2 (\citealt{Brown:2018}) and applied Gaia color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources.
+Out of the total APOGEE DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of Gaia DR2 (\citealt{Brown:2018}) and applied Gaia color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources.
 From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.

### 2018-08-08 paper/draft.tex
commit all

@@
-\documentclass[preprint]{aastex62}
+\documentclass[modern]{aastex62}
 
@@
 \usepackage{morefloats}
+
+\newcommand{\apogee}{\textsl{APOGEE}}
+\newcommand{\thecannon}{\textsl{The Cannon}}
+\newcommand{\cannon}{\textsl{Cannon}}
+\newcommand{\aspcap}{\textsl{ASPCAP}}
+\newcommand{\gaia}{\textsl{Gaia}}
+\newcommand{\zmass}{\textsl{2MASS}}
+\newcommand{\apass}{\textsl{APASS}}
+\newcommand{\sdssiv}{\textsl{SDSS-IV}}
 
@@
 
-\title{APOGEE CATALOG OF 10,000+ M DWARFS MEASURED WITH DATA-DRIVEN MODELS}
+\title{TEMPERATURES AND METALLICITIES FOR 10,000+ M DWARFS IN THE \apogee\  SURVEY}
 
@@
 
-The Cannon \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (APOGEE; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mu m$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H], and chemical abundances) to precisions higher than the model-grid pipeline.
-The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars.
-Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 we train a two-parameter model precise to 68K and 0.07 dex respectively.
-We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that The Cannon can infer spectral types to a precision of 0.6 types.
-Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,084 M dwarfs in the survey and compare our results to APOGEE pipeline measurements, and find our results to be significantly more precise and accurate than ASPCAP, with scatters of $\sim80-100K$ comapred to color-temperature relations from several literature sources.
+\thecannon\ \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mathrm{\mu m}$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H]) to precisions higher than the model-grid pipeline.
+The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim3550$K \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars.
+Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$ dex we train a two-parameter model precise to 68K and 0.07 dex respectively.
+We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that \thecannon\ can infer spectral types to a precision of 0.6 types.
+Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in the survey, and compare our results to \apogee\ pipeline measurements, and find our results to be significantly more precise and less biased than \aspcap\, with scatters of $\sim80-100$K comapred to color-temperature relations from several literature sources.
 
@@
 
-The launch of the (APOGEE; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
-\color{gcolor}{What about APOGEE's wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?}\color{black}
+The launch of the (\apogee\; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
+\color{gcolor}{What about \apogee\'s wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?}\color{black}
 
-A number of M dwarf studies have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
+A number of M dwarf studies have already been conducted in \apogee\ (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
 \citealt{Gilhool:2018}; \citealt{Skinner:2018}.
 
-Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)?
+Why is there a problem labeling M dwarfs (and especially for high res \apogee\ spectra)?
 Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.
@@
 
-Why The Cannon approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}) \color{gcolor}{For Hogg to articulate...}\color{black}
+Why \thecannon\ approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}) \color{gcolor}{For Hogg to articulate...}\color{black}
 
@@
 
-\item Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all
+\item Make the point that \thecannon\ does not require us to have good M-dwarf models in the \apogee\ wavelength range at all
 
@@
 
-\subsection{APOGEE Survey}
+\subsection{\apogee\ Survey}
 % \begin{enumerate}
-% Describe the APOGEE data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).
+% Describe the \apogee\ data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).
 
-% \item[-] Some M dwarfs in APOGEE proposed as ancilliary targets (Blake/Covey sources), plus an unknown number observed unintentionally \citep{Desphande:2013}.
+% \item[-] Some M dwarfs in \apogee\ proposed as ancilliary targets (Blake/Covey sources), plus an unknown number observed unintentionally \citep{Desphande:2013}.
 % \end{enumerate}
 
-The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (SDSS-III/IV \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
+The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (SDSS-III/IV \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
+Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
 
-APOGEE is primarily targeted for bright stellar populations, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution.
-However numerous cool, main sequence sources have also been observed either as targets proposed by the M dwarf ancilliary survey (1,350 sources selected by proper motions from the  \citealt{Desphande:2013}), or randomly by APOGEE, which we quantify in this study to be over 10,000 in total.
+\apogee\ is primarily targeted for bright stellar populations, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution.
+However numerous cool, main sequence sources have also been observed either as targets proposed by the M dwarf ancilliary survey (1,350 sources selected by proper motions from the  \citealt{Desphande:2013}), or randomly by \apogee\, which we quantify in this study to be over 10,000 in total.
 
@@
 
-The Cannon is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:cannon}).
+\thecannon\ is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:cannon}).
 The model in this sense is used to \emph{transfer} labels from spectra which we know parameters for to those which we do not.
@@
 
-APOGEE data products: {\tt\string ap1D}, {\tt\string apVisit}, {\tt\string apStar}, and {\tt\string aspcapStar}.
+\apogee\ data products: {\tt\string ap1D}, {\tt\string apVisit}, {\tt\string apStar}, and {\tt\string aspcapStar}.
 We use the {\tt\string aspcapStar} spectra, which contains the continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
@@
 
-\subsection{The Cannon: A Data-Driven Approach}
+\subsection{\thecannon\: A Data-Driven Approach}
 
-The Cannon is a fully empirical model, which applies no use of line lists or radiative transfer models.
+\thecannon\ is a fully empirical model, which applies no use of line lists or radiative transfer models.
 Describe the assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label.
 
-The Cannon model can in principle be trained on any physical or empirical labels available.
-However, limitations to consider are that test labels from The Cannon are only accurate if the training labels accurate, and only precise if the training labels are measured using a consistent method.
-The Cannon does not extrapolate well outside the parameter space of the training sample.
+\thecannon\ model can in principle be trained on any physical or empirical labels available.
+However, limitations to consider are that test labels from \thecannon\ are only accurate if the training labels accurate, and only precise if the training labels are measured using a consistent method.
+\thecannon\ does not extrapolate well outside the parameter space of the training sample.
 
@@
 
-Another mode of analysis we can look at with The Cannon is how the derivative of the model changes with respect to given training parameter, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters.
+Another mode of analysis we can look at with \thecannon\ is how the derivative of the model changes with respect to given training parameter, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters.
 As discussed further in sections \ref{subsec:mann_results} and \ref{subsec:west_results}, we show the derivative plots of each of our two models and compute the error of the derivative at each pixel using a jackknife statistic:
@@
 \subsubsection{Temperature/Metallicity Model \label{subsec:mann_results}}
-For the physical parameter model we trained The Cannon on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 68K/0.07dex respectively, similar to the original training sample uncertainties of 60K/0.08dex.
+For the physical parameter model we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
 We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr) and likely slowly rotating sources.
@@
 \subsubsection{Spectral Type Model \label{subsec:west_results}}
-We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
+We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
 We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
-As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates The Cannon does a poor job at extrapolating to labels outside of the training sample space.
+As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates \thecannon\ does a poor job at extrapolating to labels outside of the training sample space.
 
@@
 \subsection{Test Sample \label{subsec:test_selection}} 
-Out of the total APOGEE DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of Gaia DR2 (\citealt{Brown:2018}) and applied Gaia color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources.
-From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.
-Although our catalog reports the temperatures, metallicities, and spectral types for XXX sources out of 14,000 with $\chi^2<200,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
+Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \textsl{Gaia DR2} (\citealt{Brown:2018}) and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources.
+From there we applied the following selection procedure to identify a sample of ``safe'' single, main-sequence M stars, with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
 
@@
 
-\item \textbf{K dwarf cut:} In 2MASS photometry we select $M_K > 4.3$ to remove likely K dwarfs, which appear reddened in Gaia $BP$, $RP$ photometry.
+\item \textbf{K dwarf cut:} In \zmass\ photometry we select $M_K > 4.3$ to remove likely K dwarfs, which appear reddened in \gaia\ $BP$, $RP$ photometry.
 
-\item \textbf{Pre-main sequence, multiple system and subdwarf (?) cut:} in Gaia and 2MASS we apply $.7<J-K<1.1$ and points in the selction region within the points $\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ of ($BP-RP$, $M_G$) to cut remaining likely pre-main sequence, multiple sources and subdwarfs.
+\item \textbf{Pre-main sequence, multiple system and subdwarf (?) cut:} in \gaia\ and \zmass\ we apply $.7<J-K<1.1$ and points in the selction region within the points $\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ of ($BP-RP$, $M_G$) to cut remaining likely pre-main sequence, multiple sources and subdwarfs.
 
@@
 
-\item \textbf{Training bound selection:} Because extrapolation with The Cannon is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<[Fe/H]<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
+\item \textbf{Training bound selection:} Because extrapolation with \thecannon\ is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<[Fe/H]<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
 \end{enumerate}
@@
 
-In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of ASPCAP temperature values, using a combination of 2MASS and visual band photometries.
+In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of \aspcap\ temperature values using a combination of \zmass\ and visual band photometries.
 \color{gcolor}{Brief description of the color-teff relations from \citealt{Casagrande:2008}, \citealt{Boyajian:2013}, and \citealt{Mann:2015}...}\color{black}.
+To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (\apass; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by \apass, and $JHK$ photometries from \apogee.
+In Figure \ref{fig:aspcap_cannon_validation} we repeat the temperature validation tests for each of the three relations and compare the \aspcap\ temperatures for the same sources.
+\color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}
 
-To obtain visual band magnitudes for a set of sources, we cross-matched the 10,084 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from APOGEE.
-In Figure \ref{fig:aspcap_cannon_validation} we repeat the temperature validation tests for each of the three relations and compare the ASPCAP temperatures for the same sources.
-\color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}
+The gradients in the color magnitude diagrams of Figure \ref{fig:safe_selection} demonstrate the expected negative correlation between empirical \gaia\ $BP-RP$ colors and West-trained spectral classifications with the physical Mann-trained temperatures.
+Spectral type decreases nonlinearly with temperature...
+negatively correlated with metallcity for spectral types hotter than M3
 
@@
 \begin{enumerate}
-\item[-] Brief summary of what we did and found{}
+\item[-] Brief summary of what we did and found...
 
@@
 
-\item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: The Cannon can produce precise results, but is only as accurate as its training set.}\color{black}
+\item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: \thecannon\ can produce precise results, but is only as accurate as its training set.}\color{black}
 
@@
 
-\item[-] The Cannon as a test of evaluating training parameter accuracy?
-\color{gcolor}{HOGG: Should this be precision, not accuracy?
-For example adding a constant factor (say adding +1000 to every metallicity training value) does not change the scatter of the test labels}\color{black}
+\item[-] \thecannon\ as a test of evaluating training parameter accuracy?
 
@@
 \caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities.
-\textit{Third panel of each plot:} Derivative of the Cannon model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}.
+\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}.
 \label{fig:demo_teff}
@@
 \caption{\textit{Top two panels of each plot:} Mann-trained model for varying metallicities and similar temperatures.
-\textit{Third panel:} Derivative of the Cannon model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$.} \label{fig:demo_feh}
+\textit{Third panel:} Derivative of \thecannon\ model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$.} \label{fig:demo_feh}
 \end{figure}
@@
 \end{center}
-\caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
+\caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of \apogee\ spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
 \end{figure}
@@
 \plotone{figures/cannon_chi_dist.png}
-\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE-Gaia cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$.
+\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee\-\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$.
 We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as "safe".
@@
 \plotone{figures/cmd_selection.png}
-\caption{Gaia and 2MASS color-magnitude cuts for 12,037 sources with $\chi^2<100,000$.
-Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by Gaia.
+\caption{\gaia\ and \zmass\ color-magnitude cuts for 12,037 sources with $\chi^2<100,000$.
+Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia.
 \label{fig:cmd_selection}}
@@
 \begin{figure}[ht]
-\plotone{figures/safe_selection.png}
-\caption{Final sample selection with 10,084 M dwarfs based on criteria described in Section \ref{subsec:test_selection}. \label{fig:safe_selection}}
+\gridline{\fig{figures/safe_selection_teff.png}{0.5\textwidth}{}
+          \fig{figures/safe_selection_feh.png}{0.5\textwidth}{}}
+\gridline{\fig{figures/safe_selection_spt.png}{0.5\textwidth}{}}
+\caption{Final sample selection with 10,311 M dwarfs based on criteria described in Section \ref{subsec:test_selection}. \label{fig:safe_selection}}
+\end{figure}
+
+\begin{figure}[ht]
+\plotone{figures/spt_vs_teff_vs_feh.png}
+\caption{West-trained spectral types vs.
+Mann-trained metallcities and temperatures.
+For each integer spectral type bin the mean temperatures and metallicities are shown as red error boxes.
+\label{fig:west_vs_mann_labels}}
 \end{figure}
@@
 \plotone{figures/aspcap_cannon_label_hist.png}
-\caption{Distribution of temperatures and metallicities for 10,084 sources from The Cannon compared to the ASPCAP pipeline. \label{fig:aspcap_cannon_label_hist}}
+\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the \aspcap\ pipeline. \label{fig:aspcap_cannon_label_hist}}
 \end{figure}
@@
 
-% \startlongtable{}
-% \begin{deluxetable*}{ccc|ccc|ccc}
-% \tablecaption{Cannon results for Temperature/Metallicity model \label{mann_results}}
-% \tabletypesize{\scriptsize}
-% \tablehead{
-% \multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Temperature (K)}} & \multicolumn{3}{c}{\underline{Metallicity (dex)}} \\
-% 	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
-%     & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}
-%     & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}  
-% } 
-% \startdata
-% 2M00182256+4401222 & 4.59542   & 44.02278  & 3603        & 3532       & 3519        & -0.3       & -0.27     & -0.24      \\
-% 2M00182549+4401376 & 4.60779   & 44.02734  & 3218        & 3527       & 3554        & -0.3       & -0.28     & -0.27      \\
-% 2M00285391+5022330 & 7.22488   & 50.37588  & 3207        & 3185       & 3188        & 0.11       & 0.05      & 0.02       \\
-% 2M00401001+0308050 & 10.04169  & 3.13473   & 3725        & 3783       & 3778        & 0.04       & 0.11      & 0.13       \\
-% 2M00580115+3919111 & 14.50482  & 39.31977  & 3157        & 3094       & 3101        & -0.07      & -0.01     & 0.03       \\
-% 2M01232542+1638384 & 20.8559   & 16.64401  & 3272        & 3218       & 3217        & 0.1        & 0.03      & 0.0        \\
-% 2M02001278+1303112 & 30.05402  & 13.05196  & 3080        & 3057       & 3074        & -0.16      & -0.1      & -0.01      \\
-% 2M02361535+0652191 & 39.06358  & 6.87167   & 3284        & 3249       & 3253        & -0.12      & -0.29     & -0.29      \\
-% 2M03044335+6144097 & 46.18104  & 61.73583  & 3500        & 3468       & 3467        & -0.12      & -0.26     & -0.27      \\
-% 2M03553688+5214291 & 58.90373  & 52.2414   & 3435        & 3381       & 3369        & -0.35      & -0.25     & -0.18      \\
-% 2M04125880+5236421 & 63.24499  & 52.61165  & 3100        & 3181       & 3200        & -0.04      & -0.01     & 0.01       \\
-% 2M04310001+3647548 & 67.75003  & 36.79855  & 3419        & 3371       & 3369        & 0.08       & 0.11      & 0.1        \\
-% 2M05030563+2122362 & 75.77354  & 21.37664  & 3155        & 3290       & 3503        & 0.06       & -0.15     & -0.36      \\
-% 2M05222053+3031097 & 80.58561  & 30.51933  & 3389        & 3417       & 3426        & 0.28       & 0.2       & 0.14       \\
-% 2M05312734-0340356 & 82.86417  & -3.67722  & 3801        & 3820       & 3854        & 0.49       & 0.49      & 0.37       \\
-% 2M05413073+5329239 & 85.37804  & 53.48987  & 3765        & 3752       & 3755        & 0.19       & 0.15      & 0.13       \\
-% 2M05420897+1229252 & 85.53833  & 12.48956  & 3250        & 3226       & 3235        & -0.22      & -0.34     & -0.31      \\
-% 2M06000351+0242236 & 90.01458  & 2.70657   & 3214        & 3164       & 3165        & 0.07       & 0.05      & 0.03       \\
-% 2M06011106+5935508 & 90.2961   & 59.59713  & 3340        & 3258       & 3255        & -0.09      & -0.05     & -0.03      \\
-% 2M06112610+1032599 & 92.8588   & 10.54998  & 3636        & 3711       & 3698        & -0.39      & -0.42     & -0.39      \\
-% 2M06544902+3316058 & 103.70411 & 33.26823  & 3448        & 3364       & 3359        & -0.02      & 0.04      & 0.06       \\
-% 2M07171706-0501031 & 109.32108 & -5.01754  & 3193        & 3159       & 3168        & -0.09      & -0.12     & -0.09      \\
-% 2M07272450+0513329 & 111.852   & 5.2263    & 3317        & 3285       & 3285        & -0.11      & -0.2      & -0.19      \\
-% 2M07444018+0333089 & 116.16744 & 3.55252   & 3217        & 3176       & 3174        & 0.23       & 0.26      & 0.15       \\
-% 2M08031949+5250387 & 120.83131 & 52.84402  & 3508        & 3622       & 3626        & -0.26      & -0.24     & -0.24      \\
-% 2M08083284+5304377 & 122.13679 & 53.07712  & 3398        & 3280       & 3273        & 0.0        & 0.08      & 0.09       \\
-% 2M08103429-1348514 & 122.64277 & -13.81427 & 3544        & 3517       & 3515        & 0.0        & 0.0       & 0.0        \\
-% 2M08155393+3136392 & 123.97459 & 31.61077  & 3753        & 3810       & 3821        & 0.16       & 0.07      & 0.03       \\
-% 2M08323599+4510175 & 128.14998 & 45.17153  & 3494        & 3505       & 3506        & -0.14      & -0.21     & -0.21      \\
-% 2M08370799+1507475 & 129.28319 & 15.12944  & 3596        & 3553       & 3552        & -0.17      & -0.26     & -0.27      \\
-% 2M08524084+2818589 & 133.17018 & 28.31634  & 3166        & 3213       & 3239        & 0.31       & 0.5       & 0.27       \\
-% 2M08585633+0828259 & 134.73479 & 8.47384   & 3200        & 3142       & 3147        & -0.04      & -0.09     & -0.08      \\
-% 2M09142298+5241125 & 138.595   & 52.68639  & 3920        & 3900       & 3879        & -0.01      & -0.02     & 0.01       \\
-% 2M09174473+4612246 & 139.4363  & 46.20685  & 3478        & 3548       & 3556        & 0.04       & -0.02     & -0.03      \\
-% 2M09214911+4330284 & 140.45447 & 43.50784  & 3170        & 3196       & 3206        & 0.16       & 0.23      & 0.19       \\
-% 2M09285333-0722148 & 142.22221 & -7.37095  & 3508        & 3445       & 3440        & -0.27      & -0.23     & -0.22      \\
-% 2M10112218+4927153 & 152.84254 & 49.45431  & 4131        & 4007       & 3946        & 0.24       & 0.22      & 0.23       \\
-% 2M10121768-0344441 & 153.07364 & -3.74563  & 3623        & 3619       & 3620        & 0.13       & 0.16      & 0.16       \\
-% 2M10193634+1952122 & 154.90117 & 19.87002  & 3370        & 3384       & 3389        & 0.15       & 0.24      & 0.21       \\
-% 2M10285555+0050275 & 157.23155 & 0.8411    & 3548        & 3517       & 3515        & -0.18      & -0.16     & -0.15      \\
-% 2M10331367+3409120 & 158.30687 & 34.15333  & 3500        & 3518       & 3518        & -0.15      & -0.19     & -0.2       \\
-% 2M10350859+3349499 & 158.7858  & 33.83055  & 3816        & 3906       & 3907        & -0.04      & 0.01      & 0.03       \\
-% 2M10453795+1833111 & 161.4081  & 18.55306  & 3673        & 3773       & 3780        & 0.14       & 0.13      & 0.13       \\
-% 2M10505201+0648292 & 162.71704 & 6.80826   & 3238        & 3258       & 3265        & 0.16       & 0.18      & 0.12       \\
-% 2M10520440+1359509 & 163.01762 & 13.99756  & 3298        & 3230       & 3228        & -0.13      & -0.11     & -0.07      \\
-% 2M11000432+2249592 & 165.01781 & 22.833    & 3463        & 3426       & 3423        & -0.11      & -0.1      & -0.09      \\
-% 2M11032023+3558117 & 165.83423 & 35.97054  & 3563        & 3682       & 3551        & -0.38      & -0.58     & -0.39      \\
-% 2M11052903+4331357 & 166.36992 & 43.52665  & 3619        & 3547       & 3525        & -0.37      & -0.34     & -0.3       \\
-% 2M11054316+1014093 & 166.42966 & 10.23605  & 3439        & 3395       & 3393        & -0.03      & -0.06     & -0.07      \\
-% 2M11091225-0436249 & 167.30106 & -4.60695  & 3661        & 3743       & 3748        & -0.03      & -0.09     & -0.1       \\
-% 2M11263270+2145262 & 171.63622 & 21.75719  & 3610        & 3642       & 3642        & -0.18      & -0.16     & -0.15      \\
-% 2M11273856+0358359 & 171.91071 & 3.97664   & 3920        & 3932       & 3963        & 0.43       & 0.44      & 0.31       \\
-% 2M11474074+0015201 & 176.9198  & 0.25553   & 3261        & 3235       & 3236        & 0.1        & 0.23      & 0.22       \\
-% 2M11474440+0048164 & 176.93491 & 0.80473   & 3192        & 3155       & 3157        & -0.02      & -0.06     & -0.06      \\
-% 2M11530522+1855480 & 178.27184 & 18.93002  & 3773        & 3760       & 3752        & -0.14      & -0.08     & -0.06      \\
-% 2M12192028+1323524 & 184.83449 & 13.39793  & 3926        & 3954       & 3945        & 0.0        & 0.13      & 0.18       \\
-% 2M12274471-0315006 & 186.93625 & -3.25018  & 3355        & 3399       & 3407        & 0.19       & 0.2       & 0.17       \\
-% 2M12390461+4702234 & 189.76958 & 47.03974  & 3492        & 3447       & 3445        & 0.05       & 0.06      & 0.06       \\
-% 2M12470102+4637334 & 191.75412 & 46.62587  & 3571        & 3543       & 3544        & 0.22       & 0.23      & 0.22       \\
-% 2M13113464+1306461 & 197.89433 & 13.11282  & 3459        & 3407       & 3402        & -0.1       & -0.03     & -0.01      \\
-% 2M13160127+1415504 & 199.00528 & 14.26403  & 3686        & 3768       & 3775        & -0.06      & -0.16     & -0.17      \\
-% 2M13295979+1022376 & 202.49895 & 10.37731  & 3727        & 3674       & 3669        & -0.09      & -0.13     & -0.13      \\
-% 2M13345147+3746195 & 203.71442 & 37.7721   & 3273        & 3247       & 3250        & 0.14       & 0.13      & 0.08       \\
-% 2M13400879+4346380 & 205.03676 & 43.77729  & 3332        & 3322       & 3325        & -0.19      & -0.18     & -0.14      \\
-% 2M13454354+1453317 & 206.43251 & 14.89139  & 3649        & 3688       & 3677        & -0.31      & -0.37     & -0.36      \\
-% 2M15290936+5724418 & 232.28886 & 57.41167  & 3660        & 3719       & 3723        & -0.09      & -0.14     & -0.14      \\
-% 2M15594729+4403595 & 239.94707 & 44.06654  & 3526        & 3614       & 3632        & 0.05       & 0.02      & 0.01       \\
-% 2M16252459+5418148 & 246.35249 & 54.30413  & 3475        & 3407       & 3400        & -0.35      & -0.25     & -0.22      \\
-% 2M16541202+1154529 & 253.5502  & 11.91464  & 3834        & 3760       & 3697        & -0.48      & -0.38     & -0.29      \\
-% 2M17093153+4340531 & 257.38138 & 43.68139  & 3284        & 3242       & 3242        & 0.2        & 0.25      & 0.17       \\
-% 2M17362594+6820220 & 264.10803 & 68.33932  & 3439        & 3487       & 3492        & 0.05       & -0.21     & -0.23      \\
-% 2M18161819+0131277 & 274.07575 & 1.5243    & 3391        & 3350       & 3347        & -0.25      & -0.19     & -0.17      \\
-% 2M18244689-0620311 & 276.1954  & -6.342    & 3412        & 3386       & 3384        & 0.01       & 0.07      & 0.08       \\
-% 2M18424666+5937499 & 280.69489 & 59.63015  & 3441        & 3386       & 3385        & -0.23      & -0.32     & -0.32      \\
-% 2M18424688+5937374 & 280.6958  & 59.62659  & 3345        & 3391       & 3400        & -0.3       & -0.35     & -0.33      \\
-% 2M18451027+0620158 & 281.29282 & 6.33774   & 3674        & 3684       & 3688        & 0.02       & 0.03      & 0.03       \\
-% 2M19142541+1755081 & 288.6059  & 17.91894  & 3725        & 3830       & 3835        & -0.08      & -0.09     & -0.09      \\
-% 2M19165526+0510086 & 289.23032 & 5.16909   & 3558        & 3536       & 3538        & 0.1        & 0.08      & 0.07       \\
-% 2M19500252+3235012 & 297.51022 & 32.58347  & 3385        & 3214       & 3215        & -0.18      & -0.31     & -0.28      \\
-% 2M19535443+4424541 & 298.47977 & 44.41508  & 2859        & 2801       & 2889        & -0.05      & 0.08      & 0.19       \\
-% 2M20032651+2952000 & 300.86075 & 29.86652  & 3144        & 3164       & 3178        & 0.06       & -0.03     & -0.09      \\
-% 2M20531977+6209156 & 313.33246 & 62.1545   & 3791        & 3728       & 3712        & -0.06      & 0.04      & 0.08       \\
-% 2M21065473+3844265 & 316.73013 & 38.74202  & 4021        & 4004       & 3922        & -0.22      & -0.23     & -0.11      \\
-% 2M22464980+4420030 & 341.70737 & 44.33405  & 3298        & 3236       & 3235        & 0.03       & 0.03      & 0.02       \\
-% 2M22563497+1633130 & 344.14517 & 16.55347  & 3720        & 3689       & 3688        & 0.21       & 0.29      & 0.28       \\
-% 2M23060482+6355339 & 346.52011 & 63.92622  & 3700        & 3770       & 3780        & 0.19       & 0.14      & 0.11       \\
-% 2M23315208+1956142 & 352.96732 & 19.93727  & 3353        & 3278       & 3274        & 0.03       & 0.05      & 0.05       \\
-% 2M23315244+1956138 & 352.9689  & 19.93719  & 3072        & 3266       & 3296        & 0.03       & 0.04      & 0.03       \\
-% 2M23491255+0224037 & 357.30206 & 2.40136   & 3646        & 3604       & 3560        & -0.45      & -0.46     & -0.4        
-% \enddata
-% \end{deluxetable*}
-% % \end{longrotatetable}
+\begin{longrotatetable}
+\begin{deluxetable*}{ccc|ccc|ccc|c}
+\tablecaption{Cannon results for Temperature/Metallicity model; see online journal for full table. \label{mann_results}}
+\tabletypesize{\scriptsize}
+\tablehead{
+\multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Temperature (K)}} & \multicolumn{3}{c}{\underline{Metallicity (dex)}} & \multicolumn{1}{c}{\underline{Model Fit}} \\
+	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
+    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}
+    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV} 
+    & \colhead{Test $\chi^2$} 
+} 
+\startdata
+2M00182256+4401222 & 4.59542   & 44.02278  & 3603        & 3538       & 3525        & -0.3       & -0.28     & -0.26      & 15304   \\
+2M00182549+4401376 & 4.60779   & 44.02734  & 3218        & 3528       & 3560        & -0.3       & -0.29     & -0.29      & 21273   \\
+2M00285391+5022330 & 7.22488   & 50.37588  & 3207        & 3190       & 3192        & 0.11       & 0.04      & 0.01       & 31790   \\
+2M00401001+0308050 & 10.04169  & 3.13473   & 3725        & 3777       & 3772        & 0.04       & 0.12      & 0.14       & 9830    \\
+2M00580115+3919111 & 14.50482  & 39.31977  & 3157        & 3100       & 3107        & -0.07      & -0.02     & 0.01       & 13337   \\
+2M01232542+1638384 & 20.8559   & 16.64401  & 3272        & 3225       & 3223        & 0.1        & 0.02      & -0.01      & 34797   \\
+2M02001278+1303112 & 30.05402  & 13.05196  & 3080        & 3059       & 3077        & -0.16      & -0.1      & -0.02      & 33210   \\
+2M02361535+0652191 & 39.06358  & 6.87167   & 3284        & 3241       & 3243        & -0.12      & -0.28     & -0.28      & 48445   \\
+2M03044335+6144097 & 46.18104  & 61.73583  & 3500        & 3466       & 3466        & -0.12      & -0.26     & -0.28      & 40739   \\
+2M03553688+5214291 & 58.90373  & 52.2414   & 3435        & 3386       & 3375        & -0.35      & -0.28     & -0.2       & 53870   \\
+2M04125880+5236421 & 63.24499  & 52.61165  & 3100        & 3183       & 3204        & -0.04      & -0.01     & 0.0        & 46007   \\
+2M04310001+3647548 & 67.75003  & 36.79855  & 3419        & 3371       & 3369        & 0.08       & 0.11      & 0.1        & 20602   \\
+2M05222053+3031097 & 80.58561  & 30.51933  & 3389        & 3423       & 3431        & 0.28       & 0.18      & 0.13       & 19072   \\
+2M05312734-0340356 & 82.86417  & -3.67722  & 3801        & 3814       & 3849        & 0.49       & 0.5       & 0.36       & 28209   \\
+2M05413073+5329239 & 85.37804  & 53.48987  & 3765        & 3751       & 3753        & 0.19       & 0.15      & 0.14       & 17842   \\
+2M05420897+1229252 & 85.53833  & 12.48956  & 3250        & 3220       & 3229        & -0.22      & -0.33     & -0.29      & 33868   \\
+2M06000351+0242236 & 90.01458  & 2.70657   & 3214        & 3170       & 3170        & 0.07       & 0.03      & 0.01       & 21044   \\
+2M06011106+5935508 & 90.2961   & 59.59713  & 3340        & 3259       & 3255        & -0.09      & -0.05     & -0.04      & 23329   \\
+2M06112610+1032599 & 92.8588   & 10.54998  & 3636        & 3720       & 3706        & -0.39      & -0.43     & -0.4       & 36898   \\
+2M06544902+3316058 & 103.70411 & 33.26823  & 3448        & 3368       & 3363        & -0.02      & 0.03      & 0.04       & 19412   \\
+2M07171706-0501031 & 109.32108 & -5.01754  & 3193        & 3175       & 3188        & -0.09      & -0.15     & -0.13      & 103000  \\
+2M07272450+0513329 & 111.852   & 5.2263    & 3317        & 3279       & 3279        & -0.11      & -0.18     & -0.17      & 23615   \\
+2M07444018+0333089 & 116.16744 & 3.55252   & 3217        & 3174       & 3167        & 0.23       & 0.25      & 0.17       & 34181   \\
+2M08031949+5250387 & 120.83131 & 52.84402  & 3508        & 3617       & 3620        & -0.26      & -0.23     & -0.23      & 17491   \\
+\nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata
+\enddata
+\end{deluxetable*}
+\end{longrotatetable}
 
-% \newpage
+%====================================
+\newpage
 
-% \startlongtable
-% \begin{deluxetable*}{ccc|ccc}
-% \tablecaption{Cannon results for Spectral Type model \label{west_results}}
-% \tabletypesize{\scriptsize}
-% \tablehead{
-% % \colhead{Designation} & \colhead{RA} & \colhead{DEC} & \colhead{$T_{\rm eff}$} & \colhead{[Fe/H]} \\
-% % \colhead{(2MASS)} & \colhead{(deg)} & \colhead{(deg)} & \colhead{(K)} & \colhead{(dex)}
-% \multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Spectral Type}} \\
-% 	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
-%     & \colhead{Training} & \colhead{Test} & \colhead{LOOCV} 
-% } 
-% \startdata
-% 2M03114974+0115158 & 47.95722  & 1.25442  & 1          & 0.5       & 0.8        \\
-% 2M03122509+0021585 & 48.1047   & 0.36621  & 7          & 7.3       & 6.8        \\
-% 2M03153109+3955161 & 48.87842  & 39.92346 & 2          & 0.3       & 0.2        \\
-% 2M03423963+0012102 & 55.66512  & 0.20284  & 4          & 4.8       & 4.9        \\
-% 2M03440466+0049332 & 56.01972  & 0.82795  & 2          & 0.4       & 0.4        \\
-% 2M04254701+1732407 & 66.44614  & 17.54455 & 5          & 5.7       & 5.7        \\
-% 2M04262170+1800009 & 66.59065  & 18.00024 & 5          & 5.7       & 5.8        \\
-% 2M06030979+2233260 & 90.78794  & 22.55808 & 1          & 1.5       & 1.6        \\
-% 2M07230160+4055426 & 110.75797 & 40.92653 & 4          & 1.6       & 1.4        \\
-% 2M08043605+5330565 & 121.14987 & 53.51342 & 1          & 0.5       & 0.6        \\
-% 2M08460531+1035309 & 131.5221  & 10.59194 & 7          & 7.1       & 6.9        \\
-% 2M08494430+1135327 & 132.43433 & 11.59268 & 0          & 0.6       & 0.9        \\
-% 2M09013404+0326287 & 135.39185 & 3.44134  & 0          & 1.7       & 2.1        \\
-% 2M09152918+4407461 & 138.87157 & 44.12966 & 3          & 4.5       & 4.6        \\
-% 2M09183649+2207022 & 139.65205 & 22.1173  & 3          & 3.0       & 3.0        \\
-% 2M09332262+2749021 & 143.34409 & 27.81724 & 3          & 3.4       & 3.3        \\
-% 2M09373349+5534057 & 144.38957 & 55.56828 & 6          & 6.8       & 6.7        \\
-% 2M10222276+1353383 & 155.59618 & 13.89607 & 3          & 2.3       & 2.2        \\
-% 2M10313413+3441535 & 157.89214 & 34.69814 & 1          & 2.5       & 2.6        \\
-% 2M10545293+4402506 & 163.72188 & 44.04674 & 4          & 4.9       & 5.0        \\
-% 2M11010030+2035056 & 165.25085 & 20.58252 & 2          & 1.0       & 1.0        \\
-% 2M11194647+0820356 & 169.94386 & 8.34307  & 8          & 7.7       & 7.1        \\
-% 2M11203609+0704135 & 170.15036 & 7.07021  & 6          & 6.5       & 6.5        \\
-% 2M11215827+0646252 & 170.4927  & 6.77355  & 5          & 6.1       & 6.1        \\
-% 2M11570299+2028436 & 179.26254 & 20.47885 & 2          & 3.6       & 3.7        \\
-% 2M12052751+1811555 & 181.36514 & 18.19978 & 1          & 3.0       & 3.2        \\
-% 2M12080810+3520281 & 182.03253 & 35.34103 & 7          & 7.1       & 7.0        \\
-% 2M12102152+4026185 & 182.5888  & 40.44021 & 4          & 3.1       & 2.9        \\
-% 2M12115216+3525458 & 182.9689  & 35.43142 & 4          & 1.5       & 1.4        \\
-% 2M12132318+3900529 & 183.34637 & 39.01468 & 7          & 7.4       & 6.9        \\
-% 2M12202436+2508225 & 185.1015  & 25.1396  & 6          & 6.6       & 6.5        \\
-% 2M12203634+2505351 & 185.15141 & 25.09318 & 4          & 3.6       & 3.4        \\
-% 2M12212701-0030560 & 185.36256 & -0.5156  & 0          & 1.7       & 1.9        \\
-% 2M12252076+2517082 & 186.3364  & 25.28559 & 7          & 7.7       & 7.6        \\
-% 2M12264289+0009059 & 186.67876 & 0.15164  & 0          & 2.3       & 2.5        \\
-% 2M12294016+2619567 & 187.4174  & 26.33245 & 7          & 4.4       & 4.1        \\
-% 2M12423245-0646077 & 190.63527 & -6.76885 & 2          & 3.0       & 3.0        \\
-% 2M12464541-0312524 & 191.68922 & -3.2146  & 4          & 4.3       & 4.3        \\
-% 2M12471099+1109566 & 191.79576 & 11.16566 & 3          & 3.8       & 3.9        \\
-% 2M12471599+2817425 & 191.81609 & 28.29519 & 1          & 0.9       & 1.0        \\
-% 2M12492657-0312032 & 192.36092 & -3.20097 & 3          & 3.7       & 3.8        \\
-% 2M12503440+4309482 & 192.64331 & 43.16345 & 4          & 4.9       & 4.9        \\
-% 2M12523816+1240586 & 193.15907 & 12.68297 & 4          & 4.8       & 4.8        \\
-% 2M12552141+4150425 & 193.83911 & 41.84522 & 3          & 3.7       & 3.8        \\
-% 2M12564117+4233175 & 194.17155 & 42.55485 & 3          & 3.7       & 3.8        \\
-% 2M13032161+4220407 & 195.83998 & 42.34466 & 2          & 3.4       & 3.4        \\
-% 2M13282079+3859398 & 202.08669 & 38.99444 & 2          & 3.5       & 3.6        \\
-% 2M13415860+1852278 & 205.49413 & 18.87437 & 4          & 4.7       & 4.8        \\
-% 2M13442970+5625445 & 206.12355 & 56.42916 & 5          & 6.4       & 6.4        \\
-% 2M13452385+1844222 & 206.34949 & 18.73947 & 4          & 4.3       & 4.3        \\
-% 2M13542316+2543152 & 208.59673 & 25.72078 & 4          & 5.0       & 5.1        \\
-% 2M13565847+3200294 & 209.24355 & 32.00805 & 4          & 4.7       & 4.7        \\
-% 2M14003808+3217085 & 210.15862 & 32.28572 & 3          & 4.4       & 4.5        \\
-% 2M14005977+3226109 & 210.2487  & 32.43644 & 7          & 6.9       & 6.6        \\
-% 2M14030704+5403356 & 210.77922 & 54.05996 & 5          & 5.8       & 5.9        \\
-% 2M14055603+0140310 & 211.48352 & 1.67535  & 2          & -0.4      & -0.5       \\
-% 2M14102952+4349477 & 212.62604 & 43.83093 & 3          & 3.1       & 3.1        \\
-% 2M14213414+0451188 & 215.39229 & 4.85523  & 0          & -1.7      & -1.3       \\
-% 2M14275590+5817281 & 216.98328 & 58.29095 & 3          & 3.9       & 4.0        \\
-% 2M14290377+4917403 & 217.26582 & 49.29448 & 4          & 5.5       & 5.5        \\
-% 2M14350795+4954080 & 218.78323 & 49.90218 & 0          & 1.1       & 1.4        \\
-% 2M14402293+1339230 & 220.09528 & 13.65579 & 7          & 7.6       & 7.5        \\
-% 2M14575608+2140337 & 224.48373 & 21.67605 & 1          & 1.6       & 1.7        \\
-% 2M14592059+3219293 & 224.83575 & 32.32482 & 1          & 0.4       & 0.5        \\
-% 2M15004819+2050499 & 225.20079 & 20.84723 & 3          & 2.6       & 2.5        \\
-% 2M15010818+2250020 & 225.28406 & 22.83383 & 9          & 8.5       & 6.1        \\
-% 2M15034431+2730327 & 225.93346 & 27.50773 & 3          & 0.3       & 0.2        \\
-% 2M15064250+3246099 & 226.6774  & 32.76876 & 2          & 4.3       & 4.4        \\
-% 2M15141711+0044474 & 228.5713  & 0.74654  & 7          & 5.8       & 5.4        \\
-% 2M15195523+0254040 & 229.98016 & 2.90112  & 0          & 0.4       & 0.8        \\
-% 2M15232142+3453119 & 230.8392  & 34.88651 & 4          & 4.8       & 4.9        \\
-% 2M15424779+4534214 & 235.69858 & 45.57528 & 3          & 2.1       & 2.1        \\
-% 2M15575183+0657068 & 239.4659  & 6.9519   & 4          & 5.5       & 5.5        \\
-% 2M15591175+0637577 & 239.79903 & 6.63268  & 3          & 4.4       & 4.4        \\
-% 2M16022228+0616424 & 240.59286 & 6.27846  & 3          & 2.3       & 2.3        \\
-% 2M16055380+2303058 & 241.47423 & 23.05165 & 7          & 6.9       & 6.7        \\
-% 2M16084282+2327406 & 242.17833 & 23.4614  & 4          & 4.0       & 4.0        \\
-% 2M16101653+4511112 & 242.56818 & 45.18571 & 0          & 0.3       & 0.6        \\
-% 2M16103232+2249116 & 242.63477 & 22.81992 & 7          & 7.3       & 7.2        \\
-% 2M16112965+2403267 & 242.87363 & 24.05741 & 2          & 3.1       & 3.2        \\
-% 2M16113384+2429072 & 242.89096 & 24.4854  & 4          & 5.5       & 5.6        \\
-% 2M16114429+2253294 & 242.93452 & 22.89149 & 5          & 4.3       & 4.2        \\
-% 2M16282719-0050428 & 247.11214 & -0.84372 & 3          & 2.1       & 2.1        \\
-% 2M16351077-0102156 & 248.79636 & -1.03922 & 3          & 0.5       & 0.4        \\
-% 2M16355440+3716239 & 248.97668 & 37.27332 & 3          & 3.4       & 3.5        \\
-% 2M16363188+3655364 & 249.13287 & 36.92676 & 2          & 1.1       & 1.1        \\
-% 2M16383117+3746233 & 249.62987 & 37.77329 & 4          & 4.8       & 4.9        \\
-% 2M16393137+3722527 & 249.88008 & 37.38122 & 0          & 3.2       & 3.3        \\
-% 2M16393824+3550337 & 249.90933 & 35.84268 & 2          & 3.1       & 3.2        \\
-% 2M16410962+3535407 & 250.29008 & 35.59465 & 3          & 3.0       & 2.9        \\
-% 2M16485665+3639205 & 252.23611 & 36.65564 & 3          & 3.0       & 2.9        \\
-% 2M17223434+5946359 & 260.64276 & 59.77846 & 4          & 2.5       & 2.4        \\
-% 2M19203934+3748048 & 290.16394 & 37.80135 & 0          & -2.0      & -2.1       \\
-% 2M19210112+3742134 & 290.25467 & 37.70374 & 0          & -2.2      & -2.4       \\
-% 2M21342546+0021006 & 323.60614 & 0.35243  & 2          & 1.1       & 1.0        \\
-% 2M23295172+4825211 & 352.4656  & 48.42256 & 1          & 0.3       & 0.4     
-% \enddata
-% \end{deluxetable*}
+\begin{longrotatetable}
+\begin{deluxetable*}{ccc|ccc|c}
+\tablecaption{Cannon results for Spectral Type model. \label{west_results}}
+\tabletypesize{\scriptsize}
+\tablehead{
+\multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Spectral Type}}
+	& \multicolumn{1}{c}{\underline{Model Fit}} \\
+	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
+    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV} 
+    & \colhead{Test $\chi^2$}
+} 
+\startdata
+2M03114974+0115158 & 47.957262  & 1.254404  & 1          & 0.9       & 1.4        & 78236   \\
+2M03122509+0021585 & 48.104563  & 0.366251  & 7          & 7.3       & 5.5        & 56504   \\
+2M03423963+0012102 & 55.66513   & 0.202859  & 4          & 3.9       & 3.9        & 29598   \\
+2M04262170+1800009 & 66.590421  & 18.000265 & 5          & 5.0       & 5.1        & 14084   \\
+2M09152918+4407461 & 138.871602 & 44.129498 & 3          & 3.6       & 3.7        & 17921   \\
+2M09183649+2207022 & 139.652051 & 22.117298 & 3          & 2.4       & 2.4        & 10591   \\
+2M09332262+2749021 & 143.344279 & 27.817253 & 3          & 2.8       & 2.7        & 17388   \\
+2M09373349+5534057 & 144.389577 & 55.568275 & 6          & 6.5       & 6.4        & 15335   \\
+2M10313413+3441535 & 157.892222 & 34.698212 & 1          & 2.0       & 2.3        & 21134   \\
+2M11194647+0820356 & 169.943658 & 8.343246  & 8          & 7.5       & 6.5        & 37803   \\
+2M11203609+0704135 & 170.1504   & 7.070432  & 6          & 6.1       & 5.9        & 18677   \\
+2M11570299+2028436 & 179.262465 & 20.4788   & 2          & 2.9       & 3.0        & 16205   \\
+2M12203634+2505351 & 185.15143  & 25.093107 & 4          & 3.4       & 3.1        & 85745   \\
+2M12212701-0030560 & 185.362567 & -0.515566 & 0          & 1.4       & 1.8        & 22646   \\
+2M12423245-0646077 & 190.635249 & -6.768827 & 2          & 2.4       & 2.5        & 18410   \\
+2M12464541-0312524 & 191.689212 & -3.214578 & 4          & 3.6       & 3.5        & 11561   \\
+2M12471099+1109566 & 191.795795 & 11.165737 & 3          & 3.1       & 3.0        & 12738   \\
+2M12492657-0312032 & 192.360749 & -3.200903 & 3          & 3.3       & 3.5        & 37756   \\
+2M12503440+4309482 & 192.643353 & 43.163414 & 4          & 4.0       & 4.0        & 19726   \\
+2M12523816+1240586 & 193.159002 & 12.682945 & 4          & 4.1       & 4.1        & 15526   \\
+2M12552141+4150425 & 193.839213 & 41.845161 & 3          & 3.1       & 3.1        & 29274   \\
+2M12564117+4233175 & 194.171558 & 42.554871 & 3          & 3.1       & 3.1        & 35083   \\
+2M13032161+4220407 & 195.840051 & 42.344654 & 2          & 2.7       & 2.8        & 14657   \\
+2M13415860+1852278 & 205.494169 & 18.874393 & 4          & 3.9       & 3.9        & 9475    \\
+2M13442970+5625445 & 206.123779 & 56.429039 & 5          & 5.9       & 6.0        & 16017   \\
+\nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata
+\enddata
+\end{deluxetable*}
+\end{longrotatetable}
+
+%====================================
+\newpage
 

### 2018-08-09 paper/draft.tex
test
physical parameter description

@@
 \newcommand{\sdssiv}{\textsl{SDSS-IV}}
+\newcommand{\teff}{T_{\rm eff}}
 
@@
 \affil{Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY 10027, USA}
+\altaffiltext{5}{Department of Physics and Astronomy, The University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
 
@@
 
-The physical parameter sample consists of 87 sources with reference labels distributed over $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample.
-The M15 catalog in total contains 183 sources selected from CONCH-SHELL \citep{Gaidos:2013} and with infrared spectra taken in follow up from the SpeX Spectograph.
-Effective temperatures are measured by comparison of BT-SETTL atmospheric models \citep{Allard:2011} to optical spectra.
-A subsample of 29 sources with measured angular diameters and bolometric fluxes from long-baseline optical interferometry are used to callibrate effective temperature measurements.
-Relative uncertainty from the callibration sample is measured to be \color{red}{$\sim$30K (check this)}\color{black}, which added in quadrature to the scatter from model comparison gives a total reported uncertainty of $\sim$60K in T$_{\rm eff}$.
-Derivation of [Fe/H] and reported uncertainty are...
+The physical parameter sample consists of 87 sources with reference labels distributed over $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
+The M15 catalog in total contains 183 sources and the extension sample another 500 primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog.
+All targets have low-res optical spectra from the SNIFS specgrograph \citep{Lantz:2004} and infrared spectra taken with the SpeX Spectograph \citep{Rayner:2003}, which have been combined to estaimte largely empirical bolometric fluxes.
+Effective temperatures have been estimated by comparing the SNIFS spectra to BT-SETTL atmospheric models \citep{Allard:2011}.
+A subsample of 29 sources with measured angular diameters from long-baseline optical interferometry \citep{Boyajian:2012} are used to calibrate the model comparison, including masking of spectral regions poorly reproduced by the model spectra \citep{Mann:2013c}.
+Based on difference between assigned \teff\ values and those from angular diameters, absolute uncertainty on \teff\ is esitmated to be 60K in T$_{\rm eff}$, although the relative uncertainty is likely a factor of $\simeq$2 better.
+%AWM: I realized I've never actaully sat down and rigorously measured the relative Teff uncertainty.
+The results from the Cannon actually suggest ~30 is not bad.
+I just did some quick tests, and I'm not prepared to put an absolute number on it.
+A \simeq 30 here should be safeish.
+
+Iron abundances ([Fe/H]) are assignied to the physical parameter sample based on the strenngth of metal-sensitive lines in the near-infrared SpeX spectra \citep{RojasAyala:2010} using the calibration from \citet{Mann:2013a}.
+The relation between these lines and an absolute [Fe/H] scale is calibrated using wide binaries containing and F, G, or K-type primary and an M dwarf companion, under the assumption that binaries formed from the same molecular cloud and therefore have the same metallicity \citep{Bonfils:2005}.
+Uncertainties are estimated to be $\simeq$0.08\,dex based on irreducable scatter in the empirical relation between selected lines and the assigned [Fe/H] from the primary star.
+As with \teff\ relative errors on [Fe/H] are smaller, estimated to be 0.04-0.06\,dex over most of the temperature and metallicity range considered here.
 
@@
 
-We note that surface gravity (logg) and rotational velocity (vsini) parameters are not included as training labels because...
-\color{gcolor}{MANN: comments to add? particularly about the extension sample?}\color{black}
+We note that surface gravity is not inclided as a training label.
+The reason for this is that for main-sequence M dwarfs, the parameter is almost entirely redundant with metallicity.
+Unlike their higher-mass counterparts, M dwarf properties do not change measurably over the age of the Universe.
+Hence perfect knowledge of abundances and \teff\ for an M dwarf should uniquely determine its position on a color-magnitude diagram as well as its other fundamental properties.
+While we only have [Fe/H], for the uncertainties considered here, lack of information about other abundances will only be important for stars with extreme abundance patters (e.g., Carbon stars).
+Since we have no such targets in the training set, we could not handle such objects.
+
+%%AWM: Feel free to edit this for clarity. Let me know if anything makes no sense. 
+%We note that rotational velocity (vsini) is not a parameter are not included as training labels because that would be hard... also it's probably not needed... I'll let someone else write that. -AWM
+%\color{gcolor}{}\color{black}
 

### 2018-08-10 paper/draft.tex
add some outline points

--- 2018-08-08 paper/draft.tex
@@
 \newcommand{\zmass}{\textsl{2MASS}}
-\newcommand{\apass}{\textsl{APASS}}
 \newcommand{\sdssiv}{\textsl{SDSS-IV}}
@@
 
-\thecannon\ \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mathrm{\mu m}$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H]) to precisions higher than the model-grid pipeline.
-The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim3550$K \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars.
-Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$ dex we train a two-parameter model precise to 68K and 0.07 dex respectively.
+\thecannon\ \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; near-infrared) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H]) to precisions higher than the model-grid pipeline.
+The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars.
+Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$ we train a two-parameter model precise to 68K and 0.07 dex respectively.
 We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that \thecannon\ can infer spectral types to a precision of 0.6 types.
-Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in the survey, and compare our results to \apogee\ pipeline measurements, and find our results to be significantly more precise and less biased than \aspcap\, with scatters of $\sim80-100$K comapred to color-temperature relations from several literature sources.
+Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in the survey and compare our results to \apogee\ pipeline measurements, and find our results to be significantly more precise and accurate than \aspcap, with scatters of $\sim80-100$K comapred to color-temperature relations from several literature sources.
 
@@
 % \begin{enumerate}
-Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}.
+Very low mass stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}.
+Relevance to exoplanet studies, and other reasons for scientific interest...
+
+Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,0841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.
+Citations to any other large M dwarf catalogs that have
+
+The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
+\color{gcolor}{What about \apogee\'s wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?
+What }\color{black}
 
 Define what parameters/classifications are important for M dwarfs.
-We want to estimate empirical parameters/classifications (SPT, etc.), physical parameters (such as Teff, logg, metallicity), which we will refer to as \emph{stellar labels}.
 
@@
 
-Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,0841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.
-
-The launch of the (\apogee\; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
-\color{gcolor}{What about \apogee\'s wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?}\color{black}
-
-A number of M dwarf studies have already been conducted in \apogee\ (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
-\citealt{Gilhool:2018}; \citealt{Skinner:2018}.
+A number of M dwarf studies have already been conducted in \apogee\, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements:
+\begin{enumerate}
+\item \citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs
+\item \citealt{Souto:2017} and \citealt{Souto:2018} -- modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining Teff/logg/metallicity + 13 elemental abundances
+\item \citealt{Desphande:2013} -- modeled rotational and radial velocities for 200+ M dwarfs, later extended by \citealt{Gilhool:2018} to 700+ M dwarfs
+\item \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
+\item \citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries
+\end{enumerate}
 
@@
 
-Why \thecannon\ approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}) \color{gcolor}{For Hogg to articulate...}\color{black}
+Why \thecannon\ approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016})
 
 \begin{enumerate}
+\item This paper will refer to any physical or empirical measurement of a star as \emph{stellar label}
+
 \item Can be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects;
@@
 Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \textsl{Gaia DR2} (\citealt{Brown:2018}) and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources.
-From there we applied the following selection procedure to identify a sample of ``safe'' single, main-sequence M stars, with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
+From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.
+Although our catalog reports the temperatures, metallicities, and spectral types for XXX sources out of 14,827 with $\chi^2<200,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
 
@@
 \color{gcolor}{Brief description of the color-teff relations from \citealt{Casagrande:2008}, \citealt{Boyajian:2013}, and \citealt{Mann:2015}...}\color{black}.
-To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (\apass; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by \apass, and $JHK$ photometries from \apogee.
+
+To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
 In Figure \ref{fig:aspcap_cannon_validation} we repeat the temperature validation tests for each of the three relations and compare the \aspcap\ temperatures for the same sources.
 \color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}
-
-The gradients in the color magnitude diagrams of Figure \ref{fig:safe_selection} demonstrate the expected negative correlation between empirical \gaia\ $BP-RP$ colors and West-trained spectral classifications with the physical Mann-trained temperatures.
-Spectral type decreases nonlinearly with temperature...
-negatively correlated with metallcity for spectral types hotter than M3
 
@@
 \begin{enumerate}
-\item[-] Brief summary of what we did and found...
+\item[-] Brief summary of what we did and found
 
@@
 For each integer spectral type bin the mean temperatures and metallicities are shown as red error boxes.
-\label{fig:west_vs_mann_labels}}
+\label{fig:west_vs_mann}}
 \end{figure}

### 2018-08-11 paper/draft.tex
fix affiliation format

@@
 \affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
-\affiliation{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
-\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726
+\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
+\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726
 Broadway, New York, NY 10003, USA}
@@
 \affil{Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY 10027, USA}
-\altaffiltext{5}{Department of Physics and Astronomy, The University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
+\affil{Department of Physics and Astronomy, The University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
 

### 2018-08-11 paper/draft.tex
some math text fixes

@@
 \newcommand{\sdssiv}{\textsl{SDSS-IV}}
-\newcommand{\teff}{T_{\rm eff}}
+\newcommand{\teff}{$T_{\rm eff}$}
 
@@
 
-\thecannon\ \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; near-infrared) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H]) to precisions higher than the model-grid pipeline.
-The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars.
+\thecannon\ \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; near-infrared) spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H]) to precisions higher than the model-grid pipeline.
+The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim$3550K \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars.
 Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$ we train a two-parameter model precise to 68K and 0.07 dex respectively.
@@
 The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (SDSS-III/IV \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
+Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (\teff, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
 
@@
 
-The physical parameter sample consists of 87 sources with reference labels distributed over $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
+The physical parameter sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
 The M15 catalog in total contains 183 sources and the extension sample another 500 primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog.
@@
 A subsample of 29 sources with measured angular diameters from long-baseline optical interferometry \citep{Boyajian:2012} are used to calibrate the model comparison, including masking of spectral regions poorly reproduced by the model spectra \citep{Mann:2013c}.
-Based on difference between assigned \teff\ values and those from angular diameters, absolute uncertainty on \teff\ is esitmated to be 60K in T$_{\rm eff}$, although the relative uncertainty is likely a factor of $\simeq$2 better.
+Based on difference between assigned \teff\ values and those from angular diameters, absolute uncertainty on \teff\ is esitmated to be 60K in \teff, although the relative uncertainty is likely a factor of $\simeq$2 better.
 %AWM: I realized I've never actaully sat down and rigorously measured the relative Teff uncertainty.

### 2018-08-11 paper/draft.tex
add boyajian ref

@@
 \newcommand{\zmass}{\textsl{2MASS}}
+\newcommand{\sdss}{\textsl{SDSS}}
 \newcommand{\sdssiv}{\textsl{SDSS-IV}}
@@
 
-The spectral type sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from SDSS.
+The spectral type sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from \sdss.
 For each source in the catalog, spectral types were determined both through an automated routine using The Hammer \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.

### 2018-08-13 paper/draft.tex
Update draft.tex
aspcap model details

@@
 \newcommand{\sdss}{\textsl{SDSS}}
+\newcommand{\lamost}{\textsl{LAMOST}}
 \newcommand{\sdssiv}{\textsl{SDSS-IV}}
@@
 
-\author[0000-0003-3654-1602]{Andrew Mann}
+\author[0000-0003-3654-1602]{Andrew W. Mann}
 \affil{Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY 10027, USA}
@@
 
-The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (SDSS-III/IV \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (\teff, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
+The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (\textsl{SDSS-III/IV} \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
+Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (\aspcap; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (\teff, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
+For low temperatures (\teff\ $<3500$K), the pipeline uses \textsl{MARCS} plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures (\teff\ $\geq3500$K) \textsl{ATLAS9} \citep{Castelli:2004} plane-parallel models are used.
 

### 2018-08-13 paper/draft.tex
add test result table

@@
 
+\begin{longrotatetable}
+\begin{deluxetable*}{cccccccccc}
+\tabletypesize{\scriptsize}
+\tablecaption{Temperatures/Metallicities (from Mann-trained model), and spectral types (from West-trained model) for the sample 10,311 APOGEE M dwarfs. \label{test_results_table}}
+\tablehead{\colhead{ID} & \colhead{GAIA ID} & \colhead{RA}  & \colhead{DEC} & \colhead{\teff} 
+& \colhead{[FE/H]}  & \colhead{SPT} & \colhead{$\chi^2$ Mann} & \colhead{$\chi^2$ West} }   
+\startdata
+2M00432365+0332534 & 413814124681430528  & 10.848967  & 3.548273   & 3417 & 0.18  & 2.965025337011196  & 7910  & 7912   & \nodata \\
+2M10382316+2457051 & 808931427773946112  & 159.596215 & 24.951373  & 3358 & -0.25 & 3.3883746019230694 & 19051 & 25518  & \nodata \\
+2M09364698+5602332 & 598955557617293440  & 144.195513 & 56.042461  & 3819 & 0.01  & 1.502939477318542  & 7981  & 6894   & \nodata \\
+2M12165176+1543232 & 1538883006881701120 & 184.215466 & 15.723067  & 3948 & 0.16  & 0.7043011669552124 & 25354 & 19983  & \nodata \\
+2M18501195+4809327 & 2037435595069990016 & 282.549868 & 48.15976   & 3899 & 0.24  & 0.9072125957953339 & 7205  & 7151   & \nodata \\
+2M07572656+4703318 & 682241540493118080  & 119.360502 & 47.05852   & 3636 & -0.17 & 2.415889813904016  & 38430 & 45625  & \nodata \\
+2M11083992+1112503 & 1536026200794333696 & 167.166344 & 11.213979  & 3949 & 0.1   & 0.838691017104972  & 21165 & 17506  & \nodata \\
+2M12135300+1419452 & 3695183295027782656 & 183.470871 & 14.32924   & 3824 & 0.13  & 1.3034741941684376 & 35018 & 21010  & \nodata \\
+2M12273816-0049437 & 3682549562826659456 & 186.909269 & -0.828949  & 3715 & 0.04  & 1.8358346554321001 & 14179 & 10656  & \nodata \\
+2M10513639-0150064 & 777824835073481984  & 162.901465 & -1.835273  & 3997 & -0.15 & 1.1662375752059775 & 80915 & 75157  & \nodata \\
+2M02442390-1352593 & 447040266556331008  & 41.099824  & -13.883403 & 3676 & -0.28 & 2.3153986696257274 & 41375 & 56704  & \nodata \\
+2M06281785+0940439 & 65792100044278400   & 97.074558  & 9.678234   & 3659 & -0.29 & 2.3027385897344397 & 78827 & 118883 & \nodata \\
+2M13441275+5538012 & 1551387099791031552 & 206.053693 & 55.633288  & 3856 & -0.1  & 1.60734737714035   & 50070 & 45895  & \nodata \\
+2M12473762+1107449 & 1251277737449848064 & 191.906593 & 11.129187  & 3244 & -0.42 & 3.977729233180596  & 30184 & 47020  & \nodata \\
+2M01274162+8435576 & 523638809262601216  & 21.924655  & 84.599233  & 3534 & 0.07  & 2.5591246570653587 & 15529 & 15113  & \nodata \\
+2M13464494+4034158 & 1247733770955275008 & 206.686943 & 40.571056  & 3731 & -0.16 & 2.0561869363413123 & 53196 & 50765  & \nodata \\
+2M00504161+0219408 & 2376325179183866240 & 12.673481  & 2.327846   & 3745 & 0.4   & 1.4277181066664462 & 76966 & 52035  & \nodata \\
+2M14494090+3201268 & 1292097793826595328 & 222.420454 & 32.023994  & 3228 & 0.0   & 3.8941824890924712 & 28733 & 38112  & \nodata \\
+2M16434969-0123117 & 1406510018874952960 & 250.957083 & -1.386602  & 3669 & 0.03  & 2.0451626775241616 & 22346 & 17186  & \nodata \\
+2M13444356+1655384 & 1565587017583214464 & 206.18144  & 16.927287  & 3652 & 0.13  & 2.0792998403425385 & 47828 & 40933  & \nodata \\
+2M08162613+3250515 & 3053990410555781120 & 124.108745 & 32.84736   & 3837 & -0.15 & 1.7923250253907956 & 30389 & 35103  & \nodata \\
+2M10414608+2435412 & 3758656623830055168 & 160.441928 & 24.594731  & 4020 & -0.19 & 1.0546884549827298 & 41857 & 36934  & \nodata \\
+2M14373854+5014502 & 1188221157913600384 & 219.410555 & 50.247523  & 3380 & -0.13 & 3.224522283084183  & 35397 & 41989  & \nodata \\
+2M13444572+0012349 & 1475140469529542656 & 206.19035  & 0.209419   & 3782 & -0.28 & 1.9568926444705497 & 12155 & 12707  & \nodata \\
+2M21275902+1332166 & 1817731181599810944 & 321.995889 & 13.537918  & 3997 & 0.18  & 0.6483601588618648 & 78906 & 67503 & \nodata \\
+\nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata
+\enddata
+\end{deluxetable*}
+\end{longrotatetable}
+
 %%==========================================================================================

### 2018-08-13 paper/draft.tex
add kinematic + phot data tables

--- 2018-08-13 paper/draft.tex
@@
 
+\begin{longrotatetable}
+\begin{deluxetable*}{cccccccccc}
+\tabletypesize{\scriptsize}
+\tablecaption{Temperatures/Metallicities (from Mann-trained model), and spectral types (from West-trained model) for the sample 10,311 APOGEE M dwarfs, as well as goodness of fit for the respective models.
+\label{test_results_table}}
+\tablehead{\colhead{APOGEE ID} & \colhead{Gaia\ ID} & \colhead{RA}  & \colhead{DEC} & \colhead{\teff} 
+& \colhead{[Fe/H]}  & \colhead{SPT} & \colhead{$\chi^2$ Mann} & \colhead{$\chi^2$ West} & \colhead{\nodata}}   
+\startdata
+2M00432365+0332534 & 413814124681430528  & 10.848967  & 3.548273   & 3417 & 0.18  & 3.0 & 7910  & 7912   & \nodata \\
+2M10382316+2457051 & 808931427773946112  & 159.596215 & 24.951373  & 3358 & -0.25 & 3.4 & 19051 & 25518  & \nodata \\
+2M09364698+5602332 & 598955557617293440  & 144.195513 & 56.042461  & 3819 & 0.01  & 1.5 & 7981  & 6894   & \nodata \\
+2M12165176+1543232 & 1538883006881701120 & 184.215466 & 15.723067  & 3948 & 0.16  & 0.7 & 25354 & 19983  & \nodata \\
+2M18501195+4809327 & 2037435595069990016 & 282.549868 & 48.15976   & 3899 & 0.24  & 0.9 & 7205  & 7151   & \nodata \\
+2M07572656+4703318 & 682241540493118080  & 119.360502 & 47.05852   & 3636 & -0.17 & 2.4 & 38430 & 45625  & \nodata \\
+2M11083992+1112503 & 1536026200794333696 & 167.166344 & 11.213979  & 3949 & 0.1   & 0.8 & 21165 & 17506  & \nodata \\
+2M12135300+1419452 & 3695183295027782656 & 183.470871 & 14.32924   & 3824 & 0.13  & 1.3 & 35018 & 21010  & \nodata \\
+2M12273816-0049437 & 3682549562826659456 & 186.909269 & -0.828949  & 3715 & 0.04  & 1.8 & 14179 & 10656  & \nodata \\
+2M10513639-0150064 & 777824835073481984  & 162.901465 & -1.835273  & 3997 & -0.15 & 1.2 & 80915 & 75157  & \nodata \\
+2M02442390-1352593 & 447040266556331008  & 41.099824  & -13.883403 & 3676 & -0.28 & 2.3 & 41375 & 56704  & \nodata \\
+2M06281785+0940439 & 65792100044278400   & 97.074558  & 9.678234   & 3659 & -0.29 & 2.3 & 78827 & 118883 & \nodata \\
+2M13441275+5538012 & 1551387099791031552 & 206.053693 & 55.633288  & 3856 & -0.1  & 1.6 & 50070 & 45895  & \nodata \\
+2M12473762+1107449 & 1251277737449848064 & 191.906593 & 11.129187  & 3244 & -0.42 & 4.0 & 30184 & 47020  & \nodata \\
+2M01274162+8435576 & 523638809262601216  & 21.924655  & 84.599233  & 3534 & 0.07  & 2.6 & 15529 & 15113  & \nodata \\
+2M13464494+4034158 & 1247733770955275008 & 206.686943 & 40.571056  & 3731 & -0.16 & 2.1 & 53196 & 50765  & \nodata \\
+2M00504161+0219408 & 2376325179183866240 & 12.673481  & 2.327846   & 3745 & 0.4   & 1.4 & 76966 & 52035  & \nodata \\
+2M14494090+3201268 & 1292097793826595328 & 222.420454 & 32.023994  & 3228 & 0.0   & 3.9 & 28733 & 38112  & \nodata \\
+2M16434969-0123117 & 1406510018874952960 & 250.957083 & -1.386602  & 3669 & 0.03  & 2.0 & 22346 & 17186  & \nodata \\
+2M13444356+1655384 & 1565587017583214464 & 206.18144  & 16.927287  & 3652 & 0.13  & 2.1 & 47828 & 40933  & \nodata \\
+2M08162613+3250515 & 3053990410555781120 & 124.108745 & 32.84736   & 3837 & -0.15 & 1.8 & 30389 & 35103  & \nodata \\
+2M10414608+2435412 & 3758656623830055168 & 160.441928 & 24.594731  & 4020 & -0.19 & 1.1 & 41857 & 36934  & \nodata \\
+2M14373854+5014502 & 1188221157913600384 & 219.410555 & 50.247523  & 3380 & -0.13 & 3.2 & 35397 & 41989  & \nodata \\
+2M13444572+0012349 & 1475140469529542656 & 206.19035  & 0.209419   & 3782 & -0.28 & 2.0 & 12155 & 12707  & \nodata \\
+2M21275902+1332166 & 1817731181599810944 & 321.995889 & 13.537918  & 3997 & 0.18  & 0.6 & 78906 & 67503 & \nodata \\
+\nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata
+\enddata
+\end{deluxetable*}
+\end{longrotatetable}
+
+%====================================
+\newpage
+
+\begin{longrotatetable}
+\begin{deluxetable*}{cccccccccc}
+\tabletypesize{\scriptsize}
+\tablecaption{Positions and 3D kinematics for 10,311 M dwarfs; parallaxes and proper motions from \gaia\ + radial velocities from \apogee.}
+\tablehead{\colhead{\nodata} & \colhead{$\varpi$} & \colhead{$\sigma_\varpi$} & \colhead{$\mu_\alpha$}  & \colhead{$\sigma_{\mu\alpha}$} & \colhead{$\mu_\delta$} & \colhead{$\sigma_{\mu\delta}$}  & \colhead{$v_{los}$} & \colhead{\nodata}}
+\startdata
+\nodata & 0.4095971279793963  & 0.018344701025907368 & 94.6092250382206    & 0.2326543993399578  & 21.139312054788142  & 0.10645732514148187  & -14.8169    & \nodata \\
+\nodata & 0.7907713253583666  & 0.05017993250995411  & -61.145608504420785 & 0.19267235934981508 & -9.292283439334902  & 0.191605833991544    & 30.0646     & \nodata \\
+\nodata & 2.437411282195209   & 0.06484152349097437  & -40.89370712477501  & 0.12447374577648344 & -27.01010381099313  & 0.12376849807276755  & -25.7591    & \nodata \\
+\nodata & 4.748695834113324   & 0.025884382476497582 & -39.804579591840216 & 0.0930132037970028  & -10.36148239898068  & 0.057746591921984974 & -7.28313    & \nodata \\
+\nodata & 0.3830610102451543  & 0.020679199019227986 & 8.768467380162365   & 0.03404660389361428 & 142.96097527921128  & 0.03222662578521854  & -67.5318    & \nodata \\
+\nodata & 1.2156916470683867  & 0.05850523666981387  & -34.02970707374436  & 0.5249923050897877  & -60.82222958247146  & 0.41961695655028974  & 27.5802     & \nodata \\
+\nodata & 3.2562530226439863  & 0.12635148166126747  & -14.18737757924032  & 0.08323990329349258 & 3.839844523005405   & 0.08709819792316187  & 1.19813     & \nodata \\
+\nodata & 1.1214563311692185  & 0.029138557090469957 & 3.8612047002174377  & 0.10975162725538208 & -6.6212605579145185 & 0.0625749242163525   & -12.022     & \nodata \\
+\nodata & 0.4337056924769619  & 0.03519634806417813  & 63.58287541662625   & 0.12216317236007974 & -30.844101370573288 & 0.06988132634903554  & 5.3206      & \nodata \\
+\nodata & 4.860871276791149   & 0.05422112071607727  & -38.48480428052528  & 0.06681092801786047 & -46.03340135581453  & 0.05013526442486122  & 5.74919     & \nodata \\
+\nodata & 0.3900296037356137  & 0.022457415816365067 & 48.72202372576884   & 0.09931020159590297 & -41.61843538517975  & 0.0939261318266816   & 56.5766     & \nodata \\
+\nodata & 0.5323417078167814  & 0.050227726038051694 & 37.19385535484577   & 0.15032160219430055 & -145.05520943675262 & 0.1318726138984278   & 59.6754     & \nodata \\
+\nodata & 6.37232292017776    & 0.0137946901051907   & 68.79697898216794   & 0.04264188775089677 & -84.07707033295478  & 0.03680789911821181  & -4.03993    & \nodata \\
+\nodata & 0.5021240985953683  & 0.029189592695681964 & -57.9436465188258   & 0.1911215986654146  & 16.919634543207177  & 0.10679177694312522  & -58.6918    & \nodata \\
+\nodata & 5.237224518979301   & 0.029752252992649626 & 23.771764671058826  & 0.07292367996213067 & -28.25823161266089  & 0.06344587895614055  & -19.6517    & \nodata \\
+\nodata & 3.3089435737776687  & 0.04711084402088657  & -41.45242638019648  & 0.3418156551313586  & -4.510012473794788  & 0.4388646698761767   & -19.7339    & \nodata \\
+\nodata & 5.072750731615924   & 0.04188258021899252  & 13.125640321529255  & 0.08925571521229253 & -27.369888362552718 & 0.05289115275902435  & -7.76221    & \nodata \\
+\nodata & 0.3185097583003641  & 0.014007935590281349 & 6.6397581995189     & 0.27566940326796724 & -20.42262436999549  & 0.4456232897623583   & -27.3047    & \nodata \\
+\nodata & 0.7753680397807614  & 0.01945512547484866  & 7.104977799189683   & 0.16560847226787015 & -5.534895098477748  & 0.11471267191919594  & 15.3108     & \nodata \\
+\nodata & 1.4924247624828302  & 0.028700623137541913 & -22.86896549011412  & 0.13153650570900005 & -6.5044933933639575 & 0.0918536954846762   & -1.93386    & \nodata \\
+\nodata & 0.3373361076141824  & 0.020908617609888186 & -23.51284581625556  & 0.1281836138293499  & -51.21200709697348  & 0.07993530283388152  & 67.9022     & \nodata \\
+\nodata & 0.9508792557786656  & 0.04214796908087648  & -23.92315869160129  & 0.10866864291003138 & -18.19346642794714  & 0.11087251384043836  & -0.941346   & \nodata \\
+\nodata & 2.6649412215771706  & 0.02575193557459232  & -11.654902067800947 & 0.0765638732657513  & 55.80843779696036   & 0.07016094187926156  & 15.9426     & \nodata \\
+\nodata & 1.0588602017295168  & 0.03531208113851025  & -33.94690128799879  & 0.14827664116823172 & -67.52530689866221  & 0.10435398589835812  & 46.9949     & \nodata \\
+\nodata & 0.18385583304779304 & 0.027656036948018537 & -11.332297533362624 & 0.04670812215256213 & -12.924309762298892 & 0.0502953286981639   & -13.1057   & \nodata \\
+\nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata
+\enddata
+\end{deluxetable*}
+\end{longrotatetable}
+
+%====================================
+\newpage
+
+\begin{longrotatetable}
+\begin{deluxetable*}{cccccccccc}
+\tabletypesize{\scriptsize}
+\tablecaption{\apogee\ $JHK$ and \gaia\ $G$, $BP$, $RP$ photometric measurements.}
+\tablehead{\colhead{\nodata} & \colhead{$J$} & \colhead{$H$} & \colhead{$K$} & \colhead{$G$} & \colhead{$BP$} & \colhead{$RP$}}   
+\startdata
+\nodata & 12.83199977874756  & 12.175999641418455 & 11.96500015258789  & 13.415744          & 14.226433           & 12.53435            \\
+\nodata & 14.255999565124512 & 13.72599983215332  & 13.40999984741211  & 12.228349          & 12.702468           & 11.614723           \\
+\nodata & 11.791999816894533 & 11.104999542236328 & 10.967000007629396 & 10.691218          & 11.139357           & 10.104778           \\
+\nodata & 13.706000328063965 & 13.013999938964846 & 12.895999908447266 & 12.420532          & 12.90665            & 11.800449           \\
+\nodata & 11.616999626159668 & 10.946000099182127 & 10.751999855041504 & 12.908026          & 13.56541            & 12.142124           \\
+\nodata & 11.312999725341795 & 10.663999557495117 & 10.508000373840332 & 10.826382          & 10.85745            & 10.774274           \\
+\nodata & 13.128999710083008 & 12.498000144958494 & 12.32699966430664  & 14.534789          & 15.185692           & 13.744579           \\
+\nodata & 13.538000106811525 & 12.902999877929688 & 12.762999534606935 & 14.004284          & 14.160915           & 13.288598           \\
+\nodata & 12.69499969482422  & 12.036999702453613 & 11.86299991607666  & 14.93699           & 15.395644           & 14.318204           \\
+\nodata & 12.53499984741211  & 11.843999862670898 & 11.732000350952148 & 10.015749          & 10.34129            & 9.559692            \\
+\nodata & 12.536999702453613 & 11.928000450134276 & 11.682000160217285 & 13.735478          & 14.700515           & 12.765027           \\
+\nodata & 11.274999618530273 & 10.717000007629396 & 10.515000343322754 & 11.817871          & 12.467115           & 11.063624           \\
+\nodata & 10.21399974822998  & 9.600000381469727  & 9.407999992370604  & 13.178638          & 13.876391           & 12.384415           \\
+\nodata & 14.201000213623047 & 13.718000411987305 & 13.46500015258789  & 12.388862          & 12.901998           & 11.731835           \\
+\nodata & 13.104000091552734 & 12.43000030517578  & 12.256999969482422 & 8.78063            & 8.920993            & 8.569135            \\
+\nodata & 11.628999710083008 & 10.996000289916992 & 10.802000045776367 & 15.246564          & 16.136223           & 14.321473           \\
+\nodata & 12.312999725341795 & 11.618000030517578 & 11.38599967956543  & 11.829312          & 12.26109            & 11.258054           \\
+\nodata & 12.07800006866455  & 11.48799991607666  & 11.237000465393066 & 13.222831          & 13.763299           & 12.547942           \\
+\nodata & 13.682000160217285 & 12.944000244140625 & 12.72700023651123  & 12.266772          & 12.872325           & 11.548589           \\
+\nodata & 13.501999855041506 & 12.85099983215332  & 12.64900016784668  & 10.416036          & 10.922305           & 9.781987            \\
+\nodata & 13.95300006866455  & 13.29300022125244  & 13.10099983215332  & 13.099073          & 13.74368            & 12.342266           \\
+\nodata & 14.246000289916992 & 13.595000267028807 & 13.397000312805176 & 11.854821          & 12.39487            & 11.183031           \\
+\nodata & 13.241999626159668 & 12.670999526977539 & 12.432000160217285 & 13.945971          & 14.459273           & 13.295053           \\
+\nodata & 14.10200023651123  & 13.41800022125244  & 13.29800033569336  & 11.466022          & 12.018276           & 10.803698           \\
+\nodata & 12.576000213623047 & 11.892000198364258 & 11.701000213623047 & 14.51524           & 15.042348           & 13.839335    \\
+\nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata
+\enddata
+\end{deluxetable*}
+\end{longrotatetable}
+
 %%==========================================================================================

### 2018-08-13 paper/draft.tex
add table units

@@
 \label{test_results_table}}
-\tablehead{\colhead{APOGEE ID} & \colhead{Gaia\ ID} & \colhead{RA}  & \colhead{DEC} & \colhead{\teff} 
-& \colhead{[Fe/H]}  & \colhead{SPT} & \colhead{$\chi^2$ Mann} & \colhead{$\chi^2$ West} & \colhead{\nodata}}   
+\tablehead{\colhead{APOGEE ID} & \colhead{Gaia\ ID} & \colhead{RA}  & \colhead{DEC} & \colhead{\teff} & \colhead{[Fe/H]}  & \colhead{SPT} & \colhead{$\chi^2$ Mann} & \colhead{$\chi^2$ West} & \colhead{\nodata} \\
+ &  & (deg)  & (deg) & (K) & (dex) &  &  &  & }   
 \startdata
@@
 2M14373854+5014502 & 1188221157913600384 & 219.410555 & 50.247523  & 3380 & -0.13 & 3.2 & 35397 & 41989  & \nodata \\
-2M13444572+0012349 & 1475140469529542656 & 206.19035  & 0.209419   & 3782 & -0.28 & 2.0 & 12155 & 12707  & \nodata \\
-2M21275902+1332166 & 1817731181599810944 & 321.995889 & 13.537918  & 3997 & 0.18  & 0.6 & 78906 & 67503 & \nodata \\
 \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata
@@
 \tablecaption{Positions and 3D kinematics for 10,311 M dwarfs; parallaxes and proper motions from \gaia\ + radial velocities from \apogee.}
-\tablehead{\colhead{\nodata} & \colhead{$\varpi$} & \colhead{$\sigma_\varpi$} & \colhead{$\mu_\alpha$}  & \colhead{$\sigma_{\mu\alpha}$} & \colhead{$\mu_\delta$} & \colhead{$\sigma_{\mu\delta}$}  & \colhead{$v_{los}$} & \colhead{\nodata}}
+\tablehead{\colhead{\nodata} & \colhead{$\varpi$} & \colhead{$\sigma_\varpi$} & \colhead{$\mu_\alpha$}  & \colhead{$\sigma_{\mu\alpha}$} & \colhead{$\mu_\delta$} & \colhead{$\sigma_{\mu\delta}$}  & \colhead{$v_{los}$} & \colhead{\nodata} \\
+ & (mas) & (mas) & (mas/yr)  & (mas/yr) & (mas/yr) & (mas/yr)  & (km/s) & \colhead{\nodata}}   
 \startdata
@@
 \nodata & 2.6649412215771706  & 0.02575193557459232  & -11.654902067800947 & 0.0765638732657513  & 55.80843779696036   & 0.07016094187926156  & 15.9426     & \nodata \\
-\nodata & 1.0588602017295168  & 0.03531208113851025  & -33.94690128799879  & 0.14827664116823172 & -67.52530689866221  & 0.10435398589835812  & 46.9949     & \nodata \\
-\nodata & 0.18385583304779304 & 0.027656036948018537 & -11.332297533362624 & 0.04670812215256213 & -12.924309762298892 & 0.0502953286981639   & -13.1057   & \nodata \\
 \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata
@@
 \nodata & 13.241999626159668 & 12.670999526977539 & 12.432000160217285 & 13.945971          & 14.459273           & 13.295053           \\
-\nodata & 14.10200023651123  & 13.41800022125244  & 13.29800033569336  & 11.466022          & 12.018276           & 10.803698           \\
-\nodata & 12.576000213623047 & 11.892000198364258 & 11.701000213623047 & 14.51524           & 15.042348           & 13.839335    \\
 \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata

### 2018-08-13 paper/draft.tex
change table description
title abstract acks

@@
 \newcommand{\sdssiv}{\textsl{SDSS-IV}}
+\newcommand{\tess}{\textsl{TESS}}
 \newcommand{\teff}{$T_{\rm eff}$}
@@
 
-\title{TEMPERATURES AND METALLICITIES FOR 10,000+ M DWARFS IN THE \apogee\  SURVEY}
+\title{TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY}
 
@@
 \begin{abstract}
-
-\thecannon\ \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; near-infrared) spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H]) to precisions higher than the model-grid pipeline.
-The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim$3550K \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars.
-Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$ we train a two-parameter model precise to 68K and 0.07 dex respectively.
-We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that \thecannon\ can infer spectral types to a precision of 0.6 types.
-Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in the survey and compare our results to \apogee\ pipeline measurements, and find our results to be significantly more precise and accurate than \aspcap, with scatters of $\sim80-100$K comapred to color-temperature relations from several literature sources.
-
+\thecannon\ is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; near-infrared) spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H]) to precisions higher than the model-grid pipeline.
+The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim$3500K motivates the extension of this approach to M dwarf stars, but with new training labels.
+Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 68\,K and 0.07\,dex respectively.
+We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, and demonstrate that \thecannon\ acting on \apogee\ spectra can infer spectral types with predictive accuracy of 0.6 types.
+The trained models are employed to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in \apogee.
+We validate our results using photometric temperature indicators and the \gaia\ color--magnitude diagram.
+As a side result, we determine empirical relationships between spectral type and temperature and metallicity.
+These results have implications for the \tess\ Mission and other projects that involve large samples of M dwarfs.
 \end{abstract}
 
-% \keywords{stars: Mdwarfs -- stellar parameters}
+\keywords{
+surveys
+ ---
+stars: Mdwarfs
+ ---
+stellar parameters
+ ---
+methods:~data~analysis
+}
 
@@
 We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
-\color{gcolor}{HOGG: SDSS acknowledgments...
-Grants and stuff...}\color{black}
+
+\color{gcolor}{HOGG: grants etc}
+
+\color{gcolor}{JB: SDSS acknowledgment}
+
+\color{gcolor}{JB: Gaia acknowledgment}
 
@@
 \tabletypesize{\scriptsize}
-\tablecaption{\apogee\ $JHK$ and \gaia\ $G$, $BP$, $RP$ photometric measurements.}
+\tablecaption{\apogee\ $JHK$ and \gaia\ $G$, $BP$, $RP$ relative photometric magnitudes.}
 \tablehead{\colhead{\nodata} & \colhead{$J$} & \colhead{$H$} & \colhead{$K$} & \colhead{$G$} & \colhead{$BP$} & \colhead{$RP$}}   

### 2018-08-13 paper/draft.tex
shorttitle
typo
sdss acknowlegment

@@
 
-\color{gcolor}{HOGG: grants etc}
+\color{gcolor}{HOGG: grants etc}\color{black}
 
-\color{gcolor}{JB: SDSS acknowledgment}
+Funding for the Sloan Digital Sky Survey IV has been provided by the Alfred P.
+Sloan Foundation, the U.S.
+Department of Energy Office of Science, and the Participating Institutions.
+SDSS-IV acknowledges support and resources from the Center for High-Performance Computing at
+the University of Utah. The SDSS web site is www.sdss.org.
 
-\color{gcolor}{JB: Gaia acknowledgment}
+SDSS-IV is managed by the Astrophysical Research Consortium for the 
+Participating Institutions of the SDSS Collaboration including the 
+Brazilian Participation Group, the Carnegie Institution for Science, 
+Carnegie Mellon University, the Chilean Participation Group, the French Participation Group, Harvard-Smithsonian Center for Astrophysics, 
+Instituto de Astrof\'isica de Canarias, The Johns Hopkins University, 
+Kavli Institute for the Physics and Mathematics of the Universe (IPMU) / 
+University of Tokyo, the Korean Participation Group, Lawrence Berkeley National Laboratory, 
+Leibniz Institut f\"ur Astrophysik Potsdam (AIP),  
+Max-Planck-Institut f\"ur Astronomie (MPIA Heidelberg), 
+Max-Planck-Institut f\"ur Astrophysik (MPA Garching), 
+Max-Planck-Institut f\"ur Extraterrestrische Physik (MPE), 
+National Astronomical Observatories of China, New Mexico State University, 
+New York University, University of Notre Dame, 
+Observat\'ario Nacional / MCTI, The Ohio State University, 
+Pennsylvania State University, Shanghai Astronomical Observatory, 
+United Kingdom Participation Group,
+Universidad Nacional Aut\'onoma de M\'exico, University of Arizona, 
+University of Colorado Boulder, University of Oxford, University of Portsmouth, 
+University of Utah, University of Virginia, University of Washington, University of Wisconsin, 
+Vanderbilt University, and Yale University.
+
+\color{gcolor}{JB: Gaia acknowledgment}\color{black}
 

### 2018-08-13 paper/draft.tex
gaia acknowledgements

@@
 
-\shorttitle{Birky et al.}
-\shortauthors{Birky et al.}
+\shorttitle{temperatures and metallicities for m dwarfs}
+\shortauthors{birky et al.}
 
@@
 
-\color{gcolor}{JB: Gaia acknowledgment}\color{black}
+This work has made use of data from the European Space Agency (ESA) mission
+{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
+Data Processing and Analysis Consortium (DPAC,
+\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
+has been provided by national institutions, in particular the institutions
+participating in the {\it Gaia} Multilateral Agreement.
 

### 2018-08-13 paper/draft.tex
keywords

--- 2018-08-13 paper/draft.tex
@@
 We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, and demonstrate that \thecannon\ acting on \apogee\ spectra can infer spectral types with predictive accuracy of 0.6 types.
-The trained models are employed to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in \apogee.
+The trained models are employed to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in \apogee, selected in part on \gaia\ photometry and astrometry.
 We validate our results using photometric temperature indicators and the \gaia\ color--magnitude diagram.
@@
 \keywords{
-surveys
- ---
-stars: Mdwarfs
- ---
-stellar parameters
+infrared:~stars
  ---
 methods:~data~analysis
+ ---
+stars:~abundances
+ ---
+stars:~fundamental~parameters
+ ---
+stars:~late-type
+ ---
+surveys
 }

### 2018-08-13 paper/draft.tex
fix number

@@
 
-Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,0841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.
+Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.
 Citations to any other large M dwarf catalogs that have

### 2018-08-13 paper/draft.tex
adam note
abstract tweaks

--- 2018-08-13 paper/draft.tex
@@
 \begin{abstract}
-\thecannon\ is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; near-infrared) spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H]) to precisions higher than the model-grid pipeline.
-The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim$3500K motivates the extension of this approach to M dwarf stars, but with new training labels.
+\thecannon\ is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution, near-infrared \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
+The lack of reliable stellar models in the near infrared for temperatures $<3500$\,K motivates the extension of this approach to M-dwarf stars, but with new training labels from optical surveys that overlap \sdssiv\ \apogee.
 Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 68\,K and 0.07\,dex respectively.
@@
 As a side result, we determine empirical relationships between spectral type and temperature and metallicity.
-These results have implications for the \tess\ Mission and other projects that involve large samples of M dwarfs.
+These results have implications for the NASA \tess\ Mission and other projects that involve large samples of M dwarfs.
 \end{abstract}
@@
 
-% \begin{enumerate}
 Very low mass stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}.

### 2018-08-13 paper/draft.tex

@@
 
-\color{gcolor}{HOGG: grants etc}\color{black}
-
 Funding for the Sloan Digital Sky Survey IV has been provided by the Alfred P.
@@
 participating in the {\it Gaia} Multilateral Agreement.
+
+\color{gcolor}{HOGG: grants etc}\color{black}
+
+\color{gcolor}{ADAM: grants etc, FAST acknowledgement?}\color{black}
 

### 2018-08-13 paper/draft.tex
use crv scatters

@@
 The lack of reliable stellar models in the near infrared for temperatures $<3500$\,K motivates the extension of this approach to M-dwarf stars, but with new training labels from optical surveys that overlap \sdssiv\ \apogee.
-Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 68\,K and 0.07\,dex respectively.
-We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, and demonstrate that \thecannon\ acting on \apogee\ spectra can infer spectral types with predictive accuracy of 0.6 types.
+Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
+We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, and demonstrate that \thecannon\ acting on \apogee\ spectra can infer spectral types with predictive accuracy of 0.9 types.
 The trained models are employed to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in \apogee, selected in part on \gaia\ photometry and astrometry.

### 2018-08-13 paper/draft.tex
use crv scatters

@@
 \subsubsection{Spectral Type Model \label{subsec:west_results}}
-We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
+We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
 We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.

### 2018-08-13 paper/draft.tex
notes; some restructuring

--- 2018-08-13 paper/draft.tex
@@
 %%==========================================================================================
-\section{Experimental Results} \label{sec:results}
+\section{Experiments and Results} \label{sec:results}
 
-\subsection{Training Sample}
+\subsection{Temperature/Metallicity Model \label{subsec:mann_results}}
+For the physical parameter model we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
+We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr) and likely slowly rotating sources.
+While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.
 
-For each model, two tests of consistency were used to assess the validity of our derived test labels: (1) the self-test comparing training labels and test labels, and (2) the leave-one-out cross validation test, in which we train a model on all sources but \emph{n}, then apply the \emph{N-1} source trained model to get the labels for \emph{n}.
+Two tests of consistency were used to assess the validity of our derived test labels: (1) the self-test comparing training labels and test labels, and (2) the leave-one-out cross validation test, in which we train a model on all sources but \emph{n}, then apply the \emph{N-1} source trained model to get the labels for \emph{n}.
 Precision (scatter) and bias of the model for each test are calculated as the standard deviation and mean of the difference in training and test labels respectively.
+JB: POINT TO THESE RESULTS FOR THIS MODEL IN THE FIGURES
 
@@
 
-\subsubsection{Temperature/Metallicity Model \label{subsec:mann_results}}
-For the physical parameter model we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
-We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr) and likely slowly rotating sources.
-While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.
-
 \color{gcolor}{Some note on what lines vary significantly with parameter...}\color{black}
@@
 
-\subsubsection{Spectral Type Model \label{subsec:west_results}}
+\subsection{Spectral Type Model \label{subsec:west_results}}
 We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
@@
 As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates \thecannon\ does a poor job at extrapolating to labels outside of the training sample space.
+
+JB: NOW POINT HERE TO THE RESULTS OF THE CROSS-VALIDATION AND ETC.
 
@@
 
+JB: NOW POINT TO RESULTS FIGURES AND SUMMARIZE THEM, IF THERE IS ANYTHING ELSE?
+
+\subsection{Validation of parameters}
+
 In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of \aspcap\ temperature values using a combination of \zmass\ and visual band photometries.
@@
 \color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}
+
+JB ALSO POINT TO FIGURES THAT CONNECT TYPES TO PARAMETERS. AND DISCUSS.
+
+JB DO WE WANT A FIGURE THAT SHOWS ABUNDANCES AS A FUNCTION OF MILKY WAY POSITION OR ACTIONS?
 
@@
 \item[-] Discussion of validation tests and ground truth
+
+\item[-] HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?
 \end{enumerate}

### 2018-08-13 paper/draft.tex

@@
 \subsection{Spectral Type Model \label{subsec:west_results}}
-We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
+We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
 We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.

### 2018-08-13 paper/draft.tex
captions

--- 2018-08-13 paper/draft.tex
@@
 \caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities.
-\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}.
+\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K}.
 \label{fig:demo_teff}
@@
 \caption{\textit{Top two panels of each plot:} Mann-trained model for varying metallicities and similar temperatures.
-\textit{Third panel:} Derivative of \thecannon\ model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$.} \label{fig:demo_feh}
+\textit{Third panel:} Derivative of \thecannon\ model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$ dex.} \label{fig:demo_feh}
 \end{figure}
@@
 \plotone{figures/aspcap_cannon_validation.png}
-\caption{Validation test \label{fig:aspcap_cannon_validation}}
+\caption{Validation test of \cannon\ temperatures (left), as compared to \aspcap\ temperatures (right) using three different color-temperature relations derived from \citealt{Mann:2015}, \citealt{Casagrande:2008}, and \citealt{Boyajian:2013}.
+For all three relations  \label{fig:aspcap_cannon_validation}}
 \end{figure}

### 2018-08-13 paper/draft.tex
validation description
no subsection
more JB to-do items

@@
 \affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
-\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
 \affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726
@@
 \affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
+\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
 
@@
 %%==========================================================================================
-\section{Methods} \label{sec:cannon}
-
-\subsection{\thecannon\: A Data-Driven Approach}
+\section{Method}\label{sec:cannon}
 
@@
 
-Quadratic label vectors:
-	\[\ell_{n} = [1, \, SPT, \, SPT^{2}] \]
-	\[\ell_{n} = [1, \, T_{\rm eff}, \, [Fe/H], \, T_{\rm eff} \cdot [Fe/H], \, [Fe/H]^{2}] \]
+Quadratic label vectors for the two models that we will build:
+\begin{eqnarray}
+\ell_{n} &=& [1, \, SPT, \, SPT^{2}]
+\\
+\ell_{n} &=& [1, \, T_{\rm eff}, \, [Fe/H], \, T_{\rm eff} \cdot [Fe/H], \, [Fe/H]^{2}]
+,
+\end{eqnarray}
+where the first is for JB WHAT and the second is for JB WHAT?
 
-The training step description...
+JB: The training step description: OPTIMIZE THETA AT FIXED LABELS.
 
-Test step description...
+JB: Test step description: OPTIMIZE LABELS AT FIXED THETA.
+
+Thus \thecannon\ is both a generative model for the stellar spectra, and a method for obtaining
+labels from spectra. In the training step, the regression is designed to predict spectral pixels
+given labels, by learning zeroth, first, and second derivatives of the data with respect to
+the labels. In the test step, the regression is designed to predict labels given the spectral
+derivatives.
 
@@
 \caption{Validation test of \cannon\ temperatures (left), as compared to \aspcap\ temperatures (right) using three different color-temperature relations derived from \citealt{Mann:2015}, \citealt{Casagrande:2008}, and \citealt{Boyajian:2013}.
-For all three relations  \label{fig:aspcap_cannon_validation}}
+We find \cannon\ temperatures to be significantly more consistent with all three relations in terms of bias and scatter.
+\label{fig:aspcap_cannon_validation}}
 \end{figure}

### 2018-08-13 paper/draft.tex
plx cut note

@@
 \subsection{Test Sample \label{subsec:test_selection}} 
-Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \textsl{Gaia DR2} (\citealt{Brown:2018}) and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources.
+Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \textsl{Gaia DR2} (\citealt{Brown:2018}) and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources.
 From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.
@@
 \plotone{figures/cannon_chi_dist.png}
-\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee\-\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$.
+\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee\-\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
 We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as "safe".

### 2018-08-13 paper/draft.tex
code format

@@
 The spectral type sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from \sdss.
-For each source in the catalog, spectral types were determined both through an automated routine using The Hammer \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
+For each source in the catalog, spectral types were determined both through an automated routine using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
 

### 2018-08-13 paper/draft.tex
intro text

--- 2018-08-13 paper/draft.tex
@@
 
-Why \thecannon\ approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016})
+In this work we build a data-driven model for M dwarf spectra with \thecannon (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}).
+\thecannon is a regression (typically quadratic) for the spectral pixels given a training
+set of stars with reliable or accurate or assumed-accurate labels.
+The model predicts spectrum given labels, but it can be turned around and also
+used to predict labels given spectrum.
+Thus it is a generative model for stellar spectra that can be used for parameter inference.
 
-\begin{enumerate}
-\item This paper will refer to any physical or empirical measurement of a star as \emph{stellar label}
+The data-driven approach of \thecannon\ makes sense in various contexts:
+One is the context in which stellar labels are only known for a small number of stars
+but there are spectra taken for many more.
+Another is the context in which it is very expensive to obtain labels for a star, but
+there are many stars that need labels.
+Another is the context in which there are spectral models that work in one wavelength
+range or resolution but not in another.
+The context of this paper is the latter:
+There are working and reliable models of M dwarf stars in the optical, which have been
+used successfully, but (as we note above) there isn't as much success with such modeling in the infrared,
+and not at \apogee\ resolution.
 
-\item Can be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects;
-
-\item Make the point that \thecannon\ does not require us to have good M-dwarf models in the \apogee\ wavelength range at all
-
-\item better than template matching in that it is using less sparsely sampled known mdwarfs
-\end{enumerate}
-
-%photometric data, HR digram, comparison to evolutionary models
+Here we take M-dwarf labels from samples of optically observed stars that overlap the
+\sdssiv\ \apogee\ sample, and use those labels to train a model and label all of the
+\sdssiv\ \apogee\ M dwarfs.
+One set of labels are physical parameters (effective temperatures and metallicities)
+and the other set of labels are spectral types.
+In each case, \thecannon\ produces a predictive model for \apogee\ spectra, but without
+any input of stellar physics beyond whatever went in to producing the labels in the first
+place, which was in the context of optical spectroscopy.
+That is, the model to be applied to the \apogee\ data is learned from the \apogee\ data alone
+This obviates physical models in the infrared, and capitalizes on the excellent data quality
+delivered by the \apogee\ spectrograph and project.
 

### 2018-08-13 paper/draft.tex
quadratic label description

--- 2018-08-13 paper/draft.tex
@@
 
-Quadratic label vectors for the two models that we will build:
+We apply a quadratic parameterization of the model such that the label vectors for the two models that we will build are:
 \begin{eqnarray}
-\ell_{n} &=& [1, \, SPT, \, SPT^{2}]
+\ell_{n} &=& [1, \, SPT, \, SPT^{2}] \label{eqn:spt}
 \\
-\ell_{n} &=& [1, \, T_{\rm eff}, \, [Fe/H], \, T_{\rm eff} \cdot [Fe/H], \, [Fe/H]^{2}]
+\ell_{n} &=& [1, \, T_{\rm eff}, \, [Fe/H], \, T_{\rm eff}^2, \, T_{\rm eff} \cdot [Fe/H], \, [Fe/H]^{2}] \label{eqn:physical}
 ,
 \end{eqnarray}
-where the first is for JB WHAT and the second is for JB WHAT?
+Equation (\ref{eqn:spt}) is the label vector for the spectral type model with a highest order of 2, and equation (\ref{eqn:physical}) is the label vector for the physical parameter model of order 2 including a cross term between the two labels.
 

### 2018-08-13 paper/draft.tex
accuracy in the data space

--- 2018-08-13 paper/draft.tex
@@
 Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
+These are parameter-space precision statements, but the model is accurate in the data space: It accurately predicts \apogee\ spectral pixels at the percent level.
 We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, and demonstrate that \thecannon\ acting on \apogee\ spectra can infer spectral types with predictive accuracy of 0.9 types.
@@
 
-In this work we build a data-driven model for M dwarf spectra with \thecannon (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}).
-\thecannon is a regression (typically quadratic) for the spectral pixels given a training
+In this work we build a data-driven model for M dwarf spectra with \thecannon\ (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}).
+\thecannon\ is a regression (typically quadratic) for the spectral pixels given a training
 set of stars with reliable or accurate or assumed-accurate labels.
@@
 
-\color{gcolor}{HOGG: Some commentary on why $\chi^2$ is higher than expected.}\color{black}
+\color{gcolor}{HOGG: Some commentary on why $\chi^2$ is higher than expected. Also explain why this means that the model is good to percent level.}\color{black}
 

### 2018-08-13 paper/draft.tex
extension sample

@@
 The physical parameter sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
-The M15 catalog in total contains 183 sources and the extension sample another 500 primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog.
+The major difference between the extension sample is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample.
+
+The M15 catalog in total contains 183 sources and the extension sample another 500 stars, both primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog.
 All targets have low-res optical spectra from the SNIFS specgrograph \citep{Lantz:2004} and infrared spectra taken with the SpeX Spectograph \citep{Rayner:2003}, which have been combined to estaimte largely empirical bolometric fluxes.

### 2018-08-13 paper/draft.tex
clarify gravity

@@
 Unlike their higher-mass counterparts, M dwarf properties do not change measurably over the age of the Universe.
-Hence perfect knowledge of abundances and \teff\ for an M dwarf should uniquely determine its position on a color-magnitude diagram as well as its other fundamental properties.
-While we only have [Fe/H], for the uncertainties considered here, lack of information about other abundances will only be important for stars with extreme abundance patters (e.g., Carbon stars).
-Since we have no such targets in the training set, we could not handle such objects.
+Hence perfect knowledge of abundances and \teff\ for an M dwarf should uniquely determine its position its surface gravity, position on a color-magnitude diagram, and overall luminosity.
+While we only had [Fe/H] for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important for extreme cases (e.g., Carbon stars).
+Highly active stars
 

### 2018-08-14 paper/draft.tex
typo
add to cmd caption

@@
 While we only had [Fe/H] for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important for extreme cases (e.g., Carbon stars).
-Highly active stars
 
 %%AWM: Feel free to edit this for clarity. Let me know if anything makes no sense.
+
 %We note that rotational velocity (vsini) is not a parameter are not included as training labels because that would be hard... also it's probably not needed... I'll let someone else write that. -AWM
@@
 \gridline{\fig{figures/safe_selection_spt.png}{0.5\textwidth}{}}
-\caption{Final sample selection with 10,311 M dwarfs based on criteria described in Section \ref{subsec:test_selection}. \label{fig:safe_selection}}
+\caption{Final sample selection with 10,311 M dwarfs based on criteria described in Section \ref{subsec:test_selection}; temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases with increasing absolute magnitude as expected.
+Deviations from these gradients seen at the upper slice of the main sequence is due to binaries.
+\label{fig:safe_selection}}
 \end{figure}

### 2018-08-14 paper/draft.tex
hogg note
training description

@@
 \subsection{\apogee\ Survey}
-% \begin{enumerate}
-% Describe the \apogee\ data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).
-
-% \item[-] Some M dwarfs in \apogee\ proposed as ancilliary targets (Blake/Covey sources), plus an unknown number observed unintentionally \citep{Desphande:2013}.
-% \end{enumerate}
 
@@
 \thecannon\ does not extrapolate well outside the parameter space of the training sample.
+\color{gcolor}{HOGG: More notes to add here?}\color{black}
 
-Inferring the label of a star with such a model requires two steps: first the \emph{training step} in which a generative model is constructed at each pixel from the set of reference labels and fluxes, and second the \emph{test step} in which the model is applied to determine the labels of a spectrum.
+Inferring the label of a star with such a model requires two steps: first the \emph{training step} in which a generative model describing the probability density function (pdf) of the flux is constructed at each pixel from the set of spectrum with known reference labels, and second the \emph{test step} in which the model is applied to determine the labels of a spectrum.
 
@@
 
-JB: The training step description: OPTIMIZE THETA AT FIXED LABELS.
-
-JB: Test step description: OPTIMIZE LABELS AT FIXED THETA.
+% JB: The training step description: OPTIMIZE THETA AT FIXED LABELS.
+% JB: Test step description: OPTIMIZE LABELS AT FIXED THETA.
 

### 2018-08-14 paper/draft.tex
move paragraph

@@
 
+\thecannon\ model can in principle be trained on any physical or empirical labels available, giving a wide range of flexibility to the model, however choosing a training sample with high quality (accurate and precise) is critical to the performance of \thecannon.
+Limitations to consider are that test labels from \thecannon\ are only accurate if the training labels accurate, and only precise if the training labels are measured using a consistent method.
+Also choosing a sample with large dynamic range spanning the entire parameter space of interest is important, as \thecannon\ does not extrapolate well outside the parameter space of the training sample.
+\color{gcolor}{HOGG: More notes to add here?}\color{black}
+
 For the purpose of this study we have constructed two different training samples: first a one-dimensional \emph{spectral type model}, and second a two-dimensional \emph{physical parameter model} which describes the temperature and metallicity.
@@
 Describe the assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label.
-
-\thecannon\ model can in principle be trained on any physical or empirical labels available.
-However, limitations to consider are that test labels from \thecannon\ are only accurate if the training labels accurate, and only precise if the training labels are measured using a consistent method.
-\thecannon\ does not extrapolate well outside the parameter space of the training sample.
-\color{gcolor}{HOGG: More notes to add here?}\color{black}
 

### 2018-08-14 paper/draft.tex
test dist comment
jackknife error caption

@@
 \caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities.
-\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K}.
-\label{fig:demo_teff}
+\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.} \label{fig:demo_teff}
 \end{figure}
@@
 \caption{\textit{Top two panels of each plot:} Mann-trained model for varying metallicities and similar temperatures.
-\textit{Third panel:} Derivative of \thecannon\ model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$ dex.} \label{fig:demo_feh}
+\textit{Third panel:} Derivative of \thecannon\ model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$ dex; jackknife computed error at each pixel is shown in red.} \label{fig:demo_feh}
 \end{figure}
@@
 \end{center}
-\caption{Derivative plots for spectral type model taken at the median training spectral type, SPT$=3$.} \label{fig:west_derivative}
+\caption{Derivative plots for spectral type model taken at the median training spectral type, SPT$=3$; jackknife computed error at each pixel is shown in red.} \label{fig:west_derivative}
 \end{figure}
@@
 \plotone{figures/aspcap_cannon_label_hist.png}
-\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the \aspcap\ pipeline. \label{fig:aspcap_cannon_label_hist}}
+\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the \aspcap\ pipeline.
+Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<[Fe/H]<0.5$ because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars \aspcap\ predicts temperatures with slightly higher temperatures and metallicities.
+\label{fig:aspcap_cannon_label_hist}}
 \end{figure}

### 2018-08-14 paper/draft.tex
quad label

@@
 \end{eqnarray}
-Equation (\ref{eqn:spt}) is the label vector for the spectral type model with a highest order of 2, and equation (\ref{eqn:physical}) is the label vector for the physical parameter model of order 2 including a cross term between the two labels.
+Equation (\ref{eqn:spt}) is the label vector for the spectral type model, and equation (\ref{eqn:physical}) is the label vector for the physical parameter model, and we find that a second order parameterization is sufficient for describing the flux of each spectrum to one percent accuracy, as discussed further in Section \ref{subsec:mann_results}.
+
+The training step consists of
 

### 2018-08-14 paper/draft.tex
notes

@@
 
-\item model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (\citealt{Casagrande:2008} estimated effective temperatures, bolometric luminosities, and metallicities using optical and infrared photometries) -- however we want to leverage the full wavelength range and resolution of our data.
+\item model narrow wavelegth region where models are good, look at specific features -- however we want to leverage the full wavelength range and resolution of our data.
 
@@
 
-The training step consists of
+The \emph{training step} consists of optimizing the coefficient vector $\theta$ for fixed labels \emph{$\ell_n$} (the reference labels for a training sample): given a spectrum (a wavelength, flux, and flux uncertainty array), for \emph{N} stars and their corresponding training labels
+
+The \emph{test step} is an inversion of the
 
@@
 
+\color{gcolor}{What are all of the parameters we want to report and what do we want to emphasize? Are there other parameters (such as mass) that might be straightforward to estimate?}\color{black}
+
 JB ALSO POINT TO FIGURES THAT CONNECT TYPES TO PARAMETERS. AND DISCUSS.
+
+JB REDO 3 CMD PLOTS SHOWING BEFORE AND AFTER SELECTION.
 
@@
 
-% \begin{figure}[ht]
-% \begin{center}
-% \includegraphics[width=16cm]{figures/derivative_jackknife_teff.png}
-% \includegraphics[width=16cm]{figures/derivative_jackknife_feh.png}
-% \end{center}
-% \caption{Derivative plots for physical parameter model.
-\textit{Top:} Derivative of flux with repsect to temperature taken at the median training temperature, T$_{\rm eff}=3463K$.
-\textit{Bottom:} Derivative of flux with repsect to metallicity taken at the median training metallicity, [Fe/H]$=-0.03$ dex.} \label{fig:mann_derivative}
-% \end{figure}
-
 \begin{figure}[ht]
@@
 \caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee\-\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
-We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as "safe".
+We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as ``safe''.
 \label{fig:chi_dist}}

### 2018-08-15 paper/draft.tex
crv caption

@@
 \caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (left), and leave-one-out cross validation (right).
-Sources with strong rotational broadening (2M07171706−0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.} \label{fig:mann_validation}
+Sources with strong rotational broadening (2M07171706−0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.
+Predictive accuracy, as computed by the scatter in cross validation is 77K in temperature, and 0.09 dex in metallicity.} \label{fig:mann_validation}
 \end{figure}
@@
 \end{center}
-\caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).} \label{fig:west_validation}
+\caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).
+Predictive accuracy, as computed by the scatter in cross validation is 0.9 subtypes.} \label{fig:west_validation}
 \end{figure}

### 2018-08-15 paper/draft.tex
notes
aspcap #

@@
 
-The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (\textsl{SDSS-III/IV} \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
+The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (\textsl{SDSS-III/IV} \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 250,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
 Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (\aspcap; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (\teff, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
@@
 
-\color{gcolor}{Further assessment of the reliability of our model for late types?...}\color{black}
-
 \subsection{Test Sample \label{subsec:test_selection}} 
@@
 \item[-] Brief summary of what we did and found
+
+\item[-] Is it interesting to point out the comoving pairs we found? (33 FGK-M, 13 WD-M, and 174 M-M pairs)?
 

### 2018-08-15 paper/draft.tex
selection before/after caption

@@
 \begin{figure}[ht]
-\gridline{\fig{figures/safe_selection_teff.png}{0.5\textwidth}{}
-          \fig{figures/safe_selection_feh.png}{0.5\textwidth}{}}
-\gridline{\fig{figures/safe_selection_spt.png}{0.5\textwidth}{}}
-\caption{Final sample selection with 10,311 M dwarfs based on criteria described in Section \ref{subsec:test_selection}; temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases with increasing absolute magnitude as expected.
-Deviations from these gradients seen at the upper slice of the main sequence is due to binaries.
+% \gridline{\fig{figures/safe_selection_teff.png}{0.5\textwidth}{}
+%           \fig{figures/safe_selection_feh.png}{0.5\textwidth}{}}
+% \gridline{\fig{figures/safe_selection_spt.png}{0.5\textwidth}{}}
+\plottwo{figures/cmd_teff_full.png}{figures/cmd_teff_safe.png}
+\plottwo{figures/cmd_feh_full.png}{figures/cmd_feh_safe.png}
+\plottwo{figures/cmd_spt_full.png}{figures/cmd_spt_safe.png}
+\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, binaries, and K dwarfs).
+Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases with increasing absolute magnitude as expected.
+Deviations from these gradients seen at the upper slice of the main sequence is likely remaining contamination from the binary sequence.
 \label{fig:safe_selection}}

### 2018-08-16 paper/draft.tex
add to caption
open questions

@@
 These results have implications for the NASA \tess\ Mission and other projects that involve large samples of M dwarfs.
+
+\color{gcolor}{HOGG: just a reflection, I think the scope of this paper has shifted a lot from just being a technical paper about \thecannon\ working on M dwarfs, to drifting more towards M dwarf science (partly because I've learned a lot more about the properties/characteristics of M dwarfs, and the Gaia has given us a lot of intuition that we didn't have before).
+Anyways we should decide whether the paper is more of the former or the latter and maybe re-write the abstract.
+Right now it just feels a bit hacked-together, since this is basically just my AAS abstract that we kept adding some sentences to as we went, and I think it needs a better frame.
+I'm still trying to get a sense of what is scientifically interesting about this sample, and I've been struggling to pin point what the real relevance of this work is: it's not exactly just that the Cannon is precise for M dwarfs (this original scope, like an extension of Ness, is too boring I think, given what we've learned), and it's not really about M dwarf physics or why we are better than models (this is too beyond us to argue).
+
+I think maybe the biggest angle is that are illustrating what can be leveraged out of two very high quality datasets (\apogee\ and \gaia), and we are the first to identify and collect a large (atmospheric parameters, 6D phase space) that have in the past been very difficult to measure.
+Is this what we want to emphasize?
+
+Also an open question along these lines is whether we want to estimate any additional physical parameters for our sample, for example maybe adding radii (from the \citealt{Mann:2015} \teff-radius or $M_K$-radius relations), or masses from his $M_K$-mass relation he's publishing, or maybe there are some other fundamental parameters that would be interesting (and straightforward) to add?
+It seems like just because we have \gaia\ parallaxes we should go ahead and do it?
+Maybe also to do a full-circle comparison with our test sample labels.
+
+Another open question is about the [Fe/H] validation, and whether we are happy with just the cmd plot of metallicity. I think our ideas for other tests were:
+\begin{itemize}
+\item Looking at M+M binary pairs: there are 174 pairs and it looks by eye that most of the pair lines are parallel to the main sequence.
+I don't know if there's some more clever way to see if the pairs lie in the same metallicity space of the cmd, so this is probably not that useful
+\item Overplotting some isochrones at varying metallicities--which I'm still working on
+\end{itemize}
+}\color{black}
+
+Also I still have to make the vertical action plot.
 \end{abstract}
@@
 
-A number of M dwarf studies have already been conducted in \apogee\, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements:
+A number of M dwarf studies have already been conducted in \apogee, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements:
 \begin{enumerate}
@@
 and not at \apogee\ resolution.
+\color{gcolor}{HOGG: I think we can make a more specific set of points here that would be more interesting to M dwarf people: temperatures and (especially) metallicities are difficult to measure accurately for an M dwarf, but people have done a lot of work to build calibrated empirical relations for metallicities and temperatures that are very precise.
+For example \citealt{Terrien:2012}, \citealt{RojasAyala:2010}, Mann 2013 (and there's probably some other examples (Newton, etc.) and I have more literature to look at...) have a metallicity callibration based off of an FGK+M star samples, from J, H, K or optical band spectra using the equivalent width of some strong line/index (like CaI, KI, etc).
+However these metallicity callibrations are inaccesible for APOGEE spectra, because the wavelength coverage does not include the relevant callibration indices.
+\thecannon\ bridges this gap: it doesn't require using specific lines (which would probably be a mess to deconvolve at high resolution with other blends even if we had the right lines); it allows us to use a lot of low level metallicity information present in thousands of lines; and (as we demonstrate) it does so with very good precision.}\color{black}
 
@@
 \caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, binaries, and K dwarfs).
+Overplotted with orange triangles are the M15 and W11 training samples, for their respective \cannon\ test labels.
 Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases with increasing absolute magnitude as expected.

### 2018-08-16 paper/draft.tex
data file description

@@
 
-The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (\textsl{SDSS-III/IV} \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 250,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
+The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (\textsl{SDSS-III/IV} \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
 Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (\aspcap; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (\teff, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
@@
 However numerous cool, main sequence sources have also been observed either as targets proposed by the M dwarf ancilliary survey (1,350 sources selected by proper motions from the  \citealt{Desphande:2013}), or randomly by \apogee\, which we quantify in this study to be over 10,000 in total.
+
+The \aspcap\ pipeline releases several types of data files, with various levels of processing: {\tt\string ap1D} (the raw one-dimensional spectra for individual visits), {\tt\string apVisit} (the individual visit spectra with telluric subtraction), {\tt\string apStar} (the co-added {\tt\string apVisit} spectra), and most processed {\tt\string aspcapStar} spectra.
+For this work we use {\tt\string aspcapStar} files, which contains the continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
+\color{gcolor}{HOGG: do you want to comment on continuum normalization issues?}\color{black}
 
@@
 %\color{gcolor}{}\color{black}
-
-\subsection{Data Processing}
-
-\apogee\ data products: {\tt\string ap1D}, {\tt\string apVisit}, {\tt\string apStar}, and {\tt\string aspcapStar}.
-We use the {\tt\string aspcapStar} spectra, which contains the continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
-SNR of stars in our sample...
 

### 2018-08-17 paper/draft.tex
note
table header description

@@
 \end{itemize}
+
+Also I still have to make the vertical action plot.
 }\color{black}
 
-Also I still have to make the vertical action plot.
 \end{abstract}
@@
 
-\begin{longrotatetable}
-\begin{deluxetable*}{cccccccccc}
-\tabletypesize{\scriptsize}
-\tablecaption{Temperatures/Metallicities (from Mann-trained model), and spectral types (from West-trained model) for the sample 10,311 APOGEE M dwarfs, as well as goodness of fit for the respective models.
-\label{test_results_table}}
-\tablehead{\colhead{APOGEE ID} & \colhead{Gaia\ ID} & \colhead{RA}  & \colhead{DEC} & \colhead{\teff} & \colhead{[Fe/H]}  & \colhead{SPT} & \colhead{$\chi^2$ Mann} & \colhead{$\chi^2$ West} & \colhead{\nodata} \\
- &  & (deg)  & (deg) & (K) & (dex) &  &  &  & }   
+\begin{deluxetable*}{rcl}
+\tablecaption{Test results table header description.}
+\tablehead{\colhead{Column} & \colhead{Unit} & \colhead{Description}}
 \startdata
-2M00432365+0332534 & 413814124681430528  & 10.848967  & 3.548273   & 3417 & 0.18  & 3.0 & 7910  & 7912   & \nodata \\
-2M10382316+2457051 & 808931427773946112  & 159.596215 & 24.951373  & 3358 & -0.25 & 3.4 & 19051 & 25518  & \nodata \\
-2M09364698+5602332 & 598955557617293440  & 144.195513 & 56.042461  & 3819 & 0.01  & 1.5 & 7981  & 6894   & \nodata \\
-2M12165176+1543232 & 1538883006881701120 & 184.215466 & 15.723067  & 3948 & 0.16  & 0.7 & 25354 & 19983  & \nodata \\
-2M18501195+4809327 & 2037435595069990016 & 282.549868 & 48.15976   & 3899 & 0.24  & 0.9 & 7205  & 7151   & \nodata \\
-2M07572656+4703318 & 682241540493118080  & 119.360502 & 47.05852   & 3636 & -0.17 & 2.4 & 38430 & 45625  & \nodata \\
-2M11083992+1112503 & 1536026200794333696 & 167.166344 & 11.213979  & 3949 & 0.1   & 0.8 & 21165 & 17506  & \nodata \\
-2M12135300+1419452 & 3695183295027782656 & 183.470871 & 14.32924   & 3824 & 0.13  & 1.3 & 35018 & 21010  & \nodata \\
-2M12273816-0049437 & 3682549562826659456 & 186.909269 & -0.828949  & 3715 & 0.04  & 1.8 & 14179 & 10656  & \nodata \\
-2M10513639-0150064 & 777824835073481984  & 162.901465 & -1.835273  & 3997 & -0.15 & 1.2 & 80915 & 75157  & \nodata \\
-2M02442390-1352593 & 447040266556331008  & 41.099824  & -13.883403 & 3676 & -0.28 & 2.3 & 41375 & 56704  & \nodata \\
-2M06281785+0940439 & 65792100044278400   & 97.074558  & 9.678234   & 3659 & -0.29 & 2.3 & 78827 & 118883 & \nodata \\
-2M13441275+5538012 & 1551387099791031552 & 206.053693 & 55.633288  & 3856 & -0.1  & 1.6 & 50070 & 45895  & \nodata \\
-2M12473762+1107449 & 1251277737449848064 & 191.906593 & 11.129187  & 3244 & -0.42 & 4.0 & 30184 & 47020  & \nodata \\
-2M01274162+8435576 & 523638809262601216  & 21.924655  & 84.599233  & 3534 & 0.07  & 2.6 & 15529 & 15113  & \nodata \\
-2M13464494+4034158 & 1247733770955275008 & 206.686943 & 40.571056  & 3731 & -0.16 & 2.1 & 53196 & 50765  & \nodata \\
-2M00504161+0219408 & 2376325179183866240 & 12.673481  & 2.327846   & 3745 & 0.4   & 1.4 & 76966 & 52035  & \nodata \\
-2M14494090+3201268 & 1292097793826595328 & 222.420454 & 32.023994  & 3228 & 0.0   & 3.9 & 28733 & 38112  & \nodata \\
-2M16434969-0123117 & 1406510018874952960 & 250.957083 & -1.386602  & 3669 & 0.03  & 2.0 & 22346 & 17186  & \nodata \\
-2M13444356+1655384 & 1565587017583214464 & 206.18144  & 16.927287  & 3652 & 0.13  & 2.1 & 47828 & 40933  & \nodata \\
-2M08162613+3250515 & 3053990410555781120 & 124.108745 & 32.84736   & 3837 & -0.15 & 1.8 & 30389 & 35103  & \nodata \\
-2M10414608+2435412 & 3758656623830055168 & 160.441928 & 24.594731  & 4020 & -0.19 & 1.1 & 41857 & 36934  & \nodata \\
-2M14373854+5014502 & 1188221157913600384 & 219.410555 & 50.247523  & 3380 & -0.13 & 3.2 & 35397 & 41989  & \nodata \\
-\nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata
+{\tt APOGEE$\_$ID} 	& 	& \apogee\ \zmass\ designation \\
+{\tt BP$\_$MAG }		& mag	& \gaia\ $BP$ band photometry	\\
+{\tt BP$\_$RP} 		& mag	& \gaia\ $BP-RP$ color	\\
+{\tt CHI$\_$MANN} 	& 	& $\chi^2$ fit of Mann-trained \cannon\ model	\\
+{\tt CHI$\_$WEST} 	& 	& $\chi^2$ fit of West-trained \cannon\ model	\\
+{\tt DEC} 			& deg	& \apogee\ declination angle	\\
+{\tt DIST} 			& pc	& Distance (1/$\varpi$)	\\
+{\tt FE$\_$H }		& dex	& Mann-trained \cannon\ [Fe/H]	\\
+{\tt GAIA$\_$ID} 		&	& \gaia\ identification number \\
+{\tt G$\_$ABS} 		& mag	& \gaia\ $G$ band absolute magnitude	\\
+{\tt G$\_$MAG} 		& mag	& \gaia\ $G$ band photometry	\\
+{\tt H} 				& mag	& \apogee\ $H$ band photometry	\\
+{\tt H$\_$ERR} 		& mag	& \apogee\ $H$ band photometry uncertainty	\\
+{\tt J} 				& mag	& \apogee\ $J$ band photometry	\\
+{\tt J$\_$ERR} 		& mag	& \apogee\ $J$ band photometry uncertainty	\\
+{\tt K} 				& mag	& \apogee\ $K$ band photometry	\\
+{\tt K$\_$ERR} 		& mag	& \apogee\ $K$ band photometry uncertainty	\\
+{\tt PLX} 			& mas	& \gaia\ parallax	\\
+{\tt PLX$\_$ERR} 		& mas	& \gaia\ parallax uncertainty	\\
+{\tt PMDEC} 			& mas yr$^{-1}$	& \gaia\ declination proper motion	\\
+{\tt PMDEC$\_$ERR} 	& mas yr$^{-1}$	& \gaia\ declination proper motion uncertainty	\\
+{\tt PMRA} 			& mas yr$^{-1}$	& \gaia\ radial ascension proper motion	\\
+{\tt PMRA$\_$ERR} 	& mas yr$^{-1}$	& \gaia\ radial ascension proper motion uncertainty	\\
+{\tt RA} 				& deg	& \apogee\ radial ascension angle	\\
+{\tt RP$\_$MAG} 		& mag	& \gaia\ $RP$ band photometry	\\
+{\tt RV$\_$APOGEE} 	& km s$^{-1}$	& \apogee\ radial velocity	\\
+{\tt RV$\_$APOGEE$\_$ERR} 	& km s$^{-1}$	& \apogee\ radial velocity uncertainty	\\
+{\tt RV$\_$GAIA} 				& km s$^{-1}$	& \gaia\ radial velocity	\\
+{\tt RV$\_$GAIA$\_$ERR}		& km s$^{-1}$	& \gaia\ radial velocity uncertainty	\\
+{\tt SPT} 				& 	& West-trained \cannon\ spectral subtype ($M0-M9$)	\\
+{\tt TEFF} 				& K	& Mann-trained \cannon\ effective temperature	\\
+{\tt Vx} 				& km s$^{-1}$	& Cartesian $x$ velocity in Galactocentric coordnates	\\
+{\tt Vy} 				& km s$^{-1}$	& Cartesian $y$ velocity in Galactocentric coordnates	\\
+{\tt Vz} 				& km s$^{-1}$	& Cartesian $z$ velocity in Galactocentric coordnates	\\
+{\tt X} 				& pc	& Cartesian $x$ position in Galactocentric coordnates	\\
+{\tt Y} 				& pc	& Cartesian $y$ position in Galactocentric coordnates	\\
+{\tt Z} 				& pc	& Cartesian $z$ position in Galactocentric coordnates	
 \enddata
 \end{deluxetable*}
-\end{longrotatetable}
 
 %====================================
-\newpage
-
-\begin{longrotatetable}
-\begin{deluxetable*}{cccccccccc}
-\tabletypesize{\scriptsize}
-\tablecaption{Positions and 3D kinematics for 10,311 M dwarfs; parallaxes and proper motions from \gaia\ + radial velocities from \apogee.}
-\tablehead{\colhead{\nodata} & \colhead{$\varpi$} & \colhead{$\sigma_\varpi$} & \colhead{$\mu_\alpha$}  & \colhead{$\sigma_{\mu\alpha}$} & \colhead{$\mu_\delta$} & \colhead{$\sigma_{\mu\delta}$}  & \colhead{$v_{los}$} & \colhead{\nodata} \\
- & (mas) & (mas) & (mas/yr)  & (mas/yr) & (mas/yr) & (mas/yr)  & (km/s) & \colhead{\nodata}}   
-\startdata
-\nodata & 0.4095971279793963  & 0.018344701025907368 & 94.6092250382206    & 0.2326543993399578  & 21.139312054788142  & 0.10645732514148187  & -14.8169    & \nodata \\
-\nodata & 0.7907713253583666  & 0.05017993250995411  & -61.145608504420785 & 0.19267235934981508 & -9.292283439334902  & 0.191605833991544    & 30.0646     & \nodata \\
-\nodata & 2.437411282195209   & 0.06484152349097437  & -40.89370712477501  & 0.12447374577648344 & -27.01010381099313  & 0.12376849807276755  & -25.7591    & \nodata \\
-\nodata & 4.748695834113324   & 0.025884382476497582 & -39.804579591840216 & 0.0930132037970028  & -10.36148239898068  & 0.057746591921984974 & -7.28313    & \nodata \\
-\nodata & 0.3830610102451543  & 0.020679199019227986 & 8.768467380162365   & 0.03404660389361428 & 142.96097527921128  & 0.03222662578521854  & -67.5318    & \nodata \\
-\nodata & 1.2156916470683867  & 0.05850523666981387  & -34.02970707374436  & 0.5249923050897877  & -60.82222958247146  & 0.41961695655028974  & 27.5802     & \nodata \\
-\nodata & 3.2562530226439863  & 0.12635148166126747  & -14.18737757924032  & 0.08323990329349258 & 3.839844523005405   & 0.08709819792316187  & 1.19813     & \nodata \\
-\nodata & 1.1214563311692185  & 0.029138557090469957 & 3.8612047002174377  & 0.10975162725538208 & -6.6212605579145185 & 0.0625749242163525   & -12.022     & \nodata \\
-\nodata & 0.4337056924769619  & 0.03519634806417813  & 63.58287541662625   & 0.12216317236007974 & -30.844101370573288 & 0.06988132634903554  & 5.3206      & \nodata \\
-\nodata & 4.860871276791149   & 0.05422112071607727  & -38.48480428052528  & 0.06681092801786047 & -46.03340135581453  & 0.05013526442486122  & 5.74919     & \nodata \\
-\nodata & 0.3900296037356137  & 0.022457415816365067 & 48.72202372576884   & 0.09931020159590297 & -41.61843538517975  & 0.0939261318266816   & 56.5766     & \nodata \\
-\nodata & 0.5323417078167814  & 0.050227726038051694 & 37.19385535484577   & 0.15032160219430055 & -145.05520943675262 & 0.1318726138984278   & 59.6754     & \nodata \\
-\nodata & 6.37232292017776    & 0.0137946901051907   & 68.79697898216794   & 0.04264188775089677 & -84.07707033295478  & 0.03680789911821181  & -4.03993    & \nodata \\
-\nodata & 0.5021240985953683  & 0.029189592695681964 & -57.9436465188258   & 0.1911215986654146  & 16.919634543207177  & 0.10679177694312522  & -58.6918    & \nodata \\
-\nodata & 5.237224518979301   & 0.029752252992649626 & 23.771764671058826  & 0.07292367996213067 & -28.25823161266089  & 0.06344587895614055  & -19.6517    & \nodata \\
-\nodata & 3.3089435737776687  & 0.04711084402088657  & -41.45242638019648  & 0.3418156551313586  & -4.510012473794788  & 0.4388646698761767   & -19.7339    & \nodata \\
-\nodata & 5.072750731615924   & 0.04188258021899252  & 13.125640321529255  & 0.08925571521229253 & -27.369888362552718 & 0.05289115275902435  & -7.76221    & \nodata \\
-\nodata & 0.3185097583003641  & 0.014007935590281349 & 6.6397581995189     & 0.27566940326796724 & -20.42262436999549  & 0.4456232897623583   & -27.3047    & \nodata \\
-\nodata & 0.7753680397807614  & 0.01945512547484866  & 7.104977799189683   & 0.16560847226787015 & -5.534895098477748  & 0.11471267191919594  & 15.3108     & \nodata \\
-\nodata & 1.4924247624828302  & 0.028700623137541913 & -22.86896549011412  & 0.13153650570900005 & -6.5044933933639575 & 0.0918536954846762   & -1.93386    & \nodata \\
-\nodata & 0.3373361076141824  & 0.020908617609888186 & -23.51284581625556  & 0.1281836138293499  & -51.21200709697348  & 0.07993530283388152  & 67.9022     & \nodata \\
-\nodata & 0.9508792557786656  & 0.04214796908087648  & -23.92315869160129  & 0.10866864291003138 & -18.19346642794714  & 0.11087251384043836  & -0.941346   & \nodata \\
-\nodata & 2.6649412215771706  & 0.02575193557459232  & -11.654902067800947 & 0.0765638732657513  & 55.80843779696036   & 0.07016094187926156  & 15.9426     & \nodata \\
-\nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata
-\enddata
-\end{deluxetable*}
-\end{longrotatetable}
-
-%====================================
-\newpage
-
-\begin{longrotatetable}
-\begin{deluxetable*}{cccccccccc}
-\tabletypesize{\scriptsize}
-\tablecaption{\apogee\ $JHK$ and \gaia\ $G$, $BP$, $RP$ relative photometric magnitudes.}
-\tablehead{\colhead{\nodata} & \colhead{$J$} & \colhead{$H$} & \colhead{$K$} & \colhead{$G$} & \colhead{$BP$} & \colhead{$RP$}}   
-\startdata
-\nodata & 12.83199977874756  & 12.175999641418455 & 11.96500015258789  & 13.415744          & 14.226433           & 12.53435            \\
-\nodata & 14.255999565124512 & 13.72599983215332  & 13.40999984741211  & 12.228349          & 12.702468           & 11.614723           \\
-\nodata & 11.791999816894533 & 11.104999542236328 & 10.967000007629396 & 10.691218          & 11.139357           & 10.104778           \\
-\nodata & 13.706000328063965 & 13.013999938964846 & 12.895999908447266 & 12.420532          & 12.90665            & 11.800449           \\
-\nodata & 11.616999626159668 & 10.946000099182127 & 10.751999855041504 & 12.908026          & 13.56541            & 12.142124           \\
-\nodata & 11.312999725341795 & 10.663999557495117 & 10.508000373840332 & 10.826382          & 10.85745            & 10.774274           \\
-\nodata & 13.128999710083008 & 12.498000144958494 & 12.32699966430664  & 14.534789          & 15.185692           & 13.744579           \\
-\nodata & 13.538000106811525 & 12.902999877929688 & 12.762999534606935 & 14.004284          & 14.160915           & 13.288598           \\
-\nodata & 12.69499969482422  & 12.036999702453613 & 11.86299991607666  & 14.93699           & 15.395644           & 14.318204           \\
-\nodata & 12.53499984741211  & 11.843999862670898 & 11.732000350952148 & 10.015749          & 10.34129            & 9.559692            \\
-\nodata & 12.536999702453613 & 11.928000450134276 & 11.682000160217285 & 13.735478          & 14.700515           & 12.765027           \\
-\nodata & 11.274999618530273 & 10.717000007629396 & 10.515000343322754 & 11.817871          & 12.467115           & 11.063624           \\
-\nodata & 10.21399974822998  & 9.600000381469727  & 9.407999992370604  & 13.178638          & 13.876391           & 12.384415           \\
-\nodata & 14.201000213623047 & 13.718000411987305 & 13.46500015258789  & 12.388862          & 12.901998           & 11.731835           \\
-\nodata & 13.104000091552734 & 12.43000030517578  & 12.256999969482422 & 8.78063            & 8.920993            & 8.569135            \\
-\nodata & 11.628999710083008 & 10.996000289916992 & 10.802000045776367 & 15.246564          & 16.136223           & 14.321473           \\
-\nodata & 12.312999725341795 & 11.618000030517578 & 11.38599967956543  & 11.829312          & 12.26109            & 11.258054           \\
-\nodata & 12.07800006866455  & 11.48799991607666  & 11.237000465393066 & 13.222831          & 13.763299           & 12.547942           \\
-\nodata & 13.682000160217285 & 12.944000244140625 & 12.72700023651123  & 12.266772          & 12.872325           & 11.548589           \\
-\nodata & 13.501999855041506 & 12.85099983215332  & 12.64900016784668  & 10.416036          & 10.922305           & 9.781987            \\
-\nodata & 13.95300006866455  & 13.29300022125244  & 13.10099983215332  & 13.099073          & 13.74368            & 12.342266           \\
-\nodata & 14.246000289916992 & 13.595000267028807 & 13.397000312805176 & 11.854821          & 12.39487            & 11.183031           \\
-\nodata & 13.241999626159668 & 12.670999526977539 & 12.432000160217285 & 13.945971          & 14.459273           & 13.295053           \\
-\nodata & \nodata & \nodata & \nodata & \nodata & \nodata & \nodata
-\enddata
-\end{deluxetable*}
-\end{longrotatetable}
 

### 2018-08-21 paper/draft.tex
fix alignment
non-ascii character

@@
 For the physical parameter model we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
-We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr) and likely slowly rotating sources.
+We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr) and likely slowly rotating sources.
 While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.
@@
 \caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (left), and leave-one-out cross validation (right).
-Sources with strong rotational broadening (2M07171706−0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.
-Predictive accuracy, as computed by the scatter in cross validation is 77K in temperature, and 0.09 dex in metallicity.} \label{fig:mann_validation}
+Sources with strong rotational broadening (2M07171706--0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.
+Predictive accuracy, as computed by the scatter in cross validation is 77K in temperature, and 0.09 dex in metallicity.}\label{fig:mann_validation}
 \end{figure}
@@
 \caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).
-Predictive accuracy, as computed by the scatter in cross validation is 0.9 subtypes.} \label{fig:west_validation}
+Predictive accuracy, as computed by the scatter in cross validation is 0.9 subtypes.}\label{fig:west_validation}
 \end{figure}
@@
 \caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities.
-\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.} \label{fig:demo_teff}
+\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.}\label{fig:demo_teff}
 \end{figure}
@@
 \tablecaption{Test results table header description.}
-\tablehead{\colhead{Column} & \colhead{Unit} & \colhead{Description}}
+\tablehead{Column & Unit & Description}
 \startdata

### 2018-09-14 paper/draft.tex
to-do comments

@@
 
-JB REDO 3 CMD PLOTS SHOWING BEFORE AND AFTER SELECTION.
-
 JB DO WE WANT A FIGURE THAT SHOWS ABUNDANCES AS A FUNCTION OF MILKY WAY POSITION OR ACTIONS?
@@
 \caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities.
-\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.}\label{fig:demo_teff}
+\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.
+JB: ADD CHI^2 FIT TO PLOT}\label{fig:demo_teff}
 \end{figure}

### 2018-09-19 paper/draft.tex
cannon text

@@
 
-Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector \emph{l} and and coefficient vector \emph{$\theta$}.
+Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector $\ell_n$ and and coefficient vector \emph{$\theta$}.
 So for each star \emph{n}, at pixel wavelength \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
@@
 
-We apply a quadratic parameterization of the model such that the label vectors for the two models that we will build are:
+We apply a quadratic parameterization of the model such that the label vectors for the two models are all of the combinations of reference labels, up to second order:
 \begin{eqnarray}
@@
 \end{eqnarray}
-Equation (\ref{eqn:spt}) is the label vector for the spectral type model, and equation (\ref{eqn:physical}) is the label vector for the physical parameter model, and we find that a second order parameterization is sufficient for describing the flux of each spectrum to one percent accuracy, as discussed further in Section \ref{subsec:mann_results}.
+Equation (\ref{eqn:spt}) is the label vector for the spectral type model, and equation (\ref{eqn:physical}) is the label vector for the physical parameter model, and the first element ``1'' is included to allow flexibility for a linear offset to the model.
+We find that a second order parameterization is sufficient for describing the flux of each spectrum to one percent accuracy, as discussed further in Section \ref{subsec:mann_results}.
 
-The \emph{training step} consists of optimizing the coefficient vector $\theta$ for fixed labels \emph{$\ell_n$} (the reference labels for a training sample): given a spectrum (a wavelength, flux, and flux uncertainty array), for \emph{N} stars and their corresponding training labels
-
-The \emph{test step} is an inversion of the
+The \emph{training step} consists of optimizing for the coefficient vector and scatter ($\theta_{\lambda}$ and $s_{\lambda}$) at fixed labels (\emph{$\ell_n$}), and the \emph{test step} consists of optimizing for labels at fixed $\theta_{\lambda}$ and scatter.
 

### 2018-09-19 paper/draft.tex
wording

@@
 
-Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector $\ell_n$ and and coefficient vector \emph{$\theta$}.
+Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector $\ell$ and and coefficient vector \emph{$\theta$}.
 So for each star \emph{n}, at pixel wavelength \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
@@
 Equation (\ref{eqn:spt}) is the label vector for the spectral type model, and equation (\ref{eqn:physical}) is the label vector for the physical parameter model, and the first element ``1'' is included to allow flexibility for a linear offset to the model.
-We find that a second order parameterization is sufficient for describing the flux of each spectrum to one percent accuracy, as discussed further in Section \ref{subsec:mann_results}.
+We find that a second order parameterization is sufficient for reproducing the flux of each spectrum to one percent accuracy, as discussed further in Section \ref{subsec:mann_results}.
 
-The \emph{training step} consists of optimizing for the coefficient vector and scatter ($\theta_{\lambda}$ and $s_{\lambda}$) at fixed labels (\emph{$\ell_n$}), and the \emph{test step} consists of optimizing for labels at fixed $\theta_{\lambda}$ and scatter.
+The \emph{training step} consists of optimizing for the coefficient vector and scatter ($\theta_{\lambda}$ and $s_{\lambda}$) given fixed label vector (\emph{$\ell_n$}), and the \emph{test step} consists of optimizing for the labels at fixed $\theta_{\lambda}$ and $s_{\lambda}$.
 

### 2018-09-19 paper/draft.tex
minor format fix

@@
 \textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.
-JB: ADD CHI^2 FIT TO PLOT}\label{fig:demo_teff}
+JB: ADD CHI2 FIT TO PLOT}\label{fig:demo_teff}
 \end{figure}
@@
 \plotone{figures/cannon_chi_dist.png}
-\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee\-\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
+\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee --\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
 We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as ``safe''.

### 2018-09-19 paper/draft.tex
ref figure

@@
 Two tests of consistency were used to assess the validity of our derived test labels: (1) the self-test comparing training labels and test labels, and (2) the leave-one-out cross validation test, in which we train a model on all sources but \emph{n}, then apply the \emph{N-1} source trained model to get the labels for \emph{n}.
-Precision (scatter) and bias of the model for each test are calculated as the standard deviation and mean of the difference in training and test labels respectively.
-JB: POINT TO THESE RESULTS FOR THIS MODEL IN THE FIGURES
+Precision (scatter) and bias of the model for each test are calculated as the standard deviation and mean of the difference in training and test labels respectively (Figure \ref{fig:mann_validation}).
 

### 2018-09-19 paper/draft.tex
fig 3&4 description

@@
 
-Two tests of consistency were used to assess the validity of our derived test labels: (1) the self-test comparing training labels and test labels, and (2) the leave-one-out cross validation test, in which we train a model on all sources but \emph{n}, then apply the \emph{N-1} source trained model to get the labels for \emph{n}.
+Two tests of consistency were used to assess the validity of our derived test labels: (1) the self-test comparing training labels and test labels, and (2) the leave-one-out cross validation (LOOCV) test, in which we train a model on all sources but \emph{n}, then apply the \emph{N-1} source trained model to get the labels for \emph{n}.
 Precision (scatter) and bias of the model for each test are calculated as the standard deviation and mean of the difference in training and test labels respectively (Figure \ref{fig:mann_validation}).
 
-Another mode of analysis we can look at with \thecannon\ is how the derivative of the model changes with respect to given training parameter, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters.
-As discussed further in sections \ref{subsec:mann_results} and \ref{subsec:west_results}, we show the derivative plots of each of our two models and compute the error of the derivative at each pixel using a jackknife statistic:
+Another mode of analysis we can look at with \thecannon\ is how the derivative of the model changes with respect to given training parameters, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters.
+The top two panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh}
+
+The bottom panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh} show the derivative of flux with respect to temperature and metallicity at each pixel, taken at the median training values.
+In order to evaluate which spectral features show statistically significant change with respect to input label, we compute error of the derivative at each pixel using a jackknife statistic (with a 1-$\sigma$ level overplotted in red):
 \begin{equation}

### 2018-09-19 paper/draft.tex
describe cv test more

@@
 
-Two tests of consistency were used to assess the validity of our derived test labels: (1) the self-test comparing training labels and test labels, and (2) the leave-one-out cross validation (LOOCV) test, in which we train a model on all sources but \emph{n}, then apply the \emph{N-1} source trained model to get the labels for \emph{n}.
-Precision (scatter) and bias of the model for each test are calculated as the standard deviation and mean of the difference in training and test labels respectively (Figure \ref{fig:mann_validation}).
+Two tests of consistency were used to assess the validity of our model's labels: (1) the self-test comparing training label to test label values, and (2) the leave-one-out cross validation (LOOCV) test, in which we train a model on all sources but \emph{n}, then apply the \emph{N-1} source trained model to obtain the labels for star \emph{n}.
+Precision (scatter) and bias of the model for each test are calculated as the standard deviation and mean of the difference in training and test (or LOOCV) labels respectively (Figure \ref{fig:mann_validation}).
+Since the LOOCV test evaluates both how well the model reproduces the training values \emph{and} penalizes the model for over-fitting, we adopt the LOOCV scatter over the self-test scatter as the more conservative estimate of the model's precision.
 

### 2018-09-19 paper/draft.tex
ref result table

@@
 Since the LOOCV test evaluates both how well the model reproduces the training values \emph{and} penalizes the model for over-fitting, we adopt the LOOCV scatter over the self-test scatter as the more conservative estimate of the model's precision.
+The set of training, test, and LOOCV labels for each training source is reported in Table \ref{table:mann_results}.
 
 Another mode of analysis we can look at with \thecannon\ is how the derivative of the model changes with respect to given training parameters, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters.
-The top two panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh}
-
+The top two panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh} show two example spectra and model fits for two different temperatures (\ref{fig:demo_teff}), and two different temperatures (\ref{fig:demo_feh}), with atomic and molecular features identified by the abundance analysis of \citealt{Souto:2017}.
 The bottom panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh} show the derivative of flux with respect to temperature and metallicity at each pixel, taken at the median training values.
@@
 \begin{deluxetable*}{ccc|ccc|ccc|c}
-\tablecaption{Cannon results for Temperature/Metallicity model; see online journal for full table. \label{mann_results}}
+\tablecaption{Cannon results for Temperature/Metallicity model; see online journal for full table. \label{table:mann_results}}
 \tabletypesize{\scriptsize}
@@
 \begin{deluxetable*}{ccc|ccc|c}
-\tablecaption{Cannon results for Spectral Type model. \label{west_results}}
+\tablecaption{Cannon results for Spectral Type model. \label{table:west_results}}
 \tabletypesize{\scriptsize}

### 2018-09-19 paper/draft.tex
ref west figures

@@
 As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates \thecannon\ does a poor job at extrapolating to labels outside of the training sample space.
-
-JB: NOW POINT HERE TO THE RESULTS OF THE CROSS-VALIDATION AND ETC.
+Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows the self and LOOCV tests for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.
 

### 2018-09-19 paper/draft.tex
selection info

@@
 From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.
-Although our catalog reports the temperatures, metallicities, and spectral types for XXX sources out of 14,827 with $\chi^2<200,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
+Although our catalog reports the temperatures, metallicities, and spectral types for 10,311 sources out of 14,827 with $\chi^2<200,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
 
@@
 
-\item \textbf{Pre-main sequence, multiple system and subdwarf (?) cut:} in \gaia\ and \zmass\ we apply $.7<J-K<1.1$ and points in the selction region within the points $\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ of ($BP-RP$, $M_G$) to cut remaining likely pre-main sequence, multiple sources and subdwarfs.
+\item \textbf{Pre-main sequence, multiple system and subdwarf cut:} in \gaia\ and \zmass\ we apply $.7<J-K<1.1$ and points in the selction region within the points $\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ of ($BP-RP$, $M_G$) to cut remaining likely pre-main sequence, multiple sources above the main sequence and subdwarfs below the main sequence.
 
@@
 \caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities.
-\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.
-JB: ADD CHI2 FIT TO PLOT}\label{fig:demo_teff}
+\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.}\label{fig:demo_teff}
 \end{figure}

### 2018-09-20 paper/draft.tex
describe likelihood fn

@@
 Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector $\ell$ and and coefficient vector \emph{$\theta$}.
-So for each star \emph{n}, at pixel wavelength \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
+So for each star \emph{n}, at wavelength pixel \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
 \begin{equation}
-	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + {\rm noise}
+	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + {\rm noise} \label{eqn:cannon}
 \end{equation} 
 We use the noise model is ${\rm noise}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
+Equation \ref{eqn:cannon} corresponds to the single-pixel log-likelihood function:
+\begin{equation}
+	\ln p(f_{n\lambda}|\theta^T_{\lambda}, \ell_n, s^2_n) = 
+	-\frac{1}{2}\frac{[f_{n\lambda} - \theta_{\lambda}^{T} \cdot \ell_{n}]^2}{s_{\lambda}^2 + \sigma_{n\lambda}^2} - \frac{1}{2} \ln(s_{\lambda}^2 + \sigma_{n\lambda}^2) 
+	\label{eqn:likelihood}
+\end{equation}
+the probability density function of the measured flux, given the labels, coefficients and scatters.
 
@@
 
-The \emph{training step} consists of optimizing for the coefficient vector and scatter ($\theta_{\lambda}$ and $s_{\lambda}$) given fixed label vector (\emph{$\ell_n$}), and the \emph{test step} consists of optimizing for the labels at fixed $\theta_{\lambda}$ and $s_{\lambda}$.
-
-% JB: The training step description: OPTIMIZE THETA AT FIXED LABELS.
-% JB: Test step description: OPTIMIZE LABELS AT FIXED THETA.
-
+The \emph{training step} consists of optimizing the likelihood function (Equation \ref{eqn:likelihood}) for the coefficient vector and scatter ($\theta_{\lambda}$ and $s_{\lambda}$) given fixed label vector (\emph{$\ell_n$}) constructed from the reference labels, and the \emph{test step} consists of optimizing the likelihood function for the labels at fixed $\theta_{\lambda}$ and $s_{\lambda}$ obtained in the training step (see \citealt{Ness:2015} for further description).
 Thus \thecannon\ is both a generative model for the stellar spectra, and a method for obtaining 

### 2018-09-20 paper/draft.tex
model assumptions

@@
 
-\thecannon\ is a fully empirical model, which applies no use of line lists or radiative transfer models.
-Describe the assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label.
+\thecannon\ is a fully empirical model, which employs no use of line lists or radiative transfer models.
+Instead, \thecannon\ is a regression model which relies on two assumptions: first that sources with identical labels have near-identical flux at each pixel, and second that the expected flux at each pixel varies continuosly with change in label.
 

### 2018-09-20 paper/draft.tex
chi2 cut

@@
 From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.
-Although our catalog reports the temperatures, metallicities, and spectral types for 10,311 sources out of 14,827 with $\chi^2<200,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
+Although our catalog reports the temperatures, metallicities, and spectral types for 10,311 sources out of 14,827 with $\chi^2<100,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
 

### 2018-09-20 paper/draft.tex
add radius to table

@@
 
-\item \textbf{Training bound selection:} Because extrapolation with \thecannon\ is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<[Fe/H]<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
+\item \textbf{Training bound selection:} Because extrapolation with \thecannon\ is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<$[Fe/H]$<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
 \end{enumerate}
@@
 {\tt RA} 				& deg	& \apogee\ radial ascension angle	\\
+{\tt RADIUS} 			& R$_\odot$	& Radius determined by \citealt{Mann:2015} relation	\\
 {\tt RP$\_$MAG} 		& mag	& \gaia\ $RP$ band photometry	\\
@@
 {\tt RV$\_$GAIA$\_$ERR}		& km s$^{-1}$	& \gaia\ radial velocity uncertainty	\\
-{\tt SPT} 				& 	& West-trained \cannon\ spectral subtype ($M0-M9$)	\\
+{\tt SPT} 				& 	& West-trained \cannon\ spectral subtype (M0$-$M9)	\\
 {\tt TEFF} 				& K	& Mann-trained \cannon\ effective temperature	\\

### 2018-09-20 paper/draft.tex
add acknowledgments'

@@
 
-The final selection for our ``safe'' is displayed in Figure \ref{fig:safe_selection}.
-We also note that applying our model requires very little computational demand: the time to run a model on all 14,827 sources was two minutes on a personal laptop.
+The three panels of \ref{fig:safe_selection} show before and after selection of the sources in \gaia\ color-magnitude space, colored by \thecannon\ determined temperatures, metallicities, and spectral types, with their respective training samples overplotted in orange.
+Comparisons between the West-trained spectral types and Mann-trained temperatures and metallicities are shown in Figure \ref{fig:west_vs_mann}: left shows the slight negative correlation between spectral type and [Fe/H], and right shows the nonlinear decrease in temperature with visible metallicity gradient.
+We also note that applying our model requires very little computational demand: the time to train and test a model on all 14,827 sources was two minutes on a personal laptop.
 
-JB: NOW POINT TO RESULTS FIGURES AND SUMMARIZE THEM, IF THERE IS ANYTHING ELSE?
+% JB: NOW POINT TO RESULTS FIGURES AND SUMMARIZE THEM, IF THERE IS ANYTHING ELSE?
 
@@
 \acknowledgements
-We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
+We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Diogo Souto (Observatário Nacional/MCTI) and Keivan Stassun (Vanderbuilt) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
 

### 2018-09-20 paper/draft.tex
selection fig descriptions

@@
 The three panels of \ref{fig:safe_selection} show before and after selection of the sources in \gaia\ color-magnitude space, colored by \thecannon\ determined temperatures, metallicities, and spectral types, with their respective training samples overplotted in orange.
+Each plot shows the expected gradient: temperature increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly to the main sequence branch.
+Additionally we notice a slight deviation at the upper slice of the main sequence where \cannon\ temperatures, metallicities and spectral types all appear lower than the gradient trends in the rest of the branch, which is most likely the binary sequence.
 Comparisons between the West-trained spectral types and Mann-trained temperatures and metallicities are shown in Figure \ref{fig:west_vs_mann}: left shows the slight negative correlation between spectral type and [Fe/H], and right shows the nonlinear decrease in temperature with visible metallicity gradient.
 We also note that applying our model requires very little computational demand: the time to train and test a model on all 14,827 sources was two minutes on a personal laptop.
-
-% JB: NOW POINT TO RESULTS FIGURES AND SUMMARIZE THEM, IF THERE IS ANYTHING ELSE?
 
@@
 Overplotted with orange triangles are the M15 and W11 training samples, for their respective \cannon\ test labels.
-Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases with increasing absolute magnitude as expected.
+Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly up from the main sequence branch as expected.
 Deviations from these gradients seen at the upper slice of the main sequence is likely remaining contamination from the binary sequence.

### 2018-09-20 paper/draft.tex
add detail

@@
 The spectral type sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from \sdss.
-For each source in the catalog, spectral types were determined both through an automated routine using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
+For each source in the catalog, spectral types were determined both through an automated routine for comparing spectral type templates to data using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
 

### 2018-09-20 paper/draft.tex
age note

@@
 
-\color{gcolor}{HOGG: just a reflection, I think the scope of this paper has shifted a lot from just being a technical paper about \thecannon\ working on M dwarfs, to drifting more towards M dwarf science (partly because I've learned a lot more about the properties/characteristics of M dwarfs, and the Gaia has given us a lot of intuition that we didn't have before).
+% \color{gcolor}{HOGG: just a reflection, I think the scope of this paper has shifted a lot from just being a technical paper about \thecannon\ working on M dwarfs, to drifting more towards M dwarf science (partly because I've learned a lot more about the properties/characteristics of M dwarfs, and the Gaia has given us a lot of intuition that we didn't have before).
 Anyways we should decide whether the paper is more of the former or the latter and maybe re-write the abstract.
@@
 
-I think maybe the biggest angle is that are illustrating what can be leveraged out of two very high quality datasets (\apogee\ and \gaia), and we are the first to identify and collect a large (atmospheric parameters, 6D phase space) that have in the past been very difficult to measure.
+% I think maybe the biggest angle is that are illustrating what can be leveraged out of two very high quality datasets (\apogee\ and \gaia), and we are the first to identify and collect a large (atmospheric parameters, 6D phase space) that have in the past been very difficult to measure.
 Is this what we want to emphasize?
 
-Also an open question along these lines is whether we want to estimate any additional physical parameters for our sample, for example maybe adding radii (from the \citealt{Mann:2015} \teff-radius or $M_K$-radius relations), or masses from his $M_K$-mass relation he's publishing, or maybe there are some other fundamental parameters that would be interesting (and straightforward) to add?
+% Also an open question along these lines is whether we want to estimate any additional physical parameters for our sample, for example maybe adding radii (from the \citealt{Mann:2015} \teff-radius or $M_K$-radius relations), or masses from his $M_K$-mass relation he's publishing, or maybe there are some other fundamental parameters that would be interesting (and straightforward) to add?
 It seems like just because we have \gaia\ parallaxes we should go ahead and do it?
@@
 
-Another open question is about the [Fe/H] validation, and whether we are happy with just the cmd plot of metallicity. I think our ideas for other tests were:
-\begin{itemize}
-\item Looking at M+M binary pairs: there are 174 pairs and it looks by eye that most of the pair lines are parallel to the main sequence.
+% Another open question is about the [Fe/H] validation, and whether we are happy with just the cmd plot of metallicity. I think our ideas for other tests were:
+% \begin{itemize}
+% \item Looking at M+M binary pairs: there are 174 pairs and it looks by eye that most of the pair lines are parallel to the main sequence.
 I don't know if there's some more clever way to see if the pairs lie in the same metallicity space of the cmd, so this is probably not that useful
-\item Overplotting some isochrones at varying metallicities--which I'm still working on
-\end{itemize}
+% \item Overplotting some isochrones at varying metallicities--which I'm still working on
+% \end{itemize}
 
-Also I still have to make the vertical action plot.
-}\color{black}
+% Also I still have to make the vertical action plot.
+% }\color{black}
 
@@
 place, which was in the context of optical spectroscopy.
-That is, the model to be applied to the \apogee\ data is learned from the \apogee\ data alone
+That is, the model to be applied to the \apogee\ data is learned from the \apogee\ data alone.
 This obviates physical models in the infrared, and capitalizes on the excellent data quality
@@
 For the physical parameter model we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
-We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr) and likely slowly rotating sources.
+We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr for the published sample estimated by M15) and likely slowly rotating sources.
 While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.

### 2018-09-20 paper/draft.tex
completed bullet
rewrite abstract bits

@@
 \thecannon\ is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution, near-infrared \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
-The lack of reliable stellar models in the near infrared for temperatures $<3500$\,K motivates the extension of this approach to M-dwarf stars, but with new training labels from optical surveys that overlap \sdssiv\ \apogee.
+The lack of reliable atmospheric models in the near infrared for temperatures cooler than $\sim3500$\,K motivates the extension of this approach to M-dwarf stars, but with calibrated training labels from optical surveys that overlap \sdssiv\ \apogee.
 Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
-These are parameter-space precision statements, but the model is accurate in the data space: It accurately predicts \apogee\ spectral pixels at the percent level.
-We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, and demonstrate that \thecannon\ acting on \apogee\ spectra can infer spectral types with predictive accuracy of 0.9 types.
-The trained models are employed to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in \apogee, selected in part on \gaia\ photometry and astrometry.
-We validate our results using photometric temperature indicators and the \gaia\ color--magnitude diagram.
-As a side result, we determine empirical relationships between spectral type and temperature and metallicity.
+These are parameter-space precision statements, but the model is accurate in the data space: it accurately predicts \apogee\ spectral pixels at the percent level.
+We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, with predictive accuracy of 0.9 types.
+From the \textsl{APOGEE DR14} -- \textsl{Gaia DR2} overlap we select a test sample and employ our trained models to determine temperatures, metallicities and spectral subtypes for nearly all of the M dwarfs in \apogee\ (10,311 sources), and validate our \cannon\ derived \teff\ against color-temperature relations and [Fe/H] against isochrone models in \gaia\ color-magnitude space.
+Finally we use
 These results have implications for the NASA \tess\ Mission and other projects that involve large samples of M dwarfs.
+
+% The trained models are employed to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in \apogee, selected in part on \gaia\ photometry and astrometry.
+% We validate our results using photometric temperature indicators and the \gaia\ color--magnitude diagram.
+% As a side result, we determine empirical relationships between spectral type and temperature and metallicity.
 
@@
 
-JB ALSO POINT TO FIGURES THAT CONNECT TYPES TO PARAMETERS. AND DISCUSS.
-
-JB DO WE WANT A FIGURE THAT SHOWS ABUNDANCES AS A FUNCTION OF MILKY WAY POSITION OR ACTIONS?
+% JB DO WE WANT A FIGURE THAT SHOWS ABUNDANCES AS A FUNCTION OF MILKY WAY POSITION OR ACTIONS?
 

### 2018-09-20 paper/draft.tex
radii section
mention radius in abstract

@@
 These are parameter-space precision statements, but the model is accurate in the data space: it accurately predicts \apogee\ spectral pixels at the percent level.
-We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging from M0 to M9, with predictive accuracy of 0.9 types.
+We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging M0 to M9, to predictive accuracy of 0.9 types.
 From the \textsl{APOGEE DR14} -- \textsl{Gaia DR2} overlap we select a test sample and employ our trained models to determine temperatures, metallicities and spectral subtypes for nearly all of the M dwarfs in \apogee\ (10,311 sources), and validate our \cannon\ derived \teff\ against color-temperature relations and [Fe/H] against isochrone models in \gaia\ color-magnitude space.
-Finally we use
+Finally we use empirically callibrated relations to estimate radii from $M_K$ absolute magnitudes and \cannon\ temperatures.
 These results have implications for the NASA \tess\ Mission and other projects that involve large samples of M dwarfs.
@@
 We also note that applying our model requires very little computational demand: the time to train and test a model on all 14,827 sources was two minutes on a personal laptop.
+
+\subsection{Radii}
+Using empirically calibrated relations from \citealt{Mann:2015} we determine radii... some diagnostics...
 

### 2018-09-21 paper/draft.tex
metallicity paragraph

@@
 
-% \color{gcolor}{HOGG: just a reflection, I think the scope of this paper has shifted a lot from just being a technical paper about \thecannon\ working on M dwarfs, to drifting more towards M dwarf science (partly because I've learned a lot more about the properties/characteristics of M dwarfs, and the Gaia has given us a lot of intuition that we didn't have before).
+% HOGG: just a reflection, I think the scope of this paper has shifted a lot from just being a technical paper about \thecannon\ working on M dwarfs, to drifting more towards M dwarf science (partly because I've learned a lot more about the properties/characteristics of M dwarfs, and the Gaia has given us a lot of intuition that we didn't have before).
 Anyways we should decide whether the paper is more of the former or the latter and maybe re-write the abstract.
@@
 Maybe also to do a full-circle comparison with our test sample labels.
-
-% Another open question is about the [Fe/H] validation, and whether we are happy with just the cmd plot of metallicity. I think our ideas for other tests were:
-% \begin{itemize}
-% \item Looking at M+M binary pairs: there are 174 pairs and it looks by eye that most of the pair lines are parallel to the main sequence.
-I don't know if there's some more clever way to see if the pairs lie in the same metallicity space of the cmd, so this is probably not that useful
-% \item Overplotting some isochrones at varying metallicities--which I'm still working on
-% \end{itemize}
-
-% Also I still have to make the vertical action plot.
-% }\color{black}
 
@@
 The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
-\color{gcolor}{What about \apogee\'s wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?
-What }\color{black}
+What about \apogee\'s wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?
 
@@
 There are working and reliable models of M dwarf stars in the optical, which have been
-used successfully, but (as we note above) there isn't as much success with such modeling in the infrared,
-and not at \apogee\ resolution.
-\color{gcolor}{HOGG: I think we can make a more specific set of points here that would be more interesting to M dwarf people: temperatures and (especially) metallicities are difficult to measure accurately for an M dwarf, but people have done a lot of work to build calibrated empirical relations for metallicities and temperatures that are very precise.
-For example \citealt{Terrien:2012}, \citealt{RojasAyala:2010}, Mann 2013 (and there's probably some other examples (Newton, etc.) and I have more literature to look at...) have a metallicity callibration based off of an FGK+M star samples, from J, H, K or optical band spectra using the equivalent width of some strong line/index (like CaI, KI, etc).
+used successfully, but (as we note above) there isn't as much success with such modeling in the infrared, and not at \apogee\ resolution.
+
+In particular, parameters such as metallicity can be difficult to determine precisely, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths.
+A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star (under the assumption kinematically similar pairs are co-evolutionary), to developed precise empirical relations (as good as $\sim$0.07dex) between the precise adopted metallicities of the FGK primary, and strong, isolated lines in the optical spectra of the M dwarf (e.g.
+\citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
+However these metallicity callibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant callibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.
+\thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
+
+% HOGG: I think we can make a more specific set of points here that would be more interesting to M dwarf people: temperatures and (especially) metallicities are difficult to measure accurately for an M dwarf, but people have done a lot of work to build calibrated empirical relations for metallicities and temperatures that are very precise.
+% For example \citealt{Terrien:2012}, \citealt{RojasAyala:2010}, Mann 2013 (and there's probably some other examples (Newton, etc.) and I have more literature to look at...) have a metallicity callibration based off of an FGK+M star samples, from J, H, K or optical band spectra using the equivalent width of some strong line/index (like CaI, KI, etc).
 However these metallicity callibrations are inaccesible for APOGEE spectra, because the wavelength coverage does not include the relevant callibration indices.
-\thecannon\ bridges this gap: it doesn't require using specific lines (which would probably be a mess to deconvolve at high resolution with other blends even if we had the right lines); it allows us to use a lot of low level metallicity information present in thousands of lines; and (as we demonstrate) it does so with very good precision.}\color{black}
+\thecannon\ bridges this gap: it doesn't require using specific lines (which would probably be a mess to deconvolve at high resolution with other blends even if we had the right lines); it allows us to use a lot of low level metallicity information present in thousands of lines; and (as we demonstrate) it does so with very good precision.
 

### 2018-09-21 paper/draft.tex
rm old comments
add ref

@@
 
-% HOGG: just a reflection, I think the scope of this paper has shifted a lot from just being a technical paper about \thecannon\ working on M dwarfs, to drifting more towards M dwarf science (partly because I've learned a lot more about the properties/characteristics of M dwarfs, and the Gaia has given us a lot of intuition that we didn't have before).
-Anyways we should decide whether the paper is more of the former or the latter and maybe re-write the abstract.
-Right now it just feels a bit hacked-together, since this is basically just my AAS abstract that we kept adding some sentences to as we went, and I think it needs a better frame.
-I'm still trying to get a sense of what is scientifically interesting about this sample, and I've been struggling to pin point what the real relevance of this work is: it's not exactly just that the Cannon is precise for M dwarfs (this original scope, like an extension of Ness, is too boring I think, given what we've learned), and it's not really about M dwarf physics or why we are better than models (this is too beyond us to argue).
-
-% I think maybe the biggest angle is that are illustrating what can be leveraged out of two very high quality datasets (\apogee\ and \gaia), and we are the first to identify and collect a large (atmospheric parameters, 6D phase space) that have in the past been very difficult to measure.
-Is this what we want to emphasize?
-
-% Also an open question along these lines is whether we want to estimate any additional physical parameters for our sample, for example maybe adding radii (from the \citealt{Mann:2015} \teff-radius or $M_K$-radius relations), or masses from his $M_K$-mass relation he's publishing, or maybe there are some other fundamental parameters that would be interesting (and straightforward) to add?
-It seems like just because we have \gaia\ parallaxes we should go ahead and do it?
-Maybe also to do a full-circle comparison with our test sample labels.
-
 \end{abstract}
@@
 In particular, parameters such as metallicity can be difficult to determine precisely, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths.
-A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star (under the assumption kinematically similar pairs are co-evolutionary), to developed precise empirical relations (as good as $\sim$0.07dex) between the precise adopted metallicities of the FGK primary, and strong, isolated lines in the optical spectra of the M dwarf (e.g.
+A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
 \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
@@
 \thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
-
-% HOGG: I think we can make a more specific set of points here that would be more interesting to M dwarf people: temperatures and (especially) metallicities are difficult to measure accurately for an M dwarf, but people have done a lot of work to build calibrated empirical relations for metallicities and temperatures that are very precise.
-% For example \citealt{Terrien:2012}, \citealt{RojasAyala:2010}, Mann 2013 (and there's probably some other examples (Newton, etc.) and I have more literature to look at...) have a metallicity callibration based off of an FGK+M star samples, from J, H, K or optical band spectra using the equivalent width of some strong line/index (like CaI, KI, etc).
-However these metallicity callibrations are inaccesible for APOGEE spectra, because the wavelength coverage does not include the relevant callibration indices.
-\thecannon\ bridges this gap: it doesn't require using specific lines (which would probably be a mess to deconvolve at high resolution with other blends even if we had the right lines); it allows us to use a lot of low level metallicity information present in thousands of lines; and (as we demonstrate) it does so with very good precision.
 

### 2018-09-21 paper/draft.tex
wording

@@
 \apogee\ is primarily targeted for bright stellar populations, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution.
-However numerous cool, main sequence sources have also been observed either as targets proposed by the M dwarf ancilliary survey (1,350 sources selected by proper motions from the  \citealt{Desphande:2013}), or randomly by \apogee\, which we quantify in this study to be over 10,000 in total.
+However numerous cool, main sequence sources have also been observed either as targets proposed by the M dwarf ancilliary survey (1,350 sources selected by proper motions; \citealt{Desphande:2013}), or randomly by \apogee\, which we quantify in this study to be over 10,000 in total.
 

### 2018-09-21 paper/draft.tex
typo
update

@@
 
-We note that surface gravity is not inclided as a training label.
+We note that surface gravity is not included as a training label.
 The reason for this is that for main-sequence M dwarfs, the parameter is almost entirely redundant with metallicity.
@@
 
-In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of \aspcap\ temperature values using a combination of \zmass\ and visual band photometries.
-\color{gcolor}{Brief description of the color-teff relations from \citealt{Casagrande:2008}, \citealt{Boyajian:2013}, and \citealt{Mann:2015}...}\color{black}.
+As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations, which use combination of \zmass\ and visual band photometries to predict temperatures.
+SUMMARIZE \citealt{Casagrande:2008}, \citealt{Boyajian:2013}, and \citealt{Mann:2015}
 
 To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
-In Figure \ref{fig:aspcap_cannon_validation} we repeat the temperature validation tests for each of the three relations and compare the \aspcap\ temperatures for the same sources.
+Figure \ref{fig:aspcap_cannon_validation} shows the temperature validation tests for each of the three relations and compared to the \aspcap\ temperatures for the same sources.
 \color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}
-
-\color{gcolor}{What are all of the parameters we want to report and what do we want to emphasize? Are there other parameters (such as mass) that might be straightforward to estimate?}\color{black}
 

### 2018-09-24 paper/draft.tex
removed some HOGG comments; tweaks

@@
 \hypersetup{linkcolor=bcolor,citecolor=bcolor,filecolor=cyan,urlcolor=bcolor}
-
-\newcommand{\vdag}{(v)^\dagger}
-\newcommand\aastex{AAS\TeX}
-\newcommand\latex{La\TeX}
 
@@
 \begin{abstract}
-\thecannon\ is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution, near-infrared \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
-The lack of reliable atmospheric models in the near infrared for temperatures cooler than $\sim3500$\,K motivates the extension of this approach to M-dwarf stars, but with calibrated training labels from optical surveys that overlap \sdssiv\ \apogee.
+\thecannon\ is a flexible, data-driven spectral-modeling and parameter-inference framework, demonstrated on high-resolution, near-infrared \sdssiv\ \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
+The lack of reliable atmospheric models in the near infrared for temperatures cooler than $\sim3500$\,K motivates the extension of this approach to M-dwarf stars, but with calibrated training labels from optical surveys that overlap \apogee.
 Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
@@
 We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging M0 to M9, to predictive accuracy of 0.9 types.
-From the \textsl{APOGEE DR14} -- \textsl{Gaia DR2} overlap we select a test sample and employ our trained models to determine temperatures, metallicities and spectral subtypes for nearly all of the M dwarfs in \apogee\ (10,311 sources), and validate our \cannon\ derived \teff\ against color-temperature relations and [Fe/H] against isochrone models in \gaia\ color-magnitude space.
+From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we select a test sample and employ our trained models to determine temperatures, metallicities and spectral subtypes for a conservatively selected sample of M dwarfs in \apogee\ (10,311 sources), and validate our \cannon\ derived \teff\ against color--temperature relations and [Fe/H] against isochrone models in \gaia\ color-magnitude space.
 Finally we use empirically callibrated relations to estimate radii from $M_K$ absolute magnitudes and \cannon\ temperatures.
-These results have implications for the NASA \tess\ Mission and other projects that involve large samples of M dwarfs.
-
-% The trained models are employed to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in \apogee, selected in part on \gaia\ photometry and astrometry.
-% We validate our results using photometric temperature indicators and the \gaia\ color--magnitude diagram.
-% As a side result, we determine empirical relationships between spectral type and temperature and metallicity.
-
+This project has implications for the NASA \tess\ Mission and other surveys that involve large samples of M dwarfs. 
 \end{abstract}
@@
 The \aspcap\ pipeline releases several types of data files, with various levels of processing: {\tt\string ap1D} (the raw one-dimensional spectra for individual visits), {\tt\string apVisit} (the individual visit spectra with telluric subtraction), {\tt\string apStar} (the co-added {\tt\string apVisit} spectra), and most processed {\tt\string aspcapStar} spectra.
-For this work we use {\tt\string aspcapStar} files, which contains the continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
-\color{gcolor}{HOGG: do you want to comment on continuum normalization issues?}\color{black}
+For this work we use {\tt\string aspcapStar} files, which contains the pseudo-continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
+In other work we have recommended an alternative pseudo-continuum normalization \citep{Ness:2015}, but we don't find obvious issues with the normalization in what follows, so we stay with the survey pipeline outputs.
 
@@
 
-\thecannon\ model can in principle be trained on any physical or empirical labels available, giving a wide range of flexibility to the model, however choosing a training sample with high quality (accurate and precise) is critical to the performance of \thecannon.
-Limitations to consider are that test labels from \thecannon\ are only accurate if the training labels accurate, and only precise if the training labels are measured using a consistent method.
-Also choosing a sample with large dynamic range spanning the entire parameter space of interest is important, as \thecannon\ does not extrapolate well outside the parameter space of the training sample.
-\color{gcolor}{HOGG: More notes to add here?}\color{black}
+\thecannon\ model can in principle be trained on any physical or empirical labels available, giving a wide range of flexibility to the model.
+However choosing a training sample with high quality labels (precise, but also preferably accurate) is critical to the performance of \thecannon.
+Limitations of \thecannon\ include that test (output) labels are only accurate if the training labels are accurate, and only precise if the training labels are measured consistently across the training sample.
+Also, it is critical to have a training sample with the dynamic range to span the entire parameter space of interest, as \thecannon\ does not extrapolate well outside the parameter space of the training sample.
+Finally, \thecannon\ deeply assumes that the dependence of spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions---if that isn't true, there will be features that \thecannon\ cannot reproduce.
 
@@
 
-\color{gcolor}{HOGG: Some commentary on why $\chi^2$ is higher than expected. Also explain why this means that the model is good to percent level.}\color{black}
+The spectra contain roughly 8000 pixels,
+so we might expect the $\chi^2$ values to be close to 8000 in magnitude.
+They are much higher.
+This discrepancy flows from the fact that---while the spectral model is good at the percent
+level---the signal-to-noise of the typical spectra are more than 100.
+That is, the $\chi^2$ values do show that the model is not good in the frequentist
+sense; it is only good at the percent level.
+This percent-level accuracy of the model in the data domain is also demonstrated...
+\color{gcolor}{BIRKY: Can you demonstrate this by showing the RMS residuals, putting it in as a figure, and putting this paragraph back to me?
+I am imagining a figure that shows a scatter plot of T vs RMS residual, colored by metallicity.
+Or something like that.}\color{black}
 

### 2018-10-07 paper/draft.tex
wording

@@
 The lack of reliable atmospheric models in the near infrared for temperatures cooler than $\sim3500$\,K motivates the extension of this approach to M-dwarf stars, but with calibrated training labels from optical surveys that overlap \apogee.
-Using a training sample of 87 M dwarfs in the temperature range $2859 < T_{\rm eff} < 4131$\,K, and metallicty range $-0.48 < \mathrm{[Fe/H]}< 0.49$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
+Using a training sample of 87 M dwarfs in the temperature range $2860 < T_{\rm eff} < 4130$\,K, and metallicty range $-0.5 < \mathrm{[Fe/H]}< 0.5$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
 These are parameter-space precision statements, but the model is accurate in the data space: it accurately predicts \apogee\ spectral pixels at the percent level.
 We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging M0 to M9, to predictive accuracy of 0.9 types.
-From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we select a test sample and employ our trained models to determine temperatures, metallicities and spectral subtypes for a conservatively selected sample of M dwarfs in \apogee\ (10,311 sources), and validate our \cannon\ derived \teff\ against color--temperature relations and [Fe/H] against isochrone models in \gaia\ color-magnitude space.
+From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we select a test sample and employ our trained models to determine temperatures, metallicities and spectral subtypes for a conservatively selected sample of M dwarfs in \apogee\ (10,311 sources), and compare our \cannon\ derived \teff\ against color--temperature relations and [Fe/H] against isochrone models in \gaia\ color-magnitude space.
 Finally we use empirically callibrated relations to estimate radii from $M_K$ absolute magnitudes and \cannon\ temperatures.

### 2018-10-09 paper/draft.tex
isochrone fig
more caption detail

@@
 
-Iron abundances ([Fe/H]) are assignied to the physical parameter sample based on the strenngth of metal-sensitive lines in the near-infrared SpeX spectra \citep{RojasAyala:2010} using the calibration from \citet{Mann:2013a}.
+Iron abundances ([Fe/H]) are assignied to the physical parameter sample based on the strength of metal-sensitive lines in the near-infrared SpeX spectra \citep{RojasAyala:2010} using the calibration from \citet{Mann:2013a}.
 The relation between these lines and an absolute [Fe/H] scale is calibrated using wide binaries containing and F, G, or K-type primary and an M dwarf companion, under the assumption that binaries formed from the same molecular cloud and therefore have the same metallicity \citep{Bonfils:2005}.
 Uncertainties are estimated to be $\simeq$0.08\,dex based on irreducable scatter in the empirical relation between selected lines and the assigned [Fe/H] from the primary star.
-As with \teff\ relative errors on [Fe/H] are smaller, estimated to be 0.04-0.06\,dex over most of the temperature and metallicity range considered here.
+As with \teff\ relative errors on [Fe/H] are smaller, estimated to be 0.04--0.06\,dex over most of the temperature and metallicity range considered here.
 
@@
 
-\item[-] Is it interesting to point out the comoving pairs we found? (33 FGK-M, 13 WD-M, and 174 M-M pairs)?
-
 \item[-] Things we learned along the way: For example, not all training sets worked well in our experiments, probably because they were too small or had too little dynamic range.
@@
 \item[-] HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?
+\end{enumerate}
+
+Future work:
+\begin{enumerate}
+\item[-] Create a binary sample; using color/magnitude \citealt{Widmark:2018} and by spectroscopy \citealt{ElBadry:2018}
 \end{enumerate}
@@
 
+%---------------------------
+\begin{figure}[ht]
+\plotone{figures/isochrone_cmd.png}
+\caption{Comparison of Cannon [Fe/H] as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
+The four panels show five metallicity tracks (ranging $-0.5<$[M/H]$<0.5$) over varying ages (1, 3, 5, and 7 Gyr).
+\label{fig:isochrone_cmd}}
+\end{figure}
+
+\begin{figure}[ht]
+\plotone{figures/mann_radii.png}
+\caption{Comparison of radii estimated from M15 relations. \label{fig:radii}}
+\end{figure}
+
 %%==========================================================================================

### 2018-10-12 paper/draft.tex
improve captions

@@
 \plotone{figures/aspcap_cannon_validation.png}
-\caption{Validation test of \cannon\ temperatures (left), as compared to \aspcap\ temperatures (right) using three different color-temperature relations derived from \citealt{Mann:2015}, \citealt{Casagrande:2008}, and \citealt{Boyajian:2013}.
-We find \cannon\ temperatures to be significantly more consistent with all three relations in terms of bias and scatter.
-\label{fig:aspcap_cannon_validation}}
+\caption{\cannon\ (right column) and \aspcap\ (left column) temperatures as compared to temperatures estimated from visible and infrared color-temperature relations from three different literature sources for a subsample of 6,000 sources.
+Shown are \cannon\ and \aspcap\ temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and [Fe/H] dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2013} relation (bottom).
+We find \aspcap\ overestimates \teff\ on average by about 130K for the M15 and B13 relations about 220K for the C08.
+Additionally we see the scatter much higher for the \aspcap\ temperatures ($\sim220$K) compared to the scatter of \cannon\ temperatures ($\sim80-100$K), the scatter for both \aspcap\ and \cannon\ temperatures largely due to variation in metallicity.
+\label{fig:teff_comparisons}}
 \end{figure}
@@
 \plotone{figures/mann_radii.png}
-\caption{Comparison of radii estimated from M15 relations. \label{fig:radii}}
+\caption{Comparison of radii estimated from M15 relations: on the x-axis radii are derived from  K magnitudes and \cannon\ [Fe/H], on the y-axis radii are derived from \cannon\ \teff\ and [Fe/H].
+\label{fig:radii}}
 \end{figure}

### 2018-10-20 paper/draft.tex
HOGG notes

@@
 
-\color{gcolor}{Some note on what lines vary significantly with parameter...}\color{black}
+\color{gcolor}{HOGG/ADAM: Do we want some note on what lines vary significantly with parameter?
+I could summarize the which kinds of lines show increasing or decreasing trends, but have no idea how to interpret from there.}\color{black}
 
@@
 To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
-Figure \ref{fig:aspcap_cannon_validation} shows the temperature validation tests for each of the three relations and compared to the \aspcap\ temperatures for the same sources.
+Figure \ref{fig:teff_comparisons} shows the temperature validation tests for each of the three relations and compared to the \aspcap\ temperatures for the same sources.
 \color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}
@@
 
-\item[-] Some discussion of whether the derivatives are sensible or surprising.
+\item[-] Some discussion of whether the derivatives are sensible or surprising. \color{gcolor}{HOGG: I'm not sure what to say here}\color{black}
 
-\item[-] \thecannon\ as a test of evaluating training parameter accuracy?
+\item[-] \color{gcolor}{HOGG:\thecannon\ as a test of evaluating training parameter accuracy?}\color{black}
 
-\item[-] Discussion of validation tests and ground truth
+\item[-] \color{gcolor}{HOGG: Discussion of validation tests and ground truth}\color{black}
 
-\item[-] HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?
 \end{enumerate}
@@
 \item[-] Create a binary sample; using color/magnitude \citealt{Widmark:2018} and by spectroscopy \citealt{ElBadry:2018}
+
+\item[-] \color{gcolor}{HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?}\color{black}
 \end{enumerate}
+
+\color{gcolor}{HOGG: Is a conclusion section necessary?}\color{black}
 

### 2018-11-27 paper/draft.tex
add discussion content

@@
 Using a training sample of 87 M dwarfs in the temperature range $2860 < T_{\rm eff} < 4130$\,K, and metallicty range $-0.5 < \mathrm{[Fe/H]}< 0.5$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
-These are parameter-space precision statements, but the model is accurate in the data space: it accurately predicts \apogee\ spectral pixels at the percent level.
+The model is accurate in the data space: it accurately predicts \apogee\ spectral pixels at the one-percent level.
 We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging M0 to M9, to predictive accuracy of 0.9 types.
@@
 
-\subsection{Validation of parameters}
+\subsection{Color-Temperature Comparisons}
 
-As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations, which use combination of \zmass\ and visual band photometries to predict temperatures.
-SUMMARIZE \citealt{Casagrande:2008}, \citealt{Boyajian:2013}, and \citealt{Mann:2015}
+As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures.
+To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (\textsl{APASS}; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by \textsl{APASS}, and $JHK$ photometries from \apogee.
+Figure \ref{fig:teff_comparisons} shows \cannon\ vs.
+photometric temperatures on the right, and \aspcap\ vs.
+photometric temperatures on the left for each of the 6,257 sources, colored by their respective spectroscopic metallicities.
 
-To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
-Figure \ref{fig:teff_comparisons} shows the temperature validation tests for each of the three relations and compared to the \aspcap\ temperatures for the same sources.
-\color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}
+% SUMMARIZE \citealt{Casagrande:2008}, \citealt{Boyajian:2012}, and \citealt{Mann:2015} 
+Shown in the top panels of \ref{fig:teff_comparisons}
+
+The scatter of each relation being largely due to metallicity variation, both for the \cannon\ and \aspcap\ metallicities.
+\cannon\ temperatures show the highest agreement with the photometric relations from \citealt{Mann:2015} and \citealt{Boyajian:2012}, with a scatter of 80K. Same calibration scale
+
+While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average \aspcap\ temperatures are much hotter, and scatters are much higher skewed by 500 sources which \aspcap\ predicts to have temperatures over 4200K.
+
+\subsection{Isochrone Comparisons}
 
@@
 \plotone{figures/aspcap_cannon_validation.png}
-\caption{\cannon\ (right column) and \aspcap\ (left column) temperatures as compared to temperatures estimated from visible and infrared color-temperature relations from three different literature sources for a subsample of 6,000 sources.
-Shown are \cannon\ and \aspcap\ temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and [Fe/H] dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2013} relation (bottom).
+\caption{\cannon\ (right column) and \aspcap\ (left column) temperatures as compared to temperatures estimated from visible and infrared color-temperature relations from three different literature sources for a subsample of 6,257 sources.
+Shown are \cannon\ and \aspcap\ temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and [Fe/H] dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation (bottom).
 We find \aspcap\ overestimates \teff\ on average by about 130K for the M15 and B13 relations about 220K for the C08.

### 2018-11-28 paper/draft.tex
adams intro revisions

@@
 In this work we build a data-driven model for M dwarf spectra with \thecannon\ (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}).
-\thecannon\ is a regression (typically quadratic) for the spectral pixels given a training
-set of stars with reliable or accurate or assumed-accurate labels.
-The model predicts spectrum given labels, but it can be turned around and also
-used to predict labels given spectrum.
+\thecannon\ is a pixel-by-pixel regression for spectral data given a training
+set of stars with reliable (assumed) accurate labels.
+The model can both predict a spectrum given labels, and also predict labels given a spectrum.
 Thus it is a generative model for stellar spectra that can be used for parameter inference.
 
-The data-driven approach of \thecannon\ makes sense in various contexts:
-One is the context in which stellar labels are only known for a small number of stars
-but there are spectra taken for many more.
-Another is the context in which it is very expensive to obtain labels for a star, but
-there are many stars that need labels.
-Another is the context in which there are spectral models that work in one wavelength
-range or resolution but not in another.
-The context of this paper is the latter:
+The data-driven approach of \thecannon\ is ideal in certain cases:
+if stellar labels are only known for a small number of stars
+but there are spectra taken for many more;
+if it is computationally expensive to obtain labels for a star, but
+there are many stars that need labels;
+if there are spectral models that work in one wavelength range or resolution but not in another.
+The context of this paper is the last:
 There are working and reliable models of M dwarf stars in the optical, which have been
@@
 
-In particular, parameters such as metallicity can be difficult to determine precisely, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths.
+In particular, parameters such as metallicity can be difficult to determine in M dwarfs, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths.
 A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
@@
 
-The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (\textsl{SDSS-III/IV} \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
+The \apogee\ survey 
+% \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (\textsl{SDSS-III/IV} \citealt{Eisenstein:2011}; \citealt{Blanton:2017})
+is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
 Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (\aspcap; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (\teff, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
-For low temperatures (\teff\ $<3500$K), the pipeline uses \textsl{MARCS} plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures (\teff\ $\geq3500$K) \textsl{ATLAS9} \citep{Castelli:2004} plane-parallel models are used.
+For low temperatures ($2800<$ \teff\ $<3500$K), the pipeline uses \textsl{MARCS} plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures (\teff\ $\geq3500$K) \textsl{ATLAS9} \citep{Castelli:2004} plane-parallel models are used.
 
-\apogee\ is primarily targeted for bright stellar populations, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution.
-However numerous cool, main sequence sources have also been observed either as targets proposed by the M dwarf ancilliary survey (1,350 sources selected by proper motions; \citealt{Desphande:2013}), or randomly by \apogee\, which we quantify in this study to be over 10,000 in total.
+\apogee\ is primarily targeted for bright stellar populations particularly red giants, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution.
+However numerous cool, main sequence sources have also been observed either as targets proposed by M dwarf ancilliary surveys, or serendipitously.
+We determine in this study at least 10,000 M dwarfs have been observed to date.
 
@@
 For this work we use {\tt\string aspcapStar} files, which contains the pseudo-continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
-In other work we have recommended an alternative pseudo-continuum normalization \citep{Ness:2015}, but we don't find obvious issues with the normalization in what follows, so we stay with the survey pipeline outputs.
+In previous work it has been recommended to use an alternative pseudo-continuum normalization \citep{Ness:2015}, but we don't find obvious issues with the normalization in our analysis, so we retain the survey pipeline outputs.
 
@@
 \thecannon\ is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:cannon}).
-The model in this sense is used to \emph{transfer} labels from spectra which we know parameters for to those which we do not.
-Hence this data-driven approach effectively removes the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (ideally very accurately measured) \emph{reference labels} that can be measured from other surveys/methods that are easier or more accurate to infer from.
+The model in this sense is used to \emph{transfer} labels from spectra for which we know parameters to those which we do not.
+Hence this data-driven approach effectively circumvents the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (and very accurately measured) \emph{reference labels} possibly measured from other data.
 
 \thecannon\ model can in principle be trained on any physical or empirical labels available, giving a wide range of flexibility to the model.
-However choosing a training sample with high quality labels (precise, but also preferably accurate) is critical to the performance of \thecannon.
+However choosing a training sample with high quality labels is critical to the performance of \thecannon.
 Limitations of \thecannon\ include that test (output) labels are only accurate if the training labels are accurate, and only precise if the training labels are measured consistently across the training sample.
@@
 As with \teff\ relative errors on [Fe/H] are smaller, estimated to be 0.04--0.06\,dex over most of the temperature and metallicity range considered here.
-
-Spectral types reported by M15 for the sources using optical molecular band indices (by the method of \citealt{Lepine:2013}) and ranged K7.3--M4.9 for the subsample we used, although these labels were not included in the model training.
 

### 2018-11-28 paper/draft.tex
move caption details to text

@@
 Also, it is critical to have a training sample with the dynamic range to span the entire parameter space of interest, as \thecannon\ does not extrapolate well outside the parameter space of the training sample.
-Finally, \thecannon\ deeply assumes that the dependence of spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions---if that isn't true, there will be features that \thecannon\ cannot reproduce.
+Finally, \thecannon\ assumes that the dependence of spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions---if that isn't true, there will be features that \thecannon\ cannot reproduce.
 
@@
 Shown in the top panels of \ref{fig:teff_comparisons}
+
+We find \aspcap\ overestimates \teff\ on average by $130-170$K, while \thecannon\ underestimates \teff\ by $\sim25$K .
+We also see the scatter is much higher for the \aspcap\ temperatures ($\sim230$K) compared to the scatter of \cannon\ temperatures ($\sim85$K).
 
@@
 \plotone{figures/aspcap_cannon_validation.png}
-\caption{\cannon\ (right column) and \aspcap\ (left column) temperatures as compared to temperatures estimated from visible and infrared color-temperature relations from three different literature sources for a subsample of 6,257 sources.
-Shown are \cannon\ and \aspcap\ temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and [Fe/H] dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation (bottom).
-We find \aspcap\ overestimates \teff\ on average by about 130K for the M15 and B13 relations about 220K for the C08.
-Additionally we see the scatter much higher for the \aspcap\ temperatures ($\sim220$K) compared to the scatter of \cannon\ temperatures ($\sim80-100$K), the scatter for both \aspcap\ and \cannon\ temperatures largely due to variation in metallicity.
+\caption{\cannon\ (right column) and \aspcap\ (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 6,257 sources with $V,J,K$ colors.
+Shown are \cannon\ and \aspcap\ temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and [Fe/H] dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation $V-K$ and [Fe/H] (bottom).
 \label{fig:teff_comparisons}}

### 2018-12-07 paper/draft.tex
adam's edits

@@
 The \emph{training step} consists of optimizing the likelihood function (Equation \ref{eqn:likelihood}) for the coefficient vector and scatter ($\theta_{\lambda}$ and $s_{\lambda}$) given fixed label vector (\emph{$\ell_n$}) constructed from the reference labels, and the \emph{test step} consists of optimizing the likelihood function for the labels at fixed $\theta_{\lambda}$ and $s_{\lambda}$ obtained in the training step (see \citealt{Ness:2015} for further description).
-Thus \thecannon\ is both a generative model for the stellar spectra, and a method for obtaining 
-labels from spectra. In the training step, the regression is designed to predict spectral pixels
+In the training step, the regression is designed to predict spectral pixels
 given labels, by learning zeroth, first, and second derivatives of the data with respect to
@@
 This percent-level accuracy of the model in the data domain is also demonstrated...
+\color{gcolor}{HOGG: Adam says he doesn't think this is right: if $\chi^2/n>>1$ then the model is NOT an accurate representation of the data.}\color{black} \\
 \color{gcolor}{BIRKY: Can you demonstrate this by showing the RMS residuals, putting it in as a figure, and putting this paragraph back to me?
@@
 
-The scatter of each relation being largely due to metallicity variation, both for the \cannon\ and \aspcap\ metallicities.
-\cannon\ temperatures show the highest agreement with the photometric relations from \citealt{Mann:2015} and \citealt{Boyajian:2012}, with a scatter of 80K. Same calibration scale
+% The scatter of each relation being largely due to metallicity variation, both for the \cannon\ and \aspcap\ metallicities.
+\cannon\ temperatures show the highest agreement with the photometric relations from \citealt{Mann:2015} and \citealt{Boyajian:2012}, with a scatter of 80K.
 
@@
 \end{enumerate}
-
-\color{gcolor}{HOGG: Is a conclusion section necessary?}\color{black}
 

### 2019-01-21 paper/draft.tex
make figures more readable

@@
 \usepackage{morefloats}
+
+% \usepackage{draftwatermark}
+% \SetWatermarkScale{3}
+% \SetWatermarkLightness{.7}
 
@@
 This percent-level accuracy of the model in the data domain is also demonstrated...
-\color{gcolor}{HOGG: Adam says he doesn't think this is right: if $\chi^2/n>>1$ then the model is NOT an accurate representation of the data.}\color{black} \\
+% \color{gcolor}{HOGG: Adam says he doesn't think this is right: if $\chi^2/n>>1$ then the model is NOT an accurate representation of the data.}\color{black} \\
 \color{gcolor}{BIRKY: Can you demonstrate this by showing the RMS residuals, putting it in as a figure, and putting this paragraph back to me?
@@
 \subsection{Radii}
-Using empirically calibrated relations from \citealt{Mann:2015} we determine radii... some diagnostics...
+% Using empirically calibrated relations from \citealt{Mann:2015} we determine radii... some diagnostics...
+TBD
 
@@
 
-As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures.
+TBD
+% As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures.
 To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (\textsl{APASS}; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by \textsl{APASS}, and $JHK$ photometries from \apogee.
@@
 
-% SUMMARIZE \citealt{Casagrande:2008}, \citealt{Boyajian:2012}, and \citealt{Mann:2015} 
-Shown in the top panels of \ref{fig:teff_comparisons}
+% % SUMMARIZE \citealt{Casagrande:2008}, \citealt{Boyajian:2012}, and \citealt{Mann:2015} 
+% Shown in the top panels of \ref{fig:teff_comparisons}
 
-We find \aspcap\ overestimates \teff\ on average by $130-170$K, while \thecannon\ underestimates \teff\ by $\sim25$K .
+% We find \aspcap\ overestimates \teff\ on average by $130-170$K, while \thecannon\ underestimates \teff\ by $\sim25$K .
 We also see the scatter is much higher for the \aspcap\ temperatures ($\sim230$K) compared to the scatter of \cannon\ temperatures ($\sim85$K).
 
-% The scatter of each relation being largely due to metallicity variation, both for the \cannon\ and \aspcap\ metallicities.
-\cannon\ temperatures show the highest agreement with the photometric relations from \citealt{Mann:2015} and \citealt{Boyajian:2012}, with a scatter of 80K.
+% % The scatter of each relation being largely due to metallicity variation, both for the \cannon\ and \aspcap\ metallicities.
+% \cannon\ temperatures show the highest agreement with the photometric relations from \citealt{Mann:2015} and \citealt{Boyajian:2012}, with a scatter of 80K.
 
-While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average \aspcap\ temperatures are much hotter, and scatters are much higher skewed by 500 sources which \aspcap\ predicts to have temperatures over 4200K.
+% While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average \aspcap\ temperatures are much hotter, and scatters are much higher skewed by 500 sources which \aspcap\ predicts to have temperatures over 4200K.
 
 \subsection{Isochrone Comparisons}
+TBD
 
@@
 \section{Discussion} \label{sec:discussion}
+TBD
 
-\begin{enumerate}
-\item[-] Brief summary of what we did and found
+% \begin{enumerate}
+% \item[-] Brief summary of what we did and found
 
-\item[-] Things we learned along the way: For example, not all training sets worked well in our experiments, probably because they were too small or had too little dynamic range.
+% \item[-] Things we learned along the way: For example, not all training sets worked well in our experiments, probably because they were too small or had too little dynamic range.
 
-\item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: \thecannon\ can produce precise results, but is only as accurate as its training set.}\color{black}
+% \item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: \thecannon\ can produce precise results, but is only as accurate as its training set.}\color{black}
 
-\item[-] Some discussion of whether the derivatives are sensible or surprising. \color{gcolor}{HOGG: I'm not sure what to say here}\color{black}
+% \item[-] Some discussion of whether the derivatives are sensible or surprising. \color{gcolor}{HOGG: I'm not sure what to say here}\color{black}
 
-\item[-] \color{gcolor}{HOGG:\thecannon\ as a test of evaluating training parameter accuracy?}\color{black}
+% \item[-] \color{gcolor}{HOGG:\thecannon\ as a test of evaluating training parameter accuracy?}\color{black}
 
-\item[-] \color{gcolor}{HOGG: Discussion of validation tests and ground truth}\color{black}
+% \item[-] \color{gcolor}{HOGG: Discussion of validation tests and ground truth}\color{black}
 
-\end{enumerate}
+% \end{enumerate}
 
-Future work:
-\begin{enumerate}
-\item[-] Create a binary sample; using color/magnitude \citealt{Widmark:2018} and by spectroscopy \citealt{ElBadry:2018}
+% Future work:
+% \begin{enumerate}
+% \item[-] Create a binary sample; using color/magnitude \citealt{Widmark:2018} and by spectroscopy \citealt{ElBadry:2018}
 
-\item[-] \color{gcolor}{HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?}\color{black}
-\end{enumerate}
+% \item[-] \color{gcolor}{HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?}\color{black}
+% \end{enumerate}
 
@@
 \includegraphics[width=16cm]{figures/demo_derivatives_teff2.png}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
 \includegraphics[width=16cm]{figures/demo_derivatives_teff3.png}
@@
 \includegraphics[width=16cm]{figures/demo_derivatives_feh2.png}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
 \includegraphics[width=16cm]{figures/demo_derivatives_feh3.png}

### 2019-01-23 paper/draft.tex
improved figures
add line list table

@@
 \newcommand{\teff}{$T_{\rm eff}$}
+\newcommand{\feh}{[Fe/H]}
 
@@
 \begin{figure}[ht]
-\plotone{figures/self_and_validation_test_teff_snr.png}
-\plotone{figures/self_and_validation_test_feh_snr.png}
+% \plotone{figures/self_and_validation_test_teff_snr.png}
+% \plotone{figures/self_and_validation_test_feh_snr.png}
+\plottwo{figures/crv_test_teff.png}{figures/crv_test_fe_h.png}
 \caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (left), and leave-one-out cross validation (right).
@@
 \begin{figure}[ht]
-% \gridline{\fig{figures/safe_selection_teff.png}{0.5\textwidth}{}
-%           \fig{figures/safe_selection_feh.png}{0.5\textwidth}{}}
-% \gridline{\fig{figures/safe_selection_spt.png}{0.5\textwidth}{}}
 \plottwo{figures/cmd_teff_full.png}{figures/cmd_teff_safe.png}
@@
 %====================================
+
+\begin{deluxetable*}{|c|c|c|c|c|}
+\tablecaption{Strong \teff\ correlated lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
+\tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
+\startdata
+Fe I & 15207.509 & -0.007 & 0.003 & -2.622 \\
+Fe I & 15648.478 & -0.006 & 0.002 & -3.273 \\
+Fe I & 15692.643 & -0.008 & 0.002 & -4.056 \\
+FeH & 16107.127 & -0.002 & 0.001 & -2.78 \\
+FeH & 16245.463 & -0.005 & 0.002 & -3.214 \\
+FeH & 16271.519 & -0.002 & 0.001 & -2.013 \\
+FeH & 16284.562 & -0.006 & 0.001 & -5.807 \\
+FeH & 16377.066 & -0.005 & 0.001 & -4.554 \\
+FeH & 16741.489 & -0.002 & 0.001 & -2.873 \\
+FeH & 16812.415 & -0.004 & 0.001 & -4.241 \\
+FeH & 16813.808 & -0.006 & 0.001 & -3.954 \\
+FeH & 16922.405 & -0.003 & 0.001 & -3.867 \\
+FeH & 16934.801 & -0.004 & 0.001 & -3.591 \\
+OH & 15278.267 & 0.006 & 0.003 & 2.47 \\
+OH & 15280.8 & 0.004 & 0.002 & 2.007 \\
+OH & 15391.186 & 0.006 & 0.003 & 2.127 \\
+OH & 15407.355 & 0.01 & 0.002 & 4.836 \\
+OH & 15409.271 & 0.008 & 0.002 & 4.683 \\
+OH & 15505.797 & 0.009 & 0.003 & 3.511 \\
+OH & 15560.305 & 0.008 & 0.002 & 3.865 \\
+OH & 15565.895 & 0.005 & 0.002 & 2.505 \\
+OH & 15568.691 & 0.007 & 0.001 & 4.929 \\
+OH & 15572.133 & 0.008 & 0.002 & 4.393 \\
+OH & 16052.7 & 0.008 & 0.002 & 3.517 \\
+OH & 16055.362 & 0.006 & 0.002 & 3.7 \\
+OH & 16061.796 & 0.007 & 0.003 & 2.502 \\
+OH & 16065.124 & 0.009 & 0.002 & 4.214 \\
+OH & 16069.564 & 0.008 & 0.003 & 3.275 \\
+OH & 16074.227 & 0.007 & 0.002 & 3.338 \\
+OH & 16190.345 & 0.007 & 0.003 & 2.498 \\
+OH & 16192.134 & 0.008 & 0.002 & 4.012 \\
+OH & 16203.995 & 0.006 & 0.002 & 3.039 \\
+OH & 16207.129 & 0.004 & 0.002 & 2.196 \\
+OH & 16352.196 & 0.007 & 0.002 & 3.609 \\
+OH & 16354.682 & 0.011 & 0.003 & 4.167 \\
+OH & 16364.626 & 0.007 & 0.003 & 2.661 \\
+OH & 16368.244 & 0.008 & 0.002 & 4.303 \\
+OH & 16581.281 & 0.008 & 0.002 & 4.41 \\
+OH & 16581.74 & 0.004 & 0.001 & 4.842 \\
+OH & 16866.621 & 0.005 & 0.001 & 3.169 \\
+OH & 16871.982 & 0.012 & 0.002 & 4.892 \\
+OH & 16878.976 & 0.006 & 0.002 & 3.807 \\
+OH & 16884.573 & 0.005 & 0.002 & 3.066 \\
+OH & 16886.206 & 0.006 & 0.001 & 5.131 \\
+OH & 16895.074 & 0.005 & 0.002 & 2.197 \\
+Si I & 15960.044 & -0.009 & 0.004 & -2.523 \\
+Si I & 16094.893 & -0.014 & 0.003 & -4.368 \\
+Si I & 16680.77 & -0.01 & 0.003 & -3.475 \\
+K I & 15162.824 & -0.006 & 0.002 & -2.294 \\
+Ti I & 15715.641 & 0.006 & 0.002 & 2.894 \\
+Ti I & 16634.973 & 0.003 & 0.001 & 4.224 \\
+V I & 15923.924 & 0.003 & 0.001 & 3.6 \\
+Mn I & 15159.054 & -0.006 & 0.002 & -3.123 \\
+\enddata
+\end{deluxetable*}
+
+%====================================
+
+\begin{deluxetable*}{|c|c|c|c|c|}
+\tablecaption{Strong \feh\ correlated lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
+\tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
+\startdata
+Fe I & 15490.381 & -0.004 & 0.001 & -2.722 \\
+Fe I & 15692.643 & -0.005 & 0.002 & -2.759 \\
+Fe I & 16009.512 & -0.006 & 0.003 & -2.143 \\
+FeH & 16114.027 & 0.004 & 0.001 & 2.673 \\
+FeH & 16574.64 & 0.003 & 0.001 & 2.356 \\
+FeH & 16694.372 & 0.003 & 0.001 & 2.193 \\
+FeH & 16814.041 & 0.004 & 0.002 & 2.513 \\
+OH & 16052.7 & 0.005 & 0.002 & 2.034 \\
+OH & 16055.362 & 0.006 & 0.002 & 2.805 \\
+OH & 16065.124 & 0.007 & 0.003 & 2.364 \\
+OH & 16871.982 & 0.007 & 0.003 & 2.306 \\
+OH & 16886.206 & 0.004 & 0.001 & 2.52 \\
+Si I & 15960.044 & -0.007 & 0.003 & -2.112 \\
+Si I & 16680.77 & -0.006 & 0.003 & -2.156 \\
+K I & 15163.033 & 0.006 & 0.003 & 2.043 \\
+V I & 15923.924 & 0.002 & 0.001 & 2.017 \\
+Cr I & 15680.073 & -0.005 & 0.002 & -2.811 \\
+Mn I & 15159.054 & -0.008 & 0.002 & -3.298 \\
+Mn I & 15262.023 & -0.005 & 0.002 & -2.996 \\
+\enddata
+\end{deluxetable*}
+
+%====================================
 \newpage

### 2019-01-29 paper/draft.tex
improved figures
point to table

@@
 where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample, indexed by \emph{n}, $\theta$ is the coeffient vector trainined on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.
-
-\color{gcolor}{HOGG/ADAM: Do we want some note on what lines vary significantly with parameter?
-I could summarize the which kinds of lines show increasing or decreasing trends, but have no idea how to interpret from there.}\color{black}
+A summary of identified lines with derivative values greater than $2\sigma_{\rm jackkife}$ is given in Table \ref{table:teff_derivative} and Table \ref{table:feh_derivative}.
 
@@
 \begin{center}
-\plotone{figures/self_and_validation_test_spt_snr.png}
+% \plotone{figures/self_and_validation_test_spt_snr.png}
+\plottwo{figures/self_test_spt.png}{figures/crv_test_spt.png}
 \end{center}
@@
 \enddata
-\end{deluxetable*}
+\end{deluxetable*} \label{table:teff_derivative}
 
@@
 \enddata
-\end{deluxetable*}
+\end{deluxetable*} \label{table:feh_derivative}
 

### 2019-03-11 paper/draft.tex
restructure & add text to intro

@@
 \newcommand{\sdss}{\textsl{SDSS}}
+\newcommand{\kepler}{\textsl{Kepler}}
 \newcommand{\lamost}{\textsl{LAMOST}}
@@
 \newcommand{\tess}{\textsl{TESS}}
+\newcommand{\msun}{$M_{\odot}$ \,}
+\newcommand{\rsun}{$R_{\odot}$ \,}
 \newcommand{\teff}{$T_{\rm eff}$}
@@
 
-Very low mass stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}.
-Relevance to exoplanet studies, and other reasons for scientific interest...
+Low-mass stars (with masses $M_{*} < 0.7\,$\msun and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and XX of the population by mass.
+With nuclear fusion timescales of $\tau>10^{12}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010}.
+
+Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations (e.g.
+\kepler, \tess), and shorter orbital periods allow for new planet discovery in less observation time than their higher mass counterparts.
+For these reasons, M dwarfs are the primary candidates for exoplanet detection searches, making detailed and precise knowledge of M dwarf chemical compositions key to constraining the atmospheric conditions of potentially habitable exoplanets through exoplanet spectra obtained through transit spectroscopy.
+
+Precise measurements of fundamental stellar parameters (including temperatures, masses, radii, and abundances) are key to studying M dwarfs on both galactic and planetary scales.
+Methods for determining these parameters...
+temperatures (models, photometry), masses (dynamical)
+
+Why is there a problem determining M dwarf metallicities (and especially for high res \apogee\ spectra)?
+Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.
 
 Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.
-Citations to any other large M dwarf catalogs that have
-
 The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
-What about \apogee\'s wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?
-
-Define what parameters/classifications are important for M dwarfs.
-
-State why fundamental physical parameters are dificult to measure for M dwarfs.
-When temperatures and metallicities for a star are unavailable, it is often useful to classify a star in terms of spectral type.
-Define spectral type (this is a classification that is based on how a spectra looks, something to do with TiO band strengths, and it correlates with the temperature of a star).
-
-A number of M dwarf studies have already been conducted in \apogee, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements:
-\begin{enumerate}
-\item \citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs
-\item \citealt{Souto:2017} and \citealt{Souto:2018} -- modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining Teff/logg/metallicity + 13 elemental abundances
-\item \citealt{Desphande:2013} -- modeled rotational and radial velocities for 200+ M dwarfs, later extended by \citealt{Gilhool:2018} to 700+ M dwarfs
-\item \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
-\item \citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries
-\end{enumerate}
-
-Why is there a problem labeling M dwarfs (and especially for high res \apogee\ spectra)?
-Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.
-Several approaches to to determining physical parameters and the limitations:
-
-\begin{enumerate}
-\item model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\citealt{Rajpurohit:2014}, \citealt{Rajpurohit:2018}) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\citealt{Allard:2013})--models don't accurately reproduce spectra plus unquantified systematic errors;
-
-\item model narrow wavelegth region where models are good, look at specific features -- however we want to leverage the full wavelength range and resolution of our data.
-
-\item using empirical classifications/templates/standards - SPT, luminosity class - this is a relative system but does not provide an absolute standard for measuring physical parameters
-\end{enumerate}
 
@@
 In previous work it has been recommended to use an alternative pseudo-continuum normalization \citep{Ness:2015}, but we don't find obvious issues with the normalization in our analysis, so we retain the survey pipeline outputs.
+
+A number of M dwarf studies have already been conducted in \apogee, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements:
+\begin{enumerate}
+\item \citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs
+\item \citealt{Souto:2017} and \citealt{Souto:2018} -- modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining Teff/logg/metallicity + 13 elemental abundances
+\item \citealt{Desphande:2013} -- modeled rotational and radial velocities for 200+ M dwarfs, later extended by \citealt{Gilhool:2018} to 700+ M dwarfs
+\item \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
+\item \citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries
+\end{enumerate}
 
@@
 
-\subsection{Radii}
-% Using empirically calibrated relations from \citealt{Mann:2015} we determine radii... some diagnostics...
-TBD
+\subsection{Masses \& Radii}
+Masses and radii for our test sample are estimated using photometric empirical relations from \citealt{Mann:2019} and \citealt{Mann:2015} respectively.
 
@@
 
-TBD
-% As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures.
+As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures.
 To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (\textsl{APASS}; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by \textsl{APASS}, and $JHK$ photometries from \apogee.
@@
 
-% We find \aspcap\ overestimates \teff\ on average by $130-170$K, while \thecannon\ underestimates \teff\ by $\sim25$K .
+We find \aspcap\ overestimates \teff\ on average by $130-170$K, while \thecannon\ underestimates \teff\ by $\sim25$K.
 We also see the scatter is much higher for the \aspcap\ temperatures ($\sim230$K) compared to the scatter of \cannon\ temperatures ($\sim85$K).
@@
 
-% While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average \aspcap\ temperatures are much hotter, and scatters are much higher skewed by 500 sources which \aspcap\ predicts to have temperatures over 4200K.
+While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average \aspcap\ temperatures are much hotter, and scatters are much higher skewed by 500 sources which \aspcap\ predicts to have temperatures over 4200K.
 
 \subsection{Isochrone Comparisons}
-TBD
 
@@
 \section{Discussion} \label{sec:discussion}
-TBD
 
-% \begin{enumerate}
-% \item[-] Brief summary of what we did and found
+\begin{enumerate}
+\item[-] Brief summary of what we did and found
 
-% \item[-] Things we learned along the way: For example, not all training sets worked well in our experiments, probably because they were too small or had too little dynamic range.
+\item[-] Things we learned along the way: For example, not all training sets worked well in our experiments, probably because they were too small or had too little dynamic range.
 
-% \item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: \thecannon\ can produce precise results, but is only as accurate as its training set.}\color{black}
+\item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: \thecannon\ can produce precise results, but is only as accurate as its training set.}\color{black}
 
-% \item[-] Some discussion of whether the derivatives are sensible or surprising. \color{gcolor}{HOGG: I'm not sure what to say here}\color{black}
+\item[-] Some discussion of whether the derivatives are sensible or surprising. \color{gcolor}{HOGG: I'm not sure what to say here}\color{black}
 
-% \item[-] \color{gcolor}{HOGG:\thecannon\ as a test of evaluating training parameter accuracy?}\color{black}
+\item[-] \color{gcolor}{HOGG:\thecannon\ as a test of evaluating training parameter accuracy?}\color{black}
 
-% \item[-] \color{gcolor}{HOGG: Discussion of validation tests and ground truth}\color{black}
+\item[-] \color{gcolor}{HOGG: Discussion of validation tests and ground truth}\color{black}
 
-% \end{enumerate}
+\end{enumerate}
 
-% Future work:
-% \begin{enumerate}
-% \item[-] Create a binary sample; using color/magnitude \citealt{Widmark:2018} and by spectroscopy \citealt{ElBadry:2018}
+Future work:
+\begin{enumerate}
+\item[-] Create a binary sample; using color/magnitude \citealt{Widmark:2018} and by spectroscopy \citealt{ElBadry:2018}
 
-% \item[-] \color{gcolor}{HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?}\color{black}
-% \end{enumerate}
+\item[-] \color{gcolor}{HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?}\color{black}
+\end{enumerate}
 

### 2019-03-13 paper/draft.tex
restructure

@@
 \begin{abstract}
-\thecannon\ is a flexible, data-driven spectral-modeling and parameter-inference framework, demonstrated on high-resolution, near-infrared \sdssiv\ \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
-The lack of reliable atmospheric models in the near infrared for temperatures cooler than $\sim3500$\,K motivates the extension of this approach to M-dwarf stars, but with calibrated training labels from optical surveys that overlap \apogee.
+We present a catalog of spectroscopic temperatures, metallicities, spectral types for 10,311 M dwarfs in the \apogee\ survey using \thecannon: a flexible, data-driven spectral-modeling and parameter-inference framework, demonstrated on high-resolution, near-infrared \sdssiv\ \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
 Using a training sample of 87 M dwarfs in the temperature range $2860 < T_{\rm eff} < 4130$\,K, and metallicty range $-0.5 < \mathrm{[Fe/H]}< 0.5$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
-The model is accurate in the data space: it accurately predicts \apogee\ spectral pixels at the one-percent level.
 We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging M0 to M9, to predictive accuracy of 0.9 types.
-From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we select a test sample and employ our trained models to determine temperatures, metallicities and spectral subtypes for a conservatively selected sample of M dwarfs in \apogee\ (10,311 sources), and compare our \cannon\ derived \teff\ against color--temperature relations and [Fe/H] against isochrone models in \gaia\ color-magnitude space.
-Finally we use empirically callibrated relations to estimate radii from $M_K$ absolute magnitudes and \cannon\ temperatures.
+From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we compare our \cannon\ derived \teff\ against color--temperature relations and \cannon\ derived [Fe/H] against isochrone models in \gaia\ color-magnitude space.
+Finally we use empirically callibrated relations to estimate masses radii from $M_K$ absolute magnitudes and \cannon\ temperatures.
 This project has implications for the NASA \tess\ Mission and other surveys that involve large samples of M dwarfs. 
@@
 
-Low-mass stars (with masses $M_{*} < 0.7\,$\msun and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and XX of the population by mass.
+Low-mass stars (with masses $M_{*} < 0.7\,$\msun and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010}.
 With nuclear fusion timescales of $\tau>10^{12}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010}.
@@
 
-Precise measurements of fundamental stellar parameters (including temperatures, masses, radii, and abundances) are key to studying M dwarfs on both galactic and planetary scales.
+Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.
+The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
+
+Precise measurements of fundamental stellar parameters (including temperatures, masses, radii, abundances and kinematics) are key to studying M dwarfs on both galactic and planetary scales.
 Methods for determining these parameters...
@@
 
-Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.
-The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
+In particular, parameters such as metallicity can be difficult to determine in M dwarfs, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths.
+A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
+\citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
+However these metallicity callibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant callibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.
 
@@
 used successfully, but (as we note above) there isn't as much success with such modeling in the infrared, and not at \apogee\ resolution.
-
-In particular, parameters such as metallicity can be difficult to determine in M dwarfs, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths.
-A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
-\citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
-However these metallicity callibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant callibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.
 \thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
@@
 However numerous cool, main sequence sources have also been observed either as targets proposed by M dwarf ancilliary surveys, or serendipitously.
-We determine in this study at least 10,000 M dwarfs have been observed to date.
+We determine in this study at least 10,000 M dwarfs have been observed by \apogee\ to date.
 

### 2019-03-13 paper/draft.tex
modify wording

@@
 Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.
+Methods for determining metallicity of M dwarfs: (1) from atmospheric models of M dwarfs, (2) empirical methods
 
-In particular, parameters such as metallicity can be difficult to determine in M dwarfs, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths.
-A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
+In particular, the metallicity abundance can be difficult to determine in M dwarfs, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths.
+A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
 \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
@@
 The context of this paper is the last:
-There are working and reliable models of M dwarf stars in the optical, which have been
+there are precise callibrations for metallicities of M dwarf stars in the optical, which have been
 used successfully, but (as we note above) there isn't as much success with such modeling in the infrared, and not at \apogee\ resolution.
@@
 
-A number of M dwarf studies have already been conducted in \apogee, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements:
-\begin{enumerate}
-\item \citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs
-\item \citealt{Souto:2017} and \citealt{Souto:2018} -- modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining Teff/logg/metallicity + 13 elemental abundances
-\item \citealt{Desphande:2013} -- modeled rotational and radial velocities for 200+ M dwarfs, later extended by \citealt{Gilhool:2018} to 700+ M dwarfs
-\item \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
-\item \citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries
-\end{enumerate}
+A number of M dwarf studies have already been conducted in \apogee, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements.
+\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs.
+\citealt{Souto:2017} and \citealt{Souto:2018} -- modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining Teff/logg/metallicity + 13 elemental abundances.
+\citealt{Desphande:2013} -- modeled rotational and radial velocities for 200+ M dwarfs, later extended by \citealt{Gilhool:2018} to 700+ M dwarfs.
+\citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
+\citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries.
 

### 2019-03-13 paper/draft.tex
reorder table columns

@@
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 % \plotone{figures/self_and_validation_test_teff_snr.png}
@@
 Predictive accuracy, as computed by the scatter in cross validation is 77K in temperature, and 0.09 dex in metallicity.}\label{fig:mann_validation}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \begin{center}
@@
 Predictive accuracy, as computed by the scatter in cross validation is 0.9 subtypes.}\label{fig:west_validation}
-\end{figure}
+\end{figure*}
 
 %---------------------------
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \begin{center}
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \begin{center}
@@
 \textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.}\label{fig:demo_teff}
-\end{figure}
+\end{figure*}
 
 %---------------------------
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \begin{center}
@@
 \textit{Third panel:} Derivative of \thecannon\ model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$ dex; jackknife computed error at each pixel is shown in red.} \label{fig:demo_feh}
-\end{figure}
+\end{figure*}
 
 %---------------------------
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \begin{center}
@@
 \caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of \apogee\ spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
-\end{figure}
+\end{figure*}
 
 %---------------------------
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \begin{center}
@@
 \caption{Derivative plots for spectral type model taken at the median training spectral type, SPT$=3$; jackknife computed error at each pixel is shown in red.} \label{fig:west_derivative}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \plotone{figures/cannon_chi_dist.png}
@@
 \label{fig:chi_dist}}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \plotone{figures/cmd_selection.png}
@@
 \label{fig:cmd_selection}}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \plottwo{figures/cmd_teff_full.png}{figures/cmd_teff_safe.png}
@@
 \label{fig:safe_selection}}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \plotone{figures/spt_vs_teff_vs_feh.png}
@@
 \label{fig:west_vs_mann}}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \plotone{figures/aspcap_cannon_label_hist.png}
@@
 \label{fig:aspcap_cannon_label_hist}}
-\end{figure}
+\end{figure*}
 
 %---------------------------
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \plotone{figures/aspcap_cannon_validation.png}
@@
 \label{fig:teff_comparisons}}
-\end{figure}
+\end{figure*}
 
 %---------------------------
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \plotone{figures/isochrone_cmd.png}
@@
 \label{fig:isochrone_cmd}}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \plotone{figures/mann_radii.png}
@@
 \label{fig:radii}}
-\end{figure}
+\end{figure*}
 
@@
 {\tt APOGEE$\_$ID} 	& 	& \apogee\ \zmass\ designation \\
-{\tt BP$\_$MAG }		& mag	& \gaia\ $BP$ band photometry	\\
-{\tt BP$\_$RP} 		& mag	& \gaia\ $BP-RP$ color	\\
+{\tt GAIA$\_$ID} 		&	& \gaia\ identification number \\
+{\tt TEFF} 				& K	& Mann-trained \cannon\ effective temperature	\\
+{\tt FE$\_$H }		& dex	& Mann-trained \cannon\ [Fe/H]	\\
+{\tt SPT} 				& 	& West-trained \cannon\ spectral subtype (M0$-$M9)	\\
 {\tt CHI$\_$MANN} 	& 	& $\chi^2$ fit of Mann-trained \cannon\ model	\\
 {\tt CHI$\_$WEST} 	& 	& $\chi^2$ fit of West-trained \cannon\ model	\\
-{\tt DEC} 			& deg	& \apogee\ declination angle	\\
-{\tt DIST} 			& pc	& Distance (1/$\varpi$)	\\
-{\tt FE$\_$H }		& dex	& Mann-trained \cannon\ [Fe/H]	\\
-{\tt GAIA$\_$ID} 		&	& \gaia\ identification number \\
-{\tt G$\_$ABS} 		& mag	& \gaia\ $G$ band absolute magnitude	\\
-{\tt G$\_$MAG} 		& mag	& \gaia\ $G$ band photometry	\\
+{\tt RADIUS} 			& R$_\odot$	& Radius determined by \citealt{Mann:2015} relation	\\
+{\tt MASS} 			& M$_\odot$	& Mass determined by \citealt{Mann:2019} relation	\\
+{\tt J} 				& mag	& \apogee\ $J$ band photometry	\\
+{\tt J$\_$ERR} 		& mag	& \apogee\ $J$ band photometry uncertainty	\\
 {\tt H} 				& mag	& \apogee\ $H$ band photometry	\\
 {\tt H$\_$ERR} 		& mag	& \apogee\ $H$ band photometry uncertainty	\\
-{\tt J} 				& mag	& \apogee\ $J$ band photometry	\\
-{\tt J$\_$ERR} 		& mag	& \apogee\ $J$ band photometry uncertainty	\\
 {\tt K} 				& mag	& \apogee\ $K$ band photometry	\\
 {\tt K$\_$ERR} 		& mag	& \apogee\ $K$ band photometry uncertainty	\\
+{\tt BP$\_$MAG }		& mag	& \gaia\ $BP$ band photometry	\\
+{\tt RP$\_$MAG} 		& mag	& \gaia\ $RP$ band photometry	\\
+{\tt G$\_$MAG} 		& mag	& \gaia\ $G$ band photometry	\\
+{\tt G$\_$ABS} 		& mag	& \gaia\ $G$ band absolute magnitude	\\
+{\tt BP$\_$RP} 		& mag	& \gaia\ $BP-RP$ color	\\
+{\tt RA} 				& deg	& \apogee\ radial ascension angle	\\
+{\tt DEC} 			& deg	& \apogee\ declination angle	\\
+{\tt PMRA} 			& mas yr$^{-1}$	& \gaia\ radial ascension proper motion	\\
+{\tt PMRA$\_$ERR} 	& mas yr$^{-1}$	& \gaia\ radial ascension proper motion uncertainty	\\
+{\tt PMDEC} 			& mas yr$^{-1}$	& \gaia\ declination proper motion	\\
+{\tt PMDEC$\_$ERR} 	& mas yr$^{-1}$	& \gaia\ declination proper motion uncertainty	\\
 {\tt PLX} 			& mas	& \gaia\ parallax	\\
 {\tt PLX$\_$ERR} 		& mas	& \gaia\ parallax uncertainty	\\
-{\tt PMDEC} 			& mas yr$^{-1}$	& \gaia\ declination proper motion	\\
-{\tt PMDEC$\_$ERR} 	& mas yr$^{-1}$	& \gaia\ declination proper motion uncertainty	\\
-{\tt PMRA} 			& mas yr$^{-1}$	& \gaia\ radial ascension proper motion	\\
-{\tt PMRA$\_$ERR} 	& mas yr$^{-1}$	& \gaia\ radial ascension proper motion uncertainty	\\
-{\tt RA} 				& deg	& \apogee\ radial ascension angle	\\
-{\tt RADIUS} 			& R$_\odot$	& Radius determined by \citealt{Mann:2015} relation	\\
-{\tt RP$\_$MAG} 		& mag	& \gaia\ $RP$ band photometry	\\
+{\tt DIST} 			& pc	& Distance (1/$\varpi$)	\\
 {\tt RV$\_$APOGEE} 	& km s$^{-1}$	& \apogee\ radial velocity	\\
@@
 {\tt RV$\_$GAIA$\_$ERR}		& km s$^{-1}$	& \gaia\ radial velocity uncertainty	\\
-{\tt SPT} 				& 	& West-trained \cannon\ spectral subtype (M0$-$M9)	\\
-{\tt TEFF} 				& K	& Mann-trained \cannon\ effective temperature	\\
 {\tt Vx} 				& km s$^{-1}$	& Cartesian $x$ velocity in Galactocentric coordnates	\\

### 2019-03-13 paper/draft.tex
improve wording

@@
 Low-mass stars (with masses $M_{*} < 0.7\,$\msun and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010}.
-With nuclear fusion timescales of $\tau>10^{12}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010}.
+With nuclear fusion timescales of $\tau>10^{11}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010}.
 
@@
 \kepler, \tess), and shorter orbital periods allow for new planet discovery in less observation time than their higher mass counterparts.
-For these reasons, M dwarfs are the primary candidates for exoplanet detection searches, making detailed and precise knowledge of M dwarf chemical compositions key to constraining the atmospheric conditions of potentially habitable exoplanets through exoplanet spectra obtained through transit spectroscopy.
+For these reasons, M dwarfs are the primary candidates for exoplanet detection searches, making detailed and precise knowledge of M dwarf chemical compositions key to constraining the atmospheric conditions of potentially habitable exoplanets.
 
-Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.
-The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
-
-Precise measurements of fundamental stellar parameters (including temperatures, masses, radii, abundances and kinematics) are key to studying M dwarfs on both galactic and planetary scales.
-Methods for determining these parameters...
-temperatures (models, photometry), masses (dynamical)
+Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade have dramatically increased the sample of known M dwarfs, and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
+The largest of such catalogs include the \citealt{West:2011} sample of 70,841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of optical molecular band index measurements.
+The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
 
@@
 Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.
-Methods for determining metallicity of M dwarfs: (1) from atmospheric models of M dwarfs, (2) empirical methods
+Methods for determining metallicity of M dwarfs: (1) from atmospheric models of M dwarfs, (2) empirical methods: spectroscopic or photometric calibrations with FGK pairs.
 

### 2019-03-13 paper/draft.tex
add refs

@@
 The largest of such catalogs include the \citealt{West:2011} sample of 70,841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of optical molecular band index measurements.
-The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.
+The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy, yet pipeline measurements \citep{Perez:2016} of \teff\ and \feh\ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K \citep{Schmidt:2016}.
 
@@
 Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.
-Methods for determining metallicity of M dwarfs: (1) from atmospheric models of M dwarfs, (2) empirical methods: spectroscopic or photometric calibrations with FGK pairs.
+Methods for determining metallicity of M dwarfs: (1) from atmospheric models of M dwarfs, (2) empirical methods: photometric calibrations (\citealt{Johnson:2009}; \citealt{Schlaufman:2010}) or spectroscopic calibrations (\citealt{Terrien:2012}; \citealt{Rojas-Ayala:2012}; \citealt{Mann:2013a}; \citealt{Newton:2014}) with FGK pairs.
+Our approach is the latter.
 

### 2019-04-26 paper/draft.tex
change teff, logg, feh, etc definitions

@@
 \newcommand{\tess}{\textsl{TESS}}
-\newcommand{\msun}{$M_{\odot}$ \,}
-\newcommand{\rsun}{$R_{\odot}$ \,}
-\newcommand{\teff}{$T_{\rm eff}$}
-\newcommand{\feh}{[Fe/H]}
+\newcommand{\msun}{M_{\odot}}
+\newcommand{\teff}{T_{\mathrm{eff}}}
+\newcommand{\logg}{\log g}
+\newcommand{\feh}{[{\mathrm{Fe}/\mathrm{H}}]}
 
@@
 
-\begin{document}
+\begin{document}\raggedbottom\sloppy\sloppypar\frenchspacing
 
@@
 \begin{abstract}
-We present a catalog of spectroscopic temperatures, metallicities, spectral types for 10,311 M dwarfs in the \apogee\ survey using \thecannon: a flexible, data-driven spectral-modeling and parameter-inference framework, demonstrated on high-resolution, near-infrared \sdssiv\ \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
-Using a training sample of 87 M dwarfs in the temperature range $2860 < T_{\rm eff} < 4130$\,K, and metallicty range $-0.5 < \mathrm{[Fe/H]}< 0.5$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
+We present a catalog of spectroscopic temperatures, metallicities,
+spectral types for 10,311 M dwarfs in the \apogee\ survey using
+\thecannon, which is a flexible, data-driven spectral-modeling and
+parameter-inference framework demonstrated to estimate
+stellar-parameter labels ($\teff$, $\logg$, $\feh$, and detailed
+abundances) to high precision.
+
+Using a training sample of 87 M dwarfs in the temperature range $2860 < \teff < 4130$\,K, and metallicty range $-0.5 < \feh < 0.5$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
+
 We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging M0 to M9, to predictive accuracy of 0.9 types.
-From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we compare our \cannon\ derived \teff\ against color--temperature relations and \cannon\ derived [Fe/H] against isochrone models in \gaia\ color-magnitude space.
+From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we compare our \cannon\ derived $\teff$ against color--temperature relations and \cannon\ derived $\feh$ against isochrone models in \gaia\ color-magnitude space.
 Finally we use empirically callibrated relations to estimate masses radii from $M_K$ absolute magnitudes and \cannon\ temperatures.
@@
 
-Low-mass stars (with masses $M_{*} < 0.7\,$\msun and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010}.
+Low-mass stars (with masses $M_{*} < 0.7\,\msun$ and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010}.
 With nuclear fusion timescales of $\tau>10^{11}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010}.
@@
 The largest of such catalogs include the \citealt{West:2011} sample of 70,841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of optical molecular band index measurements.
-The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy, yet pipeline measurements \citep{Perez:2016} of \teff\ and \feh\ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K \citep{Schmidt:2016}.
+The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K \citep{Schmidt:2016}.
 
@@
 is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (\aspcap; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (\teff, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
-For low temperatures ($2800<$ \teff\ $<3500$K), the pipeline uses \textsl{MARCS} plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures (\teff\ $\geq3500$K) \textsl{ATLAS9} \citep{Castelli:2004} plane-parallel models are used.
+Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (\aspcap; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
+For low temperatures ($2800<\teff<3500$\,K), the pipeline uses \textsl{MARCS} plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures ($\teff\geq3500$\,K) \textsl{ATLAS9} \citep{Castelli:2004} plane-parallel models are used.
 
@@
 
-The physical parameter sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
+The physical parameter sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 < \feh < 0.49$\,dex, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
 The major difference between the extension sample is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample.
@@
 A subsample of 29 sources with measured angular diameters from long-baseline optical interferometry \citep{Boyajian:2012} are used to calibrate the model comparison, including masking of spectral regions poorly reproduced by the model spectra \citep{Mann:2013c}.
-Based on difference between assigned \teff\ values and those from angular diameters, absolute uncertainty on \teff\ is esitmated to be 60K in \teff, although the relative uncertainty is likely a factor of $\simeq$2 better.
+Based on difference between assigned $\teff$ values and those from angular diameters, absolute uncertainty on $\teff$ is esitmated to be 60K in $\teff$, although the relative uncertainty is likely a factor of $\simeq$2 better.
 %AWM: I realized I've never actaully sat down and rigorously measured the relative Teff uncertainty.
@@
 
-Iron abundances ([Fe/H]) are assignied to the physical parameter sample based on the strength of metal-sensitive lines in the near-infrared SpeX spectra \citep{RojasAyala:2010} using the calibration from \citet{Mann:2013a}.
-The relation between these lines and an absolute [Fe/H] scale is calibrated using wide binaries containing and F, G, or K-type primary and an M dwarf companion, under the assumption that binaries formed from the same molecular cloud and therefore have the same metallicity \citep{Bonfils:2005}.
-Uncertainties are estimated to be $\simeq$0.08\,dex based on irreducable scatter in the empirical relation between selected lines and the assigned [Fe/H] from the primary star.
-As with \teff\ relative errors on [Fe/H] are smaller, estimated to be 0.04--0.06\,dex over most of the temperature and metallicity range considered here.
+Iron abundances ($\feh$) are assignied to the physical parameter sample based on the strength of metal-sensitive lines in the near-infrared SpeX spectra \citep{RojasAyala:2010} using the calibration from \citet{Mann:2013a}.
+The relation between these lines and an absolute $\feh$scale is calibrated using wide binaries containing and F, G, or K-type primary and an M dwarf companion, under the assumption that binaries formed from the same molecular cloud and therefore have the same metallicity \citep{Bonfils:2005}.
+Uncertainties are estimated to be $\simeq$0.08\,dex based on irreducable scatter in the empirical relation between selected lines and the assigned $\feh$ from the primary star.
+As with $\teff$ relative errors on $\feh$ are smaller, estimated to be 0.04--0.06\,dex over most of the temperature and metallicity range considered here.
 
@@
 Unlike their higher-mass counterparts, M dwarf properties do not change measurably over the age of the Universe.
-Hence perfect knowledge of abundances and \teff\ for an M dwarf should uniquely determine its position its surface gravity, position on a color-magnitude diagram, and overall luminosity.
-While we only had [Fe/H] for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important for extreme cases (e.g., Carbon stars).
+Hence perfect knowledge of abundances and $\teff$ for an M dwarf should uniquely determine its position its surface gravity, position on a color-magnitude diagram, and overall luminosity.
+While we only had $\feh$ for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important for extreme cases (e.g., Carbon stars).
 
@@
 \\
-\ell_{n} &=& [1, \, T_{\rm eff}, \, [Fe/H], \, T_{\rm eff}^2, \, T_{\rm eff} \cdot [Fe/H], \, [Fe/H]^{2}] \label{eqn:physical}
+\ell_{n} &=& [1, \, \teff, \, \feh, \, \teff^2, \, \teff \cdot \feh, \, \feh^{2}] \label{eqn:physical}
 ,
@@
 
-\item \textbf{Training bound selection:} Because extrapolation with \thecannon\ is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<$[Fe/H]$<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
+\item \textbf{Training bound selection:} Because extrapolation with \thecannon\ is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
 \end{enumerate}
@@
 Additionally we notice a slight deviation at the upper slice of the main sequence where \cannon\ temperatures, metallicities and spectral types all appear lower than the gradient trends in the rest of the branch, which is most likely the binary sequence.
-Comparisons between the West-trained spectral types and Mann-trained temperatures and metallicities are shown in Figure \ref{fig:west_vs_mann}: left shows the slight negative correlation between spectral type and [Fe/H], and right shows the nonlinear decrease in temperature with visible metallicity gradient.
+Comparisons between the West-trained spectral types and Mann-trained temperatures and metallicities are shown in Figure \ref{fig:west_vs_mann}: left shows the slight negative correlation between spectral type and $\feh$, and right shows the nonlinear decrease in temperature with visible metallicity gradient.
 We also note that applying our model requires very little computational demand: the time to train and test a model on all 14,827 sources was two minutes on a personal laptop.
@@
 
-We find \aspcap\ overestimates \teff\ on average by $130-170$K, while \thecannon\ underestimates \teff\ by $\sim25$K.
+We find \aspcap\ overestimates $\teff$ on average by $130-170$K, while \thecannon\ underestimates $\teff$ by $\sim25$K.
 We also see the scatter is much higher for the \aspcap\ temperatures ($\sim230$K) compared to the scatter of \cannon\ temperatures ($\sim85$K).
@@
 \caption{\textit{Top two panels of each plot:} Mann-trained model for varying metallicities and similar temperatures.
-\textit{Third panel:} Derivative of \thecannon\ model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$ dex; jackknife computed error at each pixel is shown in red.} \label{fig:demo_feh}
+\textit{Third panel:} Derivative of \thecannon\ model with respect to metallicity, taken at the median training metallicity, $\feh=-0.03$\,dex; jackknife computed error at each pixel is shown in red.} \label{fig:demo_feh}
 \end{figure*}
@@
 \caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the \aspcap\ pipeline.
-Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<[Fe/H]<0.5$ because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars \aspcap\ predicts temperatures with slightly higher temperatures and metallicities.
+Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars \aspcap\ predicts temperatures with slightly higher temperatures and metallicities.
 \label{fig:aspcap_cannon_label_hist}}
@@
 \caption{\cannon\ (right column) and \aspcap\ (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 6,257 sources with $V,J,K$ colors.
-Shown are \cannon\ and \aspcap\ temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and [Fe/H] dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation $V-K$ and [Fe/H] (bottom).
+Shown are \cannon\ and \aspcap\ temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and $\feh$ dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation $V-K$ and $\feh$ (bottom).
 \label{fig:teff_comparisons}}
@@
 \plotone{figures/isochrone_cmd.png}
-\caption{Comparison of Cannon [Fe/H] as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
+\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
 The four panels show five metallicity tracks (ranging $-0.5<$[M/H]$<0.5$) over varying ages (1, 3, 5, and 7 Gyr).
@@
 \plotone{figures/mann_radii.png}
-\caption{Comparison of radii estimated from M15 relations: on the x-axis radii are derived from  K magnitudes and \cannon\ [Fe/H], on the y-axis radii are derived from \cannon\ \teff\ and [Fe/H].
+\caption{Comparison of radii estimated from M15 relations: on the x-axis radii are derived from  K magnitudes and \cannon\ $\feh$, on the y-axis radii are derived from \cannon\ $\teff$ and $\feh$.
 \label{fig:radii}}
@@
 \begin{deluxetable*}{|c|c|c|c|c|}
-\tablecaption{Strong \teff\ correlated lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
+\tablecaption{Strong $\teff$-correlated lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
 \tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
@@
 \begin{deluxetable*}{|c|c|c|c|c|}
-\tablecaption{Strong \feh\ correlated lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
+\tablecaption{Strong $\feh$-correlated lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
 \tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
@@
 {\tt TEFF} 				& K	& Mann-trained \cannon\ effective temperature	\\
-{\tt FE$\_$H }		& dex	& Mann-trained \cannon\ [Fe/H]	\\
+{\tt FE$\_$H }		& dex	& Mann-trained \cannon\ $\feh$	\\
 {\tt SPT} 				& 	& West-trained \cannon\ spectral subtype (M0$-$M9)	\\

### 2019-04-26 paper/draft.tex
abstract

@@
 abundances) to high precision.
-
-Using a training sample of 87 M dwarfs in the temperature range $2860 < \teff < 4130$\,K, and metallicty range $-0.5 < \feh < 0.5$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
-
-We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging M0 to M9, to predictive accuracy of 0.9 types.
-From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we compare our \cannon\ derived $\teff$ against color--temperature relations and \cannon\ derived $\feh$ against isochrone models in \gaia\ color-magnitude space.
-Finally we use empirically callibrated relations to estimate masses radii from $M_K$ absolute magnitudes and \cannon\ temperatures.
-This project has implications for the NASA \tess\ Mission and other surveys that involve large samples of M dwarfs. 
+Using a training sample of 87 M dwarfs in the temperature range $2860
+< \teff < 4130$\,K, and metallicty range $-0.5 < \feh < 0.5$\,dex
+[WITH LABELS FROM WHERE??], we train a two-parameter model with
+predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex
+respectively.
+We also train a one-dimensional spectral type model using 51 M dwarfs
+with known spectral types ranging M0 to M9, to predictive accuracy of
+0.9 types.
+From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we
+compare our \cannon\ derived $\teff$ against color--temperature
+relations and \cannon\ derived $\feh$ against isochrone models in
+\gaia\ color-magnitude space.
+Finally we use empirically callibrated relations to estimate masses
+radii from $M_K$ absolute magnitudes and \cannon\ temperatures.
+This project has implications for the NASA \tess\ Mission and other
+surveys that depend on having good parameters for M dwarfs.
 \end{abstract}

### 2019-04-26 paper/draft.tex
note to JB

@@
 We also train a one-dimensional spectral type model using 51 M dwarfs
-with known spectral types ranging M0 to M9, to predictive accuracy of
+with known [FROM WHERE??] spectral types ranging M0 to M9, to predictive accuracy of
 0.9 types.

### 2019-04-26 paper/draft.tex
abstract; typos

@@
 \gaia\ color-magnitude space.
-Finally we use empirically callibrated relations to estimate masses
-radii from $M_K$ absolute magnitudes and \cannon\ temperatures.
+Finally we use our results to
+estimate stellar radii from absolute magnitudes and \cannon\ temperatures [AND WE FIND WHAT??].
 This project has implications for the NASA \tess\ Mission and other
@@
 \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
-However these metallicity callibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant callibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.
+However these metallicity calibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant calibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.
 
@@
 The context of this paper is the last:
-there are precise callibrations for metallicities of M dwarf stars in the optical, which have been
+there are precise calibrations for metallicities of M dwarf stars in the optical, which have been
 used successfully, but (as we note above) there isn't as much success with such modeling in the infrared, and not at \apogee\ resolution.

### 2019-04-26 paper/draft.tex
rearrange citation

--- 2019-04-26 paper/draft.tex
@@
 
-Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade have dramatically increased the sample of known M dwarfs, and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
-The largest of such catalogs include the \citealt{West:2011} sample of 70,841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of optical molecular band index measurements.
+Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade have dramatically increased the sample of known M dwarfs, and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level (e.g.
+SDSS \citealt{West:2011}; LAMOST \citealt{Yi:2014} and \citealt{Guo:2015}).
 The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K \citep{Schmidt:2016}.

### 2019-05-05 paper/draft.tex
add comments

@@
 
+\color{gcolor}{JB: I've moved this paragraph from the intro to here.
+Will expand this...
 A number of M dwarf studies have already been conducted in \apogee, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements.
@@
 \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
-\citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries.
+\citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries.}\color{black}
 
@@
 sense; it is only good at the percent level.
-This percent-level accuracy of the model in the data domain is also demonstrated...
-% \color{gcolor}{HOGG: Adam says he doesn't think this is right: if $\chi^2/n>>1$ then the model is NOT an accurate representation of the data.}\color{black} \\
+This percent-level accuracy of the model in the data domain is also demonstrated... \\
+
 \color{gcolor}{BIRKY: Can you demonstrate this by showing the RMS residuals, putting it in as a figure, and putting this paragraph back to me?
 I am imagining a figure that shows a scatter plot of T vs RMS residual, colored by metallicity.
-Or something like that.}\color{black}
+Or something like that.}\color{black} \\
+
+\color{gcolor}{HOGG: Adam's comment: if $\chi^2/n>>1$ then the model is NOT an accurate representation of the data.}\color{black} \\
 

### 2019-05-06 paper/draft.tex
Update draft.tex

@@
 
-The M15 catalog in total contains 183 sources and the extension sample another 500 stars, both primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog.
-All targets have low-res optical spectra from the SNIFS specgrograph \citep{Lantz:2004} and infrared spectra taken with the SpeX Spectograph \citep{Rayner:2003}, which have been combined to estaimte largely empirical bolometric fluxes.
+The M15 catalog in total contains 183 sources and the extension sample another 500 stars, both samples were primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog.
+All targets have low-res optical spectra from the SNIFS spectrograph \citep{Lantz:2004} and infrared spectra taken with the SpeX Spectrograph \citep{Rayner:2003}, which have been combined to estimate largely empirical bolometric fluxes.
 Effective temperatures have been estimated by comparing the SNIFS spectra to BT-SETTL atmospheric models \citep{Allard:2011}.
 A subsample of 29 sources with measured angular diameters from long-baseline optical interferometry \citep{Boyajian:2012} are used to calibrate the model comparison, including masking of spectral regions poorly reproduced by the model spectra \citep{Mann:2013c}.
-Based on difference between assigned $\teff$ values and those from angular diameters, absolute uncertainty on $\teff$ is esitmated to be 60K in $\teff$, although the relative uncertainty is likely a factor of $\simeq$2 better.
-%AWM: I realized I've never actaully sat down and rigorously measured the relative Teff uncertainty.
-The results from the Cannon actually suggest ~30 is not bad.
-I just did some quick tests, and I'm not prepared to put an absolute number on it.
-A \simeq 30 here should be safeish.
+Based on difference between assigned $\teff$ values and those from angular diameters, absolute uncertainty on $\teff$ is estimated to be 60K in $\teff$, although the relative uncertainty is likely a factor of $\simeq$2 better.
 
-Iron abundances ($\feh$) are assignied to the physical parameter sample based on the strength of metal-sensitive lines in the near-infrared SpeX spectra \citep{RojasAyala:2010} using the calibration from \citet{Mann:2013a}.
+Iron abundances ($\feh$) are assigned to the physical parameter sample based on the strength of metal-sensitive lines in the near-infrared SpeX spectra \citep{RojasAyala:2010} using the calibration from \citet{Mann:2013a}.
 The relation between these lines and an absolute $\feh$scale is calibrated using wide binaries containing and F, G, or K-type primary and an M dwarf companion, under the assumption that binaries formed from the same molecular cloud and therefore have the same metallicity \citep{Bonfils:2005}.
-Uncertainties are estimated to be $\simeq$0.08\,dex based on irreducable scatter in the empirical relation between selected lines and the assigned $\feh$ from the primary star.
+Uncertainties are estimated to be $\simeq$0.08\,dex based on irreducible scatter in the empirical relation between selected lines and the assigned $\feh$ from the primary star.
 As with $\teff$ relative errors on $\feh$ are smaller, estimated to be 0.04--0.06\,dex over most of the temperature and metallicity range considered here.
@@
 The reason for this is that for main-sequence M dwarfs, the parameter is almost entirely redundant with metallicity.
-Unlike their higher-mass counterparts, M dwarf properties do not change measurably over the age of the Universe.
-Hence perfect knowledge of abundances and $\teff$ for an M dwarf should uniquely determine its position its surface gravity, position on a color-magnitude diagram, and overall luminosity.
-While we only had $\feh$ for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important for extreme cases (e.g., Carbon stars).
-
-%%AWM: Feel free to edit this for clarity. Let me know if anything makes no sense.
+Unlike their higher-mass counterparts, M dwarf properties do not change measurably over the age of the Universe after arriving at the zero-age main sequence.
+Hence perfect knowledge of abundances and $\teff$ for an M dwarf should uniquely determine its surface gravity, position on a color-magnitude diagram, and overall luminosity.
+While we only had $\feh$ for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important compared to other uncertainties in extreme cases (e.g., Carbon stars).
 

### 2019-06-06 paper/draft.tex
minor details

--- 2019-05-05 paper/draft.tex
@@
 \kepler, \tess), and shorter orbital periods allow for new planet discovery in less observation time than their higher mass counterparts.
-For these reasons, M dwarfs are the primary candidates for exoplanet detection searches, making detailed and precise knowledge of M dwarf chemical compositions key to constraining the atmospheric conditions of potentially habitable exoplanets.
+For these reasons, M dwarfs are the primary candidates for exoplanet detection searches, making detailed and precise knowledge of M dwarf chemical compositions key to constraining the formation scenarios and atmospheric conditions of potentially habitable exoplanets.
 
@@
 SDSS \citealt{West:2011}; LAMOST \citealt{Yi:2014} and \citealt{Guo:2015}).
-The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K \citep{Schmidt:2016}.
+The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
 
-Why is there a problem determining M dwarf metallicities (and especially for high res \apogee\ spectra)?
+\color{gcolor}Why is there a problem determining M dwarf metallicities (and especially for high res \apogee\ spectra)?
 Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.
 Methods for determining metallicity of M dwarfs: (1) from atmospheric models of M dwarfs, (2) empirical methods: photometric calibrations (\citealt{Johnson:2009}; \citealt{Schlaufman:2010}) or spectroscopic calibrations (\citealt{Terrien:2012}; \citealt{Rojas-Ayala:2012}; \citealt{Mann:2013a}; \citealt{Newton:2014}) with FGK pairs.
-Our approach is the latter.
+Our approach is the latter.\color{black}
 

### 2019-06-06 paper/draft.tex
describe sections

@@
 
-In particular, the metallicity abundance can be difficult to determine in M dwarfs, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths.
 A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
@@
 delivered by the \apogee\ spectrograph and project.
+
+This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the APOGEE survey, as well as our sample selection and derivation of training parameters.
+Section \ref{sec:cannon} describes our model implementation using \thecannon framework.
+In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>10,000$ sources, and examine the validity of our parameters against color-temperature relations and isochrone grids.
+Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results in the context of present/upcoming stellar and exoplanet surveys.
 

### 2019-06-06 paper/draft.tex
description of data sec

@@
 
-This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the APOGEE survey, as well as our sample selection and derivation of training parameters.
+This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the APOGEE survey and previous studies of M dwarfs in APOGEE, as well as our sample selection and derivation of training parameters.
 Section \ref{sec:cannon} describes our model implementation using \thecannon framework.

### 2019-06-06 paper/draft.tex
format tweaks

@@
 This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the APOGEE survey and previous studies of M dwarfs in APOGEE, as well as our sample selection and derivation of training parameters.
-Section \ref{sec:cannon} describes our model implementation using \thecannon framework.
-In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>10,000$ sources, and examine the validity of our parameters against color-temperature relations and isochrone grids.
+Section \ref{sec:cannon} describes our model implementation using \thecannon\ framework.
+In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$10,000 sources, and examine the validity of our parameters against color-temperature relations and isochrone grids.
 Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results in the context of present/upcoming stellar and exoplanet surveys.

### 2019-06-06 paper/draft.tex
fix acronyms and refs

@@
 
-Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade have dramatically increased the sample of known M dwarfs, and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level (e.g.
-SDSS \citealt{West:2011}; LAMOST \citealt{Yi:2014} and \citealt{Guo:2015}).
-The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
+Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015}) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
+The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
 

### 2019-06-07 paper/draft.tex
change email
add ref

@@
 \correspondingauthor{Jessica Birky}
-\email{jbirky@ucsd.edu}
+\email{jbirky@uw.edu}
 
@@
 
-Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations (e.g.
-\kepler, \tess), and shorter orbital periods allow for new planet discovery in less observation time than their higher mass counterparts.
-For these reasons, M dwarfs are the primary candidates for exoplanet detection searches, making detailed and precise knowledge of M dwarf chemical compositions key to constraining the formation scenarios and atmospheric conditions of potentially habitable exoplanets.
+Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods allow for new planet discovery in less observation time than their higher mass counterparts.
+For these reasons, M dwarfs are the primary candidates for exoplanet detection searches (e.g.
+the NASA \kepler\ and Transiting Exoplanet Survey Satellite (\tess) missions), making detailed and precise knowledge of M dwarf chemical compositions key to constraining the formation scenarios and atmospheric conditions of potentially habitable exoplanets.
 
-Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015}) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
+Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
 The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.

### 2019-06-07 paper/draft.tex
wording tweaks

@@
 In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$10,000 sources, and examine the validity of our parameters against color-temperature relations and isochrone grids.
-Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results in the context of present/upcoming stellar and exoplanet surveys.
+Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results in the context of present and upcoming stellar and exoplanetary surveys.
 

### 2019-06-07 paper/draft.tex
atmospheres

@@
 
-\color{gcolor}Why is there a problem determining M dwarf metallicities (and especially for high res \apogee\ spectra)?
-Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.
-Methods for determining metallicity of M dwarfs: (1) from atmospheric models of M dwarfs, (2) empirical methods: photometric calibrations (\citealt{Johnson:2009}; \citealt{Schlaufman:2010}) or spectroscopic calibrations (\citealt{Terrien:2012}; \citealt{Rojas-Ayala:2012}; \citealt{Mann:2013a}; \citealt{Newton:2014}) with FGK pairs.
-Our approach is the latter.\color{black}
+While elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precisions (down to errors in the range of $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by the improvement of atmospheric models (citations),
+the determination of precise metallicities for M dwarfs has remained a long standing challenge, due to the formation of complex molecular features at temperatures, particularly TiO and VO which dominate the optical, H$_2$O and CO which dominate the infrared, and hydride molecules (FeH, CaH, CrH, MgH, etc.) which form at late spectral types \citep{Allard1997}.
 

### 2019-06-07 paper/draft.tex
add cannon ref

@@
 While elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precisions (down to errors in the range of $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by the improvement of atmospheric models (citations),
-the determination of precise metallicities for M dwarfs has remained a long standing challenge, due to the formation of complex molecular features at temperatures, particularly TiO and VO which dominate the optical, H$_2$O and CO which dominate the infrared, and hydride molecules (FeH, CaH, CrH, MgH, etc.) which form at late spectral types \citep{Allard1997}.
+the determination of precise metallicities for M dwarfs has remained a long standing challenge, due to the formation of complex diatomic and triatomic molecular features at M dwarf temperature regimes, particularly TiO and VO which dominate the optical, H$_2$O and CO which dominate the infrared, and hydride molecules (FeH, CaH, CrH, MgH, etc.) which form at late spectral types \citep{Allard1997}.
 
@@
 
-In this work we build a data-driven model for M dwarf spectra with \thecannon\ (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}).
+In this work we build a data-driven model for M dwarf spectra with \thecannon\ (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}, \citealt{Behmard2019}).
 \thecannon\ is a pixel-by-pixel regression for spectral data given a training

### 2019-06-07 paper/draft.tex
abundances

@@
 
-While elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precisions (down to errors in the range of $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by the improvement of atmospheric models (citations),
-the determination of precise metallicities for M dwarfs has remained a long standing challenge, due to the formation of complex diatomic and triatomic molecular features at M dwarf temperature regimes, particularly TiO and VO which dominate the optical, H$_2$O and CO which dominate the infrared, and hydride molecules (FeH, CaH, CrH, MgH, etc.) which form at late spectral types \citep{Allard1997}.
+While elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precisions (down to errors in the range of $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by improvements in atmospheric models towards realistic 3D local thermodynamic equilibrium (LTE) assumptions \citep{Asplund2005}, and differential abundance techniques using equivalent widths \citep{Bedell2014},
+the determination of precise metallicities for M dwarfs has remained a long standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperature regimes, particularly TiO and VO which dominate the optical, H$_2$O and CO which dominate the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) which form at late spectral types \citep{Allard1997}.
+In particular, atmospheric models fail to reproduce spectral features due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions and lack of continuum in a spectrum make it difficult to deconvolve individual features and extract line strengths from equivalent widths,
+creating fundamental limitations in efforts towards exploiting the information content available out of high resolution datasets using traditional modeling and empirical methods.
 
-A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
+A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically associated wide binaries are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
 \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).

### 2019-06-07 paper/draft.tex
describe previous m dwarf studies

@@
 \apogee\ is primarily targeted for bright stellar populations particularly red giants, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution.
-However numerous cool, main sequence sources have also been observed either as targets proposed by M dwarf ancilliary surveys, or serendipitously.
-We determine in this study at least 10,000 M dwarfs have been observed by \apogee\ to date.
+However numerous cool, main sequence sources have also been observed either as targets proposed by the \apogee\ M dwarf ancilliary survey ($\sim$1,200 sources; \citealt{Desphande:2013}), or serendipitously.
+
+A number of studies out of the M dwarf ancilliary survey have already been conducted towards measuring reliable fundamental atmospheric parameters and kinematic measurements using spectral synthesis of atmospheric model grids.
+These studies include \citealt{Desphande:2013} and \citealt{Gilhool:2018} which have studied the radial and rotational kinematics for 700+ sources,
+\citealt{Souto:2017} and \citealt{Souto:2018} which have modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining $\teff$/$\logg$/metallicity + 13 elemental abundances,
+\citealt{Rajpurohit:2018} which tested BTSettl model grids \citep{Allard:2012} on 45 M dwarfs to estimate $\teff$/$\logg$/metallicity,
+and \citealt{Skinner:2018} which has identified and measured mass ratios and radial velocities for 44 M dwarf spectroscopic binaries.
+This work complements existing studies by producing model-independent catalog of spectroscopic temperatures and metallicites to test against model predictions for the entire \apogee\ M dwarf sample, which we quantify to contain at least 10,000 sources to date (DR14).
 
@@
 In previous work it has been recommended to use an alternative pseudo-continuum normalization \citep{Ness:2015}, but we don't find obvious issues with the normalization in our analysis, so we retain the survey pipeline outputs.
-
-\color{gcolor}{JB: I've moved this paragraph from the intro to here.
-Will expand this...
-A number of M dwarf studies have already been conducted in \apogee, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements.
-\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs.
-\citealt{Souto:2017} and \citealt{Souto:2018} -- modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining Teff/logg/metallicity + 13 elemental abundances.
-\citealt{Desphande:2013} -- modeled rotational and radial velocities for 200+ M dwarfs, later extended by \citealt{Gilhool:2018} to 700+ M dwarfs.
-\citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity).
-\citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries.}\color{black}
 

### 2019-06-09 paper/draft.tex
add software refs
scipy ref
reference table 5
start expanding discussion

@@
 
+\newcommand{\code}[1]{\texttt{\detokenize{#1}}}
 \newcommand{\apogee}{\textsl{APOGEE}}
@@
 
-\subsection{Masses \& Radii}
-Masses and radii for our test sample are estimated using photometric empirical relations from \citealt{Mann:2019} and \citealt{Mann:2015} respectively.
+Table \ref{table:test_results} presents...
 
@@
 
-% % The scatter of each relation being largely due to metallicity variation, both for the \cannon\ and \aspcap\ metallicities.
-% \cannon\ temperatures show the highest agreement with the photometric relations from \citealt{Mann:2015} and \citealt{Boyajian:2012}, with a scatter of 80K.
-
 While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average \aspcap\ temperatures are much hotter, and scatters are much higher skewed by 500 sources which \aspcap\ predicts to have temperatures over 4200K.
@@
 
-\begin{enumerate}
-\item[-] Brief summary of what we did and found
+We trained and evaluated the performance of \thecannon\ at deriving atmospheric temperatures and metallicities from high resolution near-infrared spectra... summarize results.
 
-\item[-] Things we learned along the way: For example, not all training sets worked well in our experiments, probably because they were too small or had too little dynamic range.
+While \thecannon\ generally achieves excellent precision at predicting labels and reproducing the spectral features as demonstrated in this study and previous work on \thecannon\ (\citealt{Ness:2015}, \citealt{Casey:2016}), the accuracy of labels from model application to new sources is limited by the accuracy, relative precision, size, dynamic range and representation of the training sample.
 
-\item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: \thecannon\ can produce precise results, but is only as accurate as its training set.}\color{black}
+Comment on the importance of each property of training sample: \\
+accuracy -  \\
+precision -  \\
+size -  \\
+dynamic range - training sample must span \\
+representation - \\
 
-\item[-] Some discussion of whether the derivatives are sensible or surprising. \color{gcolor}{HOGG: I'm not sure what to say here}\color{black}
+Talk about the particular limitations we have found in this training sample.
 
-\item[-] \color{gcolor}{HOGG:\thecannon\ as a test of evaluating training parameter accuracy?}\color{black}
+% \begin{enumerate}
+% \item[-] Brief summary of what we did and found
 
-\item[-] \color{gcolor}{HOGG: Discussion of validation tests and ground truth}\color{black}
+% \item[-] Things we learned along the way: For example, not all training sets worked well in our experiments, probably because they were too small or had too little dynamic range.
 
-\end{enumerate}
+% \item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: \thecannon\ can produce precise results, but is only as accurate as its training set.}\color{black}
 
-Future work:
-\begin{enumerate}
-\item[-] Create a binary sample; using color/magnitude \citealt{Widmark:2018} and by spectroscopy \citealt{ElBadry:2018}
+% \item[-] Some discussion of whether the derivatives are sensible or surprising.
 
-\item[-] \color{gcolor}{HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?}\color{black}
-\end{enumerate}
+% \item[-] \color{gcolor}{HOGG:\thecannon\ as a test of evaluating training parameter accuracy?}\color{black}
+
+% \item[-] \color{gcolor}{HOGG: Discussion of validation tests and ground truth}\color{black}
+% \end{enumerate}
+
+% Future work:
+% \begin{enumerate}
+% \item[-] Create a binary sample; using color/magnitude \citealt{Widmark:2018} and by spectroscopy \citealt{ElBadry:2018}
+
+% \item[-] \color{gcolor}{HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?}\color{black}
+% \end{enumerate}
 
@@
 participating in the {\it Gaia} Multilateral Agreement.
+
+\facilities{
+	\sdssiv(\apogee),
+	\gaia}
+
+\software{
+	\code{Astropy} \citep{astropy,astropy2}, \,
+	\code{matplotlib} \citep{Hunter:2007}, \,
+	\code{numpy} \citep{vanderWalt:2011}, \,
+	\code{scipy} \citep{Jones:2001}, \,
+	\code{The Cannon} \citep{Ness:2015}}
 
@@
 \enddata
-\end{deluxetable*}
+\end{deluxetable*} \label{table:test_results}
 

### 2019-06-09 paper/draft.tex
intro P2

@@
 
-Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods allow for new planet discovery in less observation time than their higher mass counterparts.
-For these reasons, M dwarfs are the primary candidates for exoplanet detection searches (e.g.
-the NASA \kepler\ and Transiting Exoplanet Survey Satellite (\tess) missions), making detailed and precise knowledge of M dwarf chemical compositions key to constraining the formation scenarios and atmospheric conditions of potentially habitable exoplanets.
+Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods (for a fixed stellar insulation flux) allow for new planet discovery in less observation time than their higher mass counterparts.
+For these reasons, M dwarfs are primary candidates for exoplanet searches, including by the NASA \kepler\ \citep[e.g.,][]{Dressing2015} and Transiting Exoplanet Survey Satellite (\tess) \citep[e.g.,][]{Muirhead:2018} missions.
+As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with {\it JWST}.
 
-Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
+Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
 The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.

### 2019-06-09 paper/draft.tex
intro P3

@@
 
-Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
-The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
+Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources), enabling studies of M dwarf abundances on both exoplanet and Galactic scales.
+The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}.
+\apogee\ pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have only been determined to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
 
-While elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precisions (down to errors in the range of $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by improvements in atmospheric models towards realistic 3D local thermodynamic equilibrium (LTE) assumptions \citep{Asplund2005}, and differential abundance techniques using equivalent widths \citep{Bedell2014},
-the determination of precise metallicities for M dwarfs has remained a long standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperature regimes, particularly TiO and VO which dominate the optical, H$_2$O and CO which dominate the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) which form at late spectral types \citep{Allard1997}.
+Elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precision (down to $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by improvements in atmospheric models, including realistic 3D local thermodynamic equilibrium (LTE) assumptions \citep{Asplund2005}, and differential abundance techniques using equivalent widths \citep{Bedell2014}.
+However, the determination of precise metallicities for M dwarfs has remained a long-standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperature regimes, particularly TiO and VO which dominate the optical, H$_2$O and CO which dominate the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) which form at late spectral types \citep{Allard1997}.
 In particular, atmospheric models fail to reproduce spectral features due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions and lack of continuum in a spectrum make it difficult to deconvolve individual features and extract line strengths from equivalent widths,

### 2019-06-09 paper/draft.tex
intro P4_5

@@
 Low-mass stars (with masses $M_{*} < 0.7\,\msun$ and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010}.
-With nuclear fusion timescales of $\tau>10^{11}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010}.
+With nuclear fusion timescales of $\tau>10^{11}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010,Woolf:2012}.
 
@@
 Elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precision (down to $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by improvements in atmospheric models, including realistic 3D local thermodynamic equilibrium (LTE) assumptions \citep{Asplund2005}, and differential abundance techniques using equivalent widths \citep{Bedell2014}.
-However, the determination of precise metallicities for M dwarfs has remained a long-standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperature regimes, particularly TiO and VO which dominate the optical, H$_2$O and CO which dominate the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) which form at late spectral types \citep{Allard1997}.
-In particular, atmospheric models fail to reproduce spectral features due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions and lack of continuum in a spectrum make it difficult to deconvolve individual features and extract line strengths from equivalent widths,
-creating fundamental limitations in efforts towards exploiting the information content available out of high resolution datasets using traditional modeling and empirical methods.
+However, the determination of precise metallicities for M dwarfs has remained a long-standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperature regimes, particularly TiO and VO dominating the optical spectrum, H$_2$O and CO for the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) present in the spectra of the latest spectral types \citep{Allard1997}.
+In particular, atmospheric models fail to reproduce spectral features \citep[e.g.,][]{Mann:2013c} due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions, and the lack of continuum in a spectrum, all make it difficult to deconvolve individual features and extract line strengths from equivalent widths.
+The combination of these effects creates fundamental limitations to explore the information content available out of high resolution spectra using a traditional comparison to atmosphere models.
 
-A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically associated wide binaries are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
+A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically associated wide binaries are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the spectra of the M dwarf (e.g.
 \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
-However these metallicity calibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant calibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.
+However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band or include important information about the spectral shape needed to determine \teff.
+Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that would fail to utilize the additional spectral information provided by \apogee's resolution.
+
+%the relevant calibration indices used by existing studies (such as Ca~I or K~I lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.
+%AWM: I'm not really sure this is true.
+The major lines I and Babs use are in the K-band, very few studies use optical bands.
+Ryan includes ones in the H-band, some of which are in Apogee.
+I think the bigger point is that this fails to utilize the high-resolution information of the spectra, and relies instead of temperature calibration from spectral shape, which is impossible at this resolution.
+
+%%AWM: would it be useful to have some comment on data-driven models in general? Their virtues should probably be explained before invoking them
 

### 2019-06-10 paper/draft.tex
intro P6
reference derivative tables

--- 2019-06-09 paper/draft.tex
@@
 
-Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods (for a fixed stellar insulation flux) allow for new planet discovery in less observation time than their higher mass counterparts.
-For these reasons, M dwarfs are primary candidates for exoplanet searches, including by the NASA \kepler\ \citep[e.g.,][]{Dressing2015} and Transiting Exoplanet Survey Satellite (\tess) \citep[e.g.,][]{Muirhead:2018} missions.
-As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with {\it JWST}.
+Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods allow for new planet discovery in less observation time than their higher mass counterparts.
+For these reasons, M dwarfs are the primary candidates for exoplanet detection searches (e.g.
+the NASA \kepler\ and Transiting Exoplanet Survey Satellite (\tess) missions), making detailed and precise knowledge of M dwarf chemical compositions key to constraining the formation scenarios and atmospheric conditions of potentially habitable exoplanets.
 
-Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
+Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
 The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
@@
 
-We trained and evaluated the performance of \thecannon\ at deriving atmospheric temperatures and metallicities from high resolution near-infrared spectra... summarize results.
+We trained and evaluated the performance of \thecannon\ at deriving atmospheric temperatures and metallicities from M dwarfs using high resolution near-infrared spectra, and compiled a catalog of 10,311 M dwarfs with \cannon\ temperatures, metallicities, and spectral types, 6D kinematics (\gaia\ parallaxes and proper motions + \apogee\ radial velocities) shown in Table \ref{table:test_results}.
+
+Table \ref{table:teff_derivative} and Table \ref{table:feh_derivative}, summarize which features are found to be strong temperature and metallicity indicators.
 

### 2019-06-10 paper/draft.tex
make doc more readable

@@
 Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
-The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
+The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III/IV mission \citep{Eisenstein:2011,Blanton:2017} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
 
@@
 A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically associated wide binaries are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
-\citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
+\citealt{Terrien:2012}; \citealt{Rojas-Ayala:2012}; \citealt{Newton:2014}).
 However these metallicity calibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant calibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.
 
-In this work we build a data-driven model for M dwarf spectra with \thecannon\ (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}, \citealt{Behmard2019}).
+In this work we build a data-driven model for M dwarf spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a}.
 \thecannon\ is a pixel-by-pixel regression for spectral data given a training
@@
 
-The \apogee\ survey 
-% \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (\textsl{SDSS-III/IV} \citealt{Eisenstein:2011}; \citealt{Blanton:2017})
-is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
+The \apogee\ survey is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
 Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (\aspcap; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
@@
 
+% ----------------------------------------
 \subsection{Training Sample Selection}
@@
 \subsection{Temperature/Metallicity Model \label{subsec:mann_results}}
+
 For the physical parameter model we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
@@
 
+% ----------------------------------------
 \subsection{Spectral Type Model \label{subsec:west_results}}
+
 We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
@@
 
+% ----------------------------------------
 \subsection{Color-Temperature Comparisons}
@@
 
+% ----------------------------------------
 \subsection{Isochrone Comparisons}
+
+Figure \ref{fig:isochrone_cmd}
 
@@
 \begin{figure*}[ht]
-% \plotone{figures/self_and_validation_test_teff_snr.png}
-% \plotone{figures/self_and_validation_test_feh_snr.png}
 	\plottwo{figures/crv_test_teff.png}{figures/crv_test_fe_h.png}
@@
 	\begin{center}
-% \plotone{figures/self_and_validation_test_spt_snr.png}
 	\plottwo{figures/self_test_spt.png}{figures/crv_test_spt.png}

### 2019-06-10 paper/draft.tex
discussion notes

@@
 
-While \thecannon\ generally achieves excellent precision at predicting labels and reproducing the spectral features as demonstrated in this study and previous work on \thecannon\ (\citealt{Ness:2015}, \citealt{Casey:2016}), the accuracy of labels from model application to new sources is limited by the accuracy, relative precision, size, dynamic range and representation of the training sample.
+While \thecannon\ generally achieves excellent precision at predicting labels and reproducing the spectral features as demonstrated in this study and previous work on \thecannon\ \citep{Ness:2015,Casey:2016}, the accuracy of labels from model application to new sources is limited by the accuracy, relative precision, size, dynamic range and representation of the training sample.
 
 Comment on the importance of each property of training sample: \\
-accuracy -  \\
-precision -  \\
+accuracy - some discussion on ground truth of stellar labels... \\
+precision - some discussion on how consistently training parameters are derived... \\
 size -  \\
-dynamic range - training sample must span \\
-representation - \\
+dynamic range - training sample must span the desired parameter range of the test sample. \thecannon\ does not extrapolate well. \\
+representation - model cannot account for physics not represented by training sample (i.e. this training sample is limited to old, slowly rotating, not magnetically active, singular sources...) \\
 
 Talk about the particular limitations we have found in this training sample.
+Talk about model improvements: incorporate broadening \citep{Behmard2019}; spectroscopic binary model \citep{ElBadry:2018}...
+\\
 
-% \begin{enumerate}
-% \item[-] Brief summary of what we did and found
-
-% \item[-] Things we learned along the way: For example, not all training sets worked well in our experiments, probably because they were too small or had too little dynamic range.
-
-% \item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: \thecannon\ can produce precise results, but is only as accurate as its training set.}\color{black}
-
-% \item[-] Some discussion of whether the derivatives are sensible or surprising.
-
-% \item[-] \color{gcolor}{HOGG:\thecannon\ as a test of evaluating training parameter accuracy?}\color{black}
-
-% \item[-] \color{gcolor}{HOGG: Discussion of validation tests and ground truth}\color{black}
-% \end{enumerate}
-
-% Future work:
-% \begin{enumerate}
-% \item[-] Create a binary sample; using color/magnitude \citealt{Widmark:2018} and by spectroscopy \citealt{ElBadry:2018}
-
-% \item[-] \color{gcolor}{HOGG: Discussion of the implications of these results for the NASA \tess\ Mission and how the future might change..?}\color{black}
-% \end{enumerate}
+Talk future work/implications with other stellar surveys (e.g. LAMOST), and exoplanet surveys (e.g. TESS)
 

### 2019-06-10 paper/draft.tex

@@
 Low-mass stars (with masses $M_{*} < 0.7\,\msun$ and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010}.
-With nuclear fusion timescales of $\tau>10^{11}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010}.
+With nuclear fusion timescales of $\tau>10^{11}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010,Woolf:2012}.
 
-Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods allow for new planet discovery in less observation time than their higher mass counterparts.
-For these reasons, M dwarfs are the primary candidates for exoplanet detection searches (e.g.
-the NASA \kepler\ and Transiting Exoplanet Survey Satellite (\tess) missions), making detailed and precise knowledge of M dwarf chemical compositions key to constraining the formation scenarios and atmospheric conditions of potentially habitable exoplanets.
+Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods (for a fixed stellar insulation flux) allow for new planet discovery in less observation time than their higher mass counterparts.
+For these reasons, M dwarfs are primary candidates for exoplanet searches, including by the NASA \kepler\ \citep[e.g.,][]{Dressing2015} and Transiting Exoplanet Survey Satellite (\tess) \citep[e.g.,][]{Muirhead:2018} missions.
+As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with {\it JWST}.
 
-Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
-The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III/IV mission \citep{Eisenstein:2011,Blanton:2017} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
+Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources), enabling studies of M dwarf abundances on both exoplanet and Galactic scales.
+The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}.
+\apogee\ pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have only been determined to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
 
-While elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precisions (down to errors in the range of $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by improvements in atmospheric models towards realistic 3D local thermodynamic equilibrium (LTE) assumptions \citep{Asplund2005}, and differential abundance techniques using equivalent widths \citep{Bedell2014},
-the determination of precise metallicities for M dwarfs has remained a long standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperature regimes, particularly TiO and VO which dominate the optical, H$_2$O and CO which dominate the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) which form at late spectral types \citep{Allard1997}.
-In particular, atmospheric models fail to reproduce spectral features due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions and lack of continuum in a spectrum make it difficult to deconvolve individual features and extract line strengths from equivalent widths,
-creating fundamental limitations in efforts towards exploiting the information content available out of high resolution datasets using traditional modeling and empirical methods.
+Elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precision (down to $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by improvements in atmospheric models, including realistic 3D local thermodynamic equilibrium (LTE) assumptions \citep{Asplund2005}, and differential abundance techniques using equivalent widths \citep{Bedell2014}.
+However, the determination of precise metallicities for M dwarfs has remained a long-standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperature regimes, particularly TiO and VO dominating the optical spectrum, H$_2$O and CO for the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) present in the spectra of the latest spectral types \citep{Allard1997}.
+In particular, atmospheric models fail to reproduce spectral features \citep[e.g.,][]{Mann:2013c} due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions, and the lack of continuum in a spectrum, all make it difficult to deconvolve individual features and extract line strengths from equivalent widths.
+The combination of these effects creates fundamental limitations to explore the information content available out of high resolution spectra using a traditional comparison to atmosphere models.
 
-A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically associated wide binaries are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g.
-\citealt{Terrien:2012}; \citealt{Rojas-Ayala:2012}; \citealt{Newton:2014}).
-However these metallicity calibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant calibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.
+A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically associated wide binaries are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the spectra of the M dwarf (e.g.
+\citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
+However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band or include important information about the spectral shape needed to determine \teff.
+Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that would fail to utilize the additional spectral information provided by \apogee's resolution.
 
-In this work we build a data-driven model for M dwarf spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a}.
-\thecannon\ is a pixel-by-pixel regression for spectral data given a training
-set of stars with reliable (assumed) accurate labels.
+%the relevant calibration indices used by existing studies (such as Ca~I or K~I lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.
+%AWM: I'm not really sure this is true.
+The major lines I and Babs use are in the K-band, very few studies use optical bands.
+Ryan includes ones in the H-band, some of which are in Apogee.
+I think the bigger point is that this fails to utilize the high-resolution information of the spectra, and relies instead of temperature calibration from spectral shape, which is impossible at this resolution.
+
+%%AWM: would it be useful to have some comment on data-driven models in general? Their virtues should probably be explained before invoking them
+
+In this work we build a data-driven model for M dwarf spectra with \thecannon\ (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}, \citealt{Behmard2019}).
+\thecannon\ is a pixel-by-pixel regression for spectral data given a training set of stars with reliable (assumed) accurate labels.
 The model can both predict a spectrum given labels, and also predict labels given a spectrum.
 Thus it is a generative model for stellar spectra that can be used for parameter inference.
+%AWM: not an expert here, so I won't add anything, but this probably needs more information!
 
@@
 \thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
+%AWM: again, the metallicity metrics Babs/Ryan/Myself use are all in the NIR, and they aren't based on models.
+The temperature tools use models, but they are calibrated against LBOI measurements.
+I need to come back to editing later, but this paragraph probably needs to be tweaked a bit.
 

### 2019-06-10 paper/draft.tex
fix latex issues

@@
 \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
-However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band or include important information about the spectral shape needed to determine \teff.
+However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band or include important information about the spectral shape needed to determine $\teff$.
 Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that would fail to utilize the additional spectral information provided by \apogee's resolution.
@@
 Iron abundances ($\feh$) are assigned to the physical parameter sample based on the strength of metal-sensitive lines in the near-infrared SpeX spectra \citep{RojasAyala:2010} using the calibration from \citet{Mann:2013a}.
-The relation between these lines and an absolute $\feh$scale is calibrated using wide binaries containing and F, G, or K-type primary and an M dwarf companion, under the assumption that binaries formed from the same molecular cloud and therefore have the same metallicity \citep{Bonfils:2005}.
+The relation between these lines and an absolute $\feh$ scale is calibrated using wide binaries containing and F, G, or K-type primary and an M dwarf companion, under the assumption that binaries formed from the same molecular cloud and therefore have the same metallicity \citep{Bonfils:2005}.
 Uncertainties are estimated to be $\simeq$0.08\,dex based on irreducible scatter in the empirical relation between selected lines and the assigned $\feh$ from the primary star.

### 2019-06-10 paper/draft.tex
little tweaks

@@
 For these reasons, M dwarfs are primary candidates for exoplanet searches, including by the NASA \kepler\ \citep[e.g.,][]{Dressing2015} and Transiting Exoplanet Survey Satellite (\tess) \citep[e.g.,][]{Muirhead:2018} missions.
-As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with {\it JWST}.
+As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with {\sl JWST}.
 
 Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources), enabling studies of M dwarf abundances on both exoplanet and Galactic scales.
-The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}.
+The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III/IV mission \citep{Eisenstein:2011,Blanton:2017} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}.
 \apogee\ pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have only been determined to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.

### 2019-06-10 paper/draft.tex
P7_tweaks

@@
 Thus it is a generative model for stellar spectra that can be used for parameter inference.
-%AWM: not an expert here, so I won't add anything, but this probably needs more information!
+%AWM: not an expert here, so I won't add anything, but this probably needs more information.
 
@@
 there are many stars that need labels;
-if there are spectral models that work in one wavelength range or resolution but not in another.
-The context of this paper is the last:
-there are precise calibrations for metallicities of M dwarf stars in the optical, which have been
-used successfully, but (as we note above) there isn't as much success with such modeling in the infrared, and not at \apogee\ resolution.
-\thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
+if there are spectral models or techniques that work in one wavelength range or resolution but not in another.
+%AWM: seems to me like the context here is all 3. You want to run on ~10,000 stars, which is expensive if you use models. You have labels for a small fraction of these. 
+%The context of this paper is the last:
+There are empirically-calibrated methods to measure \teff\ and metallicity from flux-calibrated spectra in the optical and NIR, but existing methods to model M dwarf spectra in the NIR at high resolution are computationally expensive, calibrated over a narrow range of \teff\ and metallicity, and/or focus on the $J$-band \citep{Lindgren:2016, Lindgren:2017}.
+\thecannon\ fills this niche: it doesn't require using specific lines or opacity information that may be missing from the models; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
 %AWM: again, the metallicity metrics Babs/Ryan/Myself use are all in the NIR, and they aren't based on models.
@@
 I need to come back to editing later, but this paragraph probably needs to be tweaked a bit.
+%AWM: OK, let me know what you think about those changes.
 
@@
 any input of stellar physics beyond whatever went in to producing the labels in the first
-place, which was in the context of optical spectroscopy.
-That is, the model to be applied to the \apogee\ data is learned from the \apogee\ data alone.
+place, which was in the context of optical spectroscopy. That is, the model to be applied to the \apogee\ data is learned from the \apogee\ data alone.
 This obviates physical models in the infrared, and capitalizes on the excellent data quality

### 2019-06-10 paper/draft.tex
P7_comments
acks

--- 2019-06-10 paper/draft.tex
@@
 We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Diogo Souto (Observatário Nacional/MCTI) and Keivan Stassun (Vanderbuilt) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
+This project was partially supported by
+the US National Aeronautics and Space Administration (NASA grant NNX12AI50G),
+the US National Science Foundation (NSF grant AST-1517237),
+and the Moore--Sloan Data Science Environment at NYU.
 
@@
 
-\color{gcolor}{HOGG: grants etc}\color{black}
-
-\color{gcolor}{ADAM: grants etc, FAST acknowledgement?}\color{black}
+\color{gcolor}{ADAM: please add NASA and NSF grants in the parentheses above, and do you have a FAST acknowledgement?}\color{black}
 

### 2019-06-10 paper/draft.tex

--- 2019-06-10 paper/draft.tex
@@
 I need to come back to editing later, but this paragraph probably needs to be tweaked a bit.
-%AWM: OK, let me know what you think about those changes.
+%AWM: OK, let me know what you think about those changes. Feel free to edit this more, but do note the work of Lindgren modeling J-band at high resolution (I think they use CRIRES).
 

### 2019-06-10 paper/draft.tex
discussion

@@
 %The context of this paper is the last:
-There are empirically-calibrated methods to measure \teff\ and metallicity from flux-calibrated spectra in the optical and NIR, but existing methods to model M dwarf spectra in the NIR at high resolution are computationally expensive, calibrated over a narrow range of \teff\ and metallicity, and/or focus on the $J$-band \citep{Lindgren:2016, Lindgren:2017}.
+There are empirically-calibrated methods to measure $\teff$ and metallicity from flux-calibrated spectra in the optical and NIR, but existing methods to model M dwarf spectra in the NIR at high resolution are computationally expensive, calibrated over a narrow range of $\teff$ and metallicity, and/or focus on the $J$-band \citep{Lindgren:2016, Lindgren:2017}.
 \thecannon\ fills this niche: it doesn't require using specific lines or opacity information that may be missing from the models; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
@@
 
-We trained and evaluated the performance of \thecannon\ at deriving atmospheric temperatures and metallicities from M dwarfs using high resolution near-infrared spectra, and compiled a catalog of 10,311 M dwarfs with \cannon\ temperatures, metallicities, and spectral types, 6D kinematics (\gaia\ parallaxes and proper motions + \apogee\ radial velocities) shown in Table \ref{table:test_results}.
+We trained a data-driven model (\thecannon; \citealt{Ness:2015}) to deliver
+high-quality atmospheric parameters ($\teff$ and $\feh$) for M-type dwarf stars
+from high-resolution infrared spectra from \apogee.
+This work was motivated by the problem that M dwarfs stars are difficult to
+model physically; the data are better than the models in important senses.
+And indeed we find that our data-driven model is both accurate in the
+data domain (as a spectral synthesis model) and precise in the latent domain
+(as a tool for deriving physical parameters).
+This accuracy and precision is consistent with previous work with
+\thecannon\ (\citealt{Ness:2015, Casey:2016, Ho:2017a, Ness:2018}), but
+extended to a new regime in spectral type ($\teff$) and surface
+gravity.
+The primary result of this work is that
+we have compiled a catalog of 10,311 M dwarfs with \cannon\ temperatures,
+metallicities, spectral types, and six-dimensional kinematics.
+These data are provided in Table \ref{table:test_results}.
 
-Table \ref{table:teff_derivative} and Table \ref{table:feh_derivative}, summarize which features are found to be strong temperature and metallicity indicators.
-
-While \thecannon\ generally achieves excellent precision at predicting labels and reproducing the spectral features as demonstrated in this study and previous work on \thecannon\ \citep{Ness:2015,Casey:2016}, the accuracy of labels from model application to new sources is limited by the accuracy, relative precision, size, dynamic range and representation of the training sample.
+While \thecannon\ achieves excellent precision at predicting labels
+and reproducing spectral features, the
+accuracy of labels it produces is limited by
+the accuracy, relative precision, size, dynamic range and
+representation of the training sample.
+That is, being a supervised method, \thecannon\ is never any better in a mean (bias) sense
+(although it can be better in a precision or variance sense) than the input training
+data.
+The catalog we have produced is a label transfer from parameters provided
+in our input data (\citealt{Mann:2015}; M15) and it implicitly adopts all the biases and
+issues from that input data.
+It is also limited to the stellar-parameter domain of that input catalog.
+That said, what we have done does also represent some kind of external validation of the M15
+stellar parameters:
+The success of our model shows that they do a great job of parameterizing a generative
+model of the spectra.
 
@@
 Talk future work/implications with other stellar surveys (e.g. LAMOST), and exoplanet surveys (e.g. TESS)
+
+Given that physics-based spectral models of M dwarfs have issues, one of the possible
+future values of the data-driven model shown here is that it is highly interpretable:
+It contains within it first and second derivatives of the spectral expectation with
+respect to the atmospheric parameters.
+We show some of these derivatives in [BIRKY WHAT FIGURE] and deliver relevant data in
+Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}
+These Tables summarize spectral features in the \apogee\ bandpass
+that are found to be strong temperature and metallicity indicators.
+In the long run, this is the primary value of data-driven models for astronomy:
+To provide physical insights that drive physical understandings.
+It is our hope that \thecannon, and models like it, will, in the end, lead to
+new and improved physical models, which will, in turn, put \thecannon\ out of business.
 

### 2019-06-10 paper/draft.tex
minor intro changes

@@
 \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
-However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band or include important information about the spectral shape needed to determine $\teff$.
-Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that would fail to utilize the additional spectral information provided by \apogee's resolution.
+However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band and does not contain information about the overall spectral shape often used to determine $\teff$ and spectral type.
+Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the additional spectral information provided by \apogee's resolution.
 
@@
 
-Here we take M-dwarf labels from samples of optically observed stars that overlap the
-\sdssiv\ \apogee\ sample, and use those labels to train a model and label all of the
-\sdssiv\ \apogee\ M dwarfs.
-One set of labels are physical parameters (effective temperatures and metallicities)
-and the other set of labels are spectral types.
-In each case, \thecannon\ produces a predictive model for \apogee\ spectra, but without
+Here we take M-dwarf labels from samples of well-characterized stars that overlap the
+\sdssiv\ \apogee\ sample, and use those labels to train a model and label all of the M dwarfs observed by
+\sdssiv\ \apogee. One set of labels are physical parameters (effective temperatures and metallicities)
+and the other set of labels are spectral types. In each case, \thecannon\ produces a predictive model for \apogee\ spectra, but without
 any input of stellar physics beyond whatever went in to producing the labels in the first
-place, which was in the context of optical spectroscopy. That is, the model to be applied to the \apogee\ data is learned from the \apogee\ data alone.
-This obviates physical models in the infrared, and capitalizes on the excellent data quality
+place. That is, the model to be applied to the \apogee\ data is learned from the \apogee\ data alone.
+This obviates physical models that are often inaccurate for M dwarfs, and capitalizes on the excellent data quality
 delivered by the \apogee\ spectrograph and project.

### 2019-06-10 paper/draft.tex
wrestle with figure placement

@@
-\documentclass[modern]{aastex62}
+\documentclass[twocolumn]{aastex62}
 
@@
 
+\begin{figure}
+	\plotone{figures/crv_test_teff.png}
+	\plotone{figures/crv_test_fe_h.png}
+	\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (left), and leave-one-out cross validation (right).
+Sources with strong rotational broadening (2M07171706--0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.
+Predictive accuracy, as computed by the scatter in cross validation is 77K in temperature, and 0.09 dex in metallicity.}
+	\label{fig:mann_validation}
+\end{figure}
+
 For the physical parameter model we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
@@
 
+\begin{figure}
+	\begin{center}
+	\plotone{figures/self_test_spt.png}
+	\plotone{figures/crv_test_spt.png}
+	\end{center}
+	\caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).
+Predictive accuracy, as computed by the scatter in cross validation is 0.9 subtypes.}
+	\label{fig:west_validation}
+\end{figure}
+
 We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
@@
 
+% ----------------------------------------
 \subsection{Test Sample \label{subsec:test_selection}}
+
+\begin{figure*}
+	\plotone{figures/cannon_chi_dist.png}
+	\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee --\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
+We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as ``safe''.}
+	\label{fig:chi_dist}
+\end{figure*}
+
+\begin{figure*}
+	\plotone{figures/cmd_selection.png}
+	\caption{\gaia\ and \zmass\ color-magnitude cuts for 12,037 sources with $\chi^2<100,000$.
+Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia.}
+	\label{fig:cmd_selection}
+\end{figure*}
+
 Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \textsl{Gaia DR2} (\citealt{Brown:2018}) and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources.
@@
 
+\begin{figure*}[h]
+	\plottwo{figures/cmd_teff_full.png}{figures/cmd_teff_safe.png}
+	\plottwo{figures/cmd_feh_full.png}{figures/cmd_feh_safe.png}
+	\plottwo{figures/cmd_spt_full.png}{figures/cmd_spt_safe.png}
+	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, binaries, and K dwarfs).
+Overplotted with orange triangles are the M15 and W11 training samples, for their respective \cannon\ test labels.
+Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly up from the main sequence branch as expected.
+Deviations from these gradients seen at the upper slice of the main sequence is likely remaining contamination from the binary sequence.}
+	\label{fig:safe_selection}
+\end{figure*}
+
 The three panels of \ref{fig:safe_selection} show before and after selection of the sources in \gaia\ color-magnitude space, colored by \thecannon\ determined temperatures, metallicities, and spectral types, with their respective training samples overplotted in orange.
@@
 \subsection{Color-Temperature Comparisons}
+
+\begin{figure*}[h]
+	\plotone{figures/aspcap_cannon_validation.png}
+	\caption{\cannon\ (right column) and \aspcap\ (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 6,257 sources with $V,J,K$ colors.
+Shown are \cannon\ and \aspcap\ temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and $\feh$ dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation $V-K$ and $\feh$ (bottom).}
+	\label{fig:teff_comparisons}
+\end{figure*}
+
+\begin{figure*}[h]
+	\plotone{figures/aspcap_cannon_label_hist.png}
+	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the \aspcap\ pipeline.
+Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars \aspcap\ predicts temperatures with slightly higher temperatures and metallicities.}
+	\label{fig:aspcap_cannon_label_hist} 
+\end{figure*}
+
+\begin{figure*}[h]
+	\plotone{figures/isochrone_cmd.png}
+	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
+The four panels show five metallicity tracks (ranging $-0.5<$[M/H]$<0.5$) over varying ages (1, 3, 5, and 7 Gyr).}
+	\label{fig:isochrone_cmd}
+\end{figure*}
 
@@
 
-\newpage
-
-\begin{figure*}[ht]
-	\plottwo{figures/crv_test_teff.png}{figures/crv_test_fe_h.png}
-	\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (left), and leave-one-out cross validation (right).
-Sources with strong rotational broadening (2M07171706--0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.
-Predictive accuracy, as computed by the scatter in cross validation is 77K in temperature, and 0.09 dex in metallicity.}\label{fig:mann_validation}
-\end{figure*}
-
-\begin{figure*}[ht]
-	\begin{center}
-	\plottwo{figures/self_test_spt.png}{figures/crv_test_spt.png}
-	\end{center}
-	\caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).
-Predictive accuracy, as computed by the scatter in cross validation is 0.9 subtypes.}\label{fig:west_validation}
-\end{figure*}
-
-%---------------------------
-\begin{figure*}[ht]
+\begin{figure*}[h]
 	\begin{center}
@@
 
-\begin{figure*}[ht]
+\begin{figure*}[h]
 	\begin{center}
@@
 	\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities.
-\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.}\label{fig:demo_teff}
+\textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.}
+	\label{fig:demo_teff}
 \end{figure*}
@@
 %---------------------------
-\begin{figure*}[ht]
+\begin{figure*}[h]
 	\begin{center}
@@
 	\end{center}
-	\caption{Derivative plots for spectral type model taken at the median training spectral type, SPT$=3$; jackknife computed error at each pixel is shown in red.} \label{fig:west_derivative}
+	\caption{Derivative plots for spectral type model taken at the median training spectral type, SPT$=3$; jackknife computed error at each pixel is shown in red.} 
+	\label{fig:west_derivative}
 \end{figure*}
 
-\begin{figure*}[ht]
-	\plotone{figures/cannon_chi_dist.png}
-	\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee --\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
-We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as ``safe''.
-\label{fig:chi_dist}}
-\end{figure*}
-
-\begin{figure*}[ht]
-	\plotone{figures/cmd_selection.png}
-	\caption{\gaia\ and \zmass\ color-magnitude cuts for 12,037 sources with $\chi^2<100,000$.
-Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia.
-\label{fig:cmd_selection}}
-\end{figure*}
-
-\begin{figure*}[ht]
-	\plottwo{figures/cmd_teff_full.png}{figures/cmd_teff_safe.png}
-	\plottwo{figures/cmd_feh_full.png}{figures/cmd_feh_safe.png}
-	\plottwo{figures/cmd_spt_full.png}{figures/cmd_spt_safe.png}
-	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, binaries, and K dwarfs).
-Overplotted with orange triangles are the M15 and W11 training samples, for their respective \cannon\ test labels.
-Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly up from the main sequence branch as expected.
-Deviations from these gradients seen at the upper slice of the main sequence is likely remaining contamination from the binary sequence.
-\label{fig:safe_selection}}
-\end{figure*}
-
-\begin{figure*}[ht]
+\begin{figure*}[h]
 	\plotone{figures/spt_vs_teff_vs_feh.png}
@@
 
-\begin{figure*}[ht]
-	\plotone{figures/aspcap_cannon_label_hist.png}
-	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the \aspcap\ pipeline.
-Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars \aspcap\ predicts temperatures with slightly higher temperatures and metallicities.
-\label{fig:aspcap_cannon_label_hist}}
-\end{figure*}
+%---------------------------
 
-%---------------------------
-\begin{figure*}[ht]
-	\plotone{figures/aspcap_cannon_validation.png}
-	\caption{\cannon\ (right column) and \aspcap\ (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 6,257 sources with $V,J,K$ colors.
-Shown are \cannon\ and \aspcap\ temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and $\feh$ dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation $V-K$ and $\feh$ (bottom).
-\label{fig:teff_comparisons}}
-\end{figure*}
-
-%---------------------------
-\begin{figure*}[ht]
-	\plotone{figures/isochrone_cmd.png}
-	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
-The four panels show five metallicity tracks (ranging $-0.5<$[M/H]$<0.5$) over varying ages (1, 3, 5, and 7 Gyr).
-\label{fig:isochrone_cmd}}
-\end{figure*}
-
-\begin{figure*}[ht]
+\begin{figure*}[h]
 	\plotone{figures/mann_radii.png}
-	\caption{Comparison of radii estimated from M15 relations: on the x-axis radii are derived from  K magnitudes and \cannon\ $\feh$, on the y-axis radii are derived from \cannon\ $\teff$ and $\feh$.
-\label{fig:radii}}
+	\caption{Comparison of radii estimated from M15 relations: on the x-axis radii are derived from  K magnitudes and \cannon\ $\feh$, on the y-axis radii are derived from \cannon\ $\teff$ and $\feh$.}
+	\label{fig:radii}
 \end{figure*}

### 2019-06-10 paper/draft.tex
remove unnecessary figure borders

@@
 \begin{figure}
-	\plotone{figures/crv_test_teff.png}
-	\plotone{figures/crv_test_fe_h.png}
+	\includegraphics[width=\linewidth]{figures/crv_test_teff.png}
+	\includegraphics[width=\linewidth]{figures/crv_test_fe_h.png}
 	\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (left), and leave-one-out cross validation (right).
@@
 	\begin{center}
-	\plotone{figures/self_test_spt.png}
-	\plotone{figures/crv_test_spt.png}
+	\includegraphics[width=\linewidth]{figures/self_test_spt.png}
+	\includegraphics[width=\linewidth]{figures/crv_test_spt.png}
 	\end{center}
@@
 \begin{figure*}
-	\plotone{figures/cannon_chi_dist.png}
+	\includegraphics[width=\linewidth]{figures/cannon_chi_dist.png}
 	\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee --\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
@@
 \begin{figure*}
-	\plotone{figures/cmd_selection.png}
+	\includegraphics[width=\linewidth]{figures/cmd_selection.png}
 	\caption{\gaia\ and \zmass\ color-magnitude cuts for 12,037 sources with $\chi^2<100,000$.
@@
 \begin{figure*}[h]
-	\plotone{figures/aspcap_cannon_validation.png}
+	\includegraphics[width=\linewidth]{figures/aspcap_cannon_validation.png}
 	\caption{\cannon\ (right column) and \aspcap\ (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 6,257 sources with $V,J,K$ colors.
@@
 \begin{figure*}[h]
-	\plotone{figures/aspcap_cannon_label_hist.png}
+	\includegraphics[width=\linewidth]{figures/aspcap_cannon_label_hist.png}
 	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the \aspcap\ pipeline.
@@
 \begin{figure*}[h]
-	\plotone{figures/isochrone_cmd.png}
+	\includegraphics[width=\linewidth]{figures/isochrone_cmd.png}
 	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.

### 2019-06-10 paper/draft.tex
describe data-driven terminology in intro

@@
 
-In this work we build a data-driven model for M dwarf spectra with \thecannon\ (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}, \citealt{Behmard2019}).
-\thecannon\ is a pixel-by-pixel regression for spectral data given a training set of stars with reliable (assumed) accurate labels.
-The model can both predict a spectrum given labels, and also predict labels given a spectrum.
-Thus it is a generative model for stellar spectra that can be used for parameter inference.
+In this work we build a data-driven model for M dwarf spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a,Behmard2019}: a fully empirical model, which employs no use of line lists or radiative transfer models.
+\thecannon\ is instead a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:cannon}).
+The model in this sense is used to \emph{transfer} labels from spectra for which we know parameters to those which we do not.
+Hence this data-driven approach effectively circumvents the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (and very accurately measured) \emph{reference labels} possibly measured from other data.
+
 %AWM: not an expert here, so I won't add anything, but this probably needs more information.
+%JB: I've swapped some sentences from section 2.2 and section 3, which explain the terminology.
+Do you think this paragraph adds enough clarity now without being too redundant with the following sections 2 & 3 describing The Cannon?
 
@@
 
-\thecannon\ is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:cannon}).
-The model in this sense is used to \emph{transfer} labels from spectra for which we know parameters to those which we do not.
-Hence this data-driven approach effectively circumvents the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (and very accurately measured) \emph{reference labels} possibly measured from other data.
+\thecannon\ is a pixel-by-pixel regression for spectral data given a training set of stars with reliable (assumed) accurate labels.
+The model can both predict a spectrum given labels, and also predict labels given a spectrum.
+Thus it is a generative model for stellar spectra that can be used for parameter inference.
 
@@
 
-\thecannon\ is a fully empirical model, which employs no use of line lists or radiative transfer models.
-Instead, \thecannon\ is a regression model which relies on two assumptions: first that sources with identical labels have near-identical flux at each pixel, and second that the expected flux at each pixel varies continuosly with change in label.
+\thecannon\ is a regression model which relies on two assumptions: first that sources with identical labels have near-identical flux at each wavelength pixel, and second that the expected flux at each pixel varies continuosly with change in label.
 

### 2019-06-10 paper/draft.tex
restructure order of sections: (2) data, (3) methods, (4) sample selection
remove redundancies describing the cannon

@@
 
-% ----------------------------------------
-\subsection{Training Sample Selection}
-
-\thecannon\ is a pixel-by-pixel regression for spectral data given a training set of stars with reliable (assumed) accurate labels.
-The model can both predict a spectrum given labels, and also predict labels given a spectrum.
-Thus it is a generative model for stellar spectra that can be used for parameter inference.
-
-\thecannon\ model can in principle be trained on any physical or empirical labels available, giving a wide range of flexibility to the model.
-However choosing a training sample with high quality labels is critical to the performance of \thecannon.
-Limitations of \thecannon\ include that test (output) labels are only accurate if the training labels are accurate, and only precise if the training labels are measured consistently across the training sample.
-Also, it is critical to have a training sample with the dynamic range to span the entire parameter space of interest, as \thecannon\ does not extrapolate well outside the parameter space of the training sample.
-Finally, \thecannon\ assumes that the dependence of spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions---if that isn't true, there will be features that \thecannon\ cannot reproduce.
-
-For the purpose of this study we have constructed two different training samples: first a one-dimensional \emph{spectral type model}, and second a two-dimensional \emph{physical parameter model} which describes the temperature and metallicity.
-The choice of training labels, dimensionality of our data set, and requirements for a good training set we discuss further in Section \ref{sec:discussion}.
-
-The spectral type sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from \sdss.
-For each source in the catalog, spectral types were determined both through an automated routine for comparing spectral type templates to data using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
-
-The physical parameter sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 < \feh < 0.49$\,dex, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
-The major difference between the extension sample is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample.
-
-The M15 catalog in total contains 183 sources and the extension sample another 500 stars, both samples were primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog.
-All targets have low-res optical spectra from the SNIFS spectrograph \citep{Lantz:2004} and infrared spectra taken with the SpeX Spectrograph \citep{Rayner:2003}, which have been combined to estimate largely empirical bolometric fluxes.
-Effective temperatures have been estimated by comparing the SNIFS spectra to BT-SETTL atmospheric models \citep{Allard:2011}.
-A subsample of 29 sources with measured angular diameters from long-baseline optical interferometry \citep{Boyajian:2012} are used to calibrate the model comparison, including masking of spectral regions poorly reproduced by the model spectra \citep{Mann:2013c}.
-Based on difference between assigned $\teff$ values and those from angular diameters, absolute uncertainty on $\teff$ is estimated to be 60K in $\teff$, although the relative uncertainty is likely a factor of $\simeq$2 better.
-
-Iron abundances ($\feh$) are assigned to the physical parameter sample based on the strength of metal-sensitive lines in the near-infrared SpeX spectra \citep{RojasAyala:2010} using the calibration from \citet{Mann:2013a}.
-The relation between these lines and an absolute $\feh$ scale is calibrated using wide binaries containing and F, G, or K-type primary and an M dwarf companion, under the assumption that binaries formed from the same molecular cloud and therefore have the same metallicity \citep{Bonfils:2005}.
-Uncertainties are estimated to be $\simeq$0.08\,dex based on irreducible scatter in the empirical relation between selected lines and the assigned $\feh$ from the primary star.
-As with $\teff$ relative errors on $\feh$ are smaller, estimated to be 0.04--0.06\,dex over most of the temperature and metallicity range considered here.
-
-We note that surface gravity is not included as a training label.
-The reason for this is that for main-sequence M dwarfs, the parameter is almost entirely redundant with metallicity.
-Unlike their higher-mass counterparts, M dwarf properties do not change measurably over the age of the Universe after arriving at the zero-age main sequence.
-Hence perfect knowledge of abundances and $\teff$ for an M dwarf should uniquely determine its surface gravity, position on a color-magnitude diagram, and overall luminosity.
-While we only had $\feh$ for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important compared to other uncertainties in extreme cases (e.g., Carbon stars).
-
-%We note that rotational velocity (vsini) is not a parameter are not included as training labels because that would be hard... also it's probably not needed... I'll let someone else write that. -AWM
-%\color{gcolor}{}\color{black}
-
 %%==========================================================================================
@@
 derivatives.
+
+%%==========================================================================================
+\section{Sample Selection}
+
+\thecannon\ model can in principle be trained on any physical or empirical labels available, giving a wide range of flexibility to the model.
+However choosing a training sample with high quality labels is critical to the performance of \thecannon.
+Limitations of \thecannon\ include that test (output) labels are only accurate if the training labels are accurate, and only precise if the training labels are measured consistently across the training sample.
+Also, it is critical to have a training sample with the dynamic range to span the entire parameter space of interest, as \thecannon\ does not extrapolate well outside the parameter space of the training sample.
+Finally, \thecannon\ assumes that the dependence of spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions---if that isn't true, there will be features that \thecannon\ cannot reproduce.
+
+For the purpose of this study we have constructed two different training samples: first a one-dimensional \emph{spectral type model}, and second a two-dimensional \emph{physical parameter model} which describes the temperature and metallicity.
+The choice of training labels, dimensionality of our data set, and requirements for a good training set we discuss further in Section \ref{sec:discussion}.
+
+The spectral type sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from \sdss.
+For each source in the catalog, spectral types were determined both through an automated routine for comparing spectral type templates to data using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
+
+The physical parameter sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 < \feh < 0.49$\,dex, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
+The major difference between the extension sample is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample.
+
+The M15 catalog in total contains 183 sources and the extension sample another 500 stars, both samples were primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog.
+All targets have low-res optical spectra from the SNIFS spectrograph \citep{Lantz:2004} and infrared spectra taken with the SpeX Spectrograph \citep{Rayner:2003}, which have been combined to estimate largely empirical bolometric fluxes.
+Effective temperatures have been estimated by comparing the SNIFS spectra to BT-SETTL atmospheric models \citep{Allard:2011}.
+A subsample of 29 sources with measured angular diameters from long-baseline optical interferometry \citep{Boyajian:2012} are used to calibrate the model comparison, including masking of spectral regions poorly reproduced by the model spectra \citep{Mann:2013c}.
+Based on difference between assigned $\teff$ values and those from angular diameters, absolute uncertainty on $\teff$ is estimated to be 60K in $\teff$, although the relative uncertainty is likely a factor of $\simeq$2 better.
+
+Iron abundances ($\feh$) are assigned to the physical parameter sample based on the strength of metal-sensitive lines in the near-infrared SpeX spectra \citep{RojasAyala:2010} using the calibration from \citet{Mann:2013a}.
+The relation between these lines and an absolute $\feh$ scale is calibrated using wide binaries containing and F, G, or K-type primary and an M dwarf companion, under the assumption that binaries formed from the same molecular cloud and therefore have the same metallicity \citep{Bonfils:2005}.
+Uncertainties are estimated to be $\simeq$0.08\,dex based on irreducible scatter in the empirical relation between selected lines and the assigned $\feh$ from the primary star.
+As with $\teff$ relative errors on $\feh$ are smaller, estimated to be 0.04--0.06\,dex over most of the temperature and metallicity range considered here.
+
+We note that surface gravity is not included as a training label.
+The reason for this is that for main-sequence M dwarfs, the parameter is almost entirely redundant with metallicity.
+Unlike their higher-mass counterparts, M dwarf properties do not change measurably over the age of the Universe after arriving at the zero-age main sequence.
+Hence perfect knowledge of abundances and $\teff$ for an M dwarf should uniquely determine its surface gravity, position on a color-magnitude diagram, and overall luminosity.
+While we only had $\feh$ for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important compared to other uncertainties in extreme cases (e.g., Carbon stars).
+
+%We note that rotational velocity (vsini) is not a parameter are not included as training labels because that would be hard... also it's probably not needed... I'll let someone else write that. -AWM
+%\color{gcolor}{}\color{black}
 

### 2019-06-10 paper/draft.tex
gaia subsection placeholder note
add subsection headers
rewrite intro outline; rename section labels

@@
 In this work we build a data-driven model for M dwarf spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a,Behmard2019}: a fully empirical model, which employs no use of line lists or radiative transfer models.
-\thecannon\ is instead a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:cannon}).
+\thecannon\ is instead a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:methods}).
 The model in this sense is used to \emph{transfer} labels from spectra for which we know parameters to those which we do not.
@@
 
-This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the APOGEE survey and previous studies of M dwarfs in APOGEE, as well as our sample selection and derivation of training parameters.
-Section \ref{sec:cannon} describes our model implementation using \thecannon\ framework.
+This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the \apogee\ and \gaia\ surveys, as well as previous studies of M dwarfs in \apogee.
+Section \ref{sec:methods} describes our model implementation using \thecannon\ framework, and Section \ref{sec:sample_selection} describes the our sample selection and derivation of training parameters.
 In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$10,000 sources, and examine the validity of our parameters against color-temperature relations and isochrone grids.
@@
 
+% ----------------------------------------
+\subsection{\gaia\ Survey}
+
+Need to talk about \gaia\ data \& what we use this dataset for...
+
 %%==========================================================================================
-\section{Method}\label{sec:cannon}
+\section{Method} \label{sec:methods}
 
@@
 %%==========================================================================================
-\section{Sample Selection}
+\section{Sample Selection} \label{sec:sample_selection}
 
@@
 
+% ----------------------------------------
+\subsection{Spectral Type Sample}
+
 The spectral type sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from \sdss.
 For each source in the catalog, spectral types were determined both through an automated routine for comparing spectral type templates to data using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
+
+% ----------------------------------------
+\subsection{Physical Parameter Sample}
 

### 2019-06-11 paper/draft.tex
move derivative tables

@@
 
+\begin{deluxetable}{|c|c|c|c|c|}
+\tablecaption{Strong $\teff$-correlated molecular lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
+\tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
+	\startdata
+	FeH & 16107.127 & -0.002 & 0.001 & -2.78 \\
+	FeH & 16245.463 & -0.005 & 0.002 & -3.214 \\
+	FeH & 16271.519 & -0.002 & 0.001 & -2.013 \\
+	FeH & 16284.562 & -0.006 & 0.001 & -5.807 \\
+	FeH & 16377.066 & -0.005 & 0.001 & -4.554 \\
+	FeH & 16741.489 & -0.002 & 0.001 & -2.873 \\
+	FeH & 16812.415 & -0.004 & 0.001 & -4.241 \\
+	FeH & 16813.808 & -0.006 & 0.001 & -3.954 \\
+	FeH & 16922.405 & -0.003 & 0.001 & -3.867 \\
+	FeH & 16934.801 & -0.004 & 0.001 & -3.591 \\
+	OH & 15278.267 & 0.006 & 0.003 & 2.47 \\
+	OH & 15280.8 & 0.004 & 0.002 & 2.007 \\
+	OH & 15391.186 & 0.006 & 0.003 & 2.127 \\
+	OH & 15407.355 & 0.01 & 0.002 & 4.836 \\
+	OH & 15409.271 & 0.008 & 0.002 & 4.683 \\
+	OH & 15505.797 & 0.009 & 0.003 & 3.511 \\
+	OH & 15560.305 & 0.008 & 0.002 & 3.865 \\
+	OH & 15565.895 & 0.005 & 0.002 & 2.505 \\
+	OH & 15568.691 & 0.007 & 0.001 & 4.929 \\
+	OH & 15572.133 & 0.008 & 0.002 & 4.393 \\
+	OH & 16052.7 & 0.008 & 0.002 & 3.517 \\
+	OH & 16055.362 & 0.006 & 0.002 & 3.7 \\
+	OH & 16061.796 & 0.007 & 0.003 & 2.502 \\
+	OH & 16065.124 & 0.009 & 0.002 & 4.214 \\
+	OH & 16069.564 & 0.008 & 0.003 & 3.275 \\
+	OH & 16074.227 & 0.007 & 0.002 & 3.338 \\
+	OH & 16190.345 & 0.007 & 0.003 & 2.498 \\
+	OH & 16192.134 & 0.008 & 0.002 & 4.012 \\
+	OH & 16203.995 & 0.006 & 0.002 & 3.039 \\
+	OH & 16207.129 & 0.004 & 0.002 & 2.196 \\
+	OH & 16352.196 & 0.007 & 0.002 & 3.609 \\
+	OH & 16354.682 & 0.011 & 0.003 & 4.167 \\
+	OH & 16364.626 & 0.007 & 0.003 & 2.661 \\
+	OH & 16368.244 & 0.008 & 0.002 & 4.303 \\
+	OH & 16581.281 & 0.008 & 0.002 & 4.41 \\
+	OH & 16581.74 & 0.004 & 0.001 & 4.842 \\
+	OH & 16866.621 & 0.005 & 0.001 & 3.169 \\
+	OH & 16871.982 & 0.012 & 0.002 & 4.892 \\
+	OH & 16878.976 & 0.006 & 0.002 & 3.807 \\
+	OH & 16884.573 & 0.005 & 0.002 & 3.066 \\
+	OH & 16886.206 & 0.006 & 0.001 & 5.131 \\
+	OH & 16895.074 & 0.005 & 0.002 & 2.197 \\
+	\nodata & \nodata & \nodata & \nodata & \nodata
+	\enddata
+\end{deluxetable} \label{table:teff_derivative}
+
+\begin{deluxetable}{|c|c|c|c|c|}
+\tablecaption{Strong $\teff$-correlated atomic lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
+\tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
+	\startdata
+	\nodata & \nodata & \nodata & \nodata & \nodata \\
+	Fe I & 15207.509 & -0.007 & 0.003 & -2.622 \\
+	Fe I & 15648.478 & -0.006 & 0.002 & -3.273 \\
+	Fe I & 15692.643 & -0.008 & 0.002 & -4.056 \\
+	Si I & 15960.044 & -0.009 & 0.004 & -2.523 \\
+	Si I & 16094.893 & -0.014 & 0.003 & -4.368 \\
+	Si I & 16680.77 & -0.01 & 0.003 & -3.475 \\
+	K I & 15162.824 & -0.006 & 0.002 & -2.294 \\
+	Ti I & 15715.641 & 0.006 & 0.002 & 2.894 \\
+	Ti I & 16634.973 & 0.003 & 0.001 & 4.224 \\
+	V I & 15923.924 & 0.003 & 0.001 & 3.6 \\
+	Mn I & 15159.054 & -0.006 & 0.002 & -3.123 \\
+	\enddata
+\end{deluxetable}
+
+\begin{deluxetable}{|c|c|c|c|c|}
+\tablecaption{Strong $\feh$-correlated atomic and molecular lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
+\tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
+	\startdata
+	FeH & 16114.027 & 0.004 & 0.001 & 2.673 \\
+	FeH & 16574.64 & 0.003 & 0.001 & 2.356 \\
+	FeH & 16694.372 & 0.003 & 0.001 & 2.193 \\
+	FeH & 16814.041 & 0.004 & 0.002 & 2.513 \\
+	OH & 16052.7 & 0.005 & 0.002 & 2.034 \\
+	OH & 16055.362 & 0.006 & 0.002 & 2.805 \\
+	OH & 16065.124 & 0.007 & 0.003 & 2.364 \\
+	OH & 16871.982 & 0.007 & 0.003 & 2.306 \\
+	OH & 16886.206 & 0.004 & 0.001 & 2.52 \\
+	Fe I & 15490.381 & -0.004 & 0.001 & -2.722 \\
+	Fe I & 15692.643 & -0.005 & 0.002 & -2.759 \\
+	Fe I & 16009.512 & -0.006 & 0.003 & -2.143 \\
+	Si I & 15960.044 & -0.007 & 0.003 & -2.112 \\
+	Si I & 16680.77 & -0.006 & 0.003 & -2.156 \\
+	K I & 15163.033 & 0.006 & 0.003 & 2.043 \\
+	V I & 15923.924 & 0.002 & 0.001 & 2.017 \\
+	Cr I & 15680.073 & -0.005 & 0.002 & -2.811 \\
+	Mn I & 15159.054 & -0.008 & 0.002 & -3.298 \\
+	Mn I & 15262.023 & -0.005 & 0.002 & -2.996 \\
+	\enddata
+\end{deluxetable} \label{table:feh_derivative}
+
 The spectra contain roughly 8000 pixels,
@@
 
-We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
-We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
-As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates \thecannon\ does a poor job at extrapolating to labels outside of the training sample space.
-Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows the self and LOOCV tests for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.
-
-% ----------------------------------------
-\subsection{Test Sample \label{subsec:test_selection}}
-
 \begin{figure*}
@@
 \end{figure*}
+
+We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
+We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
+As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates \thecannon\ does a poor job at extrapolating to labels outside of the training sample space.
+Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows the self and LOOCV tests for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.
+
+% ----------------------------------------
+
+\subsection{Test Sample \label{subsec:test_selection}}
 
@@
 
-\begin{deluxetable*}{|c|c|c|c|c|}
-\tablecaption{Strong $\teff$-correlated lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
-\tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
-	\startdata
-	Fe I & 15207.509 & -0.007 & 0.003 & -2.622 \\
-	Fe I & 15648.478 & -0.006 & 0.002 & -3.273 \\
-	Fe I & 15692.643 & -0.008 & 0.002 & -4.056 \\
-	FeH & 16107.127 & -0.002 & 0.001 & -2.78 \\
-	FeH & 16245.463 & -0.005 & 0.002 & -3.214 \\
-	FeH & 16271.519 & -0.002 & 0.001 & -2.013 \\
-	FeH & 16284.562 & -0.006 & 0.001 & -5.807 \\
-	FeH & 16377.066 & -0.005 & 0.001 & -4.554 \\
-	FeH & 16741.489 & -0.002 & 0.001 & -2.873 \\
-	FeH & 16812.415 & -0.004 & 0.001 & -4.241 \\
-	FeH & 16813.808 & -0.006 & 0.001 & -3.954 \\
-	FeH & 16922.405 & -0.003 & 0.001 & -3.867 \\
-	FeH & 16934.801 & -0.004 & 0.001 & -3.591 \\
-	OH & 15278.267 & 0.006 & 0.003 & 2.47 \\
-	OH & 15280.8 & 0.004 & 0.002 & 2.007 \\
-	OH & 15391.186 & 0.006 & 0.003 & 2.127 \\
-	OH & 15407.355 & 0.01 & 0.002 & 4.836 \\
-	OH & 15409.271 & 0.008 & 0.002 & 4.683 \\
-	OH & 15505.797 & 0.009 & 0.003 & 3.511 \\
-	OH & 15560.305 & 0.008 & 0.002 & 3.865 \\
-	OH & 15565.895 & 0.005 & 0.002 & 2.505 \\
-	OH & 15568.691 & 0.007 & 0.001 & 4.929 \\
-	OH & 15572.133 & 0.008 & 0.002 & 4.393 \\
-	OH & 16052.7 & 0.008 & 0.002 & 3.517 \\
-	OH & 16055.362 & 0.006 & 0.002 & 3.7 \\
-	OH & 16061.796 & 0.007 & 0.003 & 2.502 \\
-	OH & 16065.124 & 0.009 & 0.002 & 4.214 \\
-	OH & 16069.564 & 0.008 & 0.003 & 3.275 \\
-	OH & 16074.227 & 0.007 & 0.002 & 3.338 \\
-	OH & 16190.345 & 0.007 & 0.003 & 2.498 \\
-	OH & 16192.134 & 0.008 & 0.002 & 4.012 \\
-	OH & 16203.995 & 0.006 & 0.002 & 3.039 \\
-	OH & 16207.129 & 0.004 & 0.002 & 2.196 \\
-	OH & 16352.196 & 0.007 & 0.002 & 3.609 \\
-	OH & 16354.682 & 0.011 & 0.003 & 4.167 \\
-	OH & 16364.626 & 0.007 & 0.003 & 2.661 \\
-	OH & 16368.244 & 0.008 & 0.002 & 4.303 \\
-	OH & 16581.281 & 0.008 & 0.002 & 4.41 \\
-	OH & 16581.74 & 0.004 & 0.001 & 4.842 \\
-	OH & 16866.621 & 0.005 & 0.001 & 3.169 \\
-	OH & 16871.982 & 0.012 & 0.002 & 4.892 \\
-	OH & 16878.976 & 0.006 & 0.002 & 3.807 \\
-	OH & 16884.573 & 0.005 & 0.002 & 3.066 \\
-	OH & 16886.206 & 0.006 & 0.001 & 5.131 \\
-	OH & 16895.074 & 0.005 & 0.002 & 2.197 \\
-	Si I & 15960.044 & -0.009 & 0.004 & -2.523 \\
-	Si I & 16094.893 & -0.014 & 0.003 & -4.368 \\
-	Si I & 16680.77 & -0.01 & 0.003 & -3.475 \\
-	K I & 15162.824 & -0.006 & 0.002 & -2.294 \\
-	Ti I & 15715.641 & 0.006 & 0.002 & 2.894 \\
-	Ti I & 16634.973 & 0.003 & 0.001 & 4.224 \\
-	V I & 15923.924 & 0.003 & 0.001 & 3.6 \\
-	Mn I & 15159.054 & -0.006 & 0.002 & -3.123 \\
-	\enddata
-\end{deluxetable*} \label{table:teff_derivative}
-
-%====================================
-
-\begin{deluxetable*}{|c|c|c|c|c|}
-\tablecaption{Strong $\feh$-correlated lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
-\tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
-	\startdata
-	Fe I & 15490.381 & -0.004 & 0.001 & -2.722 \\
-	Fe I & 15692.643 & -0.005 & 0.002 & -2.759 \\
-	Fe I & 16009.512 & -0.006 & 0.003 & -2.143 \\
-	FeH & 16114.027 & 0.004 & 0.001 & 2.673 \\
-	FeH & 16574.64 & 0.003 & 0.001 & 2.356 \\
-	FeH & 16694.372 & 0.003 & 0.001 & 2.193 \\
-	FeH & 16814.041 & 0.004 & 0.002 & 2.513 \\
-	OH & 16052.7 & 0.005 & 0.002 & 2.034 \\
-	OH & 16055.362 & 0.006 & 0.002 & 2.805 \\
-	OH & 16065.124 & 0.007 & 0.003 & 2.364 \\
-	OH & 16871.982 & 0.007 & 0.003 & 2.306 \\
-	OH & 16886.206 & 0.004 & 0.001 & 2.52 \\
-	Si I & 15960.044 & -0.007 & 0.003 & -2.112 \\
-	Si I & 16680.77 & -0.006 & 0.003 & -2.156 \\
-	K I & 15163.033 & 0.006 & 0.003 & 2.043 \\
-	V I & 15923.924 & 0.002 & 0.001 & 2.017 \\
-	Cr I & 15680.073 & -0.005 & 0.002 & -2.811 \\
-	Mn I & 15159.054 & -0.008 & 0.002 & -3.298 \\
-	Mn I & 15262.023 & -0.005 & 0.002 & -2.996 \\
-	\enddata
-\end{deluxetable*} \label{table:feh_derivative}
-
-%====================================
-\newpage
-
 \begin{deluxetable*}{rcl}

### 2019-06-11 paper/draft.tex
adjust paragraph spacing

@@
 In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$10,000 sources, and examine the validity of our parameters against color-temperature relations and isochrone grids.
-Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results in the context of present and upcoming stellar and exoplanetary surveys.
+Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results in the context of present and upcoming stellar and exoplanetary surveys. \\
 
@@
 
-Need to talk about \gaia\ data \& what we use this dataset for...
+Need to talk about \gaia\ data \& what we use this dataset for... \\
 
@@
 the labels. In the test step, the regression is designed to predict labels given the spectral
-derivatives.
+derivatives. \\
 
@@
 While we only had $\feh$ for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important compared to other uncertainties in extreme cases (e.g., Carbon stars).
+\\
 
@@
 
-\color{gcolor}{BIRKY: Can you demonstrate this by showing the RMS residuals, putting it in as a figure, and putting this paragraph back to me?
+% \color{gcolor}{BIRKY: Can you demonstrate this by showing the RMS residuals, putting it in as a figure, and putting this paragraph back to me?
 I am imagining a figure that shows a scatter plot of T vs RMS residual, colored by metallicity.
@@
 
-\color{gcolor}{HOGG: Adam's comment: if $\chi^2/n>>1$ then the model is NOT an accurate representation of the data.}\color{black} \\
+% \color{gcolor}{HOGG: Adam's comment: if $\chi^2/n>>1$ then the model is NOT an accurate representation of the data.}\color{black} \\
 
@@
 
-Figure \ref{fig:isochrone_cmd}
+Figure \ref{fig:isochrone_cmd} \\
 
@@
 It is our hope that \thecannon, and models like it, will, in the end, lead to
-new and improved physical models, which will, in turn, put \thecannon\ out of business.
+new and improved physical models, which will, in turn, put \thecannon\ out of business. \\
 

### 2019-06-11 paper/draft.tex
more figure moving
format fixes

@@
 \newcommand{\gaia}{\textsl{Gaia}}
+\newcommand{\drtwo}{\textsl{Gaia-DR2}}
 \newcommand{\zmass}{\textsl{2MASS}}
@@
 
-Need to talk about \gaia\ data \& what we use this dataset for... \\
+We cross-matched the {\sl APOGEE-DR14} catalog of  \citep{Brown:2018} \\
 
@@
 	\includegraphics[width=\linewidth]{figures/cannon_chi_dist.png}
-	\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee --\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
+	\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources \citep{ElBadry:2018,Skinner:2018}; \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee --\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
 We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as ``safe''.}
@@
 
-Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \textsl{Gaia DR2} (\citealt{Brown:2018}) and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources.
+Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources.
 From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.
@@
 
+\begin{figure*}[hp!]
+	\includegraphics[width=\linewidth]{figures/isochrone_cmd.png}
+	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
+The four panels show five metallicity tracks (ranging $-0.5<$[M/H]$<0.5$) over varying ages (1, 3, 5, and 7 Gyr).}
+	\label{fig:isochrone_cmd}
+\end{figure*}
+
 The three panels of \ref{fig:safe_selection} show before and after selection of the sources in \gaia\ color-magnitude space, colored by \thecannon\ determined temperatures, metallicities, and spectral types, with their respective training samples overplotted in orange.
@@
 	\label{fig:aspcap_cannon_label_hist} 
-\end{figure*}
-
-\begin{figure*}[h]
-	\includegraphics[width=\linewidth]{figures/isochrone_cmd.png}
-	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
-The four panels show five metallicity tracks (ranging $-0.5<$[M/H]$<0.5$) over varying ages (1, 3, 5, and 7 Gyr).}
-	\label{fig:isochrone_cmd}
 \end{figure*}

### 2019-06-11 paper/draft.tex
discussion notes

@@
 \newcommand{\cannon}{\textsl{Cannon}}
-\newcommand{\aspcap}{\textsl{ASPCAP}}
 \newcommand{\gaia}{\textsl{Gaia}}
@@
 
-The \aspcap\ pipeline releases several types of data files, with various levels of processing: {\tt\string ap1D} (the raw one-dimensional spectra for individual visits), {\tt\string apVisit} (the individual visit spectra with telluric subtraction), {\tt\string apStar} (the co-added {\tt\string apVisit} spectra), and most processed {\tt\string aspcapStar} spectra.
+The ASPCAP pipeline releases several types of data files, with various levels of processing: {\tt\string ap1D} (the raw one-dimensional spectra for individual visits), {\tt\string apVisit} (the individual visit spectra with telluric subtraction), {\tt\string apStar} (the co-added {\tt\string apVisit} spectra), and most processed {\tt\string aspcapStar} spectra.
 For this work we use {\tt\string aspcapStar} files, which contains the pseudo-continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
@@
 	\includegraphics[width=\linewidth]{figures/aspcap_cannon_validation.png}
-	\caption{\cannon\ (right column) and \aspcap\ (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 6,257 sources with $V,J,K$ colors.
-Shown are \cannon\ and \aspcap\ temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and $\feh$ dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation $V-K$ and $\feh$ (bottom).}
+	\caption{\cannon\ (right column) and ASPCAP (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 6,257 sources with $V,J,K$ colors.
+Shown are \cannon\ and ASPCAP temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and $\feh$ dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation $V-K$ and $\feh$ (bottom).}
 	\label{fig:teff_comparisons}
@@
 	\includegraphics[width=\linewidth]{figures/aspcap_cannon_label_hist.png}
-	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the \aspcap\ pipeline.
-Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars \aspcap\ predicts temperatures with slightly higher temperatures and metallicities.}
+	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the ASPCAP pipeline.
+Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars ASPCAP predicts temperatures with slightly higher temperatures and metallicities.}
 	\label{fig:aspcap_cannon_label_hist} 
@@
 
-As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures.
-To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (\textsl{APASS}; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by \textsl{APASS}, and $JHK$ photometries from \apogee.
+As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures).
+To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
 Figure \ref{fig:teff_comparisons} shows \cannon\ vs.
-photometric temperatures on the right, and \aspcap\ vs.
+photometric temperatures on the right, and ASPCAP vs.
 photometric temperatures on the left for each of the 6,257 sources, colored by their respective spectroscopic metallicities.
@@
 
-We find \aspcap\ overestimates $\teff$ on average by $130-170$K, while \thecannon\ underestimates $\teff$ by $\sim25$K.
-We also see the scatter is much higher for the \aspcap\ temperatures ($\sim230$K) compared to the scatter of \cannon\ temperatures ($\sim85$K).
+Compared to the \citealt{Casagrande:2008}, \citealt{Boyajian:2012}, and \citealt{Mann:2015} derived temperatures, we find ASPCAP overestimates $\teff$ on average by a bias of $130-170$K, while \thecannon\ underestimates $\teff$ by a bias of $\sim25$K.
+We also see the scatter is much higher for the ASPCAP temperatures ($\sim230$K) compared to the scatter of \cannon\ temperatures ($\sim85$K).
 
-While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average \aspcap\ temperatures are much hotter, and scatters are much higher skewed by 500 sources which \aspcap\ predicts to have temperatures over 4200K.
+While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average ASPCAP temperatures are much hotter, and scatters are much higher skewed by 500 sources which ASPCAP predicts to have temperatures over 4200K.
 
@@
 
-Comment on the importance of each property of training sample: \\
-accuracy - some discussion on ground truth of stellar labels... \\
-precision - some discussion on how consistently training parameters are derived... \\
-size -  \\
-dynamic range - training sample must span the desired parameter range of the test sample. \thecannon\ does not extrapolate well. \\
-representation - model cannot account for physics not represented by training sample (i.e. this training sample is limited to old, slowly rotating, not magnetically active, singular sources...) \\
-
 Talk about the particular limitations we have found in this training sample.
@@
 \\
-
-Talk future work/implications with other stellar surveys (e.g. LAMOST), and exoplanet surveys (e.g. TESS)
 
@@
 respect to the atmospheric parameters.
-We show some of these derivatives in [BIRKY WHAT FIGURE] and deliver relevant data in
-Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}
+We show some of these derivatives in Figures \ref{fig:demo_teff}, \ref{fig:demo_feh}, and \ref{fig:west_derivative} and deliver relevant data in
+Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}.
 These Tables summarize spectral features in the \apogee\ bandpass
@@
 new and improved physical models, which will, in turn, put \thecannon\ out of business. \\
+
+Talk future work/implications with other stellar surveys (e.g. LAMOST), and exoplanet surveys (e.g. TESS)
 

### 2019-06-11 paper/draft.tex
condense size of isochrone plots

@@
 
-\begin{figure*}[h]
-	\plottwo{figures/cmd_teff_full.png}{figures/cmd_teff_safe.png}
-	\plottwo{figures/cmd_feh_full.png}{figures/cmd_feh_safe.png}
-	\plottwo{figures/cmd_spt_full.png}{figures/cmd_spt_safe.png}
-	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, binaries, and K dwarfs).
-Overplotted with orange triangles are the M15 and W11 training samples, for their respective \cannon\ test labels.
-Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly up from the main sequence branch as expected.
-Deviations from these gradients seen at the upper slice of the main sequence is likely remaining contamination from the binary sequence.}
-	\label{fig:safe_selection}
-\end{figure*}
-
-\begin{figure*}[hp!]
-	\includegraphics[width=\linewidth]{figures/isochrone_cmd.png}
-	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
-The four panels show five metallicity tracks (ranging $-0.5<$[M/H]$<0.5$) over varying ages (1, 3, 5, and 7 Gyr).}
-	\label{fig:isochrone_cmd}
-\end{figure*}
-
 The three panels of \ref{fig:safe_selection} show before and after selection of the sources in \gaia\ color-magnitude space, colored by \thecannon\ determined temperatures, metallicities, and spectral types, with their respective training samples overplotted in orange.
@@
 \subsection{Color-Temperature Comparisons}
-
-\begin{figure*}[h]
-	\includegraphics[width=\linewidth]{figures/aspcap_cannon_validation.png}
-	\caption{\cannon\ (right column) and ASPCAP (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 6,257 sources with $V,J,K$ colors.
-Shown are \cannon\ and ASPCAP temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and $\feh$ dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation $V-K$ and $\feh$ (bottom).}
-	\label{fig:teff_comparisons}
-\end{figure*}
-
-\begin{figure*}[h]
-	\includegraphics[width=\linewidth]{figures/aspcap_cannon_label_hist.png}
-	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the ASPCAP pipeline.
-Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars ASPCAP predicts temperatures with slightly higher temperatures and metallicities.}
-	\label{fig:aspcap_cannon_label_hist} 
-\end{figure*}
 
@@
 
+% \begin{figure*}[hp!]
+% 	\includegraphics[width=\linewidth]{figures/isochrone_cmd.png}
+% 	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
+The four panels show five metallicity tracks (ranging $-0.5<$[M/H]$<0.5$) over varying ages (1, 3, 5, and 7 Gyr).}
+% 	\label{fig:isochrone_cmd}
+% \end{figure*}
+
+\begin{figure}
+	\includegraphics[width=\linewidth]{figures/cannon_vs_parsec.png}
+	\includegraphics[width=\linewidth]{figures/aspcap_vs_parsec1.png}
+	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
+The four panels show five metallicity tracks (ranging $-0.5<$[M/H]$<0.5$) over varying ages (3 and 7 Gyr).}
+	\label{fig:isochrone_cmd}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=\linewidth]{figures/aspcap_vs_parsec2.png}
+	\caption{}
+\end{figure}
+
+\begin{figure*}
+	\plotone{figures/aspcap_cannon_label_hist.png}
+	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the ASPCAP pipeline.
+Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars ASPCAP predicts temperatures with slightly higher temperatures and metallicities.}
+	\label{fig:aspcap_cannon_label_hist} 
+\end{figure*}
+
 Figure \ref{fig:isochrone_cmd} \\
@@
 % JB DO WE WANT A FIGURE THAT SHOWS ABUNDANCES AS A FUNCTION OF MILKY WAY POSITION OR ACTIONS?
+
+% ----------------------------------------
+
+\begin{figure*}[h]
+	\plottwo{figures/cmd_teff_full.png}{figures/cmd_teff_safe.png}
+	\plottwo{figures/cmd_feh_full.png}{figures/cmd_feh_safe.png}
+	\plottwo{figures/cmd_spt_full.png}{figures/cmd_spt_safe.png}
+	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, binaries, and K dwarfs).
+Overplotted with orange triangles are the M15 and W11 training samples, for their respective \cannon\ test labels.
+Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly up from the main sequence branch as expected.
+Deviations from these gradients seen at the upper slice of the main sequence is likely remaining contamination from the binary sequence.}
+	\label{fig:safe_selection}
+\end{figure*}
+
+\begin{figure*}[h]
+	\includegraphics[width=\linewidth]{figures/aspcap_cannon_validation.png}
+	\caption{\cannon\ (right column) and ASPCAP (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 6,257 sources with $V,J,K$ colors.
+Shown are \cannon\ and ASPCAP temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and $\feh$ dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation $V-K$ and $\feh$ (bottom).}
+	\label{fig:teff_comparisons}
+\end{figure*}
 

### 2019-06-11 paper/draft.tex
eliminate redundancy in intro

@@
 \sdssiv\ \apogee. One set of labels are physical parameters (effective temperatures and metallicities)
-and the other set of labels are spectral types. In each case, \thecannon\ produces a predictive model for \apogee\ spectra, but without
-any input of stellar physics beyond whatever went in to producing the labels in the first
-place. That is, the model to be applied to the \apogee\ data is learned from the \apogee\ data alone.
-This obviates physical models that are often inaccurate for M dwarfs, and capitalizes on the excellent data quality
-delivered by the \apogee\ spectrograph and project.
+and the other set of labels are spectral types.
+
+% In each case, \thecannon\ produces a predictive model for \apogee\ spectra, but without any input of stellar physics beyond whatever went in to producing the labels in the first place.
+That is, the model to be applied to the \apogee\ data is learned from the \apogee\ data alone.
+This obviates physical models that are often inaccurate for M dwarfs, and capitalizes on the excellent data quality delivered by the \apogee\ spectrograph and project.
+%JB: I think this part is redundant
 
@@
 The \apogee\ survey is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (\aspcap; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
+Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
 For low temperatures ($2800<\teff<3500$\,K), the pipeline uses \textsl{MARCS} plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures ($\teff\geq3500$\,K) \textsl{ATLAS9} \citep{Castelli:2004} plane-parallel models are used.

### 2019-06-19 paper/draft.tex
add model ref

@@
 The \apogee\ survey is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
-For low temperatures ($2800<\teff<3500$\,K), the pipeline uses \textsl{MARCS} plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures ($\teff\geq3500$\,K) \textsl{ATLAS9} \citep{Castelli:2004} plane-parallel models are used.
+Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
+For low temperatures ($2800<\teff<3500$\,K), the pipeline uses MARCS plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures ($\teff\geq3500$\,K) ATLAS9 \citep{Castelli:2004} plane-parallel models are used.
 
@@
 \citealt{Souto:2017} and \citealt{Souto:2018} which have modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining $\teff$/$\logg$/metallicity + 13 elemental abundances,
-\citealt{Rajpurohit:2018} which tested BTSettl model grids \citep{Allard:2012} on 45 M dwarfs to estimate $\teff$/$\logg$/metallicity,
+\citealt{Rajpurohit:2018} which tested BT-Settl \citep{Allard:2012} and MARCS \citep{Gustafsson:2008} model grids on 45 M dwarfs to estimate $\teff$/$\logg$/metallicity,
 and \citealt{Skinner:2018} which has identified and measured mass ratios and radial velocities for 44 M dwarf spectroscopic binaries.

### 2019-06-19 paper/draft.tex
topcat ref
add aitoff proj
comment out figure
caption edits

@@
 % ----------------------------------------
-\subsection{Spectral Type Model \label{subsec:west_results}}
 
@@
 \end{figure}
+
+\subsection{Spectral Type Model \label{subsec:west_results}}
+
+We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
+We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
+As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates \thecannon\ does a poor job at extrapolating to labels outside of the training sample space.
+Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows the self and LOOCV tests for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.
+
+% ----------------------------------------
+
+\begin{figure*}
+	\includegraphics[width=\linewidth]{figures/aitoff.png}
+	\caption{Aitoff projection of 14,827 sources.}
+	\label{fig:aitoff}
+\end{figure*}
 
@@
 \end{figure*}
-
-We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
-We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
-As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates \thecannon\ does a poor job at extrapolating to labels outside of the training sample space.
-Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows the self and LOOCV tests for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.
-
-% ----------------------------------------
 
@@
 	\includegraphics[width=\linewidth]{figures/aspcap_vs_parsec1.png}
-	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
-The four panels show five metallicity tracks (ranging $-0.5<$[M/H]$<0.5$) over varying ages (3 and 7 Gyr).}
+	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citep{Bressan:2012} in \gaia\ color-magnitude space.
+The two panels show five metallicity tracks (ranging $-0.5<$[M/H]$<+0.5$ in steps of 0.1 dex) at two different ages (3 and 7 Gyr).
+In the ASPCAP plot only the 8,960 out of 10,311 sources with metallcities $-0.5<$[M/H]$<+0.5$ are shown, while the remaining 1,351 sources with [M/H]$<-0.5$ and [M/H]$>+0.5$ are shown separately in Figure \ref{fig:isochrone_cmd_2}.}
 	\label{fig:isochrone_cmd}
@@
 	\caption{}
+	\label{fig:isochrone_cmd_2}
 \end{figure}
@@
 	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the ASPCAP pipeline.
-Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars ASPCAP predicts temperatures with slightly higher temperatures and metallicities.}
+Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars ASPCAP predictions are skewed higher in temperature and lower in metallicity compared to \cannon\ measurements.}
 	\label{fig:aspcap_cannon_label_hist} 
@@
 	\code{scipy} \citep{Jones:2001}, \,
+	\code{Topcat} \citep{Taylor2005}, \,
 	\code{The Cannon} \citep{Ness:2015}}
@@
 
-\begin{figure*}[h]
-	\plotone{figures/mann_radii.png}
-	\caption{Comparison of radii estimated from M15 relations: on the x-axis radii are derived from  K magnitudes and \cannon\ $\feh$, on the y-axis radii are derived from \cannon\ $\teff$ and $\feh$.}
-	\label{fig:radii}
-\end{figure*}
+% \begin{figure*}[h]
+% 	\plotone{figures/mann_radii.png}
+% 	\caption{Comparison of radii estimated from M15 relations: on the x-axis radii are derived from  K magnitudes and \cannon\ $\feh$, on the y-axis radii are derived from \cannon\ $\teff$ and $\feh$.}
+% 	\label{fig:radii}
+% \end{figure*}
 

### 2019-07-13 paper/draft.tex
ref catalog table
notes to self
more discussion

@@
 
-% \color{gcolor}{BIRKY: Can you demonstrate this by showing the RMS residuals, putting it in as a figure, and putting this paragraph back to me?
-I am imagining a figure that shows a scatter plot of T vs RMS residual, colored by metallicity.
-Or something like that.}\color{black} \\
-
-% \color{gcolor}{HOGG: Adam's comment: if $\chi^2/n>>1$ then the model is NOT an accurate representation of the data.}\color{black} \\
-
-% ----------------------------------------
-
 \begin{figure}
@@
 
-Table \ref{table:test_results} presents...
+Table \ref{table:test_results} outlines the parameters included in the test sample catalog, which can be downloaded from the online journal.
 
@@
 
-Talk about the particular limitations we have found in this training sample.
-Talk about model improvements: incorporate broadening \citep{Behmard2019}; spectroscopic binary model \citep{ElBadry:2018}...
-\\
+The model we have developed does have limitations however.
+For example, it delivers chi-squared values---goodness-of-fit measures---that
+are large; the model is not technically an accurate description of the spectra,
+especially when the spectra are observed at signal-to-noise levels above 100.
+For another, it does not include some known physical and instrumental effects.
+It does not include a term or terms for line broadening from rotation or
+convection (\eg, \citealt{Behmard2019}).
+It does not include any accounting for binarity or superposition of multiple
+stellar spectra (as in, say, \citealt{ElBadry:2010}).
+And it does not include any adjustments for instrumental variations, like for
+example the small but significant variations of \apogee\ resolution with
+spectrograph fiber number (as are included in \citealt{Ness2018}).
+
+In addition to all this, the \apogee\ instrument was designed to be sensitive to more than
+a dozen individual element abundances in stellar spectra!
+So the M-dwarf spectra analyzed here contain individual element abundance
+information that we have left on the table.
+Exploitation of that information might require better models of M dwarfs than
+we have at present.
+However, this is an important goal for the future with these data.
 
@@
 
-Talk future work/implications with other stellar surveys (e.g. LAMOST), and exoplanet surveys (e.g. TESS)
+HOGG: Talk future work/implications with other stellar surveys (e.g. LAMOST), and exoplanet surveys (e.g. TESS)
 

### 2019-07-13 paper/draft.tex

@@
 It does not include a term or terms for line broadening from rotation or
-convection (\eg, \citealt{Behmard2019}).
+convection (for exmaple, \citealt{Behmard2019}).
 It does not include any accounting for binarity or superposition of multiple
-stellar spectra (as in, say, \citealt{ElBadry:2010}).
+stellar spectra (as in, say, \citealt{ElBadry:2018}).
 And it does not include any adjustments for instrumental variations, like for
@@
 
-HOGG: Talk future work/implications with other stellar surveys (e.g. LAMOST), and exoplanet surveys (e.g. TESS)
-
 %%==========================================================================================
 \acknowledgements
-We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Diogo Souto (Observatário Nacional/MCTI) and Keivan Stassun (Vanderbuilt) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
+We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Melissa Ness (Columbia), Diogo Souto (Observatário Nacional/MCTI) and Keivan Stassun (Vanderbuilt) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
 This project was partially supported by

### 2019-07-19 paper/draft.tex
remove figs
mark places needing revision

@@
 abundances) to high precision.
-Using a training sample of 87 M dwarfs in the temperature range $2860
-< \teff < 4130$\,K, and metallicty range $-0.5 < \feh < 0.5$\,dex
-[WITH LABELS FROM WHERE??], we train a two-parameter model with
-predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex
-respectively.
+Using a training sample of 87 M dwarfs with optically derived labels spanning 
+$2860 < \teff < 4130$\,K callibrated with bolometric temperatures, 
+$-0.5 < \feh < 0.5$\,dex callibrated with FGK wide binary metallicities, 
+we train a two-parameter model with predictive accuracy (in cross-validation) 
+to 77\,K and 0.09\,dex respectively.
 We also train a one-dimensional spectral type model using 51 M dwarfs
-with known [FROM WHERE??] spectral types ranging M0 to M9, to predictive accuracy of
+with known SDSS optical spectral types ranging M0 to M9, to predictive accuracy of
 0.9 types.
-From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we
+\color{gcolor}{From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we
 compare our \cannon\ derived $\teff$ against color--temperature
@@
 This project has implications for the NASA \tess\ Mission and other
-surveys that depend on having good parameters for M dwarfs.
+surveys that depend on having good parameters for M dwarfs.}\color{black}
 \end{abstract}
@@
 Section \ref{sec:methods} describes our model implementation using \thecannon\ framework, and Section \ref{sec:sample_selection} describes the our sample selection and derivation of training parameters.
-In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$10,000 sources, and examine the validity of our parameters against color-temperature relations and isochrone grids.
-Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results in the context of present and upcoming stellar and exoplanetary surveys. \\
+\color{gcolor}{{In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$10,000 sources, and examine the validity of our parameters against color-temperature relations and isochrone grids.
+Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results in the context of present and upcoming stellar and exoplanetary surveys.}\color{black} \\
 
@@
 In previous work it has been recommended to use an alternative pseudo-continuum normalization \citep{Ness:2015}, but we don't find obvious issues with the normalization in our analysis, so we retain the survey pipeline outputs.
-
-% ----------------------------------------
-\subsection{\gaia\ Survey}
-
-We cross-matched the {\sl APOGEE-DR14} catalog of  \citep{Brown:2018} \\
 
@@
 sense; it is only good at the percent level.
-This percent-level accuracy of the model in the data domain is also demonstrated... \\
+\color{gcolor}{This percent-level accuracy of the model in the data domain is also demonstrated...}\color{black} \\
 
@@
 % ----------------------------------------
-
-\begin{figure*}
-	\includegraphics[width=\linewidth]{figures/aitoff.png}
-	\caption{Aitoff projection of 14,827 sources.}
-	\label{fig:aitoff}
-\end{figure*}
 
@@
 The two panels show five metallicity tracks (ranging $-0.5<$[M/H]$<+0.5$ in steps of 0.1 dex) at two different ages (3 and 7 Gyr).
-In the ASPCAP plot only the 8,960 out of 10,311 sources with metallcities $-0.5<$[M/H]$<+0.5$ are shown, while the remaining 1,351 sources with [M/H]$<-0.5$ and [M/H]$>+0.5$ are shown separately in Figure \ref{fig:isochrone_cmd_2}.}
+In the ASPCAP plot only the 8,960 out of 10,311 sources with metallcities $-0.5<$[M/H]$<+0.5$ are shown, while the remaining 1,351 sources with [M/H]$<-0.5$ and [M/H]$>+0.5$.}
 	\label{fig:isochrone_cmd}
@@
 
-\begin{figure}
-	\includegraphics[width=\linewidth]{figures/aspcap_vs_parsec2.png}
-	\caption{}
-	\label{fig:isochrone_cmd_2}
-\end{figure}
+% \begin{figure}
+% 	\includegraphics[width=\linewidth]{figures/aspcap_vs_parsec2.png}
+% 	\caption{}
+% 	\label{fig:isochrone_cmd_2}
+% \end{figure}
 

### 2019-07-19 paper/draft.tex
add ack

@@
 \acknowledgements
-We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Melissa Ness (Columbia), Diogo Souto (Observatário Nacional/MCTI) and Keivan Stassun (Vanderbuilt) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
+We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Melissa Ness (Columbia), Kevin Covey (WWU), Diogo Souto (Observatário Nacional/MCTI) and Keivan Stassun (Vanderbuilt) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
 This project was partially supported by

### 2019-07-19 paper/draft.tex
take out some figs

@@
 Additionally we notice a slight deviation at the upper slice of the main sequence where \cannon\ temperatures, metallicities and spectral types all appear lower than the gradient trends in the rest of the branch, which is most likely the binary sequence.
-Comparisons between the West-trained spectral types and Mann-trained temperatures and metallicities are shown in Figure \ref{fig:west_vs_mann}: left shows the slight negative correlation between spectral type and $\feh$, and right shows the nonlinear decrease in temperature with visible metallicity gradient.
+% Comparisons between the West-trained spectral types and Mann-trained temperatures and metallicities are shown in Figure \ref{fig:west_vs_mann}: left shows the slight negative correlation between spectral type and $\feh$, and right shows the nonlinear decrease in temperature with visible metallicity gradient.
 We also note that applying our model requires very little computational demand: the time to train and test a model on all 14,827 sources was two minutes on a personal laptop.
@@
 
-% \begin{figure*}[hp!]
-% 	\includegraphics[width=\linewidth]{figures/isochrone_cmd.png}
-% 	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citealt{Bressan:2012} in \gaia\ color-magnitude space.
-The four panels show five metallicity tracks (ranging $-0.5<$[M/H]$<0.5$) over varying ages (1, 3, 5, and 7 Gyr).}
-% 	\label{fig:isochrone_cmd}
-% \end{figure*}
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{figures/cannon_vs_parsec.png}
-	\includegraphics[width=\linewidth]{figures/aspcap_vs_parsec1.png}
-	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citep{Bressan:2012} in \gaia\ color-magnitude space.
+% \begin{figure}
+% 	\includegraphics[width=\linewidth]{figures/cannon_vs_parsec.png}
+% 	\includegraphics[width=\linewidth]{figures/aspcap_vs_parsec1.png}
+% 	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citep{Bressan:2012} in \gaia\ color-magnitude space.
 The two panels show five metallicity tracks (ranging $-0.5<$[M/H]$<+0.5$ in steps of 0.1 dex) at two different ages (3 and 7 Gyr).
 In the ASPCAP plot only the 8,960 out of 10,311 sources with metallcities $-0.5<$[M/H]$<+0.5$ are shown, while the remaining 1,351 sources with [M/H]$<-0.5$ and [M/H]$>+0.5$.}
-	\label{fig:isochrone_cmd}
-\end{figure}
-
-% \begin{figure}
-% 	\includegraphics[width=\linewidth]{figures/aspcap_vs_parsec2.png}
-% 	\caption{}
-% 	\label{fig:isochrone_cmd_2}
+% 	\label{fig:isochrone_cmd}
 % \end{figure}
 
-\begin{figure*}
-	\plotone{figures/aspcap_cannon_label_hist.png}
-	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the ASPCAP pipeline.
+% \begin{figure*}
+% 	\plotone{figures/aspcap_cannon_label_hist.png}
+% 	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the ASPCAP pipeline.
 Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars ASPCAP predictions are skewed higher in temperature and lower in metallicity compared to \cannon\ measurements.}
-	\label{fig:aspcap_cannon_label_hist} 
-\end{figure*}
-
-Figure \ref{fig:isochrone_cmd} \\
-
-% JB DO WE WANT A FIGURE THAT SHOWS ABUNDANCES AS A FUNCTION OF MILKY WAY POSITION OR ACTIONS?
+% 	\label{fig:aspcap_cannon_label_hist} 
+% \end{figure*}
 
@@
 
-\begin{figure*}[h]
-	\plotone{figures/spt_vs_teff_vs_feh.png}
-	\caption{West-trained spectral types vs.
+% \begin{figure*}[h]
+% 	\plotone{figures/spt_vs_teff_vs_feh.png}
+% 	\caption{West-trained spectral types vs.
 Mann-trained metallcities and temperatures.
@@
 \label{fig:west_vs_mann}}
-\end{figure*}
+% \end{figure*}
 
@@
 
-% \newpage
+\newpage
 
@@
 
+\newpage
+
 \begin{deluxetable*}{rcl}

### 2019-07-19 paper/draft.tex
summarize problem at start of abstract

@@
 \begin{abstract}
-We present a catalog of spectroscopic temperatures, metallicities,
+
+M dwarfs have enourmous potential for understanding structure and
+formation on both Galactic and exoplanetary scales
+with their properties and compositions.
+However, insufficient physical assumptions behind atmospheric models
+have limited abilities at reproducing spectral features in stars of the 
+coolest temperature regimes ($T_{\rm eff} < 4200\,$K) and further 
+exploiting the information content and statistical extent of
+current and upcoming large-scale spectroscopic surveys.
+Here we present a catalog of spectroscopic temperatures, metallicities,
 spectral types for 10,311 M dwarfs in the \apogee\ survey using
-\thecannon, which is a flexible, data-driven spectral-modeling and
+\thecannon: a flexible, data-driven spectral-modeling and
 parameter-inference framework demonstrated to estimate
@@
 Finally we use our results to
-estimate stellar radii from absolute magnitudes and \cannon\ temperatures [AND WE FIND WHAT??].
-This project has implications for the NASA \tess\ Mission and other
-surveys that depend on having good parameters for M dwarfs.}\color{black}
+estimate stellar radii from absolute magnitudes and \cannon\ temperatures [AND WE FIND WHAT??].}\color{black}
 \end{abstract}

### 2019-07-23 paper/draft.tex
explain selections

@@
 have limited abilities at reproducing spectral features in stars of the 
-coolest temperature regimes ($T_{\rm eff} < 4200\,$K) and further 
+coolest temperature regimes ($T_{\rm eff} < 4200\,$K) and fully
 exploiting the information content and statistical extent of
@@
 \color{gcolor}{From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we
-compare our \cannon\ derived $\teff$ against color--temperature
+apply our models and compare our \cannon\ derived $\teff$ against color--temperature
 relations and \cannon\ derived $\feh$ against isochrone models in
@@
 	\includegraphics[width=\linewidth]{figures/cmd_selection.png}
-	\caption{\gaia\ and \zmass\ color-magnitude cuts for 12,037 sources with $\chi^2<100,000$.
-Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia.}
+	\caption{\gaia\ and \zmass\ color-magnitude cuts for the12,037 sources with $\chi^2<100,000$.
+Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia.
+The coordinates for the selected quadrangles
+	$\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ corresponding to ($BP-RP$, $M_G$) for the \gaia\ color-magnitudes shown left, and 
+	$\left\{ (.7, 4.3), (1.1, 4.3), (.7, 9), (1.1, 9) \right\}$ corresponding to ($J-K$, $M_K$) for the \zmass\ color-magnitudes shown right. }
 	\label{fig:cmd_selection}
@@
 \begin{enumerate}
-\item \textbf{Known contaminant cut:} We remove known pre-main sequence sources from \citealt{Cottaar:2014} (280 sources), and known binaries from \citealt{ElBadry:2018} and \citealt{Skinner:2018} (58 sources).
+\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000.
 
-\item \textbf{K dwarf cut:} In \zmass\ photometry we select $M_K > 4.3$ to remove likely K dwarfs, which appear reddened in \gaia\ $BP$, $RP$ photometry.
+\item \textbf{Color-magnitude cuts:} Using \gaia\ and \zmass\ photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to cut sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and multiple), and subdwarfs below the main sequence.
 
-\item \textbf{Pre-main sequence, multiple system and subdwarf cut:} in \gaia\ and \zmass\ we apply $.7<J-K<1.1$ and points in the selction region within the points $\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ of ($BP-RP$, $M_G$) to cut remaining likely pre-main sequence, multiple sources above the main sequence and subdwarfs below the main sequence.
+\item \textbf{Model extraploation cuts:} Because extrapolation with \thecannon\ is unreliable, we select only sources with labels inside the range of the training sample: for the \cannon\ reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$, and for \cannon\ reported spectral subtypes we impose $0<SPT<9$.
 
-\item \textbf{Quality of fit cut:} We apply a $\chi^2$ cut of less than 100,000.
+\item \textbf{Astrometric cut:} \gaia\ $RUWE<1.2$
 
-\item \textbf{Training bound selection:} Because extrapolation with \thecannon\ is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
 \end{enumerate}

### 2019-07-24 paper/draft.tex
redo test sample figures
fix figure size issue

@@
 
+\item \textbf{Binary cut:}
+
 \end{enumerate}
@@
 
-As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures).
+\color{gcolor}{As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures).
 To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
@@
 
-While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average ASPCAP temperatures are much hotter, and scatters are much higher skewed by 500 sources which ASPCAP predicts to have temperatures over 4200K.
+While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average ASPCAP temperatures are much hotter, and scatters are much higher skewed by 500 sources which ASPCAP predicts to have temperatures over 4200K.}\color{black}
 
@@
 \begin{figure*}[h]
-	\plottwo{figures/cmd_teff_full.png}{figures/cmd_teff_safe.png}
-	\plottwo{figures/cmd_feh_full.png}{figures/cmd_feh_safe.png}
-	\plottwo{figures/cmd_spt_full.png}{figures/cmd_spt_safe.png}
-	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, binaries, and K dwarfs).
+	\includegraphics[width=.45\linewidth]{figures/cmd_teff_full.png}
+	\includegraphics[width=.49\linewidth]{figures/cmd_teff_safe.png} \\
+	\includegraphics[width=.45\linewidth]{figures/cmd_feh_full.png}
+	\includegraphics[width=.49\linewidth]{figures/cmd_feh_safe.png} \\
+	\includegraphics[width=.45\linewidth]{figures/cmd_spt_full.png}
+	\includegraphics[width=.49\linewidth]{figures/cmd_spt_safe.png}
+	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, spectroscopic binaries, and K dwarfs).
 Overplotted with orange triangles are the M15 and W11 training samples, for their respective \cannon\ test labels.

### 2019-07-24 paper/draft.tex
change final # sources

@@
 Here we present a catalog of spectroscopic temperatures, metallicities,
-spectral types for 10,311 M dwarfs in the \apogee\ survey using
+spectral types for 5,875 M dwarfs in the \apogee\ survey using
 \thecannon: a flexible, data-driven spectral-modeling and
@@
 From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.
-Although our catalog reports the temperatures, metallicities, and spectral types for 10,311 sources out of 14,827 with $\chi^2<100,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
+Although our catalog reports the temperatures, metallicities, and spectral types for 5,875 sources out of 14,827 with $\chi^2<100,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
 
@@
 \color{gcolor}{As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures).
-To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
+To obtain visual band magnitudes for a set of sources, we cross-matched the 5,875 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
 Figure \ref{fig:teff_comparisons} shows \cannon\ vs.
@@
 The two panels show five metallicity tracks (ranging $-0.5<$[M/H]$<+0.5$ in steps of 0.1 dex) at two different ages (3 and 7 Gyr).
-In the ASPCAP plot only the 8,960 out of 10,311 sources with metallcities $-0.5<$[M/H]$<+0.5$ are shown, while the remaining 1,351 sources with [M/H]$<-0.5$ and [M/H]$>+0.5$.}
+In the ASPCAP plot only the 8,960 out of 5,875 sources with metallcities $-0.5<$[M/H]$<+0.5$ are shown, while the remaining 1,351 sources with [M/H]$<-0.5$ and [M/H]$>+0.5$.}
 % 	\label{fig:isochrone_cmd}
@@
 % 	\plotone{figures/aspcap_cannon_label_hist.png}
-% 	\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the ASPCAP pipeline.
-Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 10,311 stars ASPCAP predictions are skewed higher in temperature and lower in metallicity compared to \cannon\ measurements.}
+% 	\caption{Distribution of temperatures and metallicities for 5,875 sources from \thecannon\ compared to the ASPCAP pipeline.
+Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 5,875 stars ASPCAP predictions are skewed higher in temperature and lower in metallicity compared to \cannon\ measurements.}
 % 	\label{fig:aspcap_cannon_label_hist} 
@@
 	\includegraphics[width=.49\linewidth]{figures/cmd_spt_safe.png}
-	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 10,311 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, spectroscopic binaries, and K dwarfs).
+	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 5,875 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, spectroscopic binaries, and K dwarfs).
 Overplotted with orange triangles are the M15 and W11 training samples, for their respective \cannon\ test labels.
@@
 The primary result of this work is that
-we have compiled a catalog of 10,311 M dwarfs with \cannon\ temperatures,
+we have compiled a catalog of 5,875 M dwarfs with \cannon\ temperatures,
 metallicities, spectral types, and six-dimensional kinematics.

### 2019-07-24 paper/draft.tex
add ruwe cut figures
add met comparison figs

@@
 
+\begin{figure*}
+\begin{center}
+	\includegraphics[width=.48\linewidth]{figures/ruwe_cut_before.png}
+	\includegraphics[width=.48\linewidth]{figures/ruwe_cut_after.png}
+	\caption{}
+	\label{fig:ruwe_cut}
+\end{center}
+\end{figure*}
+
 \subsection{Test Sample \label{subsec:test_selection}}
@@
 
-\item \textbf{Astrometric cut:} \gaia\ $RUWE<1.2$
+\item \textbf{Astrometric cut:} \gaia\ re-normalised unit weight error (RUWE) $RUWE<1.2$
 
@@
 \begin{figure*}[h]
+\begin{center}
 	\includegraphics[width=.45\linewidth]{figures/cmd_teff_full.png}
@@
 	\label{fig:safe_selection}
+\end{center}
 \end{figure*}
@@
 	\label{fig:teff_comparisons}
+\end{figure*}
+
+\begin{figure*}[h]
+\begin{center}
+	\includegraphics[width=5cm]{figures/met_comp1.png}
+	\includegraphics[width=5cm]{figures/met_comp2.png}
+	\includegraphics[width=5cm]{figures/met_comp3.png} 
+	\includegraphics[width=5cm]{figures/met_comp4.png}
+	\includegraphics[width=5cm]{figures/met_comp5.png}
+	\includegraphics[width=5cm]{figures/met_comp6.png} 
+	\includegraphics[width=5cm]{figures/met_comp7.png}
+	\includegraphics[width=5cm]{figures/met_comp8.png}
+	\includegraphics[width=5cm]{figures/met_comp9.png} 
+	\caption{A comparison of ASPCAP and Cannon metallicities for 8,335 test sample sources separated into temperature bins of 100K.
+We see the scatter and metal-poor bias of ASPCAP metallicities clearly increases at decreasing temperatures.}
+\end{center}
 \end{figure*}

### 2019-07-24 paper/draft.tex
fix latex spacing complaints
more fig size tweaks

@@
 
-\begin{figure*}[h]
+\begin{figure*}[]
 \begin{center}
 	\includegraphics[width=.45\linewidth]{figures/cmd_teff_full.png}
-	\includegraphics[width=.49\linewidth]{figures/cmd_teff_safe.png} \\
+	\includegraphics[width=.48\linewidth]{figures/cmd_teff_safe.png} \\
 	\includegraphics[width=.45\linewidth]{figures/cmd_feh_full.png}
-	\includegraphics[width=.49\linewidth]{figures/cmd_feh_safe.png} \\
+	\includegraphics[width=.48\linewidth]{figures/cmd_feh_safe.png} \\
 	\includegraphics[width=.45\linewidth]{figures/cmd_spt_full.png}
-	\includegraphics[width=.49\linewidth]{figures/cmd_spt_safe.png}
+	\includegraphics[width=.48\linewidth]{figures/cmd_spt_safe.png}
 	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 5,875 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, spectroscopic binaries, and K dwarfs).
@@
 
-\begin{figure*}[h]
-	\includegraphics[width=\linewidth]{figures/aspcap_cannon_validation.png}
+\begin{figure*}[]
+\begin{center}
+	\includegraphics[width=.9\linewidth]{figures/aspcap_cannon_validation.png}
 	\caption{\cannon\ (right column) and ASPCAP (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 6,257 sources with $V,J,K$ colors.
@@
 	\label{fig:teff_comparisons}
+\end{center}
 \end{figure*}
 
-\begin{figure*}[h]
+\begin{figure*}[]
 \begin{center}
-	\includegraphics[width=5cm]{figures/met_comp1.png}
-	\includegraphics[width=5cm]{figures/met_comp2.png}
-	\includegraphics[width=5cm]{figures/met_comp3.png} 
-	\includegraphics[width=5cm]{figures/met_comp4.png}
-	\includegraphics[width=5cm]{figures/met_comp5.png}
-	\includegraphics[width=5cm]{figures/met_comp6.png} 
-	\includegraphics[width=5cm]{figures/met_comp7.png}
-	\includegraphics[width=5cm]{figures/met_comp8.png}
-	\includegraphics[width=5cm]{figures/met_comp9.png} 
+	\includegraphics[width=.32\linewidth]{figures/met_comp1.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp2.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp3.png} 
+	\includegraphics[width=.32\linewidth]{figures/met_comp4.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp5.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp6.png} 
+	\includegraphics[width=.32\linewidth]{figures/met_comp7.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp8.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp9.png} 
 	\caption{A comparison of ASPCAP and Cannon metallicities for 8,335 test sample sources separated into temperature bins of 100K.
@@
 
-\begin{figure*}[h]
+\begin{figure*}[]
 	\begin{center}
@@
 %---------------------------
-\begin{figure*}[h]
+\begin{figure*}[]
 	\begin{center}
@@
 
-% \begin{figure*}[h]
+% \begin{figure*}[]
 % 	\plotone{figures/spt_vs_teff_vs_feh.png}
@@
 
-% \begin{figure*}[h]
+% \begin{figure*}[]
 % 	\plotone{figures/mann_radii.png}
@@
 %% Tables
-
-\newpage
 

### 2019-07-27 paper/draft.tex
remove old comments
finish sample selection description

@@
 
-% In each case, \thecannon\ produces a predictive model for \apogee\ spectra, but without any input of stellar physics beyond whatever went in to producing the labels in the first place.
-That is, the model to be applied to the \apogee\ data is learned from the \apogee\ data alone.
-This obviates physical models that are often inaccurate for M dwarfs, and capitalizes on the excellent data quality delivered by the \apogee\ spectrograph and project.
-%JB: I think this part is redundant
-
 This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the \apogee\ and \gaia\ surveys, as well as previous studies of M dwarfs in \apogee.
@@
 \\
-
-%We note that rotational velocity (vsini) is not a parameter are not included as training labels because that would be hard... also it's probably not needed... I'll let someone else write that. -AWM
-%\color{gcolor}{}\color{black}
 
@@
 	\includegraphics[width=.48\linewidth]{figures/ruwe_cut_after.png}
-	\caption{}
+	\caption{The left panel shows the test sample of 8,335 sources after applying Selections $1-3$ described in Section \ref{subsec:test_selection}.
+The right panel shows the same test sample reduced to 6,221 after applying an astrometric quality cut of ${\rm RUWE}<1.2$ (Selection 4).
+To further remove sources which were likely binary contamination (Selection 5), we cut out sources above the red line which sparsely lay above the majority of the main-sequence with temperatures and metallicities that deviated from the expected gradient, reducing the final test sample to a size of 5,875 sources.
+The red line shown is constrained by the coordinates $\left\{ (3.5,12), (1.87,7.5) \right\}$ corresponding to ($BP-RP$, $M_G$).}
 	\label{fig:ruwe_cut}
@@
 
-\item \textbf{Model extraploation cuts:} Because extrapolation with \thecannon\ is unreliable, we select only sources with labels inside the range of the training sample: for the \cannon\ reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$, and for \cannon\ reported spectral subtypes we impose $0<SPT<9$.
+\item \textbf{Model extraploation cuts:} Because \thecannon\ does not perform well under extrapolated regions of parameter space, we select only sources inside the range of our training sample with ASPCAP parameters $2800<T_{\rm eff}<4100$ and $-0.5<[M/H]<0.5$, and with \cannon\ parameters of $2850<T_{\rm eff}<4150$, $-0.5<\feh<0.5$, and $0<SPT<9$.
 
-\item \textbf{Astrometric cut:} \gaia\ re-normalised unit weight error (RUWE) $RUWE<1.2$
+\item \textbf{Astrometric cut:} Using the \gaia\ re-normalised unit weight error (RUWE), a metric of evaluating the fit of the astrometric solution described in the additional release notes, we apply a cut of ${\rm RUWE}<1.2$ to select out sources with high astrometric error or noise, such as binaries (see Figure \ref{fig:ruwe_cut}).
 
-\item \textbf{Binary cut:}
+\item \textbf{Binary cut:} To remove further contamination from binary sources by a conservative cut, we applied an additional color-magnitude selection on sources above the main-sequence with \cannon\ temperatures/metallicities that deviated from the gradient, which we visibly selected for in Figure \ref{fig:ruwe_cut}.
 
@@
 Each plot shows the expected gradient: temperature increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly to the main sequence branch.
-Additionally we notice a slight deviation at the upper slice of the main sequence where \cannon\ temperatures, metallicities and spectral types all appear lower than the gradient trends in the rest of the branch, which is most likely the binary sequence.
-% Comparisons between the West-trained spectral types and Mann-trained temperatures and metallicities are shown in Figure \ref{fig:west_vs_mann}: left shows the slight negative correlation between spectral type and $\feh$, and right shows the nonlinear decrease in temperature with visible metallicity gradient.
-We also note that applying our model requires very little computational demand: the time to train and test a model on all 14,827 sources was two minutes on a personal laptop.
+We also note that applying our model requires very little computational demand: the time to train and test a model on all 14,827 sources was two minutes on a 2.7\,GHz Intel core i7 laptop.
 
@@
 
-% % SUMMARIZE \citealt{Casagrande:2008}, \citealt{Boyajian:2012}, and \citealt{Mann:2015} 
-% Shown in the top panels of \ref{fig:teff_comparisons}
-
 Compared to the \citealt{Casagrande:2008}, \citealt{Boyajian:2012}, and \citealt{Mann:2015} derived temperatures, we find ASPCAP overestimates $\teff$ on average by a bias of $130-170$K, while \thecannon\ underestimates $\teff$ by a bias of $\sim25$K.
@@
 \subsection{Isochrone Comparisons}
-
-% \begin{figure}
-% 	\includegraphics[width=\linewidth]{figures/cannon_vs_parsec.png}
-% 	\includegraphics[width=\linewidth]{figures/aspcap_vs_parsec1.png}
-% 	\caption{Comparison of \cannon\ $\feh$ as compared to metallicity predicted by PARSEC isochrone tracks \citep{Bressan:2012} in \gaia\ color-magnitude space.
-The two panels show five metallicity tracks (ranging $-0.5<$[M/H]$<+0.5$ in steps of 0.1 dex) at two different ages (3 and 7 Gyr).
-In the ASPCAP plot only the 8,960 out of 5,875 sources with metallcities $-0.5<$[M/H]$<+0.5$ are shown, while the remaining 1,351 sources with [M/H]$<-0.5$ and [M/H]$>+0.5$.}
-% 	\label{fig:isochrone_cmd}
-% \end{figure}
-
-% \begin{figure*}
-% 	\plotone{figures/aspcap_cannon_label_hist.png}
-% 	\caption{Distribution of temperatures and metallicities for 5,875 sources from \thecannon\ compared to the ASPCAP pipeline.
-Note that \cannon\ temperatures only span $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$\,dex because we have imposed cutoffs for stars that fall outside of the training sample range, however for the same 5,875 stars ASPCAP predictions are skewed higher in temperature and lower in metallicity compared to \cannon\ measurements.}
-% 	\label{fig:aspcap_cannon_label_hist} 
-% \end{figure*}
 
@@
 
-% \begin{figure*}[]
-% 	\plotone{figures/spt_vs_teff_vs_feh.png}
-% 	\caption{West-trained spectral types vs.
-Mann-trained metallcities and temperatures.
-For each integer spectral type bin the mean temperatures and metallicities are shown as red error boxes.
-\label{fig:west_vs_mann}}
-% \end{figure*}
-
-%---------------------------
-
-% \begin{figure*}[]
-% 	\plotone{figures/mann_radii.png}
-% 	\caption{Comparison of radii estimated from M15 relations: on the x-axis radii are derived from  K magnitudes and \cannon\ $\feh$, on the y-axis radii are derived from \cannon\ $\teff$ and $\feh$.}
-% 	\label{fig:radii}
-% \end{figure*}
-
 %%==========================================================================================

### 2019-07-28 paper/draft.tex
add discussion paragraph

@@
 We see the scatter and metal-poor bias of ASPCAP metallicities clearly increases at decreasing temperatures.}
+	\label{fig:met_comp}
 \end{center}
@@
 However, this is an important goal for the future with these data.
+
+While a detailed analysis of atmospheric model limitations is beyond the scope of this paper, 
+our results provide an avenue to compare the FGK star metallicity scale 
+to the less well-understood M dwarf metallicity scale.
+These results find that atmospheric metallicities are systematically metal-poor biased
+compared to \cannon-based metallicities trained on sources with metallicities callibrated
+to those of FGK companions. 
+At the high temperature end ($\teff > 3600$\,K), the ASPCAP metallicity bias is $\sim-0.15\,$dex with a
+scatter of $\sim0.06\,$dex relative to \cannon\ metallicities, and increases to a bias of $\>-0.26\,$dex 
+and scatter of $\>0.1\,$dex at the lower temperature end ($\teff < 3300$\,K) (Figure \ref{fig:met_comp}).
+This metal-poor bias, while not explored to a great extent in this work, we suspect is due 
+to the line lists of the models: 
+an effect in which the optimizer of the pipeline may be lowering the metallicity of the fit 
+to better reproduce the spectrum to compensate for the missing lines or opacities.
+Further analysis would need to be done to determine if the metallicity bias
+is found in other model grids (besides the ATLAS/MARCS models used by the ASPCAP pipeline), 
+and if the effect is present at other wavelengths.
 

### 2019-07-29 paper/draft.tex
describe 2 catalogs
cut out panel

@@
 
-\begin{figure*}
+\begin{figure}
 \begin{center}
-	\includegraphics[width=.48\linewidth]{figures/ruwe_cut_before.png}
-	\includegraphics[width=.48\linewidth]{figures/ruwe_cut_after.png}
+	\includegraphics[width=\linewidth]{figures/ruwe_cut_before.png} \\
+	\includegraphics[width=\linewidth]{figures/ruwe_cut_after.png}
 	\caption{The left panel shows the test sample of 8,335 sources after applying Selections $1-3$ described in Section \ref{subsec:test_selection}.
@@
 	\label{fig:ruwe_cut}
+\end{center}
+\end{figure}
+
+\begin{figure*}[]
+\begin{center}
+	\includegraphics[width=.9\linewidth]{figures/aspcap_cannon_validation.png}
+	\caption{\cannon\ (right column) and ASPCAP (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 1,702 sources with $V,J,K$ colors.
+Shown are \cannon\ and ASPCAP temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and $\feh$ dependent relation (top), and the \citealt{Boyajian:2012} relation $V-K$ and $\feh$ (bottom).}
+	\label{fig:teff_comparisons}
+\end{center}
+\end{figure*}
+
+\begin{figure*}[]
+\begin{center}
+	\includegraphics[width=.32\linewidth]{figures/met_comp1.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp2.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp3.png} 
+	\includegraphics[width=.32\linewidth]{figures/met_comp4.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp5.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp6.png} 
+	\includegraphics[width=.32\linewidth]{figures/met_comp7.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp8.png}
+	\includegraphics[width=.32\linewidth]{figures/met_comp9.png} 
+	\caption{A comparison of ASPCAP and Cannon metallicities for 8,335 test sample sources separated into temperature bins of 100K.
+We see the scatter and metal-poor bias of ASPCAP metallicities clearly increases at decreasing temperatures.}
+	\label{fig:met_comp}
 \end{center}
@@
 Table \ref{table:test_results} outlines the parameters included in the test sample catalog, which can be downloaded from the online journal.
+Included are two versions of the catalog: the first containing all 14,827 sources before selection, and the second containing the 5,875 sources kept after making Selections $1-5$ described in this section.
 
 % ----------------------------------------
-\subsection{Color-Temperature Comparisons}
+\subsection{Temperature Validation}
 
-\color{gcolor}{As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures).
+As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures).
 To obtain visual band magnitudes for a set of sources, we cross-matched the 5,875 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
@@
 
-While the large bulk of points falling within the temperature ranges of the plots shown does not qualitatively show a lot more deviation from color-temperature relations, the average ASPCAP temperatures are much hotter, and scatters are much higher skewed by 500 sources which ASPCAP predicts to have temperatures over 4200K.}\color{black}
-
 % ----------------------------------------
-\subsection{Isochrone Comparisons}
+\subsection{Metallicity Validation}
 
@@
 	\label{fig:safe_selection}
-\end{center}
-\end{figure*}
-
-\begin{figure*}[]
-\begin{center}
-	\includegraphics[width=.9\linewidth]{figures/aspcap_cannon_validation.png}
-	\caption{\cannon\ (right column) and ASPCAP (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 6,257 sources with $V,J,K$ colors.
-Shown are \cannon\ and ASPCAP temperatures (x-axis) compared to temperatures estimated from the \citealt{Mann:2015} $V-J$ and $\feh$ dependent relation (top), \citealt{Casagrande:2008} $V-K$ relation (middle), and the \citealt{Boyajian:2012} relation $V-K$ and $\feh$ (bottom).}
-	\label{fig:teff_comparisons}
-\end{center}
-\end{figure*}
-
-\begin{figure*}[]
-\begin{center}
-	\includegraphics[width=.32\linewidth]{figures/met_comp1.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp2.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp3.png} 
-	\includegraphics[width=.32\linewidth]{figures/met_comp4.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp5.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp6.png} 
-	\includegraphics[width=.32\linewidth]{figures/met_comp7.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp8.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp9.png} 
-	\caption{A comparison of ASPCAP and Cannon metallicities for 8,335 test sample sources separated into temperature bins of 100K.
-We see the scatter and metal-poor bias of ASPCAP metallicities clearly increases at decreasing temperatures.}
-	\label{fig:met_comp}
 \end{center}

### 2019-07-30 paper/draft.tex
add new figures

@@
 
+\subsection{Test Sample \label{subsec:test_selection}}
+
+Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources.
+From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.
+Although our catalog reports the temperatures, metallicities, and spectral types for 5,875 sources out of 14,827 with $\chi^2<100,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
+
+\begin{enumerate}
+\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000.
+
+\item \textbf{Color-magnitude cuts:} Using \gaia\ and \zmass\ photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to cut sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and multiple), and subdwarfs below the main sequence.
+
+\item \textbf{Model extraploation cuts:} Because \thecannon\ does not perform well under extrapolated regions of parameter space, we select only sources inside the range of our training sample with ASPCAP parameters $2800<T_{\rm eff}<4100$ and $-0.5<[M/H]<0.5$, and with \cannon\ parameters of $2850<T_{\rm eff}<4150$, $-0.5<\feh<0.5$, and $0<SPT<9$.
+
+\item \textbf{Astrometric cut:} Using the \gaia\ re-normalised unit weight error (RUWE), a metric of evaluating the fit of the astrometric solution described in the additional release notes, we apply a cut of ${\rm RUWE}<1.2$ to select out sources with high astrometric error or noise, such as binaries (see Figure \ref{fig:ruwe_cut}).
+
+\item \textbf{Binary cut:} To remove further contamination from binary sources by a conservative cut, we applied an additional color-magnitude selection on sources above the main-sequence with \cannon\ temperatures/metallicities that deviated from the gradient, which we visibly selected for in Figure \ref{fig:ruwe_cut}.
+
+\end{enumerate}
+
 \begin{figure}
@@
 
-\subsection{Test Sample \label{subsec:test_selection}}
+\begin{figure*}
+\begin{center}
+	\includegraphics[height=5cm]{figures/toomre.png}
+	\includegraphics[height=5cm]{figures/met_hist.png}
+	\caption{Toomre diagram showing the distribution of the \cannon\ metallicities (left).
+Dashed lines at $v_{tot}=70$\,km/s and 180\,km/s show roughly the separation of thin/thick/halo stars.
+Histogram (right) shows the metallicity distribution of thin and thick disk stars, with blue corresponding to $v_{tot}<70$\,km/s and green corresponding to $70<v_{tot}<180$\,km/s respectively.
+The average metallicities of the distribution of the thin disk stars are $\feh=0.00\pm0.17\,$dex, and the average metallicities of the thick disk stars are $\feh=-0.14\pm0.19\,$dex.}
+	\label{fig:toomre}
+\end{center}
+\end{figure*}
 
-Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources.
-From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.
-Although our catalog reports the temperatures, metallicities, and spectral types for 5,875 sources out of 14,827 with $\chi^2<100,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
-
-\begin{enumerate}
-\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000.
-
-\item \textbf{Color-magnitude cuts:} Using \gaia\ and \zmass\ photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to cut sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and multiple), and subdwarfs below the main sequence.
-
-\item \textbf{Model extraploation cuts:} Because \thecannon\ does not perform well under extrapolated regions of parameter space, we select only sources inside the range of our training sample with ASPCAP parameters $2800<T_{\rm eff}<4100$ and $-0.5<[M/H]<0.5$, and with \cannon\ parameters of $2850<T_{\rm eff}<4150$, $-0.5<\feh<0.5$, and $0<SPT<9$.
-
-\item \textbf{Astrometric cut:} Using the \gaia\ re-normalised unit weight error (RUWE), a metric of evaluating the fit of the astrometric solution described in the additional release notes, we apply a cut of ${\rm RUWE}<1.2$ to select out sources with high astrometric error or noise, such as binaries (see Figure \ref{fig:ruwe_cut}).
-
-\item \textbf{Binary cut:} To remove further contamination from binary sources by a conservative cut, we applied an additional color-magnitude selection on sources above the main-sequence with \cannon\ temperatures/metallicities that deviated from the gradient, which we visibly selected for in Figure \ref{fig:ruwe_cut}.
-
-\end{enumerate}
+\begin{figure}
+\begin{center}
+	\includegraphics[width=\linewidth]{figures/wide_binary_met.png}
+	\caption{}
+\end{center}
+\end{figure}
 
@@
 
-Compared to the \citealt{Casagrande:2008}, \citealt{Boyajian:2012}, and \citealt{Mann:2015} derived temperatures, we find ASPCAP overestimates $\teff$ on average by a bias of $130-170$K, while \thecannon\ underestimates $\teff$ by a bias of $\sim25$K.
-We also see the scatter is much higher for the ASPCAP temperatures ($\sim230$K) compared to the scatter of \cannon\ temperatures ($\sim85$K).
+Compared to the \citealt{Mann:2015} and \citealt{Boyajian:2012} color-metallicity derived temperatures, both ASPCAP and \cannon\ temperatures show similar scatters of $\sim60\,$K, but offset by a constant.
+We find \cannon\ to be in better agreement with \citealt{Mann:2015} and \citealt{Boyajian:2012}, with ASPCAP overestimating $\teff$ on average $\sim110-140\,$K, and \thecannon\ underestimating $\teff$ on average by $\sim10-20\,$K.
 
@@
 \subsection{Metallicity Validation}
+
+As a check of internal consistency, we cross-matched our M dwarf final sample with the \citealt{Elbadry:2018b} catalog of $>$50,000 high-confidence, widely separated binaries identified by \drtwo.
 

### 2019-07-30 paper/draft.tex
wide binary metallicities

@@
 
+\begin{deluxetable*}{cccccccc}
+\tablecaption{Metallicities of \apogee\ M dwarfs with wide binaries.}
+\tabletypesize{\scriptsize}
+\tablehead{Mdwarf Gaia ID      & Companion Gaia ID   & Mdwarf [M/H] & Mdwarf [Fe/H] & Companion [Fe/H] & Companion source & Separation (AU)}
+	\startdata
+	1476768433933264128 & 1476762627137479552 & -0.18      & 0.01        & 0.10           & APOGEE Cannon M   & 16091      \\
+	1476646285062821888 & 1476646250703083264 & -0.02      & 0.18        & 0.27           & APOGEE Cannon M   & 4155       \\
+	4008743154907642496 & 4008743086188165248 & -0.20      & 0.02        & 0.09           & APOGEE Cannon M   & 11855      \\
+	3053911383155592832 & 3053907221325174144 & -0.25      & -0.15       & -0.15          & APOGEE Cannon M   & 12884      \\
+	1475867727751525632 & 1475867727751525504 & -0.25      & -0.15       & -0.13          & APOGEE Cannon M   & 5091       \\
+	933708229245319552  & 933708164821511424  & -0.23      & 0.11        & 0.01           & APOGEE Cannon M   & 2245       \\
+	3888520938618496768 & 3888520938618115456 & -0.16      & -0.15       & -0.14          & APOGEE Cannon M   & 1128       \\
+	3888309828090932608 & 3888309832385988992 & -0.24      & -0.01       & 0.04           & APOGEE Cannon M   & 2499       \\
+	374137560388179328  & 374137560388178304  & -0.34      & -0.24       & -0.34          & LAMOST FGK        & 5888       \\
+	4031346880591589248 & 4031346914951327104 & -0.33      & -0.25       & -0.18          & LAMOST FGK        & 6637       \\
+	3443427533502334592 & 3443427533502335744 & 0.06       & 0.20        & 0.32           & LAMOST FGK        & 3151       \\
+	1605991802162861696 & 1605991767803123200 & -0.16      & -0.03       & -0.06          & LAMOST FGK        & 1043       \\
+	4418850890305439872 & 4418851263967008000 & -0.18      & 0.11        & 0.27           & APOGEE ASPCAP FGK & 4920       \\
+	1381455722290898176 & 1381408782592655616 & -0.13      & -0.02       & -0.08          & APOGEE ASPCAP FGK & 2401       \\
+	2687603208839189376 & 2687603311918406528 & -0.12      & 0.00        & -0.03          & APOGEE ASPCAP FGK & 11201    
+	\enddata
+\end{deluxetable*}
+
 \begin{figure*}[]

### 2019-07-30 paper/draft.tex
wording

@@
 with their properties and compositions.
-However, insufficient physical assumptions behind atmospheric models
+However, current atmospheric models
 have limited abilities at reproducing spectral features in stars of the 
@@
 
+As expected, the Toomre diagram in Figure \ref{fig:toomre} shows that higher metallicity sources in the sample are concentrated in low velocity space corresponding to the thin disk population, while the thick disk population contains a higher concentration of lower metallicity sources.
+
 % ----------------------------------------

### 2019-07-30 paper/draft.tex
table caption

@@
 \newcommand{\feh}{[{\mathrm{Fe}/\mathrm{H}}]}
+\newcommand{\mh}{[{\mathrm{M}/\mathrm{H}}]}
 
@@
 
-\item \textbf{Model extraploation cuts:} Because \thecannon\ does not perform well under extrapolated regions of parameter space, we select only sources inside the range of our training sample with ASPCAP parameters $2800<T_{\rm eff}<4100$ and $-0.5<[M/H]<0.5$, and with \cannon\ parameters of $2850<T_{\rm eff}<4150$, $-0.5<\feh<0.5$, and $0<SPT<9$.
+\item \textbf{Model extraploation cuts:} Because \thecannon\ does not perform well under extrapolated regions of parameter space, we select only sources inside the range of our training sample with ASPCAP parameters $2800<T_{\rm eff}<4100$ and $-0.5<\mh<0.5$, and with \cannon\ parameters of $2850<T_{\rm eff}<4150$, $-0.5<\feh<0.5$, and $0<SPT<9$.
 
@@
 \begin{deluxetable*}{cccccccc}
-\tablecaption{Metallicities of \apogee\ M dwarfs with wide binaries.}
+\tablecaption{Metallicities of \apogee\ M dwarfs with wide binaries.
+Reported in this table are the ASPCAP [M/H] and \cannon\ [Fe/H] of the M dwarfs, as well as the metallicity values of their binary companions.
+The source/survey of the metallicity and spectral classes of the companions are also given.
+\label{table:binaries}}
 \tabletypesize{\scriptsize}
@@
 	\includegraphics[width=\linewidth]{figures/wide_binary_met.png}
-	\caption{}
+	\caption{Metallicities of \cannon\ M dwarfs in binaries (x-axis) compared to the metallicities of their companions (y-axis).
+Label colors show the source and spectral type of the companion metallicities.
+Metallicity values for each binary pair is compiled in Table \ref{table:binaries}.
+}
 \end{center}
@@
 
-As a check of internal consistency, we cross-matched our M dwarf final sample with the \citealt{Elbadry:2018b} catalog of $>$50,000 high-confidence, widely separated binaries identified by \drtwo.
+As an additional check of our metallicity consistency, we cross-matched our M dwarf final sample with the \citealt{Elbadry:2018b} catalog of $>$50,000 high-confidence, widely separated binaries identified by \drtwo.
+In total we found 230 of the \apogee\ M dwarfs to have binary pairs (XX FGK+M, XX M+M, and XX WD+M).
 
-As expected, the Toomre diagram in Figure \ref{fig:toomre} shows that higher metallicity sources in the sample are concentrated in low velocity space corresponding to the thin disk population, while the thick disk population contains a higher concentration of lower metallicity sources.
+As expected, the Toomre diagram in Figure \ref{fig:toomre} shows that higher metallicity sources in the sample are concentrated in low velocity space corresponding roughly to the thin disk population, while the thick disk population contains a higher concentration of lower metallicity sources.
+Separating the two populations into separate histograms (also shown in Figure \ref{fig:toomre}), we find that thick disk sources are nearly 1-$\sigma$ more metal poor than thin disk sources with the average thin disk metallicity being near-solar $\feh=0.00\pm0.17\,$dex, and the average metallicities of the thick disk stars being $\feh=-0.14\pm0.19\,$dex.
+By contrast metallicities from ASPCAP show no significant difference in the distribution with an average thin disk metallicity of $\mh=-0.16\pm0.16\,$dex and an average thick disk metallicity of $\mh=-0.23\pm0.17\,$dex.
 
@@
 
-\begin{figure*}[]
+\begin{figure*}
 \begin{center}

### 2019-07-31 paper/draft.tex
finish metallicity commentary

@@
 Section \ref{sec:methods} describes our model implementation using \thecannon\ framework, and Section \ref{sec:sample_selection} describes the our sample selection and derivation of training parameters.
-\color{gcolor}{{In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$10,000 sources, and examine the validity of our parameters against color-temperature relations and isochrone grids.
-Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results in the context of present and upcoming stellar and exoplanetary surveys.}\color{black} \\
+In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$5,800 sources, and examine the validity of our parameters against color-temperature relations and metallicities of binary pairs.
+Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results. \\
 
@@
 
+\begin{figure}
+\begin{center}
+	\includegraphics[width=\linewidth]{figures/wide_binary_met.png}
+	\caption{Metallicities of \cannon\ M dwarfs in binaries (x-axis) compared to the metallicities of their companions (y-axis).
+Label colors show the source and spectral type of the companion metallicities.
+Metallicity values for each binary pair is compiled in Table \ref{table:binaries}.
+The overall scatter between the 15 metallicity pairs is 0.08\,dex.}
+\end{center}
+\end{figure}
+
 \begin{deluxetable*}{cccccccc}
@@
 
-\begin{figure*}[]
+\begin{figure*}
 \begin{center}
-	\includegraphics[width=.32\linewidth]{figures/met_comp1.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp2.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp3.png} 
-	\includegraphics[width=.32\linewidth]{figures/met_comp4.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp5.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp6.png} 
-	\includegraphics[width=.32\linewidth]{figures/met_comp7.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp8.png}
-	\includegraphics[width=.32\linewidth]{figures/met_comp9.png} 
+	\includegraphics[width=.3\linewidth]{figures/met_comp1.png}
+	\includegraphics[width=.3\linewidth]{figures/met_comp2.png}
+	\includegraphics[width=.3\linewidth]{figures/met_comp3.png} 
+	\includegraphics[width=.3\linewidth]{figures/met_comp4.png}
+	\includegraphics[width=.3\linewidth]{figures/met_comp5.png}
+	\includegraphics[width=.3\linewidth]{figures/met_comp6.png} 
+	\includegraphics[width=.3\linewidth]{figures/met_comp7.png}
+	\includegraphics[width=.3\linewidth]{figures/met_comp8.png}
+	\includegraphics[width=.3\linewidth]{figures/met_comp9.png} 
 	\caption{A comparison of ASPCAP and Cannon metallicities for 8,335 test sample sources separated into temperature bins of 100K.
@@
 Histogram (right) shows the metallicity distribution of thin and thick disk stars, with blue corresponding to $v_{tot}<70$\,km/s and green corresponding to $70<v_{tot}<180$\,km/s respectively.
-The average metallicities of the distribution of the thin disk stars are $\feh=0.00\pm0.17\,$dex, and the average metallicities of the thick disk stars are $\feh=-0.14\pm0.19\,$dex.}
+The average metallicity of the thin disk distribution is $\feh=0.00\pm0.17\,$dex, and the average metallicity of the thick disk distribution is $\feh=-0.14\pm0.19\,$dex.}
 	\label{fig:toomre}
@@
 \end{figure*}
-
-\begin{figure}
-\begin{center}
-	\includegraphics[width=\linewidth]{figures/wide_binary_met.png}
-	\caption{Metallicities of \cannon\ M dwarfs in binaries (x-axis) compared to the metallicities of their companions (y-axis).
-Label colors show the source and spectral type of the companion metallicities.
-Metallicity values for each binary pair is compiled in Table \ref{table:binaries}.
-}
-\end{center}
-\end{figure}
 
@@
 
-As an additional check of our metallicity consistency, we cross-matched our M dwarf final sample with the \citealt{Elbadry:2018b} catalog of $>$50,000 high-confidence, widely separated binaries identified by \drtwo.
-In total we found 230 of the \apogee\ M dwarfs to have binary pairs (XX FGK+M, XX M+M, and XX WD+M).
+As a check of our test sample metallicity reliability, we cross-matched our M dwarf final sample with the \citealt{Elbadry:2018b} catalog of $>$50,000 high-confidence, widely separated binaries identified by \drtwo.
+In total we found 216 of the \apogee\ M dwarfs to have binary pairs (46 FGK+M, 155 M+M, and 15 WD+M).
+Out of the 155 M+M pairs, 8 pairs contained both pairs in \apogee.
+Cross-matching the list of FGK+M dwarf companions with several catalogs/surveys with measured stellar metallicities, we found an additional 7 sources with FGK metallicities from LAMOST \citep{Zhao:2012} and APOGEE (ASPCAP).
+The metallicity measurements for the 15 M dwarf binaries and their companions are given in Table \ref{table:binaries}, and the overall scatter is $0.08\,$dex--an improvement over the scatter of ASPCAP metallicities which is $0.15\,$dex for these 15 sources.
 
@@
 Separating the two populations into separate histograms (also shown in Figure \ref{fig:toomre}), we find that thick disk sources are nearly 1-$\sigma$ more metal poor than thin disk sources with the average thin disk metallicity being near-solar $\feh=0.00\pm0.17\,$dex, and the average metallicities of the thick disk stars being $\feh=-0.14\pm0.19\,$dex.
-By contrast metallicities from ASPCAP show no significant difference in the distribution with an average thin disk metallicity of $\mh=-0.16\pm0.16\,$dex and an average thick disk metallicity of $\mh=-0.23\pm0.17\,$dex.
+On the other hand, metallicities from ASPCAP show no significant difference in distribution, with an average thin disk metallicity of $\mh=-0.16\pm0.16\,$dex and an average thick disk metallicity of $\mh=-0.23\pm0.17\,$dex.
+
+Figure \ref{fig:met_comp} shows that ASPCAP metallicities are systematically more metal-poor than \cannon\ metallicities.
+We furthermore find that the bias is temperature-dependent: at the hottest temperatures ($\teff>3600\,$K) ASPCAP and \cannon\ metallicities are consistent to a scatter of $0.05-0.06\,$dex and offset by an average of $-0.12-0.15\,$dex, while at the lowest temperatures ($\teff<3200\,$K) ASPCAP and \cannon\ are consistent to a scatter of $\sim-0.13\,$dex and offset by an average of $-0.3\,$dex.
 

### 2019-07-31 paper/draft.tex
switch order
abstract + title

@@
 
-\title{TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY}
+\title{TEMPERATURES AND METALLICITIES OF M DWARFS IN THE \apogee\ SURVEY}
 
@@
 $2860 < \teff < 4130$\,K callibrated with bolometric temperatures, 
-$-0.5 < \feh < 0.5$\,dex callibrated with FGK wide binary metallicities, 
+$-0.5 < \feh < 0.5$\,dex callibrated with FGK binary metallicities, 
 we train a two-parameter model with predictive accuracy (in cross-validation) 
@@
 0.9 types.
-\color{gcolor}{From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we
-apply our models and compare our \cannon\ derived $\teff$ against color--temperature
-relations and \cannon\ derived $\feh$ against isochrone models in
-\gaia\ color-magnitude space.
-Finally we use our results to
-estimate stellar radii from absolute magnitudes and \cannon\ temperatures [AND WE FIND WHAT??].}\color{black}
+From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we
+apply our models to a selection of 5,875 M dwarfs. 
+We find \cannon\ temperatures to be in agreement to 60\,K
+compared to a subsample of 1,702 sources with color-derived temperatures, and 
+find \cannon\ metallicites to be in agreement to 0.08\,dex metallicity
+compared to a subsample of 15 FGK+M or M+M binaries. 
+Finally, our comparison between \cannon\ and APOGEE pipeline (ASPCAP) labels finds that 
+ASPCAP is systematically biased towards reporting higher temperatures and lower metallicities. 
 \end{abstract}
@@
 \begin{center}
+	\includegraphics[height=5cm]{figures/toomre.png}
+	\includegraphics[height=5cm]{figures/met_hist.png}
+	\caption{Toomre diagram showing the distribution of the \cannon\ metallicities (left).
+Dashed lines at $v_{tot}=70$\,km/s and 180\,km/s show roughly the separation of thin/thick/halo stars.
+Histogram (right) shows the metallicity distribution of thin and thick disk stars, with blue corresponding to $v_{tot}<70$\,km/s and green corresponding to $70<v_{tot}<180$\,km/s respectively.
+The average metallicity of the thin disk distribution is $\feh=0.00\pm0.17\,$dex, and the average metallicity of the thick disk distribution is $\feh=-0.14\pm0.19\,$dex.}
+	\label{fig:toomre}
+\end{center}
+\end{figure*}
+
+\begin{figure*}
+\begin{center}
 	\includegraphics[width=.3\linewidth]{figures/met_comp1.png}
@@
 	\label{fig:met_comp}
-\end{center}
-\end{figure*}
-
-\begin{figure*}
-\begin{center}
-	\includegraphics[height=5cm]{figures/toomre.png}
-	\includegraphics[height=5cm]{figures/met_hist.png}
-	\caption{Toomre diagram showing the distribution of the \cannon\ metallicities (left).
-Dashed lines at $v_{tot}=70$\,km/s and 180\,km/s show roughly the separation of thin/thick/halo stars.
-Histogram (right) shows the metallicity distribution of thin and thick disk stars, with blue corresponding to $v_{tot}<70$\,km/s and green corresponding to $70<v_{tot}<180$\,km/s respectively.
-The average metallicity of the thin disk distribution is $\feh=0.00\pm0.17\,$dex, and the average metallicity of the thick disk distribution is $\feh=-0.14\pm0.19\,$dex.}
-	\label{fig:toomre}
 \end{center}

### 2019-07-31 paper/draft.tex
comment
change # sources

@@
 sense; it is only good at the percent level.
-\color{gcolor}{This percent-level accuracy of the model in the data domain is also demonstrated...}\color{black} \\
+% \color{gcolor}{This percent-level accuracy of the model in the data domain is also demonstrated...}\color{black} \\
 
@@
 As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures).
-To obtain visual band magnitudes for a set of sources, we cross-matched the 5,875 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,257 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
+To obtain visual band magnitudes for a set of sources, we cross-matched the 5,875 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 1,702 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
 Figure \ref{fig:teff_comparisons} shows \cannon\ vs.
 photometric temperatures on the right, and ASPCAP vs.
-photometric temperatures on the left for each of the 6,257 sources, colored by their respective spectroscopic metallicities.
+photometric temperatures on the left for each of the 1,702 sources, colored by their respective spectroscopic metallicities.
 
@@
 We find \cannon\ to be in better agreement with \citealt{Mann:2015} and \citealt{Boyajian:2012}, with ASPCAP overestimating $\teff$ on average $\sim110-140\,$K, and \thecannon\ underestimating $\teff$ on average by $\sim10-20\,$K.
+\\
 

### 2019-07-31 paper/draft.tex
table formatting

@@
 
-\begin{deluxetable}{|c|c|c|c|c|}
+\begin{deluxetable}{ccccc}
 \tablecaption{Strong $\teff$-correlated molecular lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
-\tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
+\tablehead{\colhead{Species} & \colhead{Wavelength} & \colhead{Derivative} & \colhead{$\sigma_{\rm jackknife}$} & \colhead{Significance}}
 	\startdata
@@
 
-\begin{deluxetable}{|c|c|c|c|c|}
+\begin{deluxetable}{ccccc}
 \tablecaption{Strong $\teff$-correlated atomic lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
-\tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
+\tablehead{\colhead{Species} & \colhead{Wavelength} & \colhead{Derivative} & \colhead{$\sigma_{\rm jackknife}$} & \colhead{Significance}}
 	\startdata
@@
 
-\begin{deluxetable}{|c|c|c|c|c|}
+\begin{deluxetable}{ccccc}
 \tablecaption{Strong $\feh$-correlated atomic and molecular lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
-\tablehead{Species & Wavelength & Derivative & $\sigma_{\rm jackknife}$ & Significance}
+\tablehead{\colhead{Species} & \colhead{Wavelength} & \colhead{Derivative} & \colhead{$\sigma_{\rm jackknife}$} & \colhead{Significance}}
 	\startdata
@@
 \tabletypesize{\scriptsize} 
-\tablehead{Mdwarf Gaia ID      & Companion Gaia ID   & Mdwarf [M/H] & Mdwarf [Fe/H] & Companion [Fe/H] & Companion source & Separation (AU)}
+\tablehead{\colhead{Mdwarf Gaia ID}      & \colhead{Companion Gaia ID}   & \colhead{Mdwarf [M/H]} & \colhead{Mdwarf [Fe/H]} & \colhead{Companion [Fe/H]} & \colhead{Companion source} & \colhead{Separation (AU)}}
 	\startdata
@@
 \tablecaption{Test results table header description.}
-\tablehead{Column & Unit & Description}
+\tablehead{\colhead{Column} & \colhead{Unit} & \colhead{Description}}
 	\startdata

### 2019-07-31 paper/draft.tex
grants
abstract grammer

@@
 formation on both Galactic and exoplanetary scales
-with their properties and compositions.
+through their properties and compositions.
 However, current atmospheric models
-have limited abilities at reproducing spectral features in stars of the 
-coolest temperature regimes ($T_{\rm eff} < 4200\,$K) and fully
-exploiting the information content and statistical extent of
-current and upcoming large-scale spectroscopic surveys.
-Here we present a catalog of spectroscopic temperatures, metallicities,
-spectral types for 5,875 M dwarfs in the \apogee\ survey using
+have limited ability to reproduce spectral features in stars at the 
+coolest temperatures ($T_{\rm eff} < 4200\,$K) and to fully
+exploit the information content of current and upcoming large-scale spectroscopic surveys.
+Here we present a catalog of spectroscopic temperatures, metallicities, and
+spectral types for 5,875 M dwarfs in the \apogee+\drtwo\ surveys using
 \thecannon: a flexible, data-driven spectral-modeling and
@@
 Using a training sample of 87 M dwarfs with optically derived labels spanning 
-$2860 < \teff < 4130$\,K callibrated with bolometric temperatures, 
-$-0.5 < \feh < 0.5$\,dex callibrated with FGK binary metallicities, 
+$2860 < \teff < 4130$\,K calibrated with bolometric temperatures, 
+$-0.5 < \feh < 0.5$\,dex calibrated with FGK binary metallicities, 
 we train a two-parameter model with predictive accuracy (in cross-validation) 
 to 77\,K and 0.09\,dex respectively.
-We also train a one-dimensional spectral type model using 51 M dwarfs
-with known SDSS optical spectral types ranging M0 to M9, to predictive accuracy of
+We also train a one-dimensional spectral classification model using 51 M dwarfs
+with SDSS optical spectral types ranging from M0 to M9, to predictive accuracy of
 0.9 types. 
-From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we
-apply our models to a selection of 5,875 M dwarfs. 
-We find \cannon\ temperatures to be in agreement to 60\,K
+We find \cannon\ temperatures to be in agreement to within 60\,K
 compared to a subsample of 1,702 sources with color-derived temperatures, and 
-find \cannon\ metallicites to be in agreement to 0.08\,dex metallicity
+\cannon\ metallicites to be in agreement to within 0.08\,dex metallicity
 compared to a subsample of 15 FGK+M or M+M binaries. 
@@
 and the Moore--Sloan Data Science Environment at NYU.
+A.J.B. acknowledges funding support from the National Science Foundation under award No. AST-1517177.
+This work is supported by the SDSS Faculty and Student Team (FAST) initiative.
 
@@
 	\code{The Cannon} \citep{Ness:2015}}
-
-\color{gcolor}{ADAM: please add NASA and NSF grants in the parentheses above, and do you have a FAST acknowledgement?}\color{black}
 

### 2019-07-31 paper/draft.tex
intro grammar

@@
 
-\author[0000-0002-6523-9536]{Adam Burgasser}
+\author[0000-0002-6523-9536]{Adam J. Burgasser}
 \affil{Center for Astrophysics and Space Science, University of California San Diego, La Jolla, CA 92093, USA}
@@
 
-Low-mass stars (with masses $M_{*} < 0.7\,\msun$ and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010}.
-With nuclear fusion timescales of $\tau>10^{11}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010,Woolf:2012}.
+Low-mass stars, with masses $M_{*} < 0.7\,\msun$ and effective temperatures $T_{\rm eff} < 4000\,$K, are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the Galaxy's population by number \citep{Bochanski:2010}.
+With nuclear fusion timescales $\tau>10^{11}$ yr \citep{Laughlin:1997}, the chemical compositions of the M dwarf populations trace the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, and are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010,Woolf:2012}.
 
-Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity offset detections \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods (for a fixed stellar insulation flux) allow for new planet discovery in less observation time than their higher mass counterparts.
+Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity variability \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods (for a fixed stellar insolation flux) allow for new planet discovery in less observation time than for higher mass stars.
 For these reasons, M dwarfs are primary candidates for exoplanet searches, including by the NASA \kepler\ \citep[e.g.,][]{Dressing2015} and Transiting Exoplanet Survey Satellite (\tess) \citep[e.g.,][]{Muirhead:2018} missions.
-As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with {\sl JWST}.
+As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with the James Webb Space Telescope ({\sl JWST;} \citealt{Clampin2008}).
 
-Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources), enabling studies of M dwarf abundances on both exoplanet and Galactic scales.
-The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III/IV mission \citep{Eisenstein:2011,Blanton:2017} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}.
-\apogee\ pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have only been determined to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
+Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey \citep[SDSS;][]{Eisenstein:2011,Blanton:2017}, and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope \citep[LAMOST;][]{Zhao:2012} have dramatically increased the sample of known M dwarfs \citep{West:2011,Guo:2015} with catalogs of over 70,000 sources, enabling studies of M dwarf abundances on a Galactic scale.
+The Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the SDSS III/IV mission has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}.
+\apogee\ pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have been determined to precisions 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
 
 Elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precision (down to $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by improvements in atmospheric models, including realistic 3D local thermodynamic equilibrium (LTE) assumptions \citep{Asplund2005}, and differential abundance techniques using equivalent widths \citep{Bedell2014}.
-However, the determination of precise metallicities for M dwarfs has remained a long-standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperature regimes, particularly TiO and VO dominating the optical spectrum, H$_2$O and CO for the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) present in the spectra of the latest spectral types \citep{Allard1997}.
-In particular, atmospheric models fail to reproduce spectral features \citep[e.g.,][]{Mann:2013c} due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions, and the lack of continuum in a spectrum, all make it difficult to deconvolve individual features and extract line strengths from equivalent widths.
-The combination of these effects creates fundamental limitations to explore the information content available out of high resolution spectra using a traditional comparison to atmosphere models.
+However, the determination of precise metallicities for M dwarfs has remained a long-standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperatures, particularly TiO and VO bands dominating the optical spectrum, H$_2$O and CO bands in the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) present in the spectra of the latest spectral types \citep{Allard1997}.
+Atmospheric models often fail to reproduce these spectral features \citep[e.g.,][]{Mann:2013c} due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions, and the absence of a clear continuum, all contribute to make it difficult to deconvolve individual features and extract line strengths from equivalent widths.
+The combination of these effects limit our ability to explore the information content of high resolution spectra using traditional comparison to atmosphere models.
 
-A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically associated wide binaries are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the spectra of the M dwarf (e.g.
-\citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
-However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band and does not contain information about the overall spectral shape often used to determine $\teff$ and spectral type.
-Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the additional spectral information provided by \apogee's resolution.
+A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf (e.g.
+\citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}) to developed precise empirical relations (as good as $\sim$0.07\,dex).
+However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band nor information about the overall spectral shape often used to determine $\teff$ and spectral type.
+Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the greater spectral information provided by \apogee's resolution.
 

### 2019-07-31 paper/draft.tex
take out addressed comments
intro grammar

@@
 compared to a subsample of 15 FGK+M or M+M binaries. 
-Finally, our comparison between \cannon\ and APOGEE pipeline (ASPCAP) labels finds that 
+Finally, our comparison between \cannon\ and \apogee\ pipeline (ASPCAP) labels finds that 
 ASPCAP is systematically biased towards reporting higher temperatures and lower metallicities. 
@@
 
-%the relevant calibration indices used by existing studies (such as Ca~I or K~I lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.
-%AWM: I'm not really sure this is true.
-The major lines I and Babs use are in the K-band, very few studies use optical bands.
-Ryan includes ones in the H-band, some of which are in Apogee.
-I think the bigger point is that this fails to utilize the high-resolution information of the spectra, and relies instead of temperature calibration from spectral shape, which is impossible at this resolution.
-
-%%AWM: would it be useful to have some comment on data-driven models in general? Their virtues should probably be explained before invoking them
-
-In this work we build a data-driven model for M dwarf spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a,Behmard2019}: a fully empirical model, which employs no use of line lists or radiative transfer models.
-\thecannon\ is instead a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:methods}).
+In this work we build a data-driven model for M dwarf \apogee\ spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a,Behmard2019}: a fully empirical model, which employs no line lists or radiative transfer models.
+\thecannon\ is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:methods}).
 The model in this sense is used to \emph{transfer} labels from spectra for which we know parameters to those which we do not.
-Hence this data-driven approach effectively circumvents the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (and very accurately measured) \emph{reference labels} possibly measured from other data.
-
-%AWM: not an expert here, so I won't add anything, but this probably needs more information.
-%JB: I've swapped some sentences from section 2.2 and section 3, which explain the terminology.
-Do you think this paragraph adds enough clarity now without being too redundant with the following sections 2 & 3 describing The Cannon?
+This data-driven approach effectively circumvents the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (and very accurately measured) \emph{reference labels} possibly measured from other data.
 
 The data-driven approach of \thecannon\ is ideal in certain cases:
-if stellar labels are only known for a small number of stars
+if stellar labels are known for a small number of stars
 but there are spectra taken for many more;
-if it is computationally expensive to obtain labels for a star, but
+if it is computationally expensive to obtain labels for a star, and
 there are many stars that need labels;
-if there are spectral models or techniques that work in one wavelength range or resolution but not in another.
-%AWM: seems to me like the context here is all 3. You want to run on ~10,000 stars, which is expensive if you use models. You have labels for a small fraction of these. 
-%The context of this paper is the last:
-There are empirically-calibrated methods to measure $\teff$ and metallicity from flux-calibrated spectra in the optical and NIR, but existing methods to model M dwarf spectra in the NIR at high resolution are computationally expensive, calibrated over a narrow range of $\teff$ and metallicity, and/or focus on the $J$-band \citep{Lindgren:2016, Lindgren:2017}.
-\thecannon\ fills this niche: it doesn't require using specific lines or opacity information that may be missing from the models; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
-%AWM: again, the metallicity metrics Babs/Ryan/Myself use are all in the NIR, and they aren't based on models.
-The temperature tools use models, but they are calibrated against LBOI measurements.
-I need to come back to editing later, but this paragraph probably needs to be tweaked a bit.
-%AWM: OK, let me know what you think about those changes. Feel free to edit this more, but do note the work of Lindgren modeling J-band at high resolution (I think they use CRIRES).
+or if there are spectral models or techniques that work in one wavelength range or resolution but not in another.
+Existing methods to model M dwarf spectra in the NIR at high resolution are computationally expensive, calibrated over a narrow range of $\teff$ and metallicity, and/or focus on the $J$-band \citep{Lindgren:2016, Lindgren:2017}.
+\thecannon\ thus fills this niche: it doesn't require using specific lines or opacity information that may be missing from the models; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
 
-Here we take M-dwarf labels from samples of well-characterized stars that overlap the
+Here we take M-dwarf labels from samples of well-characterized stars that are present in the
 \sdssiv\ \apogee\ sample, and use those labels to train a model and label all of the M dwarfs observed by
-\sdssiv\ \apogee. One set of labels are physical parameters (effective temperatures and metallicities)
-and the other set of labels are spectral types.
+\sdssiv\ \apogee. One set of labels are physical parameters (effective temperatures and metallicities), the other set of labels are spectral types.
 

### 2019-07-31 paper/draft.tex
data grammar

@@
 In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$5,800 sources, and examine the validity of our parameters against color-temperature relations and metallicities of binary pairs.
-Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results. \\
+Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results.
 
@@
 
-\apogee\ is primarily targeted for bright stellar populations particularly red giants, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution.
+\apogee\ is primarily designed to target bright stellar populations, particularly red giants, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013}, with the objective of studying galactic composition and evolution.
 However numerous cool, main sequence sources have also been observed either as targets proposed by the \apogee\ M dwarf ancilliary survey ($\sim$1,200 sources; \citealt{Desphande:2013}), or serendipitously.
 
-A number of studies out of the M dwarf ancilliary survey have already been conducted towards measuring reliable fundamental atmospheric parameters and kinematic measurements using spectral synthesis of atmospheric model grids.
-These studies include \citealt{Desphande:2013} and \citealt{Gilhool:2018} which have studied the radial and rotational kinematics for 700+ sources,
-\citealt{Souto:2017} and \citealt{Souto:2018} which have modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining $\teff$/$\logg$/metallicity + 13 elemental abundances,
-\citealt{Rajpurohit:2018} which tested BT-Settl \citep{Allard:2012} and MARCS \citep{Gustafsson:2008} model grids on 45 M dwarfs to estimate $\teff$/$\logg$/metallicity,
-and \citealt{Skinner:2018} which has identified and measured mass ratios and radial velocities for 44 M dwarf spectroscopic binaries.
+A number of studies out of the M dwarf ancilliary survey have already been conducted to measure reliable fundamental atmospheric parameters and kinematic measurements using spectral synthesis of atmospheric model grids.
+These studies include \citealt{Desphande:2013} and \citealt{Gilhool:2018} which have studied the radial and rotational kinematics for 700+ sources;
+\citealt{Souto:2017} and \citealt{Souto:2018} which have modeled three exoplanet-hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining $\teff$/$\logg$/metallicity + 13 elemental abundances;
+\citealt{Rajpurohit:2018} which tested BT-Settl \citep{Allard:2012} and MARCS \citep{Gustafsson:2008} model grids on 45 M dwarfs to estimate $\teff$/$\logg$/metallicity;
+and \citealt{Skinner:2018} which identified and measured mass ratios and radial velocities for 44 M dwarf spectroscopic binaries.
 This work complements existing studies by producing model-independent catalog of spectroscopic temperatures and metallicites to test against model predictions for the entire \apogee\ M dwarf sample, which we quantify to contain at least 10,000 sources to date (DR14).
 
-The ASPCAP pipeline releases several types of data files, with various levels of processing: {\tt\string ap1D} (the raw one-dimensional spectra for individual visits), {\tt\string apVisit} (the individual visit spectra with telluric subtraction), {\tt\string apStar} (the co-added {\tt\string apVisit} spectra), and most processed {\tt\string aspcapStar} spectra.
-For this work we use {\tt\string aspcapStar} files, which contains the pseudo-continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
+The ASPCAP pipeline releases several types of data files, with various levels of processing: {\tt\string ap1D} (the raw one-dimensional spectra for individual visits), {\tt\string apVisit} (the individual visit spectra with telluric subtraction), {\tt\string apStar} (the co-added {\tt\string apVisit} spectra), and {\tt\string aspcapStar}, which contains the pseudo-continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline).
+We use the last dataset for our study.
 In previous work it has been recommended to use an alternative pseudo-continuum normalization \citep{Ness:2015}, but we don't find obvious issues with the normalization in our analysis, so we retain the survey pipeline outputs.

### 2019-07-31 paper/draft.tex
delete subsec header
methods grammar

@@
 
-\subsection{\apogee\ Survey}
-
 The \apogee\ survey is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
@@
 
-Inferring the label of a star with such a model requires two steps: first the \emph{training step} in which a generative model describing the probability density function (pdf) of the flux is constructed at each pixel from the set of spectrum with known reference labels, and second the \emph{test step} in which the model is applied to determine the labels of a spectrum.
+Inferring the label of a star with such a model requires two steps: first the \emph{training step} in which a generative model describing the probability density function (pdf) of the flux is constructed at each pixel from the set of spectra with known reference labels, and second the \emph{test step} in which the model is applied to determine the labels of a spectrum.
 
-Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector $\ell$ and and coefficient vector \emph{$\theta$}.
-So for each star \emph{n}, at wavelength pixel \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
+Following the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector $\ell$ and and coefficient vector \emph{$\theta$}.
+For each star \emph{n}, at wavelength pixel \emph{$\lambda$}, we assume that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
 \begin{equation}
-	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + {\rm noise} \label{eqn:cannon}
+	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + N_{\lambda} \label{eqn:cannon}
 \end{equation} 
-We use the noise model is ${\rm noise}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
+We use the noise model is $N_{\lambda}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
 Equation \ref{eqn:cannon} corresponds to the single-pixel log-likelihood function:
@@
 \end{equation}
-the probability density function of the measured flux, given the labels, coefficients and scatters.
+which gives the probability density function of the measured flux, given the labels, coefficients and scatters.
 
-We apply a quadratic parameterization of the model such that the label vectors for the two models are all of the combinations of reference labels, up to second order:
+We apply a quadratic parameterization of the model such that the label vectors for the two models are all combinations of reference labels up to second order:
 \begin{eqnarray}
@@
 
-The \emph{training step} consists of optimizing the likelihood function (Equation \ref{eqn:likelihood}) for the coefficient vector and scatter ($\theta_{\lambda}$ and $s_{\lambda}$) given fixed label vector (\emph{$\ell_n$}) constructed from the reference labels, and the \emph{test step} consists of optimizing the likelihood function for the labels at fixed $\theta_{\lambda}$ and $s_{\lambda}$ obtained in the training step (see \citealt{Ness:2015} for further description).
+The \emph{training step} consists of optimizing the likelihood function (Equation \ref{eqn:likelihood}) for the coefficient vector and scatter ($\theta_{\lambda}$ and $s_{\lambda}$) given the fixed label vector (\emph{$\ell_n$}) constructed from the reference labels.
+The \emph{test step} consists of optimizing the likelihood function for the labels at fixed $\theta_{\lambda}$ and $s_{\lambda}$ obtained in the training step (see \citealt{Ness:2015} for further description).
 In the training step, the regression is designed to predict spectral pixels
@@
 the labels. In the test step, the regression is designed to predict labels given the spectral
-derivatives. \\
+derivatives.
 

### 2019-07-31 paper/draft.tex
sample selection grammar

@@
 Limitations of \thecannon\ include that test (output) labels are only accurate if the training labels are accurate, and only precise if the training labels are measured consistently across the training sample.
-Also, it is critical to have a training sample with the dynamic range to span the entire parameter space of interest, as \thecannon\ does not extrapolate well outside the parameter space of the training sample.
-Finally, \thecannon\ assumes that the dependence of spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions---if that isn't true, there will be features that \thecannon\ cannot reproduce.
+It is also critical to have a training sample with the dynamic range to span the entire parameter space of interest, as \thecannon\ does not extrapolate well outside the parameter space of the training sample.
+Finally, \thecannon\ assumes that the dependence of spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions.
+If that isn't true, there will be features that \thecannon\ cannot reproduce.
 
-For the purpose of this study we have constructed two different training samples: first a one-dimensional \emph{spectral type model}, and second a two-dimensional \emph{physical parameter model} which describes the temperature and metallicity.
+For the purpose of this study, we have constructed two different training samples: first a one-dimensional \emph{spectral type model}, and second a two-dimensional \emph{physical parameter model} which describes the temperature and metallicity.
 The choice of training labels, dimensionality of our data set, and requirements for a good training set we discuss further in Section \ref{sec:discussion}.
@@
 % ----------------------------------------
-\subsection{Spectral Type Sample}
+\subsection{Spectral Type Training Sample}
 
-The spectral type sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from \sdss.
+The spectral type training sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from \sdss.
 For each source in the catalog, spectral types were determined both through an automated routine for comparing spectral type templates to data using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
@@
 % ----------------------------------------
-\subsection{Physical Parameter Sample}
+\subsection{Physical Parameter Training Sample}
 
-The physical parameter sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 < \feh < 0.49$\,dex, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
+The physical parameter training sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 < \feh < 0.49$\,dex, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
 The major difference between the extension sample is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample.
 
-The M15 catalog in total contains 183 sources and the extension sample another 500 stars, both samples were primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog.
+The M15 catalog in total contains 183 sources and the extension sample another 500 stars.
+Both samples were primarily selected from the proper-motion selected CONCH-SHELL \citep{Gaidos:2013} M dwarf catalog.
 All targets have low-res optical spectra from the SNIFS spectrograph \citep{Lantz:2004} and infrared spectra taken with the SpeX Spectrograph \citep{Rayner:2003}, which have been combined to estimate largely empirical bolometric fluxes.
@@
 Uncertainties are estimated to be $\simeq$0.08\,dex based on irreducible scatter in the empirical relation between selected lines and the assigned $\feh$ from the primary star.
-As with $\teff$ relative errors on $\feh$ are smaller, estimated to be 0.04--0.06\,dex over most of the temperature and metallicity range considered here.
+As with $\teff$, relative errors on $\feh$ are smaller, estimated to be 0.04--0.06\,dex over most of the temperature and metallicity range considered here.
 
@@
 While we only had $\feh$ for the training sample, for the uncertainties considered here, lack of information about [$\alpha$/Fe] or specific abundances will only be important compared to other uncertainties in extreme cases (e.g., Carbon stars).
-\\
 

### 2019-08-01 paper/draft.tex
remove 'self test'

@@
 	\includegraphics[width=\linewidth]{figures/crv_test_fe_h.png}
-	\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (left), and leave-one-out cross validation (right).
-Sources with strong rotational broadening (2M07171706--0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.
+	\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (top), and leave-one-out cross validation (bottom).
+Sources with strong rotational broadening (2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.
 Predictive accuracy, as computed by the scatter in cross validation is 77K in temperature, and 0.09 dex in metallicity.}
@@
 
-For the physical parameter model we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
-We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr for the published sample estimated by M15) and likely slowly rotating sources.
-While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.
+For the physical parameter model, we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
+We note for this model 5 out of 87 sources show possible rotational line broadening identified by visual inspection (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources show no obvious broadening.
+We note these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.
 
-Two tests of consistency were used to assess the validity of our model's labels: (1) the self-test comparing training label to test label values, and (2) the leave-one-out cross validation (LOOCV) test, in which we train a model on all sources but \emph{n}, then apply the \emph{N-1} source trained model to obtain the labels for star \emph{n}.
+To assess the validity of our model's labels we used a leave-one-out cross validation (LOOCV) test, in which we train a model on all sources but \emph{n}, then apply the \emph{N-1} source trained model to obtain the labels for star \emph{n}.
 Precision (scatter) and bias of the model for each test are calculated as the standard deviation and mean of the difference in training and test (or LOOCV) labels respectively (Figure \ref{fig:mann_validation}).
-Since the LOOCV test evaluates both how well the model reproduces the training values \emph{and} penalizes the model for over-fitting, we adopt the LOOCV scatter over the self-test scatter as the more conservative estimate of the model's precision.
-The set of training, test, and LOOCV labels for each training source is reported in Table \ref{table:mann_results}.
+Since the LOOCV test evaluates both how well the model reproduces the training values \emph{and} penalizes the model for over-fitting, we adopt the LOOCV scatter over as the estimate of the model's precision.
+The set of training, test, and cross-validated labels for each training source is reported in Table \ref{table:mann_results}.
 
@@
 	\begin{center}
-	\includegraphics[width=\linewidth]{figures/self_test_spt.png}
+	% \includegraphics[width=\linewidth]{figures/self_test_spt.png}
 	\includegraphics[width=\linewidth]{figures/crv_test_spt.png}
 	\end{center}
-	\caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).
-Predictive accuracy, as computed by the scatter in cross validation is 0.9 subtypes.}
+	\caption{Leave-one-out cross validation test for the West-trained spectral type model. Predictive accuracy, as computed by the scatter in cross validation is 0.9 subtypes.}
 	\label{fig:west_validation}

### 2019-08-01 paper/draft.tex
remove 'self test'

@@
 We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
-As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates \thecannon\ does a poor job at extrapolating to labels outside of the training sample space.
-Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows the self and LOOCV tests for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.
+As seen in Figure \ref{fig:west_validation}, the model performs poorly at reproducing spectral types $>$M8 because \thecannon\ does not extraploate well to labels outside of the training sample space.
+Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows LOOCV test for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.
 

### 2019-08-01 paper/draft.tex
results revisions

@@
 We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
-As seen in Figure \ref{fig:west_validation}, the model performs poorly at reproducing spectral types $>$M8 because \thecannon\ does not extraploate well to labels outside of the training sample space.
+As seen in Figure \ref{fig:west_validation}, the model performs poorly at reproducing spectral types $>$M8, which confirms that \thecannon\ does not extraploate well to labels outside of the training sample space.
 Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows LOOCV test for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.
@@
 Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources.
-From there we applied the following selection procedure to identify a sample of single, main-sequence M stars.
-Although our catalog reports the temperatures, metallicities, and spectral types for 5,875 sources out of 14,827 with $\chi^2<100,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):
+From there we applied additional selection criteria, described below, to identify a sample of single, main-sequence M stars, with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries:
 
 \begin{enumerate}
-\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000.
+\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000, chosen to remove badly fit sources (such as fast rotators), but include $\chi^2$ values close the distribution of training sample (Figure \ref{fig:chi_dist}).
 
-\item \textbf{Color-magnitude cuts:} Using \gaia\ and \zmass\ photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to cut sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and multiple), and subdwarfs below the main sequence.
+\item \textbf{Color-magnitude cuts:} Using \gaia\ and \zmass\ photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to remove sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and multiples), and subdwarfs below the main sequence.
 
@@
 
-\item \textbf{Astrometric cut:} Using the \gaia\ re-normalised unit weight error (RUWE), a metric of evaluating the fit of the astrometric solution described in the additional release notes, we apply a cut of ${\rm RUWE}<1.2$ to select out sources with high astrometric error or noise, such as binaries (see Figure \ref{fig:ruwe_cut}).
+\item \textbf{Astrometric cut:} Using the \gaia\ re-normalised unit weight error (RUWE), a metric of evaluating the fit of the astrometric solution described in the additional release notes, we apply a cut of ${\rm RUWE}<1.2$ to remove sources with high astrometric error or noise, such as binaries (see Figure \ref{fig:ruwe_cut}).
 
-\item \textbf{Binary cut:} To remove further contamination from binary sources by a conservative cut, we applied an additional color-magnitude selection on sources above the main-sequence with \cannon\ temperatures/metallicities that deviated from the gradient, which we visibly selected for in Figure \ref{fig:ruwe_cut}.
+\item \textbf{Binary cut:} To remove further contamination from binary sources, we applied an additional color-magnitude cut on sources above the main-sequence, which we visibly selected for in Figure \ref{fig:ruwe_cut}.
 
@@
 
-The three panels of \ref{fig:safe_selection} show before and after selection of the sources in \gaia\ color-magnitude space, colored by \thecannon\ determined temperatures, metallicities, and spectral types, with their respective training samples overplotted in orange.
+The three panels of Figure \ref{fig:safe_selection} show before and after selection of the sources in \gaia\ color-magnitude space, colored by \thecannon\ determined temperatures, metallicities, and spectral types, with their respective training samples overplotted in orange.
 Each plot shows the expected gradient: temperature increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly to the main sequence branch.

### 2019-08-01 paper/draft.tex
address rotational broadening bias

@@
 For the physical parameter model, we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
-We note for this model 5 out of 87 sources show possible rotational line broadening identified by visual inspection (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources show no obvious broadening.
-We note these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.
+We note for this model that 5 out of 87 sources show possible rotational line broadening identified by visual inspection (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources show no obvious broadening.
+We note these broadened sources have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), and have cross-validation labels biased by an average of $+65\,$K and $-0.08\,$dex, however removing them from the training sample does not significantly change the overall scatter and bias of the model (when removed the scatters are 79\,K/0.09\,dex and biases are +5\,K/+0.01\,dex).
+Hence we do not remove them from the training sample.
 

### 2019-08-01 paper/draft.tex
results revisions

@@
 	\begin{center}
-	% \includegraphics[width=\linewidth]{figures/self_test_spt.png}
 	\includegraphics[width=\linewidth]{figures/crv_test_spt.png}
@@
 \begin{deluxetable*}{cccccccc}
-\tablecaption{Metallicities of \apogee\ M dwarfs with wide binaries.
+\tablecaption{Metallicities of \apogee\ M dwarfs in wide binaries.
 Reported in this table are the ASPCAP [M/H] and \cannon\ [Fe/H] of the M dwarfs, as well as the metallicity values of their binary companions.
@@
 The three panels of Figure \ref{fig:safe_selection} show before and after selection of the sources in \gaia\ color-magnitude space, colored by \thecannon\ determined temperatures, metallicities, and spectral types, with their respective training samples overplotted in orange.
-Each plot shows the expected gradient: temperature increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly to the main sequence branch.
+Each plot shows the expected gradient: temperature increases with decreasing color, spectral subtype increases with increasing color, and the metallicity gradient is largely perpendicular to the main sequence branch.
 We also note that applying our model requires very little computational demand: the time to train and test a model on all 14,827 sources was two minutes on a 2.7\,GHz Intel core i7 laptop.
@@
 
-As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combination of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures).
-To obtain visual band magnitudes for a set of sources, we cross-matched the 5,875 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 1,702 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee.
+As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combinations of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures).
+To obtain visual band magnitudes for a set of sources, we cross-matched the 5,875 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 1,702 sources with both $BV$ photometries measured by APASS, and \zmass\ $JHK$ photometries from \apogee.
 Figure \ref{fig:teff_comparisons} shows \cannon\ vs.
@@
 
-Compared to the \citealt{Mann:2015} and \citealt{Boyajian:2012} color-metallicity derived temperatures, both ASPCAP and \cannon\ temperatures show similar scatters of $\sim60\,$K, but offset by a constant.
-We find \cannon\ to be in better agreement with \citealt{Mann:2015} and \citealt{Boyajian:2012}, with ASPCAP overestimating $\teff$ on average $\sim110-140\,$K, and \thecannon\ underestimating $\teff$ on average by $\sim10-20\,$K.
-\\
+Compared to the \citealt{Mann:2015} and \citealt{Boyajian:2012} color-metallicity derived temperatures, both ASPCAP and \cannon\ temperatures show similar scatters of $\sim60\,$K, but are offset by a constant.
+We find \cannon\ to be in better agreement with \citealt{Mann:2015} and \citealt{Boyajian:2012}, with ASPCAP overestimating $\teff$ on average $\sim110-140\,$K, and \thecannon\ underestimating $\teff$ on average by $\sim10-20\,$K, with the largest deviation at the lowest and highest $\teff$s in the latter.
 
@@
 As expected, the Toomre diagram in Figure \ref{fig:toomre} shows that higher metallicity sources in the sample are concentrated in low velocity space corresponding roughly to the thin disk population, while the thick disk population contains a higher concentration of lower metallicity sources.
-Separating the two populations into separate histograms (also shown in Figure \ref{fig:toomre}), we find that thick disk sources are nearly 1-$\sigma$ more metal poor than thin disk sources with the average thin disk metallicity being near-solar $\feh=0.00\pm0.17\,$dex, and the average metallicities of the thick disk stars being $\feh=-0.14\pm0.19\,$dex.
-On the other hand, metallicities from ASPCAP show no significant difference in distribution, with an average thin disk metallicity of $\mh=-0.16\pm0.16\,$dex and an average thick disk metallicity of $\mh=-0.23\pm0.17\,$dex.
+Separating the two populations into separate histograms (also shown in Figure \ref{fig:toomre}), we find that thick disk sources are more metal poor than thin disk sources, with the average metallicity of the thin disk stars being near-solar $\feh=0.00\pm0.17\,$dex, and the average metallicity of the thick disk stars being $\feh=-0.14\pm0.19\,$dex.
+Metallicities of the two populations from ASPCAP show a similar distribution, with an average thin disk metallicity of $\mh=-0.16\pm0.16\,$dex and an average thick disk metallicity of $\mh=-0.23\pm0.17\,$dex.
 

### 2019-08-01 paper/draft.tex
discussion revisions

@@
 We note for this model that 5 out of 87 sources show possible rotational line broadening identified by visual inspection (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources show no obvious broadening.
-We note these broadened sources have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), and have cross-validation labels biased by an average of $+65\,$K and $-0.08\,$dex, however removing them from the training sample does not significantly change the overall scatter and bias of the model (when removed the scatters are 79\,K/0.09\,dex and biases are +5\,K/+0.01\,dex).
+We note that these broadened sources have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), and have cross-validation labels biased by an average of $+65\,$K and $-0.08\,$dex, however removing them from the training sample does not significantly change the overall scatter and bias of the model (when removed the scatters are 79\,K/0.09\,dex and biases are +5\,K/+0.01\,dex).
 Hence we do not remove them from the training sample.
@@
 especially when the spectra are observed at signal-to-noise levels above 100.
-For another, it does not include some known physical and instrumental effects.
-It does not include a term or terms for line broadening from rotation or
-convection (for exmaple, \citealt{Behmard2019}).
-It does not include any accounting for binarity or superposition of multiple
+The model does not include some known physical and instrumental effects.
+It does not include terms for line broadening from rotation or
+convection (for example, \citealt{Behmard2019}), nor
+does it account for binarity or superposition of multiple
 stellar spectra (as in, say, \citealt{ElBadry:2018}).
-And it does not include any adjustments for instrumental variations, like for
+The model also does not include any adjustments for instrumental variations, like for
 example the small but significant variations of \apogee\ resolution with
-spectrograph fiber number (as are included in \citealt{Ness2018}).
+spectrograph fiber number (as included in \citealt{Ness2018}).
 
-In addition to all this, the \apogee\ instrument was designed to be sensitive to more than
-a dozen individual element abundances in stellar spectra!
+The \apogee\ instrument was designed to be sensitive to more than
+a dozen individual element abundances in stellar spectra.
 So the M-dwarf spectra analyzed here contain individual element abundance
-information that we have left on the table.
-Exploitation of that information might require better models of M dwarfs than
-we have at present.
-However, this is an important goal for the future with these data.
+information that we have ignored.
+Exploitation of that information requires a better training set of M dwarfs than
+we have at present, but is an important goal for the future with these data.
 
@@
 These results find that atmospheric metallicities are systematically metal-poor biased
-compared to \cannon-based metallicities trained on sources with metallicities callibrated
+compared to \cannon-based metallicities trained on sources with metallicities calibrated
 to those of FGK companions. 
@@
 and scatter of $\>0.1\,$dex at the lower temperature end ($\teff < 3300$\,K) (Figure \ref{fig:met_comp}).
-This metal-poor bias, while not explored to a great extent in this work, we suspect is due 
+We suspect that this metal-poor bias, while not explored to a great extent in this work, is due 
 to the line lists of the models: 
 an effect in which the optimizer of the pipeline may be lowering the metallicity of the fit 
-to better reproduce the spectrum to compensate for the missing lines or opacities.
+to compensate for the missing lines or opacities.
 Further analysis would need to be done to determine if the metallicity bias
@@
 To provide physical insights that drive physical understandings.
-It is our hope that \thecannon, and models like it, will, in the end, lead to
-new and improved physical models, which will, in turn, put \thecannon\ out of business. \\
+It is our hope that \thecannon, and models like it, will lead to
+new and improved physical models which will, in turn, put \thecannon\ out of business. \\
 
@@
 	\end{center}
-	\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities.
+	\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallicities.
 \textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.}

### 2019-08-01 paper/draft.tex
highlight needed changes

@@
 
-A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf (e.g.
+\color{gcolor}{A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf (e.g.
 \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}) to developed precise empirical relations (as good as $\sim$0.07\,dex).
 However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band nor information about the overall spectral shape often used to determine $\teff$ and spectral type.
-Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the greater spectral information provided by \apogee's resolution.
+Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the greater spectral information provided by \apogee's resolution.}\color{black}
 

### 2019-08-01 paper/draft.tex
add ruwe ref

@@
 
-\item \textbf{Astrometric cut:} Using the \gaia\ re-normalised unit weight error (RUWE), a metric of evaluating the fit of the astrometric solution described in the additional release notes, we apply a cut of ${\rm RUWE}<1.2$ to remove sources with high astrometric error or noise, such as binaries (see Figure \ref{fig:ruwe_cut}).
+\item \textbf{Astrometric cut:} Using the \gaia\ re-normalised unit weight error (RUWE), a metric of evaluating the fit of the astrometric solution described in the additional release notes \citep{Lindgren:2018}, we apply a cut of ${\rm RUWE}<1.2$ to remove sources with high astrometric error or noise, such as binaries (see Figure \ref{fig:ruwe_cut}).
 

### 2019-08-01 paper/draft.tex
unitallicize

@@
 
-\title{TEMPERATURES AND METALLICITIES OF M DWARFS IN THE \apogee\ SURVEY}
+\title{TEMPERATURES AND METALLICITIES OF M DWARFS IN THE APOGEE SURVEY}
 
@@
 Here we present a catalog of spectroscopic temperatures, metallicities, and
-spectral types for 5,875 M dwarfs in the \apogee+\drtwo\ surveys using
+spectral types for 5,875 M dwarfs in the APOGEE+\drtwo\ surveys using
 \thecannon: a flexible, data-driven spectral-modeling and
@@
 compared to a subsample of 15 FGK+M or M+M binaries. 
-Finally, our comparison between \cannon\ and \apogee\ pipeline (ASPCAP) labels finds that 
+Finally, our comparison between \cannon\ and APOGEE pipeline (ASPCAP) labels finds that 
 ASPCAP is systematically biased towards reporting higher temperatures and lower metallicities. 
@@
 Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey \citep[SDSS;][]{Eisenstein:2011,Blanton:2017}, and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope \citep[LAMOST;][]{Zhao:2012} have dramatically increased the sample of known M dwarfs \citep{West:2011,Guo:2015} with catalogs of over 70,000 sources, enabling studies of M dwarf abundances on a Galactic scale.
-The Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the SDSS III/IV mission has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}.
-\apogee\ pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have been determined to precisions 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
+The Apache Point Observatory Galactic Evolution Experiment (APOGEE; \citealt{Majewski:2015}) survey as part of the SDSS III/IV mission has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}.
+APOGEE pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have been determined to precisions 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
 
@@
 \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}) to developed precise empirical relations (as good as $\sim$0.07\,dex).
-However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band nor information about the overall spectral shape often used to determine $\teff$ and spectral type.
-Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the greater spectral information provided by \apogee's resolution.}\color{black}
+However these metallicity calibrations are inaccessible for analysis with APOGEE, as the wavelength coverage does not include important calibration indices in the $K$-band nor information about the overall spectral shape often used to determine $\teff$ and spectral type.
+Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the greater spectral information provided by APOGEE's resolution.}\color{black}
 
-In this work we build a data-driven model for M dwarf \apogee\ spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a,Behmard2019}: a fully empirical model, which employs no line lists or radiative transfer models.
+In this work we build a data-driven model for M dwarf APOGEE spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a,Behmard2019}: a fully empirical model, which employs no line lists or radiative transfer models.
 \thecannon\ is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:methods}).
@@
 Here we take M-dwarf labels from samples of well-characterized stars that are present in the
-\sdssiv\ \apogee\ sample, and use those labels to train a model and label all of the M dwarfs observed by
-\sdssiv\ \apogee. One set of labels are physical parameters (effective temperatures and metallicities), the other set of labels are spectral types.
+SDSS-IV APOGEE sample, and use those labels to train a model and label all of the M dwarfs observed by
+SDSS-IV APOGEE. One set of labels are physical parameters (effective temperatures and metallicities), the other set of labels are spectral types.
 
-This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the \apogee\ and \gaia\ surveys, as well as previous studies of M dwarfs in \apogee.
+This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the APOGEE and \gaia\ surveys, as well as previous studies of M dwarfs in APOGEE.
 Section \ref{sec:methods} describes our model implementation using \thecannon\ framework, and Section \ref{sec:sample_selection} describes the our sample selection and derivation of training parameters.
@@
 
-The \apogee\ survey is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
+The APOGEE survey is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
+Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
 For low temperatures ($2800<\teff<3500$\,K), the pipeline uses MARCS plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures ($\teff\geq3500$\,K) ATLAS9 \citep{Castelli:2004} plane-parallel models are used.
 
-\apogee\ is primarily designed to target bright stellar populations, particularly red giants, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013}, with the objective of studying galactic composition and evolution.
-However numerous cool, main sequence sources have also been observed either as targets proposed by the \apogee\ M dwarf ancilliary survey ($\sim$1,200 sources; \citealt{Desphande:2013}), or serendipitously.
+APOGEE is primarily designed to target bright stellar populations, particularly red giants, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013}, with the objective of studying galactic composition and evolution.
+However numerous cool, main sequence sources have also been observed either as targets proposed by the APOGEE M dwarf ancilliary survey ($\sim$1,200 sources; \citealt{Desphande:2013}), or serendipitously.
 
@@
 and \citealt{Skinner:2018} which identified and measured mass ratios and radial velocities for 44 M dwarf spectroscopic binaries.
-This work complements existing studies by producing model-independent catalog of spectroscopic temperatures and metallicites to test against model predictions for the entire \apogee\ M dwarf sample, which we quantify to contain at least 10,000 sources to date (DR14).
+This work complements existing studies by producing model-independent catalog of spectroscopic temperatures and metallicites to test against model predictions for the entire APOGEE M dwarf sample, which we quantify to contain at least 10,000 sources to date (DR14).
 
@@
 
-The spectral type training sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from \sdss.
+The spectral type training sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from SDSS.
 For each source in the catalog, spectral types were determined both through an automated routine for comparing spectral type templates to data using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
@@
 	\includegraphics[width=\linewidth]{figures/cannon_chi_dist.png}
-	\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources \citep{ElBadry:2018,Skinner:2018}; \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee --\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
+	\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources \citep{ElBadry:2018,Skinner:2018}; \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE--\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
 We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as ``safe''.}
@@
 	\includegraphics[width=\linewidth]{figures/cmd_selection.png}
-	\caption{\gaia\ and \zmass\ color-magnitude cuts for the12,037 sources with $\chi^2<100,000$.
+	\caption{\gaia\ and 2MASS color-magnitude cuts for the12,037 sources with $\chi^2<100,000$.
 Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia.
@@
 	$\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ corresponding to ($BP-RP$, $M_G$) for the \gaia\ color-magnitudes shown left, and 
-	$\left\{ (.7, 4.3), (1.1, 4.3), (.7, 9), (1.1, 9) \right\}$ corresponding to ($J-K$, $M_K$) for the \zmass\ color-magnitudes shown right. }
+	$\left\{ (.7, 4.3), (1.1, 4.3), (.7, 9), (1.1, 9) \right\}$ corresponding to ($J-K$, $M_K$) for the 2MASS color-magnitudes shown right. }
 	\label{fig:cmd_selection}
@@
 
-Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources.
+Out of the total APOGEE DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources.
 From there we applied additional selection criteria, described below, to identify a sample of single, main-sequence M stars, with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries:
@@
 
-\item \textbf{Color-magnitude cuts:} Using \gaia\ and \zmass\ photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to remove sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and multiples), and subdwarfs below the main sequence.
+\item \textbf{Color-magnitude cuts:} Using \gaia\ and 2MASS photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to remove sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and multiples), and subdwarfs below the main sequence.
 
@@
 \begin{deluxetable*}{cccccccc}
-\tablecaption{Metallicities of \apogee\ M dwarfs in wide binaries.
+\tablecaption{Metallicities of APOGEE M dwarfs in wide binaries.
 Reported in this table are the ASPCAP [M/H] and \cannon\ [Fe/H] of the M dwarfs, as well as the metallicity values of their binary companions.
@@
 
-As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combinations of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures).
-To obtain visual band magnitudes for a set of sources, we cross-matched the 5,875 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 1,702 sources with both $BV$ photometries measured by APASS, and \zmass\ $JHK$ photometries from \apogee.
+As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combinations of 2MASS and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures).
+To obtain visual band magnitudes for a set of sources, we cross-matched the 5,875 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 1,702 sources with both $BV$ photometries measured by APASS, and 2MASS $JHK$ photometries from APOGEE.
 Figure \ref{fig:teff_comparisons} shows \cannon\ vs.
@@
 As a check of our test sample metallicity reliability, we cross-matched our M dwarf final sample with the \citealt{Elbadry:2018b} catalog of $>$50,000 high-confidence, widely separated binaries identified by \drtwo.
-In total we found 216 of the \apogee\ M dwarfs to have binary pairs (46 FGK+M, 155 M+M, and 15 WD+M).
-Out of the 155 M+M pairs, 8 pairs contained both pairs in \apogee.
+In total we found 216 of the APOGEE M dwarfs to have binary pairs (46 FGK+M, 155 M+M, and 15 WD+M).
+Out of the 155 M+M pairs, 8 pairs contained both pairs in APOGEE.
 Cross-matching the list of FGK+M dwarf companions with several catalogs/surveys with measured stellar metallicities, we found an additional 7 sources with FGK metallicities from LAMOST \citep{Zhao:2012} and APOGEE (ASPCAP).
@@
 high-quality atmospheric parameters ($\teff$ and $\feh$) for M-type dwarf stars
-from high-resolution infrared spectra from \apogee.
+from high-resolution infrared spectra from APOGEE.
 This work was motivated by the problem that M dwarfs stars are difficult to
@@
 The model also does not include any adjustments for instrumental variations, like for
-example the small but significant variations of \apogee\ resolution with
+example the small but significant variations of APOGEE resolution with
 spectrograph fiber number (as included in \citealt{Ness2018}).
 
-The \apogee\ instrument was designed to be sensitive to more than
+The APOGEE instrument was designed to be sensitive to more than
 a dozen individual element abundances in stellar spectra.
@@
 Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}.
-These Tables summarize spectral features in the \apogee\ bandpass
+These Tables summarize spectral features in the APOGEE bandpass
 that are found to be strong temperature and metallicity indicators.
@@
 \facilities{
-	\sdssiv(\apogee),
+	SDSS-IV (APOGEE),
 	\gaia}
@@
 	\end{center}
-	\caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of \apogee\ spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
+	\caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
 \end{figure*}
@@
 	\startdata
-	{\tt APOGEE$\_$ID} 	& 	& \apogee\ \zmass\ designation \\
+	{\tt APOGEE$\_$ID} 	& 	& APOGEE 2MASS designation \\
 	{\tt GAIA$\_$ID} 		&	& \gaia\ identification number \\
@@
 	{\tt MASS} 			& M$_\odot$	& Mass determined by \citealt{Mann:2019} relation	\\
-	{\tt J} 				& mag	& \apogee\ $J$ band photometry	\\
-	{\tt J$\_$ERR} 		& mag	& \apogee\ $J$ band photometry uncertainty	\\
-	{\tt H} 				& mag	& \apogee\ $H$ band photometry	\\
-	{\tt H$\_$ERR} 		& mag	& \apogee\ $H$ band photometry uncertainty	\\
-	{\tt K} 				& mag	& \apogee\ $K$ band photometry	\\
-	{\tt K$\_$ERR} 		& mag	& \apogee\ $K$ band photometry uncertainty	\\
+	{\tt J} 				& mag	& APOGEE $J$ band photometry	\\
+	{\tt J$\_$ERR} 		& mag	& APOGEE $J$ band photometry uncertainty	\\
+	{\tt H} 				& mag	& APOGEE $H$ band photometry	\\
+	{\tt H$\_$ERR} 		& mag	& APOGEE $H$ band photometry uncertainty	\\
+	{\tt K} 				& mag	& APOGEE $K$ band photometry	\\
+	{\tt K$\_$ERR} 		& mag	& APOGEE $K$ band photometry uncertainty	\\
 	{\tt BP$\_$MAG }		& mag	& \gaia\ $BP$ band photometry	\\
@@
 	{\tt BP$\_$RP} 		& mag	& \gaia\ $BP-RP$ color	\\
-	{\tt RA} 				& deg	& \apogee\ radial ascension angle	\\
-	{\tt DEC} 			& deg	& \apogee\ declination angle	\\
+	{\tt RA} 				& deg	& APOGEE radial ascension angle	\\
+	{\tt DEC} 			& deg	& APOGEE declination angle	\\
 	{\tt PMRA} 			& mas yr$^{-1}$	& \gaia\ radial ascension proper motion	\\
@@
 	{\tt DIST} 			& pc	& Distance (1/$\varpi$)	\\
-	{\tt RV$\_$APOGEE} 	& km s$^{-1}$	& \apogee\ radial velocity	\\
-	{\tt RV$\_$APOGEE$\_$ERR} 	& km s$^{-1}$	& \apogee\ radial velocity uncertainty	\\
+	{\tt RV$\_$APOGEE} 	& km s$^{-1}$	& APOGEE radial velocity	\\
+	{\tt RV$\_$APOGEE$\_$ERR} 	& km s$^{-1}$	& APOGEE radial velocity uncertainty	\\
 	{\tt RV$\_$GAIA} 				& km s$^{-1}$	& \gaia\ radial velocity	\\

### 2019-08-01 paper/draft.tex
empirical relations

@@
 
-\color{gcolor}{A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf (e.g.
-\citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}) to developed precise empirical relations (as good as $\sim$0.07\,dex).
-However these metallicity calibrations are inaccessible for analysis with APOGEE, as the wavelength coverage does not include important calibration indices in the $K$-band nor information about the overall spectral shape often used to determine $\teff$ and spectral type.
-Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the greater spectral information provided by APOGEE's resolution.}\color{black}
+A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf \citep[e.g.][]{Terrien:2012,RojasAyala:2010,Newton:2014,Neves:2014,Lindgren:2016} to developed precise empirical relations (as good as $\sim$0.07\,dex).
+However these metallicity calibrations do not take advantage of the full wavelength coverage, nor information about the overall spectral shape often used to determine $\teff$ and spectral type.
+Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the greater spectral information provided by APOGEE's resolution.
 

### 2019-08-01 paper/draft.tex
fix ref

@@
 
-\item \textbf{Astrometric cut:} Using the \gaia\ re-normalised unit weight error (RUWE), a metric of evaluating the fit of the astrometric solution described in the additional release notes \citep{Lindgren:2018}, we apply a cut of ${\rm RUWE}<1.2$ to remove sources with high astrometric error or noise, such as binaries (see Figure \ref{fig:ruwe_cut}).
+\item \textbf{Astrometric cut:} Using the \gaia\ re-normalised unit weight error (RUWE), a metric of evaluating the fit of the astrometric solution described in the additional release notes \citep{Lindegren:2018}, we apply a cut of ${\rm RUWE}<1.2$ to remove sources with high astrometric error or noise, such as binaries (see Figure \ref{fig:ruwe_cut}).
 

### 2019-08-01 paper/draft.tex
explain significance in table

@@
 \begin{deluxetable}{ccccc}
-\tablecaption{Strong $\teff$-correlated molecular lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
+\tablecaption{Strong $\teff$-correlated molecular lines for all features with significance greater than $2\sigma_{\rm jackknife}$.
+`Significance' quantifies the number of standard deviations that the derivative of the line is significant to (with negative corresponding to negative derivatives).}
 \tablehead{\colhead{Species} & \colhead{Wavelength} & \colhead{Derivative} & \colhead{$\sigma_{\rm jackknife}$} & \colhead{Significance}}

### 2019-08-01 paper/draft.tex
fix table length issue

@@
 
+\startlongtable
 \begin{deluxetable}{ccccc}
-\tablecaption{Strong $\teff$-correlated molecular lines for all features with significance greater than $2\sigma_{\rm jackknife}$.
+\tablecaption{Strong $\teff$-correlated atomic and molecular lines for all features with significance greater than $2\sigma_{\rm jackknife}$.
 `Significance' quantifies the number of standard deviations that the derivative of the line is significant to (with negative corresponding to negative derivatives).}
+\label{table:teff_derivative}
 \tablehead{\colhead{Species} & \colhead{Wavelength} & \colhead{Derivative} & \colhead{$\sigma_{\rm jackknife}$} & \colhead{Significance}}
@@
 	OH & 16895.074 & 0.005 & 0.002 & 2.197 \\
-	\nodata & \nodata & \nodata & \nodata & \nodata
-	\enddata
-\end{deluxetable} \label{table:teff_derivative}
-
-\begin{deluxetable}{ccccc}
-\tablecaption{Strong $\teff$-correlated atomic lines for all features with significance greater than $2\sigma_{\rm jackknife}$.}
-\tablehead{\colhead{Species} & \colhead{Wavelength} & \colhead{Derivative} & \colhead{$\sigma_{\rm jackknife}$} & \colhead{Significance}}
-	\startdata
-	\nodata & \nodata & \nodata & \nodata & \nodata \\
 	Fe I & 15207.509 & -0.007 & 0.003 & -2.622 \\

### 2019-08-01 paper/draft.tex
fix plot annoyances

@@
 \begin{center}
-	\includegraphics[width=.45\linewidth]{figures/cmd_teff_full.png}
-	\includegraphics[width=.48\linewidth]{figures/cmd_teff_safe.png} \\
-	\includegraphics[width=.45\linewidth]{figures/cmd_feh_full.png}
-	\includegraphics[width=.48\linewidth]{figures/cmd_feh_safe.png} \\
-	\includegraphics[width=.45\linewidth]{figures/cmd_spt_full.png}
-	\includegraphics[width=.48\linewidth]{figures/cmd_spt_safe.png}
+	\includegraphics[width=.47\linewidth]{figures/cmd_teff_full.png}
+	\includegraphics[width=.47\linewidth]{figures/cmd_teff_safe.png} \\
+	\includegraphics[width=.47\linewidth]{figures/cmd_feh_full.png}
+	\includegraphics[width=.47\linewidth]{figures/cmd_feh_safe.png} \\
+	\includegraphics[width=.47\linewidth]{figures/cmd_spt_full.png}
+	\includegraphics[width=.47\linewidth]{figures/cmd_spt_safe.png}
 	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 5,875 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, spectroscopic binaries, and K dwarfs).

### 2019-08-01 paper/draft.tex
clarify metallicity dist description

@@
 Histogram (right) shows the metallicity distribution of thin and thick disk stars, with blue corresponding to $v_{tot}<70$\,km/s and green corresponding to $70<v_{tot}<180$\,km/s respectively.
-The average metallicity of the thin disk distribution is $\feh=0.00\pm0.17\,$dex, and the average metallicity of the thick disk distribution is $\feh=-0.14\pm0.19\,$dex.}
+The average/standard deviation metallicity of the thin disk distribution is $\feh=0.00\pm0.17\,$dex, and the average/standard deviation metallicity of the thick disk distribution is $\feh=-0.14\pm0.19\,$dex.}
 	\label{fig:toomre}
@@
 As expected, the Toomre diagram in Figure \ref{fig:toomre} shows that higher metallicity sources in the sample are concentrated in low velocity space corresponding roughly to the thin disk population, while the thick disk population contains a higher concentration of lower metallicity sources.
-Separating the two populations into separate histograms (also shown in Figure \ref{fig:toomre}), we find that thick disk sources are more metal poor than thin disk sources, with the average metallicity of the thin disk stars being near-solar $\feh=0.00\pm0.17\,$dex, and the average metallicity of the thick disk stars being $\feh=-0.14\pm0.19\,$dex.
-Metallicities of the two populations from ASPCAP show a similar distribution, with an average thin disk metallicity of $\mh=-0.16\pm0.16\,$dex and an average thick disk metallicity of $\mh=-0.23\pm0.17\,$dex.
+Separating the two populations into separate histograms (also shown in Figure \ref{fig:toomre}), we find that thick disk sources are more metal poor than thin disk sources, with the mean $\pm$ standard deviation being near-solar $\feh=0.00\pm0.17\,$dex for the thin disk distribution, and $\feh=-0.14\pm0.19\,$dex for the thick disk distribution.
+Metallicities of the two populations from ASPCAP show a similar distribution, with the mean $\pm$ standard deviation being $\mh=-0.16\pm0.16\,$dex for thin disk sources and $\mh=-0.23\pm0.17\,$dex for thick disk sources.
 
@@
 It is our hope that \thecannon, and models like it, will lead to
-new and improved physical models which will, in turn, put \thecannon\ out of business. \\
+new and improved physical models which will, in turn, put \thecannon\ out of business.
 

### 2019-08-21 paper/draft.tex
comment on M+M met scatter
gramatical fixes

@@
 
-M dwarfs have enourmous potential for understanding structure and
+M dwarfs have enormous potential for understanding structure and
 formation on both Galactic and exoplanetary scales
 through their properties and compositions.
-However, current atmospheric models
+However, current atmosphere models
 have limited ability to reproduce spectral features in stars at the 
@@
 Using a training sample of 87 M dwarfs with optically derived labels spanning 
-$2860 < \teff < 4130$\,K calibrated with bolometric temperatures, 
+$2860 < \teff < 4130$\,K calibrated with bolometric temperatures, and
 $-0.5 < \feh < 0.5$\,dex calibrated with FGK binary metallicities, 
@@
 Finally, our comparison between \cannon\ and APOGEE pipeline (ASPCAP) labels finds that 
-ASPCAP is systematically biased towards reporting higher temperatures and lower metallicities. 
+ASPCAP is systematically biased towards reporting higher temperatures and lower metallicities 
+for M dwarfs. 
 \end{abstract}
@@
 Low-mass stars, with masses $M_{*} < 0.7\,\msun$ and effective temperatures $T_{\rm eff} < 4000\,$K, are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the Galaxy's population by number \citep{Bochanski:2010}.
-With nuclear fusion timescales $\tau>10^{11}$ yr \citep{Laughlin:1997}, the chemical compositions of the M dwarf populations trace the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, and are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010,Woolf:2012}.
+With nuclear fusion timescales $\tau>10^{11}$ yr \citep{Laughlin:1997}, the chemical compositions of the M dwarf population traces the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, and are a unique probe for piecing together Galactic structure and evolution \citep{Bochanski:2010,Woolf:2012}.
 
@@
 For these reasons, M dwarfs are primary candidates for exoplanet searches, including by the NASA \kepler\ \citep[e.g.,][]{Dressing2015} and Transiting Exoplanet Survey Satellite (\tess) \citep[e.g.,][]{Muirhead:2018} missions.
-As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with the James Webb Space Telescope ({\sl JWST;} \citealt{Clampin2008}).
+As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with the James Webb Space Telescope ({\sl JWST;} \citealt{Clampin2008}).
 
-Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey \citep[SDSS;][]{Eisenstein:2011,Blanton:2017}, and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope \citep[LAMOST;][]{Zhao:2012} have dramatically increased the sample of known M dwarfs \citep{West:2011,Guo:2015} with catalogs of over 70,000 sources, enabling studies of M dwarf abundances on a Galactic scale.
-The Apache Point Observatory Galactic Evolution Experiment (APOGEE; \citealt{Majewski:2015}) survey as part of the SDSS III/IV mission has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}.
-APOGEE pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have been determined to precisions 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.
+Advances in instrumentation and the implementation of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey \citep[SDSS;][]{Eisenstein:2011,Blanton:2017} and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope \citep[LAMOST;][]{Zhao:2012}, have dramatically increased the sample of known M dwarfs \citep{West:2011,Guo:2015} with spectroscopic catalogs of over 70,000 sources, enabling studies of M dwarf abundances on a Galactic scale.
+The Apache Point Observatory Galactic Evolution Experiment (APOGEE; \citealt{Majewski:2015}) survey, as part of the SDSS III/IV mission, has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}.
+APOGEE pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have been determined to precisions 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmosphere models \citep{Schmidt:2016}.
 
-Elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precision (down to $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by improvements in atmospheric models, including realistic 3D local thermodynamic equilibrium (LTE) assumptions \citep{Asplund2005}, and differential abundance techniques using equivalent widths \citep{Bedell2014}.
-However, the determination of precise metallicities for M dwarfs has remained a long-standing challenge due to the formation of complex diatomic and triatomic molecular bands at M dwarf temperatures, particularly TiO and VO bands dominating the optical spectrum, H$_2$O and CO bands in the infrared, and hydride bands (FeH, CaH, CrH, MgH, etc.) present in the spectra of the latest spectral types \citep{Allard1997}.
+Elemental abundance measurements from high resolution spectra of F, G, and K stars have achieved extremely high precision (down to $0.01-0.03$\,dex; \citealt{Nissen2018}) enabled by improvements in atmosphere models including realistic 3D local thermodynamic equilibrium (LTE) assumptions \citep{Asplund2005}, and differential abundance techniques using equivalent widths \citep{Bedell2014}.
+However, the determination of precise metallicities for M dwarfs has remained a long-standing challenge due to the formation of diatomic and triatomic molecules at M dwarf temperatures, with absorption from TiO and VO in the optical, H2O and CO in the infrared, and hydrides (FeH, CaH, CrH, MgH, etc.) present in the spectra of the latest spectral types \citep{Allard1997}.
 Atmospheric models often fail to reproduce these spectral features \citep[e.g.,][]{Mann:2013c} due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions, and the absence of a clear continuum, all contribute to make it difficult to deconvolve individual features and extract line strengths from equivalent widths.
-The combination of these effects limit our ability to explore the information content of high resolution spectra using traditional comparison to atmosphere models.
+The combination of these effects limit our ability to explore the information content of high resolution spectra using traditional methods.
 
@@
 The metallicity measurements for the 15 M dwarf binaries and their companions are given in Table \ref{table:binaries}, and the overall scatter is $0.08\,$dex--an improvement over the scatter of ASPCAP metallicities which is $0.15\,$dex for these 15 sources.
+Internal consistency of the two models (the scatter of the 8 M+M pairs both in APOGEE) are 0.06\,dex and 0.12\,dex for \cannon\ and ASPCAP respectively.
 
@@
 	\includegraphics[width=16cm]{figures/demo_derivatives_teff1.png} 
+	\end{center}
+\end{figure*}
+
+\begin{figure*}[]
+	\begin{center}
 	\includegraphics[width=16cm]{figures/demo_derivatives_teff2.png}
@@
 	\includegraphics[width=16cm]{figures/demo_derivatives_feh1.png}
+	\end{center}
+\end{figure*}
+
+\begin{figure*}[]
+	\begin{center}
 	\includegraphics[width=16cm]{figures/demo_derivatives_feh2.png}

### 2019-08-21 paper/draft.tex
more grammar things

@@
 However, the determination of precise metallicities for M dwarfs has remained a long-standing challenge due to the formation of diatomic and triatomic molecules at M dwarf temperatures, with absorption from TiO and VO in the optical, H2O and CO in the infrared, and hydrides (FeH, CaH, CrH, MgH, etc.) present in the spectra of the latest spectral types \citep{Allard1997}.
-Atmospheric models often fail to reproduce these spectral features \citep[e.g.,][]{Mann:2013c} due to incomplete line lists and opacities, and the presence of millions of weak, blended transitions, and the absence of a clear continuum, all contribute to make it difficult to deconvolve individual features and extract line strengths from equivalent widths.
+Atmospheric models often fail to reproduce these spectral features \citep[e.g.,][]{Mann:2013c} due to incomplete line lists and opacities.
+The presence of millions of weak, blended transitions, and the absence of a clear continuum, contribute to make it difficult to deconvolve individual features and extract line strengths from equivalent widths.
 The combination of these effects limit our ability to explore the information content of high resolution spectra using traditional methods.
@@
 A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf \citep[e.g.][]{Terrien:2012,RojasAyala:2010,Newton:2014,Neves:2014,Lindgren:2016} to developed precise empirical relations (as good as $\sim$0.07\,dex).
-However these metallicity calibrations do not take advantage of the full wavelength coverage, nor information about the overall spectral shape often used to determine $\teff$ and spectral type.
-Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the greater spectral information provided by APOGEE's resolution.
+However these metallicity calibrations do not take advantage of the full wavelength coverage available, nor information about the overall spectral shape often used to determine $\teff$ and spectral type.
+Furthermore, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the greater spectral information provided by APOGEE's resolution.
 
-In this work we build a data-driven model for M dwarf APOGEE spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a,Behmard2019}: a fully empirical model, which employs no line lists or radiative transfer models.
+In this work we build a data-driven model for M dwarf APOGEE spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a,Behmard2019}: a fully empirical model which employs no line lists or radiative transfer models.
 \thecannon\ is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:methods}).
@@
 or if there are spectral models or techniques that work in one wavelength range or resolution but not in another.
-Existing methods to model M dwarf spectra in the NIR at high resolution are computationally expensive, calibrated over a narrow range of $\teff$ and metallicity, and/or focus on the $J$-band \citep{Lindgren:2016, Lindgren:2017}.
+Existing methods to model M dwarf spectra in the NIR at high resolution are computationally expensive, calibrated over a narrow range of $\teff$ and metallicity, and/or focused on the $J$-band \citep{Lindgren:2016, Lindgren:2017}.
 \thecannon\ thus fills this niche: it doesn't require using specific lines or opacity information that may be missing from the models; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
@@
 This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the APOGEE and \gaia\ surveys, as well as previous studies of M dwarfs in APOGEE.
-Section \ref{sec:methods} describes our model implementation using \thecannon\ framework, and Section \ref{sec:sample_selection} describes the our sample selection and derivation of training parameters.
-In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$5,800 sources, and examine the validity of our parameters against color-temperature relations and metallicities of binary pairs.
+Section \ref{sec:methods} describes our model implementation using \thecannon\ framework, and Section \ref{sec:sample_selection} describes our sample selection and derivation of training parameters.
+In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of nearly 6,000 sources, and examine the validity of our parameters against color-temperature relations and metallicities of binary pairs.
 Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results.
@@
 The APOGEE survey is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
-Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).
+Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters \citep{Meszaros:2012}.
 For low temperatures ($2800<\teff<3500$\,K), the pipeline uses MARCS plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures ($\teff\geq3500$\,K) ATLAS9 \citep{Castelli:2004} plane-parallel models are used.
@@
 A number of studies out of the M dwarf ancilliary survey have already been conducted to measure reliable fundamental atmospheric parameters and kinematic measurements using spectral synthesis of atmospheric model grids.
-These studies include \citealt{Desphande:2013} and \citealt{Gilhool:2018} which have studied the radial and rotational kinematics for 700+ sources;
-\citealt{Souto:2017} and \citealt{Souto:2018} which have modeled three exoplanet-hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining $\teff$/$\logg$/metallicity + 13 elemental abundances;
-\citealt{Rajpurohit:2018} which tested BT-Settl \citep{Allard:2012} and MARCS \citep{Gustafsson:2008} model grids on 45 M dwarfs to estimate $\teff$/$\logg$/metallicity;
-and \citealt{Skinner:2018} which identified and measured mass ratios and radial velocities for 44 M dwarf spectroscopic binaries.
+These studies include \citet{Desphande:2013} and \citet{Gilhool:2018} which have studied the radial and rotational kinematics for 700+ sources;
+\citet{Souto:2017} and \citet{Souto:2018} which have modeled three exoplanet-hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining $\teff$/$\logg$/metallicity + 13 elemental abundances;
+\citet{Rajpurohit:2018} which tested BT-Settl \citet{Allard:2012} and MARCS \citet{Gustafsson:2008} model grids on 45 M dwarfs to estimate $\teff$/$\logg$/metallicity;
+and \citet{Skinner:2018} which identified and measured mass ratios and radial velocities for 44 M dwarf spectroscopic binaries.
 This work complements existing studies by producing model-independent catalog of spectroscopic temperatures and metallicites to test against model predictions for the entire APOGEE M dwarf sample, which we quantify to contain at least 10,000 sources to date (DR14).
@@
 We use the last dataset for our study.
-In previous work it has been recommended to use an alternative pseudo-continuum normalization \citep{Ness:2015}, but we don't find obvious issues with the normalization in our analysis, so we retain the survey pipeline outputs.
+In previous work it has been recommended to use an alternative pseudo-continuum normalization \citep{Ness:2015}, but we didn't find obvious issues with the normalization in our analysis, so we retain the survey pipeline outputs.
 
@@
 
-\thecannon\ is a regression model which relies on two assumptions: first that sources with identical labels have near-identical flux at each wavelength pixel, and second that the expected flux at each pixel varies continuosly with change in label.
+\thecannon\ is a regression model which relies on two assumptions: first, that sources with identical labels have near-identical flux at each wavelength pixel; and second, that the expected flux at each pixel varies continuosly with change in label.
 
-Inferring the label of a star with such a model requires two steps: first the \emph{training step} in which a generative model describing the probability density function (pdf) of the flux is constructed at each pixel from the set of spectra with known reference labels, and second the \emph{test step} in which the model is applied to determine the labels of a spectrum.
+Inferring the label of a star with such a model requires two steps: first, the \emph{training step} in which a generative model describing the probability density function (pdf) of the flux is constructed at each pixel from the set of spectra with known reference labels; and second, the \emph{test step} in which the model is applied to determine the labels of a spectrum.
 
-Following the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector $\ell$ and and coefficient vector \emph{$\theta$}.
+Following the procedure of \citet{Ness:2015} and \citet{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector $\ell$ and and coefficient vector \emph{$\theta$}.
 For each star \emph{n}, at wavelength pixel \emph{$\lambda$}, we assume that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
@@
 \thecannon\ model can in principle be trained on any physical or empirical labels available, giving a wide range of flexibility to the model.
-However choosing a training sample with high quality labels is critical to the performance of \thecannon.
+However choosing a training sample with high quality labels is critical to its performance.
 Limitations of \thecannon\ include that test (output) labels are only accurate if the training labels are accurate, and only precise if the training labels are measured consistently across the training sample.
@@
 
-The spectral type training sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from SDSS.
+The spectral type training sample consists of 51 sources, spanning M0-M9 cross-matched from the \citet{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from SDSS.
 For each source in the catalog, spectral types were determined both through an automated routine for comparing spectral type templates to data using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
@@
 
-The physical parameter training sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 < \feh < 0.49$\,dex, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
+The physical parameter training sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 < \feh < 0.49$\,dex, 41 sources of which are drawn from \citet{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
 The major difference between the extension sample is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample.
@@
 A subsample of 29 sources with measured angular diameters from long-baseline optical interferometry \citep{Boyajian:2012} are used to calibrate the model comparison, including masking of spectral regions poorly reproduced by the model spectra \citep{Mann:2013c}.
-Based on difference between assigned $\teff$ values and those from angular diameters, absolute uncertainty on $\teff$ is estimated to be 60K in $\teff$, although the relative uncertainty is likely a factor of $\simeq$2 better.
+Based on the difference between assigned $\teff$ values and those from angular diameters, absolute uncertainty on $\teff$ is estimated to be 60K in $\teff$, although the relative uncertainty is likely a factor of $\simeq$2 better.
 
@@
 We note for this model that 5 out of 87 sources show possible rotational line broadening identified by visual inspection (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources show no obvious broadening.
-We note that these broadened sources have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), and have cross-validation labels biased by an average of $+65\,$K and $-0.08\,$dex, however removing them from the training sample does not significantly change the overall scatter and bias of the model (when removed the scatters are 79\,K/0.09\,dex and biases are +5\,K/+0.01\,dex).
+We note that these broadened sources have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), and that the labels for these 5 sources arep biased by an average of $+65\,$K and $-0.08\,$dex.
+However, removing them from the training sample does not significantly change the overall scatter and bias of the model.
+For the model overall, the cross-validation bias is +4K/+0.008 dex with the rapid rotators included in the training set, and the bias is +5K/+0.01 dex when the rapid rotators are not included.
 Hence we do not remove them from the training sample.
@@
 Another mode of analysis we can look at with \thecannon\ is how the derivative of the model changes with respect to given training parameters, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters.
-The top two panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh} show two example spectra and model fits for two different temperatures (\ref{fig:demo_teff}), and two different temperatures (\ref{fig:demo_feh}), with atomic and molecular features identified by the abundance analysis of \citealt{Souto:2017}.
+The top two panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh} show two example spectra and model fits for two different temperatures (Figure \ref{fig:demo_teff}), and two different metallicities (Figure \ref{fig:demo_feh}), with atomic and molecular features identified by the abundance analysis of \citet{Souto:2017}.
 The bottom panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh} show the derivative of flux with respect to temperature and metallicity at each pixel, taken at the median training values.
@@
 \end{equation}
-where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample, indexed by \emph{n}, $\theta$ is the coeffient vector trainined on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.
+where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample indexed by \emph{n}, $\theta$ is the coeffient vector trained on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.
 A summary of identified lines with derivative values greater than $2\sigma_{\rm jackkife}$ is given in Table \ref{table:teff_derivative} and Table \ref{table:feh_derivative}.
@@
 They are much higher.
-This discrepancy flows from the fact that---while the spectral model is good at the percent
-level---the signal-to-noise of the typical spectra are more than 100.
+This discrepancy flows from the fact that, while the spectral model is good at the percent
+level, the signal-to-noise of the typical spectra are more than 100.
 That is, the $\chi^2$ values do show that the model is not good in the frequentist
@@
 We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
-As seen in Figure \ref{fig:west_validation}, the model performs poorly at reproducing spectral types $>$M8, which confirms that \thecannon\ does not extraploate well to labels outside of the training sample space.
+As seen in Figure \ref{fig:west_validation}, the model performs poorly at reproducing spectral types $>$M8, which confirms that \thecannon\ does not extrapolate well to labels outside of the training sample space.
 Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows LOOCV test for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.
@@
 	\includegraphics[width=\linewidth]{figures/cmd_selection.png}
-	\caption{\gaia\ and 2MASS color-magnitude cuts for the12,037 sources with $\chi^2<100,000$.
+	\caption{\gaia\ and 2MASS color-magnitude cuts for the 12,037 sources with $\chi^2<100,000$.
 Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia.
-The coordinates for the selected quadrangles
-	$\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ corresponding to ($BP-RP$, $M_G$) for the \gaia\ color-magnitudes shown left, and 
-	$\left\{ (.7, 4.3), (1.1, 4.3), (.7, 9), (1.1, 9) \right\}$ corresponding to ($J-K$, $M_K$) for the 2MASS color-magnitudes shown right. }
+The coordinates for the selected quadrangles are
+	$\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ corresponding to ($BP-RP$, $M_G$) for the \gaia\ color-magnitudes shown in the left panel, and 
+	$\left\{ (.7, 4.3), (1.1, 4.3), (.7, 9), (1.1, 9) \right\}$ corresponding to ($J-K$, $M_K$) for the 2MASS color-magnitudes shown in the right panel. }
 	\label{fig:cmd_selection}
@@
 
-\item \textbf{Model extraploation cuts:} Because \thecannon\ does not perform well under extrapolated regions of parameter space, we select only sources inside the range of our training sample with ASPCAP parameters $2800<T_{\rm eff}<4100$ and $-0.5<\mh<0.5$, and with \cannon\ parameters of $2850<T_{\rm eff}<4150$, $-0.5<\feh<0.5$, and $0<SPT<9$.
+\item \textbf{Model extrapolation cuts:} Because \thecannon\ does not perform well under extrapolated regions of parameter space, we select only sources inside the range of our training sample with ASPCAP parameters $2800<T_{\rm eff}<4100$ and $-0.5<\mh<0.5$, and with \cannon\ parameters of $2850<T_{\rm eff}<4150$, $-0.5<\feh<0.5$, and $0<SPT<9$.
 
@@
 The right panel shows the same test sample reduced to 6,221 after applying an astrometric quality cut of ${\rm RUWE}<1.2$ (Selection 4).
-To further remove sources which were likely binary contamination (Selection 5), we cut out sources above the red line which sparsely lay above the majority of the main-sequence with temperatures and metallicities that deviated from the expected gradient, reducing the final test sample to a size of 5,875 sources.
-The red line shown is constrained by the coordinates $\left\{ (3.5,12), (1.87,7.5) \right\}$ corresponding to ($BP-RP$, $M_G$).}
+To further remove sources which were likely binary contamination (Selection 5), we cut out sources above the red line which sparsely lay above the majority of the main-sequence with temperatures and metallicities that deviate from the expected gradient, reducing the final test sample to a size of 5,875 sources.
+The red line shown is constrained by the ($BP-RP$, $M_G$) coordinates $\left\{ (3.5,12), (1.87,7.5) \right\}$.}
 	\label{fig:ruwe_cut}
@@
 
-Compared to the \citealt{Mann:2015} and \citealt{Boyajian:2012} color-metallicity derived temperatures, both ASPCAP and \cannon\ temperatures show similar scatters of $\sim60\,$K, but are offset by a constant.
-We find \cannon\ to be in better agreement with \citealt{Mann:2015} and \citealt{Boyajian:2012}, with ASPCAP overestimating $\teff$ on average $\sim110-140\,$K, and \thecannon\ underestimating $\teff$ on average by $\sim10-20\,$K, with the largest deviation at the lowest and highest $\teff$s in the latter.
+Compared to the \citet{Mann:2015} and \citet{Boyajian:2012} color-metallicity derived temperatures, both ASPCAP and \cannon\ temperatures show similar scatters of $\sim60\,$K, but are offset by a constant.
+We find \cannon\ to be in better agreement with \citet{Mann:2015} and \citet{Boyajian:2012}, with ASPCAP overestimating $\teff$ on average $\sim110-140\,$K, and \thecannon\ underestimating $\teff$ on average by $\sim10-20\,$K, with the largest deviation in the latter at the lowest and highest Teffs.
 
@@
 
-As a check of our test sample metallicity reliability, we cross-matched our M dwarf final sample with the \citealt{Elbadry:2018b} catalog of $>$50,000 high-confidence, widely separated binaries identified by \drtwo.
+As a check of our test sample metallicity reliability, we cross-matched our M dwarf final sample with the \citet{Elbadry:2018b} catalog of $>$50,000 high-confidence, widely separated binaries identified by \drtwo.
 In total we found 216 of the APOGEE M dwarfs to have binary pairs (46 FGK+M, 155 M+M, and 15 WD+M).
@@
 
-As expected, the Toomre diagram in Figure \ref{fig:toomre} shows that higher metallicity sources in the sample are concentrated in low velocity space corresponding roughly to the thin disk population, while the thick disk population contains a higher concentration of lower metallicity sources.
-Separating the two populations into separate histograms (also shown in Figure \ref{fig:toomre}), we find that thick disk sources are more metal poor than thin disk sources, with the mean $\pm$ standard deviation being near-solar $\feh=0.00\pm0.17\,$dex for the thin disk distribution, and $\feh=-0.14\pm0.19\,$dex for the thick disk distribution.
+As expected, the Toomre diagram in Figure \ref{fig:toomre} shows that higher metallicity sources in the sample are concentrated in low velocity space corresponding roughly to the thin disk population; while the thick disk population contains a higher concentration of lower metallicity sources.
+Separating the two populations into separate histograms (also shown in Figure \ref{fig:toomre}), we find that thick disk sources are more metal poor than thin disk sources, with the mean $\pm$ standard deviation of $\feh=0.00\pm0.17\,$dex for the thin disk distribution, and $\feh=-0.14\pm0.19\,$dex for the thick disk distribution.
 Metallicities of the two populations from ASPCAP show a similar distribution, with the mean $\pm$ standard deviation being $\mh=-0.16\pm0.16\,$dex for thin disk sources and $\mh=-0.23\pm0.17\,$dex for thick disk sources.
@@
 Figure \ref{fig:met_comp} shows that ASPCAP metallicities are systematically more metal-poor than \cannon\ metallicities.
-We furthermore find that the bias is temperature-dependent: at the hottest temperatures ($\teff>3600\,$K) ASPCAP and \cannon\ metallicities are consistent to a scatter of $0.05-0.06\,$dex and offset by an average of $-0.12-0.15\,$dex, while at the lowest temperatures ($\teff<3200\,$K) ASPCAP and \cannon\ are consistent to a scatter of $\sim-0.13\,$dex and offset by an average of $-0.3\,$dex.
+We further find that the bias is temperature-dependent: at the hottest temperatures ($\teff>3600\,$K) ASPCAP and \cannon\ metallicities are consistent to a scatter of $0.05-0.06\,$dex and offset by an average of $-0.12-0.15\,$dex, while at the lowest temperatures ($\teff<3200\,$K) ASPCAP and \cannon\ are consistent to a scatter of $\sim-0.13\,$dex and offset by an average of $-0.3\,$dex.
 
@@
 Temperature gradient increases with decreasing color, spectral subtype increases with increasing color, and metallicity gradient increases perpendicularly up from the main sequence branch as expected.
-Deviations from these gradients seen at the upper slice of the main sequence is likely remaining contamination from the binary sequence.}
+Deviations from these gradients seen at the upper boundary of the main sequence is likely remaining contamination from the binary sequence.}
 	\label{fig:safe_selection}
@@
 model physically; the data are better than the models in important senses.
-And indeed we find that our data-driven model is both accurate in the
+Indeed we find that our data-driven model is both accurate in the
 data domain (as a spectral synthesis model) and precise in the latent domain
@@
 \thecannon\ (\citealt{Ness:2015, Casey:2016, Ho:2017a, Ness:2018}), but
-extended to a new regime in spectral type ($\teff$) and surface
-gravity.
+here extends to a new regime in spectral type ($\teff$).
 The primary result of this work is that
@@
 representation of the training sample.
-That is, being a supervised method, \thecannon\ is never any better in a mean (bias) sense
-(although it can be better in a precision or variance sense) than the input training
-data.
+That is, being a supervised method, \thecannon\ is never any better in a mean (bias) sense than the input training data, although it can be better in a precision or variance sense.
 The catalog we have produced is a label transfer from parameters provided
-in our input data (\citealt{Mann:2015}; M15) and it implicitly adopts all the biases and
+in our input data (M15) and it implicitly adopts all the biases and
 issues from that input data.
 It is also limited to the stellar-parameter domain of that input catalog.
-That said, what we have done does also represent some kind of external validation of the M15
-stellar parameters:
-The success of our model shows that they do a great job of parameterizing a generative
-model of the spectra.
+That said, this work provides and external validation of the M15 stellar parameters.
 
-The model we have developed does have limitations however.
-For example, it delivers chi-squared values---goodness-of-fit measures---that
+The model we have developed does have limitations, however.
+For example, it delivers chi-squared goodness-of-fit measures that
 are large; the model is not technically an accurate description of the spectra,
@@
 The model does not include some known physical and instrumental effects.
-It does not include terms for line broadening from rotation or
-convection (for example, \citealt{Behmard2019}), nor
-does it account for binarity or superposition of multiple
+such as line broadening from rotation or
+convection (for example, \citealt{Behmard2019}), 
+or binarity and the superposition of multiple
 stellar spectra (as in, say, \citealt{ElBadry:2018}).
-The model also does not include any adjustments for instrumental variations, like for
-example the small but significant variations of APOGEE resolution with
+The model also does not include any adjustments for instrumental variations, like 
+the small but significant variations of APOGEE resolution with
 spectrograph fiber number (as included in \citealt{Ness2018}).

### 2019-08-21 paper/draft.tex
report spt <=M6

@@
 We also train a one-dimensional spectral classification model using 51 M dwarfs
-with SDSS optical spectral types ranging from M0 to M9, to predictive accuracy of
-0.9 types. 
+with SDSS optical spectral types ranging from M0 to M6, to predictive accuracy of
+0.7 types. 
 We find \cannon\ temperatures to be in agreement to within 60\,K
@@
 
-We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
+We trained \thecannon\ on 51 M dwarfs ranging M0-M9 with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
 We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
 As seen in Figure \ref{fig:west_validation}, the model performs poorly at reproducing spectral types $>$M8, which confirms that \thecannon\ does not extrapolate well to labels outside of the training sample space.
+Because of this skew for late-type sources, we report our spectral type model to be precise to $\pm$0.7 spectral types for the range M0-M6.
 Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows LOOCV test for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.

### 2019-08-21 paper/draft.tex
add uncertainties to tables

@@
 As expected, the Toomre diagram in Figure \ref{fig:toomre} shows that higher metallicity sources in the sample are concentrated in low velocity space corresponding roughly to the thin disk population; while the thick disk population contains a higher concentration of lower metallicity sources.
-Separating the two populations into separate histograms (also shown in Figure \ref{fig:toomre}), we find that thick disk sources are more metal poor than thin disk sources, with the mean $\pm$ standard deviation of $\feh=0.00\pm0.17\,$dex for the thin disk distribution, and $\feh=-0.14\pm0.19\,$dex for the thick disk distribution.
+Separating the two populations into separate histograms (also shown in Figure \ref{fig:toomre}), we find that thick disk sources are marginally more metal poor than thin disk sources, with the mean $\pm$ standard deviation of $\feh=0.00\pm0.17\,$dex for the thin disk distribution, and $\feh=-0.14\pm0.19\,$dex for the thick disk distribution.
 Metallicities of the two populations from ASPCAP show a similar distribution, with the mean $\pm$ standard deviation being $\mh=-0.16\pm0.16\,$dex for thin disk sources and $\mh=-0.23\pm0.17\,$dex for thick disk sources.
@@
 \begin{deluxetable*}{ccc|ccc|ccc|c}
-\tablecaption{Cannon results for Temperature/Metallicity model; see online journal for full table. \label{table:mann_results}}
+\tablecaption{\cannon\ results for Temperature/Metallicity model; see online journal for full table.
+Reported uncertainties for the M15 training sample labels are $\pm$60\,K/0.08\,dex, and reported uncertainties for \thecannon\ model based on the cross-validation scatter are $\pm$77\,K/0.09\,dex.
+\label{table:mann_results}}
 \tabletypesize{\scriptsize}
@@
 \begin{deluxetable*}{ccc|ccc|c}
-\tablecaption{Cannon results for Spectral Type model. \label{table:west_results}}
+\tablecaption{Cannon results for Spectral Type model.
+Reported uncertainties for the W11 training sample labels are $\pm$1 spectral types, and reported uncertainties for \thecannon\ model based on the cross-validation scatter are $\pm$0.7 spectral types for types $\leq$M6.
+\label{table:west_results}}
 \tabletypesize{\scriptsize}

### 2019-08-21 paper/draft.tex
report consistent scatters

@@
 or if there are spectral models or techniques that work in one wavelength range or resolution but not in another.
-Existing methods to model M dwarf spectra in the NIR at high resolution are computationally expensive, calibrated over a narrow range of $\teff$ and metallicity, and/or focused on the $J$-band \citep{Lindgren:2016, Lindgren:2017}.
+Existing methods to model M dwarf spectra in the NIR at high resolution are computationally expensive, often calibrated over a narrow range of Teff and/or metallicity.
 \thecannon\ thus fills this niche: it doesn't require using specific lines or opacity information that may be missing from the models; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.
@@
 to those of FGK companions. 
-At the high temperature end ($\teff > 3600$\,K), the ASPCAP metallicity bias is $\sim-0.15\,$dex with a
-scatter of $\sim0.06\,$dex relative to \cannon\ metallicities, and increases to a bias of $\>-0.26\,$dex 
-and scatter of $\>0.1\,$dex at the lower temperature end ($\teff < 3300$\,K) (Figure \ref{fig:met_comp}).
+At the high temperature end ($\teff > 3600$\,K), the ASPCAP metallicity bias is $-0.12-0.15\,$dex with a scatter of $0.05-0.06\,$dex relative to \cannon\ metallicities, and increases to a bias of $-0.3\,$dex and scatter of $0.13\,$dex at the lower temperature end ($\teff < 3200$\,K) (Figure \ref{fig:met_comp}).
 We suspect that this metal-poor bias, while not explored to a great extent in this work, is due 
 to the line lists of the models: 
-an effect in which the optimizer of the pipeline may be lowering the metallicity of the fit 
-to compensate for the missing lines or opacities.
+an effect in which the optimizer of the pipeline may be lowering the continuum level and metallicity of the fit to compensate for the missing lines or opacities.
 Further analysis would need to be done to determine if the metallicity bias

### 2019-08-22 paper/draft.tex
tweaks

@@
 Compared to the \citet{Mann:2015} and \citet{Boyajian:2012} color-metallicity derived temperatures, both ASPCAP and \cannon\ temperatures show similar scatters of $\sim60\,$K, but are offset by a constant.
-We find \cannon\ to be in better agreement with \citet{Mann:2015} and \citet{Boyajian:2012}, with ASPCAP overestimating $\teff$ on average $\sim110-140\,$K, and \thecannon\ underestimating $\teff$ on average by $\sim10-20\,$K, with the largest deviation in the latter at the lowest and highest Teffs.
+We find \cannon\ to be in better agreement with \citet{Mann:2015} and \citet{Boyajian:2012}, with ASPCAP overestimating $\teff$ on average $\sim110-140\,$K, and \thecannon\ underestimating $\teff$ on average by $\sim10-20\,$K, with the largest deviation in the latter at the lowest and highest $\teff$s.
 
@@
 
-As expected, the Toomre diagram in Figure \ref{fig:toomre} shows that higher metallicity sources in the sample are concentrated in low velocity space corresponding roughly to the thin disk population; while the thick disk population contains a higher concentration of lower metallicity sources.
+As expected, the Toomre diagram in Figure \ref{fig:toomre} shows that higher metallicity sources in the sample are concentrated in low velocity space corresponding roughly to the thin disk population; while the thick disk population contains a slightly higher concentration of lower metallicity sources.
 Separating the two populations into separate histograms (also shown in Figure \ref{fig:toomre}), we find that thick disk sources are marginally more metal poor than thin disk sources, with the mean $\pm$ standard deviation of $\feh=0.00\pm0.17\,$dex for the thin disk distribution, and $\feh=-0.14\pm0.19\,$dex for the thick disk distribution.

### 2019-08-23 paper/draft.tex
submitted materials

@@
 \subsection{Temperature/Metallicity Model \label{subsec:mann_results}}
-
+2
 \begin{figure}
-	\includegraphics[width=\linewidth]{figures/crv_test_teff.png}
-	\includegraphics[width=\linewidth]{figures/crv_test_fe_h.png}
+	\includegraphics[width=\linewidth]{crv_test_teff.png}
+	\includegraphics[width=\linewidth]{crv_test_fe_h.png}
 	\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (top), and leave-one-out cross validation (bottom).
@@
 sense; it is only good at the percent level.
-% \color{gcolor}{This percent-level accuracy of the model in the data domain is also demonstrated...}\color{black} \\
 
@@
 	\begin{center}
-	\includegraphics[width=\linewidth]{figures/crv_test_spt.png}
+	\includegraphics[width=\linewidth]{crv_test_spt.png}
 	\end{center}
@@
 \begin{figure*}
-	\includegraphics[width=\linewidth]{figures/cannon_chi_dist.png}
+	\includegraphics[width=\linewidth]{cannon_chi_dist.png}
 	\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources \citep{ElBadry:2018,Skinner:2018}; \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE--\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
@@
 \begin{figure*}
-	\includegraphics[width=\linewidth]{figures/cmd_selection.png}
+	\includegraphics[width=\linewidth]{cmd_selection.png}
 	\caption{\gaia\ and 2MASS color-magnitude cuts for the 12,037 sources with $\chi^2<100,000$.
@@
 \begin{center}
-	\includegraphics[width=\linewidth]{figures/ruwe_cut_before.png} \\
-	\includegraphics[width=\linewidth]{figures/ruwe_cut_after.png}
+	\includegraphics[width=\linewidth]{ruwe_cut_before.png} \\
+	\includegraphics[width=\linewidth]{ruwe_cut_after.png}
 	\caption{The left panel shows the test sample of 8,335 sources after applying Selections $1-3$ described in Section \ref{subsec:test_selection}.
@@
 \begin{center}
-	\includegraphics[width=.9\linewidth]{figures/aspcap_cannon_validation.png}
+	\includegraphics[width=.9\linewidth]{aspcap_cannon_validation.png}
 	\caption{\cannon\ (right column) and ASPCAP (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 1,702 sources with $V,J,K$ colors.
@@
 \begin{center}
-	\includegraphics[width=\linewidth]{figures/wide_binary_met.png}
+	\includegraphics[width=\linewidth]{wide_binary_met.png}
 	\caption{Metallicities of \cannon\ M dwarfs in binaries (x-axis) compared to the metallicities of their companions (y-axis).
@@
 \begin{center}
-	\includegraphics[height=5cm]{figures/toomre.png}
-	\includegraphics[height=5cm]{figures/met_hist.png}
+	\includegraphics[height=5cm]{toomre.png}
+	\includegraphics[height=5cm]{met_hist.png}
 	\caption{Toomre diagram showing the distribution of the \cannon\ metallicities (left).
@@
 \begin{center}
-	\includegraphics[width=.3\linewidth]{figures/met_comp1.png}
-	\includegraphics[width=.3\linewidth]{figures/met_comp2.png}
-	\includegraphics[width=.3\linewidth]{figures/met_comp3.png} 
-	\includegraphics[width=.3\linewidth]{figures/met_comp4.png}
-	\includegraphics[width=.3\linewidth]{figures/met_comp5.png}
-	\includegraphics[width=.3\linewidth]{figures/met_comp6.png} 
-	\includegraphics[width=.3\linewidth]{figures/met_comp7.png}
-	\includegraphics[width=.3\linewidth]{figures/met_comp8.png}
-	\includegraphics[width=.3\linewidth]{figures/met_comp9.png} 
+	\includegraphics[width=.3\linewidth]{met_comp1.png}
+	\includegraphics[width=.3\linewidth]{met_comp2.png}
+	\includegraphics[width=.3\linewidth]{met_comp3.png} 
+	\includegraphics[width=.3\linewidth]{met_comp4.png}
+	\includegraphics[width=.3\linewidth]{met_comp5.png}
+	\includegraphics[width=.3\linewidth]{met_comp6.png} 
+	\includegraphics[width=.3\linewidth]{met_comp7.png}
+	\includegraphics[width=.3\linewidth]{met_comp8.png}
+	\includegraphics[width=.3\linewidth]{met_comp9.png} 
 	\caption{A comparison of ASPCAP and Cannon metallicities for 8,335 test sample sources separated into temperature bins of 100K.
@@
 \begin{center}
-	\includegraphics[width=.47\linewidth]{figures/cmd_teff_full.png}
-	\includegraphics[width=.47\linewidth]{figures/cmd_teff_safe.png} \\
-	\includegraphics[width=.47\linewidth]{figures/cmd_feh_full.png}
-	\includegraphics[width=.47\linewidth]{figures/cmd_feh_safe.png} \\
-	\includegraphics[width=.47\linewidth]{figures/cmd_spt_full.png}
-	\includegraphics[width=.47\linewidth]{figures/cmd_spt_safe.png}
+	\includegraphics[width=.47\linewidth]{cmd_teff_full.png}
+	\includegraphics[width=.47\linewidth]{cmd_teff_safe.png} \\
+	\includegraphics[width=.47\linewidth]{cmd_feh_full.png}
+	\includegraphics[width=.47\linewidth]{cmd_feh_safe.png} \\
+	\includegraphics[width=.47\linewidth]{cmd_spt_full.png}
+	\includegraphics[width=.47\linewidth]{cmd_spt_safe.png}
 	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 5,875 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, spectroscopic binaries, and K dwarfs).
@@
 	\begin{center}
-	\includegraphics[width=16cm]{figures/demo_derivatives_teff1.png} 
+	\includegraphics[width=16cm]{demo_derivatives_teff1.png} 
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{figures/demo_derivatives_teff2.png}
+	\includegraphics[width=16cm]{demo_derivatives_teff2.png}
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{figures/demo_derivatives_teff3.png}
+	\includegraphics[width=16cm]{demo_derivatives_teff3.png}
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{figures/demo_derivatives_feh1.png}
+	\includegraphics[width=16cm]{demo_derivatives_feh1.png}
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{figures/demo_derivatives_feh2.png}
+	\includegraphics[width=16cm]{demo_derivatives_feh2.png}
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{figures/demo_derivatives_feh3.png}
+	\includegraphics[width=16cm]{demo_derivatives_feh3.png}
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=12cm]{figures/Spectral_Sequence_1.png}
-	\includegraphics[width=12cm]{figures/Spectral_Sequence_2.png}
-	\includegraphics[width=12cm]{figures/Spectral_Sequence_3.png}
+	\includegraphics[width=12cm]{Spectral_Sequence_1.png}
+	\includegraphics[width=12cm]{Spectral_Sequence_2.png}
+	\includegraphics[width=12cm]{Spectral_Sequence_3.png}
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{figures/derivative_jackknife_spt.png}
+	\includegraphics[width=16cm]{derivative_jackknife_spt.png}
 	\end{center}
@@
 	{\tt SPT} 				& 	& West-trained \cannon\ spectral subtype (M0$-$M9)	\\
+	{\tt TEFF$\_$APOGEE} 				& K	& ASPCAP pipeline effective temperature	\\
+	{\tt M$\_$H$\_$APOGEE} 				& dex	& ASPCAP pipeline [M/H]	\\
 	{\tt CHI$\_$MANN} 	& 	& $\chi^2$ fit of Mann-trained \cannon\ model	\\
 	{\tt CHI$\_$WEST} 	& 	& $\chi^2$ fit of West-trained \cannon\ model	\\
-	{\tt RADIUS} 			& R$_\odot$	& Radius determined by \citealt{Mann:2015} relation	\\
-	{\tt MASS} 			& M$_\odot$	& Mass determined by \citealt{Mann:2019} relation	\\
-	{\tt J} 				& mag	& APOGEE $J$ band photometry	\\
-	{\tt J$\_$ERR} 		& mag	& APOGEE $J$ band photometry uncertainty	\\
-	{\tt H} 				& mag	& APOGEE $H$ band photometry	\\
-	{\tt H$\_$ERR} 		& mag	& APOGEE $H$ band photometry uncertainty	\\
-	{\tt K} 				& mag	& APOGEE $K$ band photometry	\\
-	{\tt K$\_$ERR} 		& mag	& APOGEE $K$ band photometry uncertainty	\\
+	{\tt J$\_$MAG} 				& mag	& APOGEE $J$ band photometry	\\
+	{\tt H$\_$MAG} 				& mag	& APOGEE $H$ band photometry	\\
+	{\tt K$\_$MAG} 				& mag	& APOGEE $K$ band photometry	\\
 	{\tt BP$\_$MAG }		& mag	& \gaia\ $BP$ band photometry	\\
@@
 	{\tt G$\_$MAG} 		& mag	& \gaia\ $G$ band photometry	\\
+	{\tt J$\_$ABS} 		& mag	& APOGEE $J$ band absolute magnitude	\\
+	{\tt H$\_$ABS} 		& mag	& APOGEE $H$ band absolute magnitude	\\
+	{\tt K$\_$ABS} 		& mag	& APOGEE $K$ band absolute magnitude	\\
 	{\tt G$\_$ABS} 		& mag	& \gaia\ $G$ band absolute magnitude	\\
@@
 	{\tt Y} 				& pc	& Cartesian $y$ position in Galactocentric coordnates	\\
-	{\tt Z} 				& pc	& Cartesian $z$ position in Galactocentric coordnates	
+	{\tt Z} 				& pc	& Cartesian $z$ position in Galactocentric coordnates  \\
+	{\tt SNR} 				& 	& Signal-to-noise ratio of the APOGEE spectrum
 \enddata

### 2019-09-10 paper/draft.tex
add acknowledgments

@@
 \usepackage{morefloats}
-
-% \usepackage{draftwatermark}
-% \SetWatermarkScale{3}
-% \SetWatermarkLightness{.7}
 
@@
 \subsection{Temperature/Metallicity Model \label{subsec:mann_results}}
-2
+
 \begin{figure}
@@
 \acknowledgements
-We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Melissa Ness (Columbia), Kevin Covey (WWU), Diogo Souto (Observatário Nacional/MCTI) and Keivan Stassun (Vanderbuilt) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
+We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Melissa Ness (Columbia), Kevin Covey (WWU), Diogo Souto (Observatário Nacional/MCTI), Keivan Stassun (Vanderbuilt), Anthony Brown (Leiden), Aida Behmard (Caltech) and Christopher Theissen (UCSD) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
 This project was partially supported by
@@
 
-\begin{figure*}[]
+\begin{figure*}
 	\begin{center}
@@
 %---------------------------
-\begin{figure*}[]
+\begin{figure*}
 	\begin{center}

### 2019-10-27 paper/draft.tex
minor edits

@@
 \accepted{\color{red} MM/DD/YY}
-\submitjournal{\color{red} JOURNAL}
+\submitjournal{\apj}
 
 \shorttitle{temperatures and metallicities for m dwarfs}
-\shortauthors{birky et al.}
 
@@
 \title{TEMPERATURES AND METALLICITIES OF M DWARFS IN THE APOGEE SURVEY}
-
-\correspondingauthor{Jessica Birky}
-\email{jbirky@uw.edu}
-
-\author[0000-0002-7961-6881]{Jessica Birky}
-\affil{Center for Astrophysics and Space Science, University of California San Diego, La Jolla, CA 92093, USA}
-\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
-
-\author[0000-0003-2866-9403]{David W. Hogg}
-\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
-\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726
-Broadway, New York, NY 10003, USA}
-\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
-\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
-
-\author[0000-0003-3654-1602]{Andrew W. Mann}
-\affil{Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY 10027, USA}
-\affil{Department of Physics and Astronomy, The University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
-
-\author[0000-0002-6523-9536]{Adam J. Burgasser}
-\affil{Center for Astrophysics and Space Science, University of California San Diego, La Jolla, CA 92093, USA}
 
@@
 Additionally, the low masses of M dwarfs make for easier detection of planets by radial velocity variability \citep{Trifonov:2018}, high planet-to-star radii ratios make for easier detection of exoplanet transits in light curve observations \citep{Nutzman:2008}, and shorter orbital periods (for a fixed stellar insolation flux) allow for new planet discovery in less observation time than for higher mass stars.
-For these reasons, M dwarfs are primary candidates for exoplanet searches, including by the NASA \kepler\ \citep[e.g.,][]{Dressing2015} and Transiting Exoplanet Survey Satellite (\tess) \citep[e.g.,][]{Muirhead:2018} missions.
-As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with the James Webb Space Telescope ({\sl JWST;} \citealt{Clampin2008}).
+For these reasons, M dwarfs are primary candidates for exoplanet searches, including by the NASA \kepler\ \citep[e.g.,][]{Dressing2015} and \textsl{Transiting Exoplanet Survey Satellite} (\tess) \citep[e.g.,][]{Muirhead:2018} missions.
+As a result, detailed and precise knowledge of M dwarf chemical compositions has become key to constraining the properties, formation scenarios, and atmospheric conditions of potentially habitable exoplanets observable with the \textsl{James Webb Space Telescope} ({\sl JWST;} \citealt{Clampin2008}).
 
@@
 
-A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf \citep[e.g.][]{Terrien:2012,RojasAyala:2010,Newton:2014,Neves:2014,Lindgren:2016} to developed precise empirical relations (as good as $\sim$0.07\,dex).
+A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf \citep[e.g.,][]{Terrien:2012,RojasAyala:2010,Newton:2014,Neves:2014,Lindgren:2016} to developed precise empirical relations (as good as $\sim$0.07\,dex).
 However these metallicity calibrations do not take advantage of the full wavelength coverage available, nor information about the overall spectral shape often used to determine $\teff$ and spectral type.
@@
 
-The APOGEE survey is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
+The APOGEE survey is a high resolution (R$\sim$22,500), $H$-band ($1.5-1.7\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}).
 Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters \citep{Meszaros:2012}.
@@
 
-Inferring the label of a star with such a model requires two steps: first, the \emph{training step} in which a generative model describing the probability density function (pdf) of the flux is constructed at each pixel from the set of spectra with known reference labels; and second, the \emph{test step} in which the model is applied to determine the labels of a spectrum.
+Inferring the label of a star with such a model requires two steps: first, the \emph{training step} in which a generative model describing the probability density function of the flux is constructed at each pixel from the set of spectra with known reference labels; and second, the \emph{test step} in which the model is applied to determine the labels of a spectrum.
 
-Following the procedure of \citet{Ness:2015} and \citet{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector $\ell$ and and coefficient vector \emph{$\theta$}.
+Following the procedure of \citet{Ness:2015} and \citet{Ho:2017a} we adopt a simple linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector $\ell$ and coefficient vector \emph{$\theta$}.
 For each star \emph{n}, at wavelength pixel \emph{$\lambda$}, we assume that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
@@
 \end{equation} 
-We use the noise model is $N_{\lambda}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
-Equation \ref{eqn:cannon} corresponds to the single-pixel log-likelihood function:
+\color{red}{Here we use the noise model}\color{black} $N_{\lambda}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
+Equation (\ref{eqn:cannon}) corresponds to the single-pixel log-likelihood function:
 \begin{equation}
@@
 \thecannon\ model can in principle be trained on any physical or empirical labels available, giving a wide range of flexibility to the model.
-However choosing a training sample with high quality labels is critical to its performance.
+However, choosing a training sample with high quality labels is critical to its performance.
 Limitations of \thecannon\ include that test (output) labels are only accurate if the training labels are accurate, and only precise if the training labels are measured consistently across the training sample.
 It is also critical to have a training sample with the dynamic range to span the entire parameter space of interest, as \thecannon\ does not extrapolate well outside the parameter space of the training sample.
-Finally, \thecannon\ assumes that the dependence of spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions.
+Finally, \thecannon\ assumes that the dependence of \color{red}{the}\color{black} spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions.
 If that isn't true, there will be features that \thecannon\ cannot reproduce.
 
-For the purpose of this study, we have constructed two different training samples: first a one-dimensional \emph{spectral type model}, and second a two-dimensional \emph{physical parameter model} which describes the temperature and metallicity.
-The choice of training labels, dimensionality of our data set, and requirements for a good training set we discuss further in Section \ref{sec:discussion}.
+For the purpose of this study, we have constructed two different training samples: first a one-dimensional \emph{spectral type model}, and second, a two-dimensional \emph{physical parameter model}, which describes the temperature and metallicity.
+The choice of training labels, dimensionality of our data set, and requirements for a good training set \color{red}{are discussed}\color{black} further in Section \ref{sec:discussion}.
 
@@
 
-The spectral type training sample consists of 51 sources, spanning M0-M9 cross-matched from the \citet{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from SDSS.
+The spectral type training sample consists of 51 sources, spanning M0$-$M9 cross-matched from the \citet{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from SDSS.
 For each source in the catalog, spectral types were determined both through an automated routine for comparing spectral type templates to data using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.
@@
 
-The physical parameter training sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 < \feh < 0.49$\,dex, 41 sources of which are drawn from \citet{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample to M15 analyzed using similar data and identical techniques to M15.
-The major difference between the extension sample is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample.
+The physical parameter training sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 < \feh < 0.49$\,dex, 41 sources of which are drawn from \citet{Mann:2015} (hereafter M15), and 46 of which are part of a previously unpublished extension sample to M15, analyzed using similar data and identical techniques to M15.
+The major difference between the extension sample \color{red}{to M15}\color{black} is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample.
 
@@
 Iron abundances ($\feh$) are assigned to the physical parameter sample based on the strength of metal-sensitive lines in the near-infrared SpeX spectra \citep{RojasAyala:2010} using the calibration from \citet{Mann:2013a}.
-The relation between these lines and an absolute $\feh$ scale is calibrated using wide binaries containing and F, G, or K-type primary and an M dwarf companion, under the assumption that binaries formed from the same molecular cloud and therefore have the same metallicity \citep{Bonfils:2005}.
+The relation between these lines and an absolute $\feh$ scale is calibrated using wide binaries containing an F, G, or K-type primary and an M dwarf companion, under the assumption that binaries formed from the same molecular cloud and therefore have the same metallicity \citep{Bonfils:2005}.
 Uncertainties are estimated to be $\simeq$0.08\,dex based on irreducible scatter in the empirical relation between selected lines and the assigned $\feh$ from the primary star.
@@
 
-For the physical parameter model, we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
+For the physical parameter model, we trained \thecannon\ on 87 M dwarfs with two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex.
 We note for this model that 5 out of 87 sources show possible rotational line broadening identified by visual inspection (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources show no obvious broadening.
-We note that these broadened sources have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), and that the labels for these 5 sources arep biased by an average of $+65\,$K and $-0.08\,$dex.
+We note that these broadened sources have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), and that the labels for these 5 sources are biased by an average of $+65\,$K and $-0.08\,$dex.
 However, removing them from the training sample does not significantly change the overall scatter and bias of the model.
@@
 Precision (scatter) and bias of the model for each test are calculated as the standard deviation and mean of the difference in training and test (or LOOCV) labels respectively (Figure \ref{fig:mann_validation}).
-Since the LOOCV test evaluates both how well the model reproduces the training values \emph{and} penalizes the model for over-fitting, we adopt the LOOCV scatter over as the estimate of the model's precision.
+Since the LOOCV test evaluates both how well the model reproduces the training values \emph{and} penalizes the model for over-fitting, we adopt the LOOCV scatter as the estimate of the model's precision.
 The set of training, test, and cross-validated labels for each training source is reported in Table \ref{table:mann_results}.
@@
 The bottom panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh} show the derivative of flux with respect to temperature and metallicity at each pixel, taken at the median training values.
-In order to evaluate which spectral features show statistically significant change with respect to input label, we compute error of the derivative at each pixel using a jackknife statistic (with a 1-$\sigma$ level overplotted in red):
+In order to evaluate which spectral features show statistically significant change with respect to input label, we compute \color{red}{the}\color{black} error of the derivative at each pixel using a jackknife statistic (with a 1-$\sigma$ level overplotted in red):
 \begin{equation}
@@
 \end{equation}
-where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample indexed by \emph{n}, $\theta$ is the coeffient vector trained on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.
+where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample indexed by \emph{n}, $\theta$ is the coeffient vector trained on all \emph{N} sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.
 A summary of identified lines with derivative values greater than $2\sigma_{\rm jackkife}$ is given in Table \ref{table:teff_derivative} and Table \ref{table:feh_derivative}.
@@
 The spectra contain roughly 8000 pixels,
-so we might expect the $\chi^2$ values to be close to 8000 in magnitude.
-They are much higher.
+so we might expect the $\chi^2$ values to be close to 8000 in magnitude,
+\color{red}{however, they}\color{black} are much higher.
 This discrepancy flows from the fact that, while the spectral model is good at the percent
@@
 
-We trained \thecannon\ on 51 M dwarfs ranging M0-M9 with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
+We trained \thecannon\ on 51 M dwarfs ranging \color{red}{between}\color{black} M0$-$M9 with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
 We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
@@
 
-Out of the total APOGEE DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources.
+Out of the total APOGEE DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross-match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources.
 From there we applied additional selection criteria, described below, to identify a sample of single, main-sequence M stars, with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries:
@@
 \begin{enumerate}
-\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000, chosen to remove badly fit sources (such as fast rotators), but include $\chi^2$ values close the distribution of training sample (Figure \ref{fig:chi_dist}).
+\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000, chosen to remove badly fit sources (such as fast rotators), but include $\chi^2$ values close \color{red}{to}\color{black} the distribution of training sample (Figure \ref{fig:chi_dist}).
 
-\item \textbf{Color-magnitude cuts:} Using \gaia\ and 2MASS photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to remove sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and multiples), and subdwarfs below the main sequence.
+\item \textbf{Color-magnitude cuts:} Using \gaia\ and 2MASS photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to remove sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and\color{red}{/or}\color{black} multiples), and subdwarfs below the main sequence.
 
@@
 Table \ref{table:test_results} outlines the parameters included in the test sample catalog, which can be downloaded from the online journal.
-Included are two versions of the catalog: the first containing all 14,827 sources before selection, and the second containing the 5,875 sources kept after making Selections $1-5$ described in this section.
+Included are two versions of the catalog: the first containing all 14,827 sources before selection, and the second containing the 5,875 sources kept after making selections $1-5$ described in this section.
 
@@
 Compared to the \citet{Mann:2015} and \citet{Boyajian:2012} color-metallicity derived temperatures, both ASPCAP and \cannon\ temperatures show similar scatters of $\sim60\,$K, but are offset by a constant.
-We find \cannon\ to be in better agreement with \citet{Mann:2015} and \citet{Boyajian:2012}, with ASPCAP overestimating $\teff$ on average $\sim110-140\,$K, and \thecannon\ underestimating $\teff$ on average by $\sim10-20\,$K, with the largest deviation in the latter at the lowest and highest $\teff$s.
+We find \cannon\ to be in better agreement with \citet{Mann:2015} and \citet{Boyajian:2012}, with ASPCAP overestimating $\teff$ on average $\sim110-140\,$K, and \thecannon\ underestimating $\teff$ on average by $\sim10-20\,$K, with the largest deviation in the latter at the lowest and highest $\teff$.
 
@@
 Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}.
-These Tables summarize spectral features in the APOGEE bandpass
+These tables summarize spectral features in the APOGEE bandpass
 that are found to be strong temperature and metallicity indicators.
@@
 \acknowledgements
-We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Melissa Ness (Columbia), Kevin Covey (WWU), Diogo Souto (Observatário Nacional/MCTI), Keivan Stassun (Vanderbuilt), Anthony Brown (Leiden), Aida Behmard (Caltech) and Christopher Theissen (UCSD) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
+We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Melissa Ness (Columbia), Kevin Covey (WWU), Diogo Souto (Observatário Nacional/MCTI), Keivan Stassun (Vanderbilt), \color{red}{Katia Cunha (NOAO), Anthony Brown (Leiden), Aida Behmard (Caltech) and Christopher Theissen (UCSD)}\color{black} for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
 This project was partially supported by

### 2019-10-30 paper/draft.tex
referee response

@@
 
-A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf \citep[e.g.,][]{Terrien:2012,RojasAyala:2010,Newton:2014,Neves:2014,Lindgren:2016} to developed precise empirical relations (as good as $\sim$0.07\,dex).
+A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf \citep[e.g.,][]{Terrien:2012,RojasAyala:2010,Newton:2014,Neves:2014,Lindgren:2016} to develope precise empirical relations (as good as $\sim$0.07\,dex).
 However these metallicity calibrations do not take advantage of the full wavelength coverage available, nor information about the overall spectral shape often used to determine $\teff$ and spectral type.
@@
 
-\thecannon\ is a regression model which relies on two assumptions: first, that sources with identical labels have near-identical flux at each wavelength pixel; and second, that the expected flux at each pixel varies continuosly with change in label.
+\thecannon\ is a regression model which relies on two assumptions: first, that sources with identical labels have near-identical flux at each wavelength pixel; and second, that the expected flux at each pixel varies continuously with change in label.
 
@@
 \end{equation} 
-\color{red}{Here we use the noise model}\color{black} $N_{\lambda}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
+\color{red}{\bf Here we use the noise model }\color{black} $N_{\lambda}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
 Equation (\ref{eqn:cannon}) corresponds to the single-pixel log-likelihood function:
@@
 
-\thecannon\ model can in principle be trained on any physical or empirical labels available, giving a wide range of flexibility to the model.
+\thecannon\ model can in principle be trained on any physical or empirical labels available \color{red}{\bf beyond those which typically parameterize theoretical atmospheric models ($\teff$, logg, [M/H], etc.), such as additional physical parameters (e.g., mass/age \citealt{Ness:2016}) or empirical proxies for physical parameters (e.g., spectral types, colors, magnitudes), }\color{black} giving a wide range of flexibility to the model.
 However, choosing a training sample with high quality labels is critical to its performance.
@@
 It is also critical to have a training sample with the dynamic range to span the entire parameter space of interest, as \thecannon\ does not extrapolate well outside the parameter space of the training sample.
-Finally, \thecannon\ assumes that the dependence of \color{red}{the}\color{black} spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions.
+Finally, \thecannon\ assumes that the dependence of \color{red}{\bf the }\color{black} spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions.
 If that isn't true, there will be features that \thecannon\ cannot reproduce.
@@
 For the purpose of this study, we have constructed two different training samples: first a one-dimensional \emph{spectral type model}, and second, a two-dimensional \emph{physical parameter model}, which describes the temperature and metallicity.
-The choice of training labels, dimensionality of our data set, and requirements for a good training set \color{red}{are discussed}\color{black} further in Section \ref{sec:discussion}.
+The choice of training labels, dimensionality of our data set, and requirements for a good training set \color{red}{\bf are discussed }\color{black} further in Section \ref{sec:discussion}.
 
@@
 The physical parameter training sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 < \feh < 0.49$\,dex, 41 sources of which are drawn from \citet{Mann:2015} (hereafter M15), and 46 of which are part of a previously unpublished extension sample to M15, analyzed using similar data and identical techniques to M15.
-The major difference between the extension sample \color{red}{to M15}\color{black} is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample.
+The major difference between the extension sample is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample \color{red}{\bf (however, all sources in the training sample were visually inspected by color-magnitude position for binarity before addition).}\color{black}
 
@@
 
-Another mode of analysis we can look at with \thecannon\ is how the derivative of the model changes with respect to given training parameters, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters.
+Another mode of analysis we can \color{red}{\bf utilize }\color{black} with \thecannon\ is how the derivative of the model changes with respect to given training parameters, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters.
 The top two panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh} show two example spectra and model fits for two different temperatures (Figure \ref{fig:demo_teff}), and two different metallicities (Figure \ref{fig:demo_feh}), with atomic and molecular features identified by the abundance analysis of \citet{Souto:2017}.
 The bottom panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh} show the derivative of flux with respect to temperature and metallicity at each pixel, taken at the median training values.
-In order to evaluate which spectral features show statistically significant change with respect to input label, we compute \color{red}{the}\color{black} error of the derivative at each pixel using a jackknife statistic (with a 1-$\sigma$ level overplotted in red):
+In order to evaluate which spectral features show statistically significant change with respect to input label, we compute \color{red}{\bf the }\color{black} error of the derivative at each pixel using a jackknife statistic (with a 1-$\sigma$ level overplotted in red):
 \begin{equation}
@@
 so we might expect the $\chi^2$ values to be close to 8000 in magnitude,
-\color{red}{however, they}\color{black} are much higher.
+\color{red}{\bf however, they }\color{black} are much higher.
 This discrepancy flows from the fact that, while the spectral model is good at the percent
@@
 
-We trained \thecannon\ on 51 M dwarfs ranging \color{red}{between}\color{black} M0$-$M9 with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
+We trained \thecannon\ on 51 M dwarfs ranging \color{red}{\bf between }\color{black} M0$-$M9 with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
 We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
@@
 \begin{enumerate}
-\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000, chosen to remove badly fit sources (such as fast rotators), but include $\chi^2$ values close \color{red}{to}\color{black} the distribution of training sample (Figure \ref{fig:chi_dist}).
+\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000, chosen to remove badly fit sources (such as fast rotators), but include $\chi^2$ values close \color{red}{\bf to }\color{black} the distribution of training sample (Figure \ref{fig:chi_dist}).
 
-\item \textbf{Color-magnitude cuts:} Using \gaia\ and 2MASS photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to remove sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and\color{red}{/or}\color{black} multiples), and subdwarfs below the main sequence.
+\item \textbf{Color-magnitude cuts:} Using \gaia\ and 2MASS photometry we apply the \color{red}{\bf additional }\color{black} color-magnitude selections shown in Figure \ref{fig:cmd_selection} to remove sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and\color{red}{\bf /or }\color{black} multiples), and subdwarfs below the main sequence.
 
@@
 The overall scatter between the 15 metallicity pairs is 0.08\,dex.}
+	\label{fig:wide_binary_met}
 \end{center}
@@
 Cross-matching the list of FGK+M dwarf companions with several catalogs/surveys with measured stellar metallicities, we found an additional 7 sources with FGK metallicities from LAMOST \citep{Zhao:2012} and APOGEE (ASPCAP).
-The metallicity measurements for the 15 M dwarf binaries and their companions are given in Table \ref{table:binaries}, and the overall scatter is $0.08\,$dex--an improvement over the scatter of ASPCAP metallicities which is $0.15\,$dex for these 15 sources.
+The metallicity measurements for the 15 M dwarf binaries and their companions are given in Table \ref{table:binaries} \color{red}{\bf and shown in Figure \ref{fig:wide_binary_met} }\color{black}, and the overall scatter is $0.08\,$dex--an improvement over the scatter of ASPCAP metallicities which is $0.15\,$dex for these 15 sources.
 Internal consistency of the two models (the scatter of the 8 M+M pairs both in APOGEE) are 0.06\,dex and 0.12\,dex for \cannon\ and ASPCAP respectively.
@@
 It is also limited to the stellar-parameter domain of that input catalog.
-That said, this work provides and external validation of the M15 stellar parameters.
+That said, this work provides an external validation of the M15 stellar parameters.
 
@@
 an effect in which the optimizer of the pipeline may be lowering the continuum level and metallicity of the fit to compensate for the missing lines or opacities.
-Further analysis would need to be done to determine if the metallicity bias
+\color{red}{\bf We also note that this analysis was completed using data from Data Release 14 of APOGEE, 
+which did include molecular lines from FeH in the pipeline at the time, which become numerous and strong for $\teff \lesssim 3600$\,K. }\color{black}
+Further analysis would need to be done to \color{red}{\bf quantify the metallicity improvement for M dwarfs in later data releases of APOGEE now available (in which FeH has been included in analysis)}\color{black}, and determine if the metallicity bias
 is found in other model grids (besides the ATLAS/MARCS models used by the ASPCAP pipeline), 
@@
 \acknowledgements
-We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Melissa Ness (Columbia), Kevin Covey (WWU), Diogo Souto (Observatário Nacional/MCTI), Keivan Stassun (Vanderbilt), \color{red}{Katia Cunha (NOAO), Anthony Brown (Leiden), Aida Behmard (Caltech) and Christopher Theissen (UCSD)}\color{black} for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
+We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Melissa Ness (Columbia), Kevin Covey (WWU), Diogo Souto (Observatário Nacional/MCTI), Keivan Stassun (Vanderbilt), \color{red}{\bf Katia Cunha (NOAO), Anthony Brown (Leiden), Aida Behmard (Caltech) and Christopher Theissen (UCSD) }\color{black} for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
 This project was partially supported by
@@
 	\end{center}
-	\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallicities.
+	\caption{\textit{Top two panels of each plot:} \color{red}{\bf APOGEE spectra (black), overlaid by the }\color{black} Mann-trained \color{red}{\bf \cannon\ }\color{black} model for \color{red}{\bf two sources of }\color{black} varying temperatures, and similar metallicities.
 \textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.}
@@
 	\end{center}
-	\caption{\textit{Top two panels of each plot:} Mann-trained model for varying metallicities and similar temperatures.
+	\caption{\textit{Top two panels of each plot:} \color{red}{\bf APOGEE spectra (black), overlaid by the }\color{black} Mann-trained \color{red}{\bf \cannon\ }\color{black} model for \color{red}{\bf two sources of }\color{black} varying metallicities and similar temperatures.
 \textit{Third panel:} Derivative of \thecannon\ model with respect to metallicity, taken at the median training metallicity, $\feh=-0.03$\,dex; jackknife computed error at each pixel is shown in red.} \label{fig:demo_feh}

### 2020-01-16 paper/paper.tex
add submitted version

--- 2019-10-30 paper/draft.tex
@@
 
-\received{\color{red} MM/DD/YY}
-\revised{\color{red} MM/DD/YY}
-\accepted{\color{red} MM/DD/YY}
 \submitjournal{\apj}
@@
 \shorttitle{temperatures and metallicities for m dwarfs}
+\shortauthors{birky et al.}
 
@@
 \title{TEMPERATURES AND METALLICITIES OF M DWARFS IN THE APOGEE SURVEY}
+
+\correspondingauthor{Jessica Birky}
+\email{jbirky@uw.edu}
+
+\author[0000-0002-7961-6881]{Jessica Birky}
+\affil{Center for Astrophysics and Space Science, University of California San Diego, La Jolla, CA 92093, USA}
+\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
+
+\author[0000-0003-2866-9403]{David W. Hogg}
+\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
+\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726
+Broadway, New York, NY 10003, USA}
+\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
+\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
+
+\author[0000-0003-3654-1602]{Andrew W. Mann}
+\affil{Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY 10027, USA}
+\affil{Department of Physics and Astronomy, The University of North Carolina at Chapel Hill, Chapel Hill, NC 27599, USA}
+
+\author[0000-0002-6523-9536]{Adam Burgasser}
+\affil{Center for Astrophysics and Space Science, University of California San Diego, La Jolla, CA 92093, USA}
 
@@
 \end{equation} 
-\color{red}{\bf Here we use the noise model }\color{black} $N_{\lambda}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
+Here we use the noise model $N_{\lambda}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.
 Equation (\ref{eqn:cannon}) corresponds to the single-pixel log-likelihood function:
@@
 
-\thecannon\ model can in principle be trained on any physical or empirical labels available \color{red}{\bf beyond those which typically parameterize theoretical atmospheric models ($\teff$, logg, [M/H], etc.), such as additional physical parameters (e.g., mass/age \citealt{Ness:2016}) or empirical proxies for physical parameters (e.g., spectral types, colors, magnitudes), }\color{black} giving a wide range of flexibility to the model.
+\thecannon\ model can in principle be trained on any physical or empirical labels available beyond those which typically parameterize theoretical atmospheric models ($\teff$, logg, [M/H], etc.), such as additional physical parameters (e.g., mass/age \citealt{Ho:2017b}) or empirical proxies for physical parameters (e.g., spectral types, colors, magnitudes), giving a wide range of flexibility to the model.
 However, choosing a training sample with high quality labels is critical to its performance.
@@
 It is also critical to have a training sample with the dynamic range to span the entire parameter space of interest, as \thecannon\ does not extrapolate well outside the parameter space of the training sample.
-Finally, \thecannon\ assumes that the dependence of \color{red}{\bf the }\color{black} spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions.
+Finally, \thecannon\ assumes that the dependence of the spectrum on labels is continuous and smooth---and in this implementation, well approximated by quadratic functions.
 If that isn't true, there will be features that \thecannon\ cannot reproduce.
@@
 For the purpose of this study, we have constructed two different training samples: first a one-dimensional \emph{spectral type model}, and second, a two-dimensional \emph{physical parameter model}, which describes the temperature and metallicity.
-The choice of training labels, dimensionality of our data set, and requirements for a good training set \color{red}{\bf are discussed }\color{black} further in Section \ref{sec:discussion}.
+The choice of training labels, dimensionality of our data set, and requirements for a good training set are discussed further in Section \ref{sec:discussion}.
 
@@
 The physical parameter training sample consists of 87 sources with reference labels distributed over $2859 <T_{\rm eff} < 4131$K, and $-0.48 < \feh < 0.49$\,dex, 41 sources of which are drawn from \citet{Mann:2015} (hereafter M15), and 46 of which are part of a previously unpublished extension sample to M15, analyzed using similar data and identical techniques to M15.
-The major difference between the extension sample is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample \color{red}{\bf (however, all sources in the training sample were visually inspected by color-magnitude position for binarity before addition).}\color{black}
+The major difference between the extension sample is that the sample had lower-quality or no parallaxes (prior to Gaia data) and hence were omitted from the M15 study and were less vetted for binarity than the M15 sample (however, all sources in the training sample were visually inspected by color-magnitude position for binarity before addition).
 
@@
 \begin{figure}
-	\includegraphics[width=\linewidth]{crv_test_teff.png}
-	\includegraphics[width=\linewidth]{crv_test_fe_h.png}
+	\includegraphics[width=\linewidth]{figs/crv_test_teff.png}
+	\includegraphics[width=\linewidth]{figs/crv_test_fe_h.png}
 	\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (top), and leave-one-out cross validation (bottom).
@@
 
-Another mode of analysis we can \color{red}{\bf utilize }\color{black} with \thecannon\ is how the derivative of the model changes with respect to given training parameters, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters.
+Another mode of analysis we can utilize with \thecannon\ is how the derivative of the model changes with respect to given training parameters, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters.
 The top two panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh} show two example spectra and model fits for two different temperatures (Figure \ref{fig:demo_teff}), and two different metallicities (Figure \ref{fig:demo_feh}), with atomic and molecular features identified by the abundance analysis of \citet{Souto:2017}.
 The bottom panels of Figures \ref{fig:demo_teff} and \ref{fig:demo_feh} show the derivative of flux with respect to temperature and metallicity at each pixel, taken at the median training values.
-In order to evaluate which spectral features show statistically significant change with respect to input label, we compute \color{red}{\bf the }\color{black} error of the derivative at each pixel using a jackknife statistic (with a 1-$\sigma$ level overplotted in red):
+In order to evaluate which spectral features show statistically significant change with respect to input label, we compute the error of the derivative at each pixel using a jackknife statistic (with a 1-$\sigma$ level overplotted in red):
 \begin{equation}
@@
 so we might expect the $\chi^2$ values to be close to 8000 in magnitude,
-\color{red}{\bf however, they }\color{black} are much higher.
+however, they are much higher.
 This discrepancy flows from the fact that, while the spectral model is good at the percent
@@
 	\begin{center}
-	\includegraphics[width=\linewidth]{crv_test_spt.png}
+	\includegraphics[width=\linewidth]{figs/crv_test_spt.png}
 	\end{center}
@@
 
-We trained \thecannon\ on 51 M dwarfs ranging \color{red}{\bf between }\color{black} M0$-$M9 with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
+We trained \thecannon\ on 51 M dwarfs ranging between M0$-$M9 with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type.
 We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source.
@@
 \begin{figure*}
-	\includegraphics[width=\linewidth]{cannon_chi_dist.png}
+	\includegraphics[width=\linewidth]{figs/cannon_chi_dist.png}
 	\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources \citep{ElBadry:2018,Skinner:2018}; \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE--\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$.
@@
 \begin{figure*}
-	\includegraphics[width=\linewidth]{cmd_selection.png}
+	\includegraphics[width=\linewidth]{figs/cmd_selection.png}
 	\caption{\gaia\ and 2MASS color-magnitude cuts for the 12,037 sources with $\chi^2<100,000$.
@@
 \begin{enumerate}
-\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000, chosen to remove badly fit sources (such as fast rotators), but include $\chi^2$ values close \color{red}{\bf to }\color{black} the distribution of training sample (Figure \ref{fig:chi_dist}).
+\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000, chosen to remove badly fit sources (such as fast rotators), but include $\chi^2$ values close to the distribution of training sample (Figure \ref{fig:chi_dist}).
 
-\item \textbf{Color-magnitude cuts:} Using \gaia\ and 2MASS photometry we apply the \color{red}{\bf additional }\color{black} color-magnitude selections shown in Figure \ref{fig:cmd_selection} to remove sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and\color{red}{\bf /or }\color{black} multiples), and subdwarfs below the main sequence.
+\item \textbf{Color-magnitude cuts:} Using \gaia\ and 2MASS photometry we apply the additional color-magnitude selections shown in Figure \ref{fig:cmd_selection} to remove sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and/or multiples), and subdwarfs below the main sequence.
 
@@
 \begin{center}
-	\includegraphics[width=\linewidth]{ruwe_cut_before.png} \\
-	\includegraphics[width=\linewidth]{ruwe_cut_after.png}
+	\includegraphics[width=\linewidth]{figs/ruwe_cut_before.png} \\
+	\includegraphics[width=\linewidth]{figs/ruwe_cut_after.png}
 	\caption{The left panel shows the test sample of 8,335 sources after applying Selections $1-3$ described in Section \ref{subsec:test_selection}.
@@
 \begin{center}
-	\includegraphics[width=.9\linewidth]{aspcap_cannon_validation.png}
+	\includegraphics[width=.9\linewidth]{figs/aspcap_cannon_validation.png}
 	\caption{\cannon\ (right column) and ASPCAP (left column) temperatures as compared to temperatures estimated color-temperature relations from three different literature sources for a subsample of 1,702 sources with $V,J,K$ colors.
@@
 \begin{center}
-	\includegraphics[width=\linewidth]{wide_binary_met.png}
+	\includegraphics[width=\linewidth]{figs/wide_binary_met.png}
 	\caption{Metallicities of \cannon\ M dwarfs in binaries (x-axis) compared to the metallicities of their companions (y-axis).
@@
 \begin{center}
-	\includegraphics[height=5cm]{toomre.png}
-	\includegraphics[height=5cm]{met_hist.png}
+	\includegraphics[height=5cm]{figs/toomre.png}
+	\includegraphics[height=5cm]{figs/met_hist.png}
 	\caption{Toomre diagram showing the distribution of the \cannon\ metallicities (left).
@@
 \begin{center}
-	\includegraphics[width=.3\linewidth]{met_comp1.png}
-	\includegraphics[width=.3\linewidth]{met_comp2.png}
-	\includegraphics[width=.3\linewidth]{met_comp3.png} 
-	\includegraphics[width=.3\linewidth]{met_comp4.png}
-	\includegraphics[width=.3\linewidth]{met_comp5.png}
-	\includegraphics[width=.3\linewidth]{met_comp6.png} 
-	\includegraphics[width=.3\linewidth]{met_comp7.png}
-	\includegraphics[width=.3\linewidth]{met_comp8.png}
-	\includegraphics[width=.3\linewidth]{met_comp9.png} 
+	\includegraphics[width=.3\linewidth]{figs/met_comp1.png}
+	\includegraphics[width=.3\linewidth]{figs/met_comp2.png}
+	\includegraphics[width=.3\linewidth]{figs/met_comp3.png} 
+	\includegraphics[width=.3\linewidth]{figs/met_comp4.png}
+	\includegraphics[width=.3\linewidth]{figs/met_comp5.png}
+	\includegraphics[width=.3\linewidth]{figs/met_comp6.png} 
+	\includegraphics[width=.3\linewidth]{figs/met_comp7.png}
+	\includegraphics[width=.3\linewidth]{figs/met_comp8.png}
+	\includegraphics[width=.3\linewidth]{figs/met_comp9.png} 
 	\caption{A comparison of ASPCAP and Cannon metallicities for 8,335 test sample sources separated into temperature bins of 100K.
@@
 Cross-matching the list of FGK+M dwarf companions with several catalogs/surveys with measured stellar metallicities, we found an additional 7 sources with FGK metallicities from LAMOST \citep{Zhao:2012} and APOGEE (ASPCAP).
-The metallicity measurements for the 15 M dwarf binaries and their companions are given in Table \ref{table:binaries} \color{red}{\bf and shown in Figure \ref{fig:wide_binary_met} }\color{black}, and the overall scatter is $0.08\,$dex--an improvement over the scatter of ASPCAP metallicities which is $0.15\,$dex for these 15 sources.
+The metallicity measurements for the 15 M dwarf binaries and their companions are given in Table \ref{table:binaries} and shown in Figure \ref{fig:wide_binary_met}, and the overall scatter is $0.08\,$dex--an improvement over the scatter of ASPCAP metallicities which is $0.15\,$dex for these 15 sources.
 Internal consistency of the two models (the scatter of the 8 M+M pairs both in APOGEE) are 0.06\,dex and 0.12\,dex for \cannon\ and ASPCAP respectively.
@@
 \begin{center}
-	\includegraphics[width=.47\linewidth]{cmd_teff_full.png}
-	\includegraphics[width=.47\linewidth]{cmd_teff_safe.png} \\
-	\includegraphics[width=.47\linewidth]{cmd_feh_full.png}
-	\includegraphics[width=.47\linewidth]{cmd_feh_safe.png} \\
-	\includegraphics[width=.47\linewidth]{cmd_spt_full.png}
-	\includegraphics[width=.47\linewidth]{cmd_spt_safe.png}
+	\includegraphics[width=.47\linewidth]{figs/cmd_teff_full.png}
+	\includegraphics[width=.47\linewidth]{figs/cmd_teff_safe.png} \\
+	\includegraphics[width=.47\linewidth]{figs/cmd_feh_full.png}
+	\includegraphics[width=.47\linewidth]{figs/cmd_feh_safe.png} \\
+	\includegraphics[width=.47\linewidth]{figs/cmd_spt_full.png}
+	\includegraphics[width=.47\linewidth]{figs/cmd_spt_safe.png}
 	\caption{Full sample of 14,827 M dwarfs colored by \cannon\ labels before selection (left) and final sample selection of 5,875 M dwarfs after applying selection criteria described in Section \ref{subsec:test_selection} (right), to reduce contamination from sources with that are not similar to the training sample (not single, main-sequence M stars, such as pre-main sequence, spectroscopic binaries, and K dwarfs).
@@
 an effect in which the optimizer of the pipeline may be lowering the continuum level and metallicity of the fit to compensate for the missing lines or opacities.
-\color{red}{\bf We also note that this analysis was completed using data from Data Release 14 of APOGEE, 
-which did include molecular lines from FeH in the pipeline at the time, which become numerous and strong for $\teff \lesssim 3600$\,K. }\color{black}
-Further analysis would need to be done to \color{red}{\bf quantify the metallicity improvement for M dwarfs in later data releases of APOGEE now available (in which FeH has been included in analysis)}\color{black}, and determine if the metallicity bias
+We also note that this analysis was completed using data from Data Release 14 of APOGEE, 
+which did include molecular lines from FeH in the pipeline at the time, which become numerous and strong for $\teff \lesssim 3600$\,K \citep{Souto:2017}.
+Further analysis would need to be done to quantify the metallicity improvement for M dwarfs in future data releases of APOGEE, and determine if the metallicity bias
 is found in other model grids (besides the ATLAS/MARCS models used by the ASPCAP pipeline), 
@@
 \acknowledgements
-We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Melissa Ness (Columbia), Kevin Covey (WWU), Diogo Souto (Observatário Nacional/MCTI), Keivan Stassun (Vanderbilt), \color{red}{\bf Katia Cunha (NOAO), Anthony Brown (Leiden), Aida Behmard (Caltech) and Christopher Theissen (UCSD) }\color{black} for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
+We would like to acknowledge Hans Walter Rix (MPIA), Derek Homeier (Universit{\"a}t Heidelberg), Wolfgang Bradner (MPIA), Anna-Christina Eilers (MPIA), Melissa Ness (Columbia), Kevin Covey (WWU), Diogo Souto (Observatário Nacional/MCTI), Keivan Stassun (Vanderbilt), Katia Cunha (NOAO), Anthony Brown (Leiden), Aida Behmard (Caltech) and Christopher Theissen (UCSD) for constructive discussions in the process of this project, as well as Bertrand Goldman (MPIA) and those who have supported the internship program at the Max Planck Institute f{\"u}r Astronomie for providing JB with funding and hospitality.
 This project was partially supported by
@@
 	\begin{center}
-	\includegraphics[width=16cm]{demo_derivatives_teff1.png} 
+	\includegraphics[width=16cm]{figs/demo_derivatives_teff1.png} 
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{demo_derivatives_teff2.png}
+	\includegraphics[width=16cm]{figs/demo_derivatives_teff2.png}
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{demo_derivatives_teff3.png}
+	\includegraphics[width=16cm]{figs/demo_derivatives_teff3.png}
 	\end{center}
-	\caption{\textit{Top two panels of each plot:} \color{red}{\bf APOGEE spectra (black), overlaid by the }\color{black} Mann-trained \color{red}{\bf \cannon\ }\color{black} model for \color{red}{\bf two sources of }\color{black} varying temperatures, and similar metallicities.
+	\caption{\textit{Top two panels of each plot:} APOGEE spectra (black), overlaid by the Mann-trained \cannon\ model for two sources of varying temperatures, and similar metallicities.
 \textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{demo_derivatives_feh1.png}
+	\includegraphics[width=16cm]{figs/demo_derivatives_feh1.png}
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{demo_derivatives_feh2.png}
+	\includegraphics[width=16cm]{figs/demo_derivatives_feh2.png}
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{demo_derivatives_feh3.png}
+	\includegraphics[width=16cm]{figs/demo_derivatives_feh3.png}
 	\end{center}
-	\caption{\textit{Top two panels of each plot:} \color{red}{\bf APOGEE spectra (black), overlaid by the }\color{black} Mann-trained \color{red}{\bf \cannon\ }\color{black} model for \color{red}{\bf two sources of }\color{black} varying metallicities and similar temperatures.
+	\caption{\textit{Top two panels of each plot:} APOGEE spectra (black), overlaid by the Mann-trained \cannon\ model for two sources of varying metallicities and similar temperatures.
 \textit{Third panel:} Derivative of \thecannon\ model with respect to metallicity, taken at the median training metallicity, $\feh=-0.03$\,dex; jackknife computed error at each pixel is shown in red.} \label{fig:demo_feh}
@@
 	\begin{center}
-	\includegraphics[width=12cm]{Spectral_Sequence_1.png}
-	\includegraphics[width=12cm]{Spectral_Sequence_2.png}
-	\includegraphics[width=12cm]{Spectral_Sequence_3.png}
+	\includegraphics[width=12cm]{figs/Spectral_Sequence_1.png}
+	\includegraphics[width=12cm]{figs/Spectral_Sequence_2.png}
+	\includegraphics[width=12cm]{figs/Spectral_Sequence_3.png}
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=16cm]{derivative_jackknife_spt.png}
+	\includegraphics[width=16cm]{figs/derivative_jackknife_spt.png}
 	\end{center}
